---
title: "![](CHOPMC_logo.pdf){width=5in}"
author: "Scott G. Daniel, Ph.D."
date: "`r Sys.Date()`"
output:
  word_document:
    toc: true
    toc_depth: '4'
  pdf_document:
    keep_tex: false
    toc: true
    toc_depth: 4
geometry: margin=3cm
---

```{r setup, echo=FALSE}
### ================
###   knitr setup
### ================
library(knitr)
opts_chunk$set(
  tidy=FALSE,
  cache=FALSE,
  echo=FALSE,
  warning=FALSE,
  message=FALSE,
  dpi=100,
  fig.width=4,
  fig.height=4,
  fig.align = "center"
  )
```

```{r libs and funcs, message=FALSE, warning=FALSE}

library(here)
library(tidyverse)
library(qiimer)
library(vegan)
library(ape)
library(usedist)

## Visualization packages
library(pander)
library(kableExtra)

library(pheatbuilder)
library(ggbeeswarm)
library(ggsci)
library(viridis)
library(RColorBrewer)

# stats packages
library(adonisplus)
library(nlme)
library(emmeans) # for lmer post-hoc tests
library(broom)
library(ghibli)
library(glue)
library(magrittr)
select <- dplyr::select
library(patchwork)
library(qualpalr)
library(randomForest)
library(microbenchmark)
library(Boruta)
library(mlbench)
library(caret)
library(pROC)
```

```{r load data}

load(file = here("Data","shotgun_data_2026-04-28.RData")) #source for raw data
load(file = here("Data","sagori_taxonomic_data.RData")) #source for fixed s2 table
load(file = here("Data","sagori_kegg_data.RData")) #source for gene abundance data
table_iterator <- function() {
  table_count <- 0
  function(x, title = "table_export", ...) {
    table_count <<- table_count + 1
    table_name <- paste0(title, table_count, ".csv")
    table_path <- here("Output_sensitivity", table_name)
    write_csv(x = x, file = table_path, ...)
    print(paste0("This table is saved as ", table_name, "."))
  }
}

table_export <- table_iterator()

```

```{r log variables and add more metadata}

sens_data <- read_csv(here("Data", "clin_variables_for_scott_family_ID.csv")) %>%
  mutate(redcap_id = as.character(redcap_id))

s2 %<>%
  mutate(npo_days_prop_min = min(npo_days_prop[npo_days_prop > 0])) %>%
  mutate(npo_days_prop = ifelse(npo_days_prop == 0, npo_days_prop_min/2, npo_days_prop)) %>%
  mutate(log_npo_days_prop = log(npo_days_prop)) %>%
  mutate(AbxHoursPerHOL_min = min(AbxHoursPerHOL[AbxHoursPerHOL > 0])) %>%
  mutate(AbxHoursPerHOL = ifelse(AbxHoursPerHOL == 0, AbxHoursPerHOL_min/2, AbxHoursPerHOL)) %>%
  mutate(log_AbxHoursPerHOL = log(AbxHoursPerHOL)) %>%
  left_join(sens_data, join_by(redcap_id))

```

```{r moar colors}

ann_colors <- list(category = c(`1` = "#66C2A5", `2` = "#FC8D62", `3` = "#8DA0CB"
),
growth_cat = c(normal = "#92BBD9FF", `mild retardation` = "#DCCA2CFF", 
`moderate to severe retardation` = "#6FB382FF"),
TimePoint = c(T0 = "#440154FF", 
T1 = "#48186AFF", T2 = "#472D7BFF", T3 = "#424086FF", T4 = "#3B528BFF", 
T5 = "#33638DFF", T6 = "#2C728EFF", T7 = "#26828EFF", T8 = "#21908CFF", 
T9 = "#1F9F88FF", T10 = "#27AD81FF", T11 = "#3EBC74FF", T12 = "#5DC863FF", 
T13 = "#82D34DFF", T14 = "#AADC32FF", T15 = "#D5E21AFF", T16 = "#FDE725FF"
),
growth_fail = c(No = "#9FCB6A", Yes = "#8367CC"),
growth_label = c(None = "#92BBD9FF", Mild = "#DCCA2CFF", 
`Moderate\nto severe` = "#6FB382FF"))

```

```{r replace zeros and filter}

# replace zeros per Taxa
max_num <- max(summed_props)
prop_cut = 0.01
summed_props_long <- summed_props %>%
  as.data.frame() %>% 
  rownames_to_column("Taxa") %>% 
  pivot_longer(-Taxa, names_to="SampleID", values_to="props") %>%
  group_by(Taxa) %>%
  mutate(Taxa_min = ifelse(props == 0, max_num, props)) %>%
  mutate(props = ifelse(props == 0, min(Taxa_min)/2, props)) %>%
  ungroup() %>%
  select(-Taxa_min) %>%
  droplevels()

props_toTest <- summed_props_long %>%
  group_by(Taxa) %>%
  mutate(mean_prop = mean(props)) %>%
  ungroup() %>%
  filter(mean_prop > prop_cut) %>%
  filter(str_detect(Taxa, "Bacteria|acnes", negate = T)) %>%
  mutate(log10_props = log10(props))

perm <- 999

```

# Growth faltering original variables

## alpha div. Linear model

```{r, results='asis'}

my_formula <- "alpha ~ bwt_z + log_npo_days_prop + WOL * growth_cat"

summaries_df <- s2 %>%
  filter(!TimePoint %in% "T0") %>%
  pivot_longer(c("Richness", "Shannon"), names_to="metric", values_to="alpha") %>%
  group_by(metric) %>%
  do(tidy_lmer(nlme::lme(as.formula(my_formula), random=~1|SubjectID, data=., na.action=na.omit))) %>%
  ungroup() %>%
  filter(!grepl("Intercept", term)) %>%
  mutate(term = str_replace(term, "WOL", "Week of life")) %>%
  mutate(term = if_else(str_detect(term, "growth_cat"), str_remove(term, "growth_cat"), term))

summaries_df %>%
  pander(digits = 2, split.table = Inf, split.cells = 20, caption = glue("Linear model of the alpha diversity measurement with growth_cat as the predictive variable. SubjectID added as a random effect. Full formula: {my_formula}"), emphasize.strong.rows = which(.$p.value<=0.05))

table_export(summaries_df, "sensitivity_analysis_")

```

## Permanova with growth_cat and time

```{r, results='asis'}

my_formula <- "distmat ~ bwt_z + log_npo_days_prop + growth_cat * WOL"

# summaries_df <-
#   s2 %>%
#     do(adonisplus(
#       ., distmat = bc, formula = as.formula(my_formula),
#       sample_id_var = SampleID, rep_meas_var = SubjectID,
#       shuffle = c(WOL = "within", growth_cat = "between"), permutations=perm)) %>%
#   mutate(metric = "Bray-Curtis") %>%
#   filter(!term %in% c("Residuals", "Total")) %>%
#   select(-one_of(c("sumsq", "meansq"))) %>%
#   select(metric, everything())

rds_file <- here("Output_sensitivity", "permanova_results_01.rds")
# write_rds(summaries_df, rds_file)
summaries_df <- read_rds(rds_file)

summaries_df %>%
  mutate(term = str_replace(term, "WOL", "Week of life")) %>%
  pander(digits = 2, split.table = Inf, split.cells = 20, caption = glue("Permanova of Bray-curtis distances between samples. SubjectID added as a random effect. Full formula: {my_formula}"), emphasize.strong.rows = which(.$p.value<=0.05))

table_export(summaries_df, "sensitivity_analysis_")

```

## Differential abundance

```{r}

my_formula <- "log10_props ~ bwt_z + log_npo_days_prop + WOL * growth_cat"

summaries_df <- run_lmer_on_props(factor_to_test = "Taxa", props_toTest = props_toTest, s_toTest = s2, form1 = my_formula, rep_mes_label = "SubjectID", p_cutoff = 0.05)

summaries_df %>%
  mutate(term = str_replace(term, "WOL", "Week of life")) %>%
  mutate(term = if_else(str_detect(term, "growth_cat"), str_remove(term, "growth_cat"), term)) %>%
  pander(digits = 2, split.table = Inf, split.cells = 20, caption = glue("Linear model of the log10-transformed relative abundance with WOL and growth categories as the predictive variables. SubjectID added as a random effect. Full formula: {my_formula}"), emphasize.strong.rows = which(.$fdr<=0.05))

table_export(summaries_df, "sensitivity_analysis_")

```


# Growth failure original variables

## alpha div. Linear model

```{r, results='asis'}

my_formula <- "alpha ~ bwt_z + log_npo_days_prop + WOL * growth_failure"

summaries_df <- s2 %>%
  filter(!TimePoint %in% "T0") %>%
  pivot_longer(c("Richness", "Shannon"), names_to="metric", values_to="alpha") %>%
  group_by(metric) %>%
  do(tidy_lmer(nlme::lme(as.formula(my_formula), random=~1|SubjectID, data=., na.action=na.omit))) %>%
  ungroup() %>%
  filter(!grepl("Intercept", term)) %>%
  mutate(term = str_replace(term, "WOL", "Week of life")) %>%
  mutate(term = str_replace(term, "growth_failureYes", "Had growth failure"))

summaries_df %>%
  pander(digits = 2, split.table = Inf, split.cells = 20, caption = glue("Linear model of the alpha diversity measurement with growth_failure as the predictive variable. SubjectID added as a random effect. Full formula: {my_formula}"), emphasize.strong.rows = which(.$p.value<=0.05))

table_export(summaries_df, "sensitivity_analysis_")

```

## Permanova with growth_failure and time

```{r, results='asis'}

my_formula <- "distmat ~ bwt_z + log_npo_days_prop + WOL * growth_failure"

# summaries_df <-
#   s2 %>%
#     do(adonisplus(
#       ., distmat = bc, formula = as.formula(my_formula),
#       sample_id_var = SampleID, rep_meas_var = SubjectID,
#       shuffle = c(WOL = "within", growth_failure = "between"), permutations=perm)) %>%
#   mutate(metric = "Bray-Curtis") %>%
#   filter(!term %in% c("Residuals", "Total")) %>%
#   select(-one_of(c("sumsq", "meansq"))) %>%
#   select(metric, everything())

rds_file <- here("Output_sensitivity", "permanova_results_02.rds")
# write_rds(summaries_df, rds_file)
summaries_df <- read_rds(rds_file)

summaries_df %>%
  mutate(term = str_replace(term, "WOL", "Week of life")) %>%
  pander(digits = 2, split.table = Inf, split.cells = 20, caption = glue("Permanova of Bray-curtis distances between samples. SubjectID added as a random effect. Full formula: {my_formula}"), emphasize.strong.rows = which(.$p.value<=0.05))

table_export(summaries_df, "sensitivity_analysis_")


```

## Differential abundance

```{r}

my_formula <- "log10_props ~ bwt_z + log_npo_days_prop + WOL * growth_failure"

summaries_df <- run_lmer_on_props(factor_to_test = "Taxa", props_toTest = props_toTest, s_toTest = s2, form1 = my_formula, rep_mes_label = "SubjectID", p_cutoff = 0.05)

summaries_df %>%
  mutate(term = str_replace(term, "WOL", "Week of life")) %>%
  mutate(term = str_replace(term, "growth_failureYes", "Had growth failure")) %>%
  pander(digits = 2, split.table = Inf, split.cells = 20, caption = glue("Linear model of the log10-transformed relative abundance with WOL and growth failure as the predictive variables. SubjectID added as a random effect. Full formula: {my_formula}"), emphasize.strong.rows = which(.$fdr<=0.05))

table_export(summaries_df, "sensitivity_analysis_")

```

# Growth faltering + abx3 + prolacta

## alpha div. Linear model

```{r, results='asis'}

my_formula <- "alpha ~ bwt_z + log_npo_days_prop + abx_lt3d + prolacta + WOL * growth_cat"

summaries_df <- s2 %>%
  filter(!TimePoint %in% "T0") %>%
  pivot_longer(c("Richness", "Shannon"), names_to="metric", values_to="alpha") %>%
  group_by(metric) %>%
  do(tidy_lmer(nlme::lme(as.formula(my_formula), random=~1|SubjectID, data=., na.action=na.omit))) %>%
  ungroup() %>%
  filter(!grepl("Intercept", term)) %>%
  mutate(term = str_replace(term, "WOL", "Week of life")) %>%
  mutate(term = if_else(str_detect(term, "growth_cat"), str_remove(term, "growth_cat"), term))

summaries_df %>%
  pander(digits = 2, split.table = Inf, split.cells = 20, caption = glue("Linear model of the alpha diversity measurement with growth_cat as the predictive variable. SubjectID added as a random effect. Full formula: {my_formula}"), emphasize.strong.rows = which(.$p.value<=0.05))

table_export(summaries_df, "sensitivity_analysis_")

```

## Permanova with growth_cat and time

```{r, results='asis'}

my_formula <- "distmat ~ abx_lt3d + prolacta + bwt_z + log_npo_days_prop + WOL * growth_cat"

# summaries_df <-
#   s2 %>%
#     do(adonisplus(
#       ., distmat = bc, formula = as.formula(my_formula),
#       sample_id_var = SampleID, rep_meas_var = SubjectID,
#       shuffle = c(WOL = "within", growth_cat = "between"), permutations=perm)) %>%
#   mutate(metric = "Bray-Curtis") %>%
#   filter(!term %in% c("Residuals", "Total")) %>%
#   select(-one_of(c("sumsq", "meansq"))) %>%
#   select(metric, everything())

rds_file <- here("Output_sensitivity", "permanova_results_03.rds")
# write_rds(summaries_df, rds_file)
summaries_df <- read_rds(rds_file)

summaries_df %>%
  mutate(term = str_replace(term, "WOL", "Week of life")) %>%
  pander(digits = 2, split.table = Inf, split.cells = 20, caption = glue("Permanova of Bray-curtis distances between samples. SubjectID added as a random effect. Full formula: {my_formula}"), emphasize.strong.rows = which(.$p.value<=0.05))

table_export(summaries_df, "sensitivity_analysis_")

```

## Differential abundance

```{r}

my_formula <- "log10_props ~ abx_lt3d + prolacta + bwt_z + log_npo_days_prop + WOL * growth_cat"

summaries_df <- run_lmer_on_props(factor_to_test = "Taxa", props_toTest = props_toTest, s_toTest = s2, form1 = my_formula, rep_mes_label = "SubjectID", p_cutoff = 0.05)

summaries_df %>%
  mutate(term = str_replace(term, "WOL", "Week of life")) %>%
  mutate(term = if_else(str_detect(term, "growth_cat"), str_remove(term, "growth_cat"), term)) %>%
  pander(digits = 2, split.table = Inf, split.cells = 20, caption = glue("Linear model of the log10-transformed relative abundance with WOL and growth categories as the predictive variables. SubjectID added as a random effect. Full formula: {my_formula}"), emphasize.strong.rows = which(.$fdr<=0.05))

table_export(summaries_df, "sensitivity_analysis_")

```


# Growth failure + abx3 + prolacta

## alpha div. Linear model

```{r, results='asis'}

my_formula <- "alpha ~ bwt_z + log_npo_days_prop + abx_lt3d + prolacta + WOL * growth_failure"

summaries_df <- s2 %>%
  filter(!TimePoint %in% "T0") %>%
  pivot_longer(c("Richness", "Shannon"), names_to="metric", values_to="alpha") %>%
  group_by(metric) %>%
  do(tidy_lmer(nlme::lme(as.formula(my_formula), random=~1|SubjectID, data=., na.action=na.omit))) %>%
  ungroup() %>%
  filter(!grepl("Intercept", term)) %>%
  mutate(term = str_replace(term, "WOL", "Week of life")) %>%
  mutate(term = str_replace(term, "growth_failureYes", "Had growth failure"))

summaries_df %>%
  pander(digits = 2, split.table = Inf, split.cells = 20, caption = glue("Linear model of the alpha diversity measurement with growth_failure as the predictive variable. SubjectID added as a random effect. Full formula: {my_formula}"), emphasize.strong.rows = which(.$p.value<=0.05))

table_export(summaries_df, "sensitivity_analysis_")

```

## Permanova with growth_failure and time

```{r, results='asis'}

my_formula <- "distmat ~ abx_lt3d + prolacta + bwt_z + log_npo_days_prop + WOL * growth_failure"

# summaries_df <-
#   s2 %>%
#     do(adonisplus(
#       ., distmat = bc, formula = as.formula(my_formula),
#       sample_id_var = SampleID, rep_meas_var = SubjectID,
#       shuffle = c(WOL = "within", growth_failure = "between"), permutations=perm)) %>%
#   mutate(metric = "Bray-Curtis") %>%
#   filter(!term %in% c("Residuals", "Total")) %>%
#   select(-one_of(c("sumsq", "meansq"))) %>%
#   select(metric, everything())

rds_file <- here("Output_sensitivity", "permanova_results_04.rds")
# write_rds(summaries_df, rds_file)
summaries_df <- read_rds(rds_file)

summaries_df %>%
  mutate(term = str_replace(term, "WOL", "Week of life")) %>%
  pander(digits = 2, split.table = Inf, split.cells = 20, caption = glue("Permanova of Bray-curtis distances between samples. SubjectID added as a random effect. Full formula: {my_formula}"), emphasize.strong.rows = which(.$p.value<=0.05))

table_export(summaries_df, "sensitivity_analysis_")


```

## Differential abundance

```{r}

my_formula <- "log10_props ~ abx_lt3d + prolacta + bwt_z + log_npo_days_prop + WOL * growth_failure"

summaries_df <- run_lmer_on_props(factor_to_test = "Taxa", props_toTest = props_toTest, s_toTest = s2, form1 = my_formula, rep_mes_label = "SubjectID", p_cutoff = 0.05)

summaries_df %>%
  mutate(term = str_replace(term, "WOL", "Week of life")) %>%
  mutate(term = str_replace(term, "growth_failureYes", "Had growth failure")) %>%
  pander(digits = 2, split.table = Inf, split.cells = 20, caption = glue("Linear model of the log10-transformed relative abundance with WOL and growth failure as the predictive variables. SubjectID added as a random effect. Full formula: {my_formula}"), emphasize.strong.rows = which(.$fdr<=0.05))

table_export(summaries_df, "sensitivity_analysis_")

```

# Family_ID for twins

Including Family_ID given the non-independant nature of twins. SubjectID is nested within Family_ID since there can never be a SubjectID that has multiple Family_ID's.

## Growth faltering

### alpha div. Linear model

```{r, results='asis'}

my_formula <- "alpha ~ bwt_z + log_npo_days_prop + WOL * growth_cat"

summaries_df <- s2 %>%
  filter(!TimePoint %in% "T0") %>%
  pivot_longer(c("Richness", "Shannon"), names_to="metric", values_to="alpha") %>%
  group_by(metric) %>%
  do(tidy_lmer(nlme::lme(as.formula(my_formula), random=~1|SubjectID/Family_ID, data=., na.action=na.omit))) %>%
  ungroup() %>%
  filter(!grepl("Intercept", term)) %>%
  mutate(term = str_replace(term, "WOL", "Week of life")) %>%
  mutate(term = if_else(str_detect(term, "growth_cat"), str_remove(term, "growth_cat"), term))

summaries_df %>%
  pander(digits = 2, split.table = Inf, split.cells = 20, caption = glue("Linear model of the alpha diversity measurement with growth_cat as the predictive variable. SubjectID/Family_ID added as a random effect. Full formula: {my_formula}"), emphasize.strong.rows = which(.$p.value<=0.05))

table_export(summaries_df, "sensitivity_analysis_")

```

### Permanova with growth_cat and time

Family_ID has to be included as a covariate because you can't do nested random variables .

```{r, results='asis'}

my_formula <- "distmat ~ Family_ID + bwt_z + log_npo_days_prop + WOL * growth_cat"

# summaries_df <-
#   s2 %>%
#     do(adonisplus(
#       ., distmat = bc, formula = as.formula(my_formula),
#       sample_id_var = SampleID, rep_meas_var = SubjectID,
#       shuffle = c(WOL = "within", growth_cat = "between"), permutations=perm)) %>%
#   mutate(metric = "Bray-Curtis") %>%
#   filter(!term %in% c("Residuals", "Total")) %>%
#   select(-one_of(c("sumsq", "meansq"))) %>%
#   select(metric, everything())

rds_file <- here("Output_sensitivity", "permanova_results_05.rds")
# write_rds(summaries_df, rds_file)
summaries_df <- read_rds(rds_file)

summaries_df %>%
  mutate(term = str_replace(term, "WOL", "Week of life")) %>%
  pander(digits = 2, split.table = Inf, split.cells = 20, caption = glue("Permanova of Bray-curtis distances between samples. SubjectID added as a random effect. Full formula: {my_formula}"), emphasize.strong.rows = which(.$p.value<=0.05))

table_export(summaries_df, "sensitivity_analysis_")

```

### Differential abundance

```{r}

my_formula <- "log10_props ~ bwt_z + log_npo_days_prop + WOL * growth_cat"

summaries_df <- run_lmer_on_props(factor_to_test = "Taxa", props_toTest = props_toTest, s_toTest = s2, form1 = my_formula, rep_mes_label = "SubjectID/Family_ID", p_cutoff = 0.05)

summaries_df %>%
  mutate(term = str_replace(term, "WOL", "Week of life")) %>%
  mutate(term = if_else(str_detect(term, "growth_cat"), str_remove(term, "growth_cat"), term)) %>%
  pander(digits = 2, split.table = Inf, split.cells = 20, caption = glue("Linear model of the log10-transformed relative abundance with WOL and growth categories as the predictive variables. SubjectID added as a random effect. Full formula: {my_formula}"), emphasize.strong.rows = which(.$fdr<=0.05))

table_export(summaries_df, "sensitivity_analysis_")

```


## Growth failure

### alpha div. Linear model

```{r, results='asis'}

my_formula <- "alpha ~ bwt_z + log_npo_days_prop + WOL * growth_failure"

summaries_df <- s2 %>%
  filter(!TimePoint %in% "T0") %>%
  pivot_longer(c("Richness", "Shannon"), names_to="metric", values_to="alpha") %>%
  group_by(metric) %>%
  do(tidy_lmer(nlme::lme(as.formula(my_formula), random=~1|SubjectID/Family_ID, data=., na.action=na.omit))) %>%
  ungroup() %>%
  filter(!grepl("Intercept", term)) %>%
  mutate(term = str_replace(term, "WOL", "Week of life")) %>%
  mutate(term = str_replace(term, "growth_failureYes", "Had growth failure"))

summaries_df %>%
  pander(digits = 2, split.table = Inf, split.cells = 20, caption = glue("Linear model of the alpha diversity measurement with growth_failure as the predictive variable. SubjectID added as a random effect. Full formula: {my_formula}"), emphasize.strong.rows = which(.$p.value<=0.05))

table_export(summaries_df, "sensitivity_analysis_")

```

### Permanova with growth_failure and time

```{r, results='asis'}

my_formula <- "distmat ~ Family_ID + bwt_z + log_npo_days_prop + WOL * growth_failure"

# summaries_df <-
#   s2 %>%
#     do(adonisplus(
#       ., distmat = bc, formula = as.formula(my_formula),
#       sample_id_var = SampleID, rep_meas_var = SubjectID,
#       shuffle = c(WOL = "within", growth_failure = "between"), permutations=perm)) %>%
#   mutate(metric = "Bray-Curtis") %>%
#   filter(!term %in% c("Residuals", "Total")) %>%
#   select(-one_of(c("sumsq", "meansq"))) %>%
#   select(metric, everything())

rds_file <- here("Output_sensitivity", "permanova_results_06.rds")
# write_rds(summaries_df, rds_file)
summaries_df <- read_rds(rds_file)

summaries_df %>%
  mutate(term = str_replace(term, "WOL", "Week of life")) %>%
  pander(digits = 2, split.table = Inf, split.cells = 20, caption = glue("Permanova of Bray-curtis distances between samples. SubjectID added as a random effect. Full formula: {my_formula}"), emphasize.strong.rows = which(.$p.value<=0.05))

table_export(summaries_df, "sensitivity_analysis_")


```

### Manual method with adonis2:

```{r}
# try doing this manually with adonis2
# use dist_subset to make the distance matrix match the s2
# then run adonis2 with strata = SubjectID (but you won't get the shuffling thing)
# load(file = "data_for_adonis.RData")
s_simple <- s2 %>% select(SampleID, Family_ID, bwt_z, log_npo_days_prop, WOL, growth_failure, SubjectID)
my_dist <- dist_subset(bc, s_simple$SampleID)
# summaries_adonis2 <- adonis2(formula = my_dist ~ Family_ID + bwt_z + log_npo_days_prop + WOL * growth_failure, data = s_simple, permutations = perm, by = "terms", strata = s_simple$SubjectID)

rds_file <- here("Output_sensitivity", "adonis2_results_01.rds")
# write_rds(summaries_adonis2, rds_file)
summaries_adonis2 <- read_rds(rds_file)

my_table <- adonisplus::tidy(summaries_adonis2)
my_table %>%
  mutate(term = str_replace(term, "WOL", "Week of life")) %>%
  pander(digits = 2, split.table = Inf, split.cells = 20, caption = glue("Permanova of Bray-curtis distances between samples using Adonis2.  Full formula: {my_formula}"), emphasize.strong.rows = which(.$p.value<=0.05))
table_export(my_table, "sensitivity_analysis_")
save(s2, bc, file = here("data_for_adonis.RData"))
```

### Adonisplus or Bust

```{r}

my_formula <- "distmat ~ bwt_z + log_npo_days_prop + WOL * growth_failure"

# summaries_df <-
#   s2 %>%
#     do(adonisplus(
#       ., distmat = bc, formula = as.formula(my_formula),
#       sample_id_var = SampleID, rep_meas_var = Family_ID,
#       shuffle = c(WOL = "within"), permutations=perm)) %>%
#   mutate(metric = "Bray-Curtis") %>%
#   filter(!term %in% c("Residuals", "Total")) %>%
#   select(-one_of(c("sumsq", "meansq"))) %>%
#   select(metric, everything())

rds_file <- here("Output_sensitivity", "special_permanova_results.rds")
# write_rds(summaries_df, rds_file)
summaries_df <- read_rds(rds_file)

summaries_df %>%
  mutate(term = str_replace(term, "WOL", "Week of life")) %>%
  pander(digits = 2, split.table = Inf, split.cells = 20, caption = glue("Permanova of Bray-curtis distances between samples. Family_ID added as a random effect. Full formula: {my_formula}"), emphasize.strong.rows = which(.$p.value<=0.05))

table_export(summaries_df, "sensitivity_analysis_")

```

### Differential abundance

```{r}

my_formula <- "log10_props ~ bwt_z + log_npo_days_prop + WOL * growth_failure"

summaries_df <- run_lmer_on_props(factor_to_test = "Taxa", props_toTest = props_toTest, s_toTest = s2, form1 = my_formula, rep_mes_label = "SubjectID/Family_ID", p_cutoff = 0.05)

summaries_df %>%
  mutate(term = str_replace(term, "WOL", "Week of life")) %>%
  mutate(term = str_replace(term, "growth_failureYes", "Had growth failure")) %>%
  pander(digits = 2, split.table = Inf, split.cells = 20, caption = glue("Linear model of the log10-transformed relative abundance with WOL and growth failure as the predictive variables. SubjectID added as a random effect. Full formula: {my_formula}"), emphasize.strong.rows = which(.$fdr<=0.05))

table_export(summaries_df, "sensitivity_analysis_")

```


# Growth failure at 10% threshold

We did 3% threshold in main results

```{r}

# have to reload the s2 table
load(file = here("Data","sagori_taxonomic_data.RData"))

s2 %<>%
  mutate(growth_failure = fct_recode(GF36_10, Yes = "1", No = "0"))

perm <- 999

```

### alpha div. Linear model

```{r, results='asis'}

my_formula <- "alpha ~ bwt_z + log_npo_days_prop + WOL * growth_failure"

summaries_df <- s2 %>%
  filter(!TimePoint %in% "T0") %>%
  pivot_longer(c("Richness", "Shannon"), names_to="metric", values_to="alpha") %>%
  group_by(metric) %>%
  do(tidy_lmer(nlme::lme(as.formula(my_formula), random=~1|SubjectID, data=., na.action=na.omit))) %>%
  ungroup() %>%
  filter(!grepl("Intercept", term)) %>%
  mutate(term = str_replace(term, "WOL", "Week of life")) %>%
  mutate(term = str_replace(term, "growth_failureYes", "Had growth failure"))

summaries_df %>%
  pander(digits = 2, split.table = Inf, split.cells = 20, caption = glue("Linear model of the alpha diversity measurement with growth_failure as the predictive variable. SubjectID added as a random effect. Full formula: {my_formula}"), emphasize.strong.rows = which(.$p.value<=0.05))

table_export(summaries_df, "sensitivity_analysis_")

```

### Permanova with growth_failure and time

```{r, results='asis'}

my_formula <- "distmat ~ bwt_z + log_npo_days_prop + WOL * growth_failure"

# summaries_df <-
#   s2 %>%
#     do(adonisplus(
#       ., distmat = bc, formula = as.formula(my_formula),
#       sample_id_var = SampleID, rep_meas_var = SubjectID,
#       shuffle = c(WOL = "within", growth_failure = "between"), permutations=perm)) %>%
#   mutate(metric = "Bray-Curtis") %>%
#   filter(!term %in% c("Residuals", "Total")) %>%
#   select(-one_of(c("sumsq", "meansq"))) %>%
#   select(metric, everything())

rds_file <- here("Output_sensitivity", "permanova_results_07.rds")
# write_rds(summaries_df, rds_file)
summaries_df <- read_rds(rds_file)

summaries_df %>%
  mutate(term = str_replace(term, "WOL", "Week of life")) %>%
  pander(digits = 2, split.table = Inf, split.cells = 20, caption = glue("Permanova of Bray-curtis distances between samples. SubjectID added as a random effect. Full formula: {my_formula}"), emphasize.strong.rows = which(.$p.value<=0.05))

table_export(summaries_df, "sensitivity_analysis_")


```

### Differential abundance

```{r}

my_formula <- "log10_props ~ bwt_z + log_npo_days_prop + WOL * growth_failure"

summaries_df <- run_lmer_on_props(factor_to_test = "Taxa", props_toTest = props_toTest, s_toTest = s2, form1 = my_formula, rep_mes_label = "SubjectID", p_cutoff = 0.05)

summaries_df %>%
  mutate(term = str_replace(term, "WOL", "Week of life")) %>%
  mutate(term = str_replace(term, "growth_failureYes", "Had growth failure")) %>%
  pander(digits = 2, split.table = Inf, split.cells = 20, caption = glue("Linear model of the log10-transformed relative abundance with WOL and growth failure as the predictive variables. SubjectID added as a random effect. Full formula: {my_formula}"), emphasize.strong.rows = which(.$fdr<=0.05))

table_export(summaries_df, "sensitivity_analysis_")

```

# End of report

```{r Generate time stamped report, eval=FALSE, include=FALSE}
#notes
#neat: you can run this following command in the console to give your reports custom names (or date-stamp them)
#rmarkdown::render('ronen_stein_Report_shotgun.Rmd',output_file = paste('CEASE.report.', Sys.Date(), '.pdf', sep=''))

rstudioapi::documentSaveAll()

script_dir <- file.path("Scripts","Rmds")
output_dir <- file.path("Output_sensitivity")

report_fp <- here::here(script_dir, "sensitivity/sensitivity_analysis.Rmd")
output_fp <- here::here(output_dir, paste('growth_paper_sensitivity_analysis_', Sys.Date(), '.pdf', sep=''))

rmarkdown::render(report_fp, output_file = output_fp, output_format = "pdf_document")

```
