---
title: |
  Sagori Mukhopadhyay - Runs 1 to 5 - Random forest analysis
  With growth failure at the 10% percentile cutoff
author: "Scott G. Daniel, PennCHOP Microbiome Program"
date: "`r Sys.Date()`"
geometry: margin=3cm
output: 
    pdf_document:
        keep_tex: false
        toc: true
        toc_depth: 4
---

```{r setup, echo=FALSE}
### ================
###   knitr setup
### ================
library(knitr)
opts_chunk$set(
  tidy=FALSE,
  cache=FALSE,
  echo=FALSE,
  warning=FALSE,
  message=FALSE,
  dpi=100,
  fig.width=8,
  fig.height=8,
  fig.align = "center"
  )
```

```{r libs and funcs, message=FALSE, warning=FALSE}

library(here)
library(tidyverse)
library(qiimer)
library(vegan)
library(ape)
library(usedist)

## Visualization packages
library(pander)
library(kableExtra)

library(pheatbuilder)
library(ggbeeswarm)
library(ggsci)
library(viridis)
library(RColorBrewer)

# stats packages
library(adonisplus)
library(nlme)
library(emmeans) # for lmer post-hoc tests
library(broom)
library(ghibli)
library(glue)
library(magrittr)
select <- dplyr::select
library(patchwork)
library(qualpalr)
library(randomForest)
library(microbenchmark)
library(Boruta)
library(mlbench)
library(caret)
library(pROC)
```

```{r load data}
# source(here("Scripts", "Functions", "common_functions.R"))
load(file = here("Data","shotgun_data_2026-04-28.RData")) #source for raw data
load(file = here("Data","sagori_taxonomic_data.RData")) #source for fixed s2 table
load(file = here("Data","sagori_kegg_data.RData")) #source for gene abundance data
table_export <- table_iterator()

```

# Random forest analysis / predicting microbiome age

```{r setup data frames}

s2 %<>%
  mutate(growth_failure = fct_recode(GF36_10, Yes = "1", No = "0"))

s_toTest <- s2 %>% filter(Keep, !isControl)

top25 <- summed_props_long %>%
  group_by(Taxa) %>%
  summarise(mean_props = mean(props)) %>%
  ungroup() %>%
  arrange(desc(mean_props)) %>%
  slice(1:29) %>%
  filter(!str_detect(Taxa, "Ralstonia|Cutibacterium|Vibrio"))

training_data <- s_toTest %>%
  filter(growth_failure %in% "No") %>%
  left_join(summed_props_long, by = "SampleID") %>%
  filter(Taxa %in% top25$Taxa) %>%
  select(SampleID, DOL, Taxa, props) %>% #SubjectID, 
  mutate(Taxa = str_replace_all(Taxa, " ", "_")) %>%
  pivot_wider(names_from = "Taxa", values_from = "props", values_fill = 0) %>%
  column_to_rownames(var = "SampleID")

set.seed(43)
validation_babies <- sample(s_toTest$SubjectID, 15)
validation_samples <- s_toTest %>%
  filter(growth_failure %in% "No") %>%
  filter(SubjectID %in% validation_babies) %>%
  pull(SampleID)

validation_data <- training_data[validation_samples,]
ind2 <- setdiff(rownames(training_data), rownames(validation_data))
training_data <- training_data[ind2,]

testing_data <- s_toTest %>%
  filter(growth_failure %in% "Yes") %>%
  left_join(summed_props_long, by = "SampleID") %>%
  filter(Taxa %in% top25$Taxa) %>%
  select(SampleID, DOL, Taxa, props) %>% #SubjectID, 
  mutate(Taxa = str_replace_all(Taxa, " ", "_")) %>%
  pivot_wider(names_from = "Taxa", values_from = "props", values_fill = 0) %>%
  column_to_rownames(var = "SampleID")

#check
# all.equal(colnames(testing_data), colnames(training_data))
# [1] TRUE

```

To use a random forest model we will do the following:

1) Use the top 25 most abundant bacterial species (excluding common contaminants like *Cutibacterium acnes*.

2) Divide the data set into those with regular growth (the training set) and those with growth failure (the test set).

5) Do a random selection of `r length(validation_babies)` babies as the validation set (which gives `r length(validation_samples)` samples).

3) Construct the random forest model on the training set (minus validation samples).

4) We will use the model to predict the microbiome age of the growth failure babies. The hypothesis being that their microbiomes are actually younger and less-developed than similarly-aged normal growth babies.

Here is a summary of the model with the % variance explained:

```{r the model}
rf_model <- randomForest(DOL ~ ., data = training_data, importance=T)
print(rf_model)

#save for later
rf_model_25taxa <- rf_model
```

## Accuracy

We can also order the predictor taxa according to how much they add to the accuracy of the model:

```{r, results='asis'}
imp <- importance(rf_model, type = 1)
imp_df <- tibble(Taxa = dimnames(imp)[[1]], MeanDecAccu = as.vector(imp)) #strata = "SubjectID",
imp_df %>%
  arrange(desc(MeanDecAccu)) %>%
  pander(digits=2)
```

### Plot

Or as a plot:

```{r, fig.width=7, fig.height=5}

varImpPlot(rf_model, type = 1)

dev.copy(pdf, here("Output_sensitivity", "importance_graph.pdf"))

dev.off()

```

## Plot of training data

Next, we can plot the predicted age of the samples vs. the actual age:

```{r}

plot_df <- tibble(class = "Regular growth", MMI = as.vector(rf_model$predicted), actual_DOL = as.vector(rf_model$y))

growth_class_colors <- c("Regular growth" = "#446590FF")

# nls = nonlinear least squares
# SSlogis = Self-starting logistic model

mod1 = nls(MMI ~ SSlogis(actual_DOL, Asym, xmid, scal), data=plot_df)
# mod1

plot_df %>%
  ggplot(mapping = aes(x=actual_DOL, y=MMI, color=class)) +
  theme_bw() +
  theme(legend.position = "none") +
  geom_point() +
  geom_smooth(
    method = "nls", formula = y ~ SSlogis(x, Asym, xmid, scal),
    se = FALSE
  ) +
  annotate(geom="text", x=10, y=70, label=glue("Logistic curve for normal growth:\n Horiz. asymptote={round(mod1$m$getPars()['Asym'],2)}\n  xmid={round(mod1$m$getPars()['xmid'],2)}\n  scale={round(mod1$m$getPars()['scal'],2)}"), color = "#446590FF", hjust = 0) +
  coord_equal() +
  scale_color_manual(values = growth_class_colors) +
  labs(
    y = "Predicted DOL",
    x = "Actual DOL",
    title = "Predicted age from microbiome abundances",
    subtitle = "Curve represents self-starting logistic model"
  )

ggsave(here("Output_sensitivity", "random forest prediction for normal growth.pdf"), height = 6, width = 5.4)

```

## Plot of all data

Finally, we can predict the microbiome age of babies with growth failure:

```{r}
p2 <- predict(rf_model, testing_data)

plot_df <- tibble(class = "Regular growth", MMI = as.vector(rf_model$predicted), actual_DOL = as.vector(rf_model$y), SampleID = names(rf_model$y)) %>%
  bind_rows(tibble(class = "Growth failure", MMI = as.vector(p2), actual_DOL = testing_data$DOL, SampleID = rownames(testing_data))) %>%
  mutate(class = fct_relevel(class, "Regular growth"))

growth_class_colors <- c("Regular growth" = "#446590FF", "Growth failure" = "#AD8152FF")

mod2 = nls(MMI ~ SSlogis(actual_DOL, Asym, xmid, scal), data=plot_df %>% filter(class %in% "Growth failure"))
# mod2

# mod2 = nls(MMI ~ SSfpl(actual_DOL, A, B, xmid, scal), data=plot_df %>% filter(class %in% "Growth failure"))
# weird error
# Error in nls(y ~ cbind(1, 1/(1 + exp((xmid - x)/exp(lscal)))), data = xy,  : 
#   step factor 0.000488281 reduced below 'minFactor' of 0.000976562
# mod2

for_save <- plot_df %>%
  ggplot(mapping = aes(x=actual_DOL, y=MMI, color=class)) +
  theme_bw() +
  geom_point() +
  annotate(geom="text", x=10, y=70, label=glue("Logistic curve for growth failure:\n Horiz. asymptote={round(mod2$m$getPars()['Asym'],2)}\n  xmid={round(mod2$m$getPars()['xmid'],2)}\n  scale={round(mod2$m$getPars()['scal'],2)}"), color = "#AD8152FF", hjust = 0) +
  geom_smooth(
    method = "nls", formula = y ~ SSlogis(x, Asym, xmid, scal),
    se = FALSE
  ) +
  coord_equal() +
  scale_color_manual(values = growth_class_colors) +
  labs(
    y = "Predicted DOL",
    x = "Actual DOL",
    title = "Predicted age from microbiome abundances",
    subtitle = "Curves represents self-starting logistic model"
  )

ggsave(plot = for_save, here("Output_sensitivity", "random forest prediction for both growth legend.pdf"), height = 6, width = 6)

for_save + theme(legend.position = "none")

ggsave(here("Output_sensitivity", "random forest prediction for both growth.pdf"), height = 6, width = 5.4)

```

## Predicted minus actual

```{r, fig.width=5, fig.height=3}

# mod1 is the SSlogis model for class %in% "Regular growth"
# mod2 is the SSlogis model for class %in% "Growth failure"

plot_df$curve_value <- predict(mod1, newdata = plot_df)

plot_df %<>%
  mutate(RMMI = MMI - curve_value)

plot_df_25taxa <- plot_df

plot_df %>%
  ggplot(aes(x = class, y = RMMI)) +
  geom_boxplot(outlier.alpha = 0) +
  geom_quasirandom(dodge.width = 0.75, alpha = 0.25)

```

### Mixed-effects model

```{r,results='asis'}

test_df <- plot_df %>%
  left_join(s_toTest %>% select(SampleID, SubjectID), join_by(SampleID))

tidy_lmer(nlme::lme(fixed = RMMI ~ class, random = ~ 1 | SubjectID, data = test_df)) %>% pander(digits=2, split.cells=15)

```

## ROC curves

```{r}

my_roc <- plot_df %>%
  mutate(class = fct_relevel(class, "Regular growth")) %>%
  roc(class ~ RMMI, data = .)

plot(my_roc)

# my_roc$specificities

with(my_roc, data_frame(sensitivities, specificities, thresholds)) %>%
  # mutate(thresholds = if_else(thresholds < 0.3, 0.3, thresholds)) %>%
  # mutate(thresholds = if_else(thresholds > 0.53, 0.55, thresholds)) %>%
  mutate(fpr = 1 - specificities, tpr = sensitivities) %>%
  select(thresholds, tpr, fpr) %>%
  gather(performance, value, -thresholds) %>%
  ggplot() +
  geom_step(aes(x=thresholds, y=value, color=performance)) +
  labs(title = "Prediction of growth class based on RMMI",
       y="False or True positive rate (fpr or tpr)",
       x="RMMI threshold for predicting growth class")

ggsave(filename = here::here("Output_sensitivity","TPFP for growth class 25 taxa.pdf"), device = "pdf")

```

### AUROC and Gini

Where AUROC is total area under the curve and Gini indexe is 2 X AUROC - 1. [See here for more](https://en.wikipedia.org/wiki/Receiver_operating_characteristic#Area_under_the_curve).

Gini is more intuitive as 1 is a perfect model while 0 is no better than chance.

```{r}

auroc <- auc(my_roc)

print(auroc)

gini <- 2 * auroc - 1

glue("Gini index is {round(gini,4)}")

```

## For validation set

```{r, fig.width=5, fig.height=3}

p3 <- predict(rf_model, validation_data)

plot_df <- tibble(class = "Training", MMI = as.vector(rf_model$predicted), actual_DOL = as.vector(rf_model$y), SampleID = names(rf_model$y)) %>%
  bind_rows(tibble(class = "Validation", MMI = as.vector(p3), actual_DOL = validation_data$DOL, SampleID = rownames(validation_data))) %>%
  mutate(class = fct_relevel(class, "Training"))

plot_df$curve_value <- predict(mod1, newdata = plot_df)

plot_df %<>%
  mutate(RMMI = MMI - curve_value)

plot_df %>%
  ggplot(aes(x = class, y = RMMI)) +
  geom_boxplot(outlier.alpha = 0) +
  geom_quasirandom(dodge.width = 0.75, alpha = 0.25)

```

### Mixed-effects model

```{r,results='asis'}

test_df <- plot_df %>%
  left_join(s_toTest %>% select(SampleID, SubjectID), join_by(SampleID))

tidy_lmer(nlme::lme(fixed = RMMI ~ class, random = ~ 1 | SubjectID, data = test_df)) %>% pander(digits=2, split.cells=15)

```

# Random forest feature trimming

```{r reselect}

min_frac_nonzero <- 0.75
mean_abund <- 0.01

top_taxa <- summed_props %>%
  as.data.frame() %>% 
  rownames_to_column("Taxa") %>% 
  pivot_longer(-Taxa, names_to="SampleID", values_to="props") %>%
  group_by(Taxa) %>%
  summarise(frac_non_zero = sum(props > 0) / length(props), mean_prop = mean(props)) %>%
  ungroup() %>%
  filter(frac_non_zero > min_frac_nonzero) %>%
  filter(!str_detect(Taxa, "Ralstonia|Cutibacterium|Vibrio|Bradyrhizobium|Rhizobiales"))

top50 <- summed_props_long %>%
  group_by(Taxa) %>%
  summarise(mean_props = mean(props)) %>%
  ungroup() %>%
  arrange(desc(mean_props)) %>%
  slice(1:54) %>%
  filter(!str_detect(Taxa, "Ralstonia|Cutibacterium|Vibrio|Bradyrhizobium|Rhizobiales"))

training_data <- s_toTest %>%
  filter(growth_failure %in% "No") %>%
  left_join(summed_props_long, by = "SampleID") %>%
  filter(Taxa %in% top_taxa$Taxa) %>%
  select(SampleID, DOL, Taxa, props) %>% #SubjectID, 
  mutate(Taxa = str_replace_all(Taxa, " ", "_")) %>%
  pivot_wider(names_from = "Taxa", values_from = "props", values_fill = 0) %>%
  column_to_rownames(var = "SampleID")

validation_data <- training_data[validation_samples,]
ind2 <- setdiff(rownames(training_data), rownames(validation_data))
training_data <- training_data[ind2,]

testing_data <- s_toTest %>%
  filter(growth_failure %in% "Yes") %>%
  left_join(summed_props_long, by = "SampleID") %>%
  filter(Taxa %in% top_taxa$Taxa) %>%
  select(SampleID, DOL, Taxa, props) %>% #SubjectID, 
  mutate(Taxa = str_replace_all(Taxa, " ", "_")) %>%
  pivot_wider(names_from = "Taxa", values_from = "props", values_fill = 0) %>%
  column_to_rownames(var = "SampleID")

#sanity check, make sure same baby is not in both validation and training
# samples_in_training <- rownames(training_data)
# training_babies <- s_toTest %>%
#   filter(SampleID %in% samples_in_training) %>%
#   pull(SubjectID)
# intersect(training_babies, validation_babies) #should be "character(0)"

```

We will use a filter of `r min_frac_nonzero` of samples being nonzero for a taxa leaving us with `r length(top_taxa$Taxa)` taxa.

## Boruta

Let's try Boruta which iteratively removes the features which are proved by a statistical test to be less relevant than random probes. Reference: Kursa, M. B., & Rudnicki, W. R. (2010). Feature selection with the Boruta package. J Stat Softw, 36(11), 1-13.

```{r}

# set.seed(42)
# boruta_model <- Boruta(DOL ~ ., data = training_data, doTrace = 2, maxRuns = 500)
# save(boruta_model, file = here("Output_sensitivity", "boruta_model_taxa_noSubjectID.RData"))
load(file = here("Output_sensitivity", "boruta_model_taxa_noSubjectID.RData"))

print(boruta_model)

```

```{r}

plot(boruta_model, las = 2, cex.axis = 0.7)

```

## Two random forests

We will do one random forest on the original `r length(top_taxa$Taxa)` variables and one on the `r length(getSelectedAttributes(boruta_model))` important variables as determined by Boruta.

```{r}

set.seed(42)
rf_model <- randomForest(DOL ~ ., data = training_data, localImp=T) #, strata = "SubjectID"
rf_model_improved <- randomForest(training_data[,getSelectedAttributes(boruta_model)], training_data$DOL, localImp=T) #, strata = "SubjectID"

print(rf_model)
print(rf_model_improved)

rf_model_improved_78taxa <- rf_model_improved

```

```{r eval=FALSE, include=FALSE}

library(randomForestExplainer)
explain_forest(rf_model_improved, interactions = F)

```

## Plot of training data 2

Next, we can plot the predicted age of the samples vs. the actual age:

```{r}

plot_df <- tibble(class = "Regular growth", MMI = as.vector(rf_model_improved$predicted), actual_DOL = as.vector(rf_model_improved$y))

growth_class_colors <- c("Regular growth" = "#446590FF")

# nls = nonlinear least squares
# SSlogis = Self-starting logistic model

mod1 = nls(MMI ~ SSlogis(actual_DOL, Asym, xmid, scal), data=plot_df)
# mod1

plot_df %>%
  ggplot(mapping = aes(x=actual_DOL, y=MMI, color=class)) +
  theme_bw() +
  theme(legend.position = "none") +
  geom_point() +
  geom_smooth(
    method = "nls", formula = y ~ SSlogis(x, Asym, xmid, scal),
    se = FALSE
  ) +
  annotate(geom="text", x=10, y=70, label=glue("Logistic curve for normal growth:\n Horiz. asymptote={round(mod1$m$getPars()['Asym'],2)}\n  xmid={round(mod1$m$getPars()['xmid'],2)}\n  scale={round(mod1$m$getPars()['scal'],2)}"), color = "#446590FF", hjust = 0) +
  coord_equal() +
  scale_color_manual(values = growth_class_colors) +
  labs(
    y = "Predicted DOL",
    x = "Actual DOL",
    title = "Predicted age from microbiome abundances",
    subtitle = "Curve represents self-starting logistic model"
  )

ggsave(here("Output_sensitivity", "random forest prediction for normal growth 2.pdf"), height = 6, width = 5.4)

```

## Plot of all data 2

Finally, we can predict the microbiome age of babies with growth failure:

```{r}
p2 <- predict(rf_model_improved, testing_data)

plot_df <- tibble(class = "Regular growth", MMI = as.vector(rf_model_improved$predicted), actual_DOL = as.vector(rf_model_improved$y), SampleID = rownames(training_data)) %>%
  bind_rows(tibble(class = "Growth failure", MMI = as.vector(p2), actual_DOL = testing_data$DOL, SampleID = rownames(testing_data))) %>%
  mutate(class = fct_relevel(class, "Regular growth"))

growth_class_colors <- c("Regular growth" = "#446590FF", "Growth failure" = "#AD8152FF")

mod2 = nls(MMI ~ SSlogis(actual_DOL, Asym, xmid, scal), data=plot_df %>% filter(class %in% "Growth failure"))
# mod2

# mod2 = nls(MMI ~ SSfpl(actual_DOL, A, B, xmid, scal), data=plot_df %>% filter(class %in% "Growth failure"))
# weird error
# Error in nls(y ~ cbind(1, 1/(1 + exp((xmid - x)/exp(lscal)))), data = xy,  : 
#   step factor 0.000488281 reduced below 'minFactor' of 0.000976562
# mod2

for_save <- plot_df %>%
  ggplot(mapping = aes(x=actual_DOL, y=MMI, color=class)) +
  theme_bw() +
  geom_point() +
  # annotate(geom="text", x=10, y=70, label=glue("Logistic curve for growth failure:\n Horiz. asymptote={round(mod2$m$getPars()['Asym'],2)}\n  xmid={round(mod2$m$getPars()['xmid'],2)}\n  scale={round(mod2$m$getPars()['scal'],2)}"), color = "#AD8152FF", hjust = 0) +
  geom_smooth(
    method = "nls", formula = y ~ SSlogis(x, Asym, xmid, scal),
    se = FALSE
  ) +
  scale_color_manual(values = growth_class_colors) +
  labs(
    y = "Predicted DOL",
    x = "Actual DOL",
    title = "",
    subtitle = ""
  )

ggsave(plot = for_save, here("Output_sensitivity", "random forest prediction for both growth legend 2.pdf"), height = 6, width = 6)

for_save + theme(legend.position = "none")

ggsave(here("Output_sensitivity", "random forest prediction for both growth 2.pdf"), height = 5.5, width = 10)

```

## Predicted minus actual

```{r, fig.width=5, fig.height=3}

# mod1 is the SSlogis model for class %in% "Regular growth"
# mod2 is the SSlogis model for class %in% "Growth failure"

plot_df$curve_value <- predict(mod1, newdata = plot_df)

plot_df %<>%
  mutate(RMMI = MMI - curve_value)

plot_df_78taxa <- plot_df

plot_df %>%
  ggplot(aes(x = class, y = RMMI)) +
  geom_boxplot(outlier.alpha = 0) +
  geom_quasirandom(dodge.width = 0.75, alpha = 0.25) +
  theme_bw() +
  labs(x = "Class", y = "Relative microbiota maturity index")

ggsave(here("Output_sensitivity", "RMMI_random_forest_feature_trimming_78_taxa.pdf"), height = 3, width = 3)

```

### Mixed-effects model

```{r, results='asis'}

test_df <- plot_df %>%
  left_join(s_toTest %>% select(SampleID, SubjectID), join_by(SampleID))

tidy_lmer(nlme::lme(fixed = RMMI ~ class, random = ~ 1 | SubjectID, data = test_df)) %>% pander(digits=2, split.cells=15)

```

## ROC curves

```{r 78 taxa roc}

my_roc <- plot_df %>%
  mutate(class = fct_relevel(class, "Regular growth")) %>%
  roc(class ~ RMMI, data = .)

ggroc(my_roc) +
  theme_bw() +
  labs(title = "ROC curve") +
  geom_abline(slope = 1, intercept = c(1,0), lty = 2)

ggsave(filename = here::here("Output_sensitivity","ROC curve for growth class 78 taxa.pdf"), device = "pdf", height = 3, width = 3)

with(my_roc, data_frame(sensitivities, specificities, thresholds)) %>%
  # mutate(thresholds = if_else(thresholds < 0.3, 0.3, thresholds)) %>%
  # mutate(thresholds = if_else(thresholds > 0.53, 0.55, thresholds)) %>%
  mutate(fpr = 1 - specificities, tpr = sensitivities) %>%
  select(thresholds, tpr, fpr) %>%
  gather(performance, value, -thresholds) %>%
  ggplot() +
  theme_bw() +
  geom_step(aes(x=thresholds, y=value, color=performance)) +
  labs(title = "",
       y="False or True positive rate (fpr or tpr)",
       x="RMMI threshold for predicting growth class")

ggsave(filename = here::here("Output_sensitivity","TPFP curve for growth class 78 taxa.pdf"), device = "pdf", height = 3, width = 4)

```

### AUROC and Gini

Where AUROC is total area under the curve and Gini indexe is 2 X AUROC - 1. [See here for more](https://en.wikipedia.org/wiki/Receiver_operating_characteristic#Area_under_the_curve).

Gini is more intuitive as 1 is a perfect model while 0 is no better than chance.

```{r}

auroc <- auc(my_roc)

print(auroc)

gini <- 2 * auroc - 1

glue("Gini index is {round(gini,4)}")

```

## For validation set

```{r, fig.width=5, fig.height=3}

p3 <- predict(rf_model_improved, validation_data)

plot_df <- plot_df_78taxa %>%
  bind_rows(tibble(class = "Validation", MMI = as.vector(p3), actual_DOL = validation_data$DOL, SampleID = rownames(validation_data))) %>%
  mutate(class = fct_relevel(class, "Regular growth", "Growth failure"))

plot_df$curve_value <- predict(mod1, newdata = plot_df)

plot_df %<>%
  mutate(RMMI = MMI - curve_value)

plot_df %>%
  ggplot(aes(x = class, y = RMMI)) +
  geom_boxplot(outlier.alpha = 0) +
  geom_quasirandom(dodge.width = 0.75, alpha = 0.25)

```

### Mixed-effects model

```{r, results='asis'}

test_df <- plot_df %>%
  left_join(s_toTest %>% select(SampleID, SubjectID), join_by(SampleID))

tidy_lmer(nlme::lme(fixed = RMMI ~ class, random = ~ 1 | SubjectID, data = test_df)) %>% pander(digits=2, split.cells=15)

```

# Random forest with a mix of variables

Since adding pathway data did not increase accuracy at all, we will try actual KEGG genes.

Here we will use the 78 taxa as before, the top 500 KEGG genes by mean abundance, log(`AbxHoursPerHOL`), and log(`npo_days_prop`) to predict microbiome maturity.

Values will have to be scaled to each measurements mean and standard deviation.

```{r scale and combine}

taxa_data <- summed_props_long %>%
  filter(Taxa %in% top_taxa$Taxa) %>%
  group_by(Taxa) %>%
  mutate(scaled_value = as.numeric(scale(props))) %>%
  ungroup()

top_kegg <- kegg_props %>%
  group_by(geneID) %>%
  summarise(frac_non_zero = sum(props > 0) / length(props), mean_prop = mean(props)) %>%
  arrange(desc(mean_prop)) %>%
  slice(1:500)

kegg_data <- kegg_props %>%
  filter(geneID %in% top_kegg$geneID) %>%
  group_by(geneID) %>%
  mutate(scaled_value = as.numeric(scale(props))) %>%
  ungroup() %>%
  select(SampleID, geneID, scaled_value) %>%
  pivot_wider(names_from = "geneID", values_from = "scaled_value", values_fill = 0)

clinical_data <- s_toTest %>%
  mutate(npo_days_prop_min = min(npo_days_prop[npo_days_prop > 0])) %>%
  mutate(npo_days_prop = ifelse(npo_days_prop == 0, npo_days_prop_min/2, npo_days_prop)) %>%
  mutate(log_npo_days_prop = log(npo_days_prop)) %>%
  mutate(AbxHoursPerHOL_min = min(AbxHoursPerHOL[AbxHoursPerHOL > 0])) %>%
  mutate(AbxHoursPerHOL = ifelse(AbxHoursPerHOL == 0, AbxHoursPerHOL_min/2, AbxHoursPerHOL)) %>%
  mutate(log_AbxHoursPerHOL = log(AbxHoursPerHOL)) %>%
  select(SampleID, log_AbxHoursPerHOL, log_npo_days_prop) %>%
  filter(!is.na(log_npo_days_prop))

clinical_data[,2:3] <- scale(clinical_data[,2:3])

training_data <- s_toTest %>%
  filter(growth_failure %in% "No") %>%
  left_join(taxa_data, by = "SampleID") %>%
  select(SampleID, DOL, Taxa, scaled_value) %>%
  mutate(Taxa = str_replace_all(Taxa, " ", "_")) %>%
  pivot_wider(names_from = "Taxa", values_from = "scaled_value", values_fill = 0)

validation_data <- training_data %>% filter(SampleID %in% validation_samples)
ind2 <- setdiff(training_data$SampleID, validation_data$SampleID)
training_data <- training_data %>% filter(SampleID %in% ind2)

expanded_training_data <- training_data %>%
  left_join(kegg_data) %>%
  inner_join(clinical_data) %>%
  column_to_rownames(var = "SampleID")
expanded_validation_data <- validation_data %>%
  left_join(kegg_data) %>%
  inner_join(clinical_data) %>%
  column_to_rownames(var = "SampleID")

testing_data <- s_toTest %>%
  filter(growth_failure %in% "Yes") %>%
  left_join(taxa_data, by = "SampleID") %>%
  select(SampleID, DOL, Taxa, scaled_value) %>%
  mutate(Taxa = str_replace_all(Taxa, " ", "_")) %>%
  pivot_wider(names_from = "Taxa", values_from = "scaled_value", values_fill = 0)

expanded_testing_data <- testing_data %>%
  left_join(kegg_data) %>%
  inner_join(clinical_data) %>%
  column_to_rownames(var = "SampleID")

```

## Boruta

We will use Boruta again:

```{r}

# set.seed(42)
# boruta_model <- Boruta(DOL ~ ., data = expanded_training_data, doTrace = 2, maxRuns = 500)
# save(boruta_model, file = here("Output_sensitivity", "boruta_model_taxa_and_more_noSubjectID.RData"))
load(file = here("Output_sensitivity", "boruta_model_taxa_and_more_noSubjectID.RData"))

print(boruta_model)

```

## The rf model

```{r}
set.seed(42)
rf_model <- randomForest(DOL ~ ., data = expanded_training_data, localImp=T)
rf_model_improved <- randomForest(expanded_training_data[,getSelectedAttributes(boruta_model)], expanded_training_data$DOL, localImp=T)

rf_model_combined <- rf_model_improved

print(rf_model)
print(rf_model_improved)
```

## Plot of training data taxa and kegg

Next, we can plot the predicted age of the samples vs. the actual age:

```{r}

plot_df <- tibble(class = "Regular growth", MMI = as.vector(rf_model_improved$predicted), actual_DOL = as.vector(rf_model_improved$y))

growth_class_colors <- c("Regular growth" = "#446590FF")

# nls = nonlinear least squares
# SSlogis = Self-starting logistic model

mod1 = nls(MMI ~ SSlogis(actual_DOL, Asym, xmid, scal), data=plot_df)
# mod1

plot_df %>%
  ggplot(mapping = aes(x=actual_DOL, y=MMI, color=class)) +
  theme_bw() +
  theme(legend.position = "none") +
  geom_point() +
  geom_smooth(
    method = "nls", formula = y ~ SSlogis(x, Asym, xmid, scal),
    se = FALSE
  ) +
  annotate(geom="text", x=10, y=70, label=glue("Logistic curve for normal growth:\n Horiz. asymptote={round(mod1$m$getPars()['Asym'],2)}\n  xmid={round(mod1$m$getPars()['xmid'],2)}\n  scale={round(mod1$m$getPars()['scal'],2)}"), color = "#446590FF", hjust = 0) +
  coord_equal() +
  scale_color_manual(values = growth_class_colors) +
  labs(
    y = "Predicted DOL",
    x = "Actual DOL",
    title = "Predicted age from microbiome abundances",
    subtitle = "Curve represents self-starting logistic model"
  )

ggsave(here("Output_sensitivity", "random forest prediction for normal growth taxa and kegg.pdf"), height = 6, width = 5.4)

```

## Plot of all data taxa and kegg

Finally, we can predict the microbiome age of babies with growth failure:

```{r}
p2 <- predict(rf_model_improved, expanded_testing_data)

plot_df <- tibble(class = "Regular growth", MMI = as.vector(rf_model_improved$predicted), actual_DOL = as.vector(rf_model_improved$y), SampleID = rownames(expanded_training_data)) %>%
  bind_rows(tibble(class = "Growth failure", MMI = as.vector(p2), actual_DOL = expanded_testing_data$DOL, SampleID = rownames(expanded_testing_data))) %>%
  mutate(class = fct_relevel(class, "Regular growth"))

growth_class_colors <- c("Regular growth" = "#446590FF", "Growth failure" = "#AD8152FF")

mod2 = nls(MMI ~ SSlogis(actual_DOL, Asym, xmid, scal), data=plot_df %>% filter(class %in% "Growth failure"))
# mod2

# mod2 = nls(MMI ~ SSfpl(actual_DOL, A, B, xmid, scal), data=plot_df %>% filter(class %in% "Growth failure"))
# weird error
# Error in nls(y ~ cbind(1, 1/(1 + exp((xmid - x)/exp(lscal)))), data = xy,  : 
#   step factor 0.000488281 reduced below 'minFactor' of 0.000976562
# mod2

for_save <- plot_df %>%
  ggplot(mapping = aes(x=actual_DOL, y=MMI, color=class)) +
  theme_bw() +
  geom_point() +
  annotate(geom="text", x=10, y=70, label=glue("Logistic curve for growth failure:\n Horiz. asymptote={round(mod2$m$getPars()['Asym'],2)}\n  xmid={round(mod2$m$getPars()['xmid'],2)}\n  scale={round(mod2$m$getPars()['scal'],2)}"), color = "#AD8152FF", hjust = 0) +
  geom_smooth(
    method = "nls", formula = y ~ SSlogis(x, Asym, xmid, scal),
    se = FALSE
  ) +
  coord_equal() +
  scale_color_manual(values = growth_class_colors) +
  labs(
    y = "Predicted DOL",
    x = "Actual DOL",
    title = "Predicted age from microbiome abundances",
    subtitle = "Curves represents self-starting logistic model"
  )

ggsave(plot = for_save, here("Output_sensitivity", "random forest prediction for both growth legend taxa and kegg.pdf"), height = 6, width = 6)

for_save + theme(legend.position = "none")

ggsave(here("Output_sensitivity", "random forest prediction for both growth taxa and kegg.pdf"), height = 6, width = 5.4)

no_formula <- plot_df %>%
  ggplot(mapping = aes(x=actual_DOL, y=MMI, color=class)) +
  theme_bw() +
  geom_point() +
  geom_smooth(
    method = "nls", formula = y ~ SSlogis(x, Asym, xmid, scal),
    se = FALSE
  ) +
  coord_equal() +
  scale_color_manual(values = growth_class_colors) +
  labs(
    y = "Predicted day of life",
    x = "Actual day of life",
    color = "Type of growth"
  )

ggsave(plot = no_formula, here("Output_sensitivity", "random forest prediction for both growth legend taxa and kegg no formula.pdf"), height = 6, width = 6)

```

## Predicted minus actual

```{r, fig.width=5, fig.height=3}

plot_df$curve_value <- predict(mod1, newdata = plot_df)

plot_df %<>%
  mutate(RMMI = MMI - curve_value)

plot_df %>%
  ggplot(aes(x = class, y = RMMI)) +
  geom_boxplot(outlier.alpha = 0) +
  geom_quasirandom(dodge.width = 0.75, alpha = 0.25) +
  labs(y = "Relative microbiome maturity index",
    x = "")

ggsave(filename = here("Output_sensitivity", "RMMI growth comparison all variables.pdf"), height = 3, width = 3)

```

### Mixed-effects model

```{r, results='asis'}

test_df <- plot_df %>%
  left_join(s_toTest %>% select(SampleID, SubjectID), join_by(SampleID))

tidy_lmer(nlme::lme(fixed = RMMI ~ class, random = ~ 1 | SubjectID, data = test_df)) %>% pander(digits=2, split.cells=15)

```

## ROC curves

```{r}

my_roc <- plot_df %>%
  mutate(class = fct_relevel(class, "Regular growth")) %>%
  roc(class ~ RMMI, data = .)

plot(my_roc)
# my_roc$specificities

with(my_roc, data_frame(sensitivities, specificities, thresholds)) %>%
  # mutate(thresholds = if_else(thresholds < 0.3, 0.3, thresholds)) %>%
  # mutate(thresholds = if_else(thresholds > 0.53, 0.55, thresholds)) %>%
  mutate(fpr = 1 - specificities, tpr = sensitivities) %>%
  select(thresholds, tpr, fpr) %>%
  gather(performance, value, -thresholds) %>%
  ggplot() +
  geom_step(aes(x=thresholds, y=value, color=performance)) +
  labs(title = "Prediction of growth class based on RMMI",
       y="False or True positive rate (fpr or tpr)",
       x="RMMI threshold for predicting growth class")

ggsave(filename = here::here("Output_sensitivity","TPFP curve for growth class mix of variables.pdf"), device = "pdf")

```

### AUROC and Gini

Where AUROC is total area under the curve and Gini indexe is 2 X AUROC - 1. [See here for more](https://en.wikipedia.org/wiki/Receiver_operating_characteristic#Area_under_the_curve).

Gini is more intuitive as 1 is a perfect model while 0 is no better than chance.

```{r}

auroc <- auc(my_roc)

print(auroc)

gini <- 2 * auroc - 1

glue("Gini index is {round(gini,4)}")

```

## For validation set

```{r, fig.width=5, fig.height=3}

p3 <- predict(rf_model_improved, expanded_validation_data)

plot_df <- tibble(class = "Training", MMI = as.vector(rf_model_improved$predicted), actual_DOL = as.vector(rf_model_improved$y), SampleID = rownames(expanded_training_data)) %>%
  bind_rows(tibble(class = "Validation", MMI = as.vector(p3), actual_DOL = expanded_validation_data$DOL, SampleID = rownames(expanded_validation_data))) %>%
  mutate(class = fct_relevel(class, "Training"))

plot_df$curve_value <- predict(mod1, newdata = plot_df)

plot_df %<>%
  mutate(RMMI = MMI - curve_value)

plot_df %>%
  ggplot(aes(x = class, y = RMMI)) +
  geom_boxplot(outlier.alpha = 0) +
  geom_quasirandom(dodge.width = 0.75, alpha = 0.25)

```

### Mixed-effects model

```{r, results='asis'}

test_df <- plot_df %>%
  left_join(s_toTest %>% select(SampleID, SubjectID), join_by(SampleID))

tidy_lmer(nlme::lme(fixed = RMMI ~ class, random = ~ 1 | SubjectID, data = test_df)) %>% pander(digits=2, split.cells=15)

```

## Important variables

```{r, fig.width=7, fig.height=5}

varImpPlot(rf_model_improved, type = 1)

dev.copy(pdf, here("Output_sensitivity", "importance_graph_kegg_taxa_clinical_var.pdf"))

```

```{r for Miren, eval=FALSE, include=FALSE}

s2 %>%
  select(SampleID, SubjectID, redcap_id, DOL) %>%
  write_csv(here("Output_sensitivity", "for_miren.csv"))

```

# Random forest with just KEGG

Here, we will just use the top 500 KEGG genes by mean abundance.

```{r}

top_kegg <- kegg_props %>%
  group_by(geneID) %>%
  summarise(frac_non_zero = sum(props > 0) / length(props), mean_prop = mean(props)) %>%
  arrange(desc(mean_prop)) %>%
  slice(1:500)

kegg_data <- kegg_props %>%
  filter(geneID %in% top_kegg$geneID)

training_data <- kegg_data %>%
  right_join(s_toTest %>% select(SampleID, growth_failure, DOL)) %>%
  filter(growth_failure %in% "No") %>%
  select(SampleID, geneID, props, DOL) %>%
  pivot_wider(names_from = "geneID", values_from = "props", values_fill = 0)

validation_data <- training_data %>% filter(SampleID %in% validation_samples) %>%
  column_to_rownames(var = "SampleID")
ind2 <- setdiff(training_data$SampleID, validation_data$SampleID)
training_data <- training_data %>% filter(SampleID %in% ind2) %>%
  column_to_rownames(var = "SampleID")

testing_data <- kegg_data %>%
  right_join(s_toTest %>% select(SampleID, growth_failure, DOL)) %>%
  filter(growth_failure %in% "Yes") %>%
  select(SampleID, geneID, props, DOL) %>%
  pivot_wider(names_from = "geneID", values_from = "props", values_fill = 0) %>%
  column_to_rownames(var = "SampleID")

```

## Boruta

We will use Boruta again:

```{r}

# set.seed(42)
# boruta_model <- Boruta(DOL ~ ., data = training_data, doTrace = 2, maxRuns = 500)
# save(boruta_model, file = here("Output_sensitivity", "boruta_model_just_KEGG_noSubjectID.RData"))
load(file = here("Output_sensitivity", "boruta_model_just_KEGG_noSubjectID.RData"))

print(boruta_model)

```

## The rf model

```{r}
set.seed(42)
rf_model <- randomForest(DOL ~ ., data = training_data, localImp=T)
rf_model_improved <- randomForest(training_data[,getSelectedAttributes(boruta_model)], training_data$DOL, localImp=T)

rf_model_just_kegg <- rf_model_improved

print(rf_model)
print(rf_model_improved)
```

## Plot of training data kegg

Next, we can plot the predicted age of the samples vs. the actual age:

```{r}

plot_df <- tibble(class = "Regular growth", MMI = as.vector(rf_model_improved$predicted), actual_DOL = as.vector(rf_model_improved$y))

growth_class_colors <- c("Regular growth" = "#446590FF")

# nls = nonlinear least squares
# SSlogis = Self-starting logistic model

mod1 = nls(MMI ~ SSlogis(actual_DOL, Asym, xmid, scal), data=plot_df)
# mod1

plot_df %>%
  ggplot(mapping = aes(x=actual_DOL, y=MMI, color=class)) +
  theme_bw() +
  theme(legend.position = "none") +
  geom_point() +
  geom_smooth(
    method = "nls", formula = y ~ SSlogis(x, Asym, xmid, scal),
    se = FALSE
  ) +
  annotate(geom="text", x=10, y=70, label=glue("Logistic curve for normal growth:\n Horiz. asymptote={round(mod1$m$getPars()['Asym'],2)}\n  xmid={round(mod1$m$getPars()['xmid'],2)}\n  scale={round(mod1$m$getPars()['scal'],2)}"), color = "#446590FF", hjust = 0) +
  coord_equal() +
  scale_color_manual(values = growth_class_colors) +
  labs(
    y = "Predicted DOL",
    x = "Actual DOL",
    title = "Predicted age from microbiome abundances",
    subtitle = "Curve represents self-starting logistic model"
  )

ggsave(here("Output_sensitivity", "random forest prediction for normal growth kegg.pdf"), height = 6, width = 5.4)

```

## Plot of all data kegg

Finally, we can predict the microbiome age of babies with growth failure:

```{r}
p2 <- predict(rf_model_improved, testing_data)

plot_df <- tibble(class = "Regular growth", MMI = as.vector(rf_model_improved$predicted), actual_DOL = as.vector(rf_model_improved$y), SampleID = rownames(training_data)) %>%
  bind_rows(tibble(class = "Growth failure", MMI = as.vector(p2), actual_DOL = testing_data$DOL, SampleID = rownames(testing_data))) %>%
  mutate(class = fct_relevel(class, "Regular growth"))

growth_class_colors <- c("Regular growth" = "#446590FF", "Growth failure" = "#AD8152FF")

mod2 = nls(MMI ~ SSlogis(actual_DOL, Asym, xmid, scal), data=plot_df %>% filter(class %in% "Growth failure"))
# mod2

# mod2 = nls(MMI ~ SSfpl(actual_DOL, A, B, xmid, scal), data=plot_df %>% filter(class %in% "Growth failure"))
# weird error
# Error in nls(y ~ cbind(1, 1/(1 + exp((xmid - x)/exp(lscal)))), data = xy,  : 
#   step factor 0.000488281 reduced below 'minFactor' of 0.000976562
# mod2

for_save <- plot_df %>%
  ggplot(mapping = aes(x=actual_DOL, y=MMI, color=class)) +
  theme_bw() +
  geom_point() +
  annotate(geom="text", x=10, y=70, label=glue("Logistic curve for growth failure:\n Horiz. asymptote={round(mod2$m$getPars()['Asym'],2)}\n  xmid={round(mod2$m$getPars()['xmid'],2)}\n  scale={round(mod2$m$getPars()['scal'],2)}"), color = "#AD8152FF", hjust = 0) +
  geom_smooth(
    method = "nls", formula = y ~ SSlogis(x, Asym, xmid, scal),
    se = FALSE
  ) +
  coord_equal() +
  scale_color_manual(values = growth_class_colors) +
  labs(
    y = "Predicted DOL",
    x = "Actual DOL",
    title = "Predicted age from microbiome abundances",
    subtitle = "Curves represents self-starting logistic model"
  )

ggsave(plot = for_save, here("Output_sensitivity", "random forest prediction for both growth legend kegg.pdf"), height = 6, width = 6)

for_save + theme(legend.position = "none")

ggsave(here("Output_sensitivity", "random forest prediction for both growth kegg.pdf"), height = 6, width = 5.4)

no_formula <- plot_df %>%
  ggplot(mapping = aes(x=actual_DOL, y=MMI, color=class)) +
  theme_bw() +
  geom_point() +
  geom_smooth(
    method = "nls", formula = y ~ SSlogis(x, Asym, xmid, scal),
    se = FALSE
  ) +
  coord_equal() +
  scale_color_manual(values = growth_class_colors) +
  labs(
    y = "Predicted day of life",
    x = "Actual day of life",
    color = "Type of growth"
  )

ggsave(plot = no_formula, here("Output_sensitivity", "random forest prediction for both growth legend kegg no formula.pdf"), height = 6, width = 6)

```

## Predicted minus actual

```{r, fig.width=5, fig.height=3}

plot_df$curve_value <- predict(mod1, newdata = plot_df)

plot_df %<>%
  mutate(RMMI = MMI - curve_value)

plot_df_kegg <- plot_df

plot_df %>%
  ggplot(aes(x = class, y = RMMI)) +
  geom_boxplot(outlier.alpha = 0) +
  geom_quasirandom(dodge.width = 0.75, alpha = 0.25) +
  labs(y = "Relative microbiome maturity index",
    x = "")

ggsave(filename = here("Output_sensitivity", "RMMI growth comparison KEGG.pdf"), height = 3, width = 3)

```

### Mixed-effects model

```{r, results='asis'}

test_df <- plot_df %>%
  left_join(s_toTest %>% select(SampleID, SubjectID), join_by(SampleID))

tidy_lmer(nlme::lme(fixed = RMMI ~ class, random = ~ 1 | SubjectID, data = test_df)) %>% pander(digits=2, split.cells=15)

```

## ROC curves

```{r}

my_roc <- plot_df %>%
  mutate(class = fct_relevel(class, "Regular growth")) %>%
  roc(class ~ RMMI, data = .)

plot(my_roc)
# my_roc$specificities

with(my_roc, data_frame(sensitivities, specificities, thresholds)) %>%
  # mutate(thresholds = if_else(thresholds < 0.3, 0.3, thresholds)) %>%
  # mutate(thresholds = if_else(thresholds > 0.53, 0.55, thresholds)) %>%
  mutate(fpr = 1 - specificities, tpr = sensitivities) %>%
  select(thresholds, tpr, fpr) %>%
  gather(performance, value, -thresholds) %>%
  ggplot() +
  geom_step(aes(x=thresholds, y=value, color=performance)) +
  labs(title = "Prediction of growth class based on RMMI",
       y="False or True positive rate (fpr or tpr)",
       x="RMMI threshold for predicting growth class")

ggsave(filename = here::here("Output_sensitivity","TPFP curve for growth class just KEGG.pdf"), device = "pdf")

```

### AUROC and Gini

Where AUROC is total area under the curve and Gini indexe is 2 X AUROC - 1. [See here for more](https://en.wikipedia.org/wiki/Receiver_operating_characteristic#Area_under_the_curve).

Gini is more intuitive as 1 is a perfect model while 0 is no better than chance.

```{r}

auroc <- auc(my_roc)

print(auroc)

gini <- 2 * auroc - 1

glue("Gini index is {round(gini,4)}")

```

## For validation set

```{r, fig.width=5, fig.height=3}

p3 <- predict(rf_model_improved, validation_data)

plot_df <- tibble(class = "Training", MMI = as.vector(rf_model_improved$predicted), actual_DOL = as.vector(rf_model_improved$y), SampleID = rownames(training_data)) %>%
  bind_rows(tibble(class = "Validation", MMI = as.vector(p3), actual_DOL = validation_data$DOL, SampleID = rownames(validation_data))) %>%
  mutate(class = fct_relevel(class, "Training"))

plot_df$curve_value <- predict(mod1, newdata = plot_df)

plot_df %<>%
  mutate(RMMI = MMI - curve_value)

plot_df %>%
  ggplot(aes(x = class, y = RMMI)) +
  geom_boxplot(outlier.alpha = 0) +
  geom_quasirandom(dodge.width = 0.75, alpha = 0.25)

```

### Mixed-effects model

```{r, results='asis'}

test_df <- plot_df %>%
  left_join(s_toTest %>% select(SampleID, SubjectID), join_by(SampleID))

tidy_lmer(nlme::lme(fixed = RMMI ~ class, random = ~ 1 | SubjectID, data = test_df)) %>% pander(digits=2, split.cells=15)

```

# End of report

```{r savepoint, include=FALSE}

save(rf_model_25taxa, rf_model_improved_78taxa, rf_model_combined, rf_model_just_kegg, plot_df_kegg, plot_df_25taxa, plot_df_78taxa, file = here("Output_sensitivity", "random_forest_models.RData"))

save.image(file = here("Output_sensitivity", "after_random_forest_analysis.RData"))

```


```{r Generate time stamped report, eval=FALSE, include=FALSE}
#notes
#neat: you can run this following command in the console to give your reports custom names (or date-stamp them)
#rmarkdown::render('ronen_stein_Report_shotgun.Rmd',output_file = paste('CEASE.report.', Sys.Date(), '.pdf', sep=''))

rstudioapi::documentSaveAll()

script_dir <- file.path("Scripts","Rmds", "sensitivity")
output_dir <- file.path("Output_sensitivity")

report_fp <- here::here(script_dir, "Shotgun_random_forest_analysis_gf10.Rmd")
output_fp <- here::here(output_dir, paste('sagori_microbiome_and_random_forests_SSlogis_gf10_', Sys.Date(), '.pdf', sep=''))

rmarkdown::render(report_fp, output_file = output_fp)

```
