---
title: |
  ![](CHOPMC_logo.pdf){width=5in}
  \break
  Sagori Mukhopadhyay - Runs 1 to 5 - Figures and tables
author: "Scott G. Daniel, PennCHOP Microbiome Program"
date: "`r Sys.Date()`"
geometry: margin=3cm
output: 
    pdf_document:
        keep_tex: false
        toc: true
        toc_depth: 4
---

```{r setup, echo=FALSE}
### ================
###   knitr setup
### ================
library(knitr)
opts_chunk$set(
  tidy=FALSE,
  cache=FALSE,
  echo=FALSE,
  warning=FALSE,
  message=FALSE,
  dpi=100,
  fig.width=4,
  fig.height=4,
  fig.align = "center"
  )
```

```{r libs and funcs, message=FALSE, warning=FALSE}

library(here)
library(tidyverse)
library(qiimer)
library(vegan)
library(ape)
library(usedist)

## Visualization packages
library(pander)
library(kableExtra)

library(pheatbuilder)
library(ggbeeswarm)
library(ggsci)
library(viridis)
library(RColorBrewer)

# stats packages
library(adonisplus)
library(nlme)
library(emmeans) # for lmer post-hoc tests
library(broom)
library(ghibli)
library(glue)
library(magrittr)
select <- dplyr::select
library(patchwork)
library(qualpalr)
library(randomForest)
library(microbenchmark)
library(Boruta)
library(mlbench)
library(caret)
library(pROC)
```

```{r load data}

load(file = here("Data","shotgun_data_2026-04-28.RData")) #source for raw data
load(file = here("Data","sagori_taxonomic_data.RData")) #source for fixed s2 table
load(file = here("Data","sagori_kegg_data.RData")) #source for gene abundance data
table_export <- table_iterator()

```

```{r log variables}

s2 %<>%
  mutate(npo_days_prop_min = min(npo_days_prop[npo_days_prop > 0])) %>%
  mutate(npo_days_prop = ifelse(npo_days_prop == 0, npo_days_prop_min/2, npo_days_prop)) %>%
  mutate(log_npo_days_prop = log(npo_days_prop)) %>%
  mutate(AbxHoursPerHOL_min = min(AbxHoursPerHOL[AbxHoursPerHOL > 0])) %>%
  mutate(AbxHoursPerHOL = ifelse(AbxHoursPerHOL == 0, AbxHoursPerHOL_min/2, AbxHoursPerHOL)) %>%
  mutate(log_AbxHoursPerHOL = log(AbxHoursPerHOL))

```

```{r moar colors}

ann_colors <- list(category = c(`1` = "#66C2A5", `2` = "#FC8D62", `3` = "#8DA0CB"
),
growth_cat = c(normal = "#92BBD9FF", `mild retardation` = "#DCCA2CFF", 
`moderate to severe retardation` = "#6FB382FF"),
TimePoint = c(T0 = "#440154FF", 
T1 = "#48186AFF", T2 = "#472D7BFF", T3 = "#424086FF", T4 = "#3B528BFF", 
T5 = "#33638DFF", T6 = "#2C728EFF", T7 = "#26828EFF", T8 = "#21908CFF", 
T9 = "#1F9F88FF", T10 = "#27AD81FF", T11 = "#3EBC74FF", T12 = "#5DC863FF", 
T13 = "#82D34DFF", T14 = "#AADC32FF", T15 = "#D5E21AFF", T16 = "#FDE725FF"
),
growth_fail = c(No = "#9FCB6A", Yes = "#8367CC"),
growth_label = c(None = "#92BBD9FF", Mild = "#DCCA2CFF", 
`Moderate\nto severe` = "#6FB382FF"))

```

```{r sra upload}

s2 %>% write_csv(here("all_the_samples.csv"))

s_sra <- s2 %>%
  filter(which_run %in% c("run4", "run5")) %>%
  rename(sample_name = SampleID, collection_date = date_collected) %>%
  mutate(collection_date = if_else(is.na(collection_date), "2019", collection_date)) %>%
  mutate(collection_date = str_remove(collection_date, ".+/")) %>%
  mutate(collection_date = if_else(str_length(collection_date) == 2, paste0("20", collection_date), collection_date)) %>%
  mutate(host = "Homo sapiens") %>%
  mutate(organism = "human gut metagenome") %>%
  mutate(env_broad_scale = "dense settlement biome", env_local_scale = "human-associated environment", env_medium = "fecal material", geo_loc_name = "USA", lat_lon = "39.9473 N 75.1958 W") %>%
  select(sample_name, organism, collection_date, env_broad_scale, env_local_scale, env_medium, geo_loc_name, host, lat_lon, WOL, host_sex = Sex, growth_failure, description = SampleType, log_npo_days_prop, log_AbxHoursPerHOL, growth_cat, bwt_z, PMA, DOL) %>%
  group_by(growth_cat) %>%
  mutate(replicate = paste0(growth_cat, row_number())) %>%
  ungroup()

s_sra %>%
  write_tsv(file = here("Data","sample_table_for_sra.tsv"))

sra_metadata <- read_csv(here("Data","sagori_run4_and_5_samples.csv"), col_names = c("sample_name","filename","filename2")) %>%
  mutate(filename = str_replace(filename, "/scr1/users/danielsg/sagori_run_[45]/data_files/", "")) %>%
  mutate(filename = str_replace(filename, "_R1", "_1")) %>%
  mutate(filename2 = str_replace(filename2, "/scr1/users/danielsg/sagori_run_[45]/data_files/", "")) %>%
  mutate(filename2 = str_replace(filename2, "_R2", "_2")) %>%
  mutate(library_ID = str_replace(filename, "_1.fastq.gz", "")) %>%
  right_join(s_sra, by = "sample_name") %>%
  mutate(title = paste("DNA sequencing of gut bacteria in babies: ",replicate)) %>%
  mutate(library_strategy = "WGS") %>%
  mutate(library_source = "METAGENOMIC") %>%
  mutate(library_selection = "RANDOM") %>%
  mutate(library_layout = "paired") %>%
  mutate(platform = "ILLUMINA") %>%
  mutate(instrument_model = "Illumina HiSeq 2500") %>%
  mutate(design_description = "Shotgun libraries were generated using the Nextera Flex Library Prep kit and sequenced on the Illumina HiSeq 2500 using 2x125 bp chemistry") %>%
  mutate(filetype = "fastq")

sra_metadata %>% select(sample_name, library_ID, title, library_strategy, library_source, library_selection, library_layout, platform, instrument_model, design_description, filetype, filename, filename2) %>%
  write_tsv(file = here("Data","sra_metadata.tsv"))

for_copying <- c(sra_metadata$filename, sra_metadata$filename2)
write_lines(for_copying, file = here("Data","file_list_for_sra_upload.txt"))

```

# FIGURE 1B - Growth velocity and birthweight

```{r}

# s2 %>%
#   write_csv(here("Output", "metadata_that_scott_has.csv"))

#fixing npo_days_prop for s2
# s2 %<>%
#     mutate(npo_days_n = ifelse(SampleID %in% "SRC.738128", 1, npo_days_n)) %>%
#   mutate(npo_days_n = ifelse(SampleID %in% "SRC.738559", 1, npo_days_n)) %>%
#   mutate(npo_days_prop = npo_days_n / DOL) %>%
#   mutate(npo_days_prop = ifelse(npo_days_prop == Inf, 1, npo_days_prop))

fig1b_data <- s2 %>%
  select(redcap_id, bwt_pct, GV_36wk, growth_cat, growth_failure) %>%
  unique() %>%
  mutate(growth_label = case_when(growth_cat %in% "normal" & growth_failure %in% "No" ~ "No failure, no faltering",
                                  growth_cat %in% "mild retardation" & growth_failure %in% "No" ~ "No failure, mild faltering",
                                  growth_cat %in% "moderate to severe retardation" & growth_failure %in% "No" ~ "No failure, sev faltering",
                                  growth_cat %in% "normal" & growth_failure %in% "Yes" ~ "Failure, no faltering",
                                  growth_cat %in% "mild retardation" & growth_failure %in% "Yes" ~ "Failure, mild faltering",
                                  growth_cat %in% "moderate to severe retardation" & growth_failure %in% "Yes" ~ "Failure, sev faltering")) %>%
  mutate(growth_label = fct_relevel(growth_label, "Failure, no faltering",
                                    "Failure, mild faltering",
                                    "Failure, sev faltering",
                                    "No failure, no faltering",
                                    "No failure, mild faltering",
                                    "No failure, sev faltering"))

growth_label_colors <- setNames(ghibli_palette(name = "PonyoMedium", n = 6, type = "discrete"), unique(fig1b_data$growth_label))

growth_label_colors <- c(`No failure, mild faltering` = "#4C413FFF", `Failure, sev faltering` = "#EEBCB1FF", 
`No failure, no faltering` = "#278B9AFF", `No failure, sev faltering` = "#B50A2AFF", 
`Failure, no faltering` = "#DE7862FF", `Failure, mild faltering` = "#D8AF39FF"
)

fig1b_data %>%
  ggplot(aes(x=bwt_pct, y=GV_36wk, colour = growth_label)) +
  geom_point() +
  scale_color_manual(values = growth_label_colors) +
  theme_bw() +
  labs(y = "Growth velocity at 36 weeks",
       x = "Birthweight percentile",
       color = "Growth failure and faltering")

ggsave(here("Output", "FIG1B_GVandbwtpct.pdf"), width = 5, height = 3, useDingbats = F)

```

# FIGURE 1C - Alpha diversity

```{r fig.width=8}

s2 %>%
  filter(!TimePoint %in% "T0") %>%
  pivot_longer(c("Richness", "Shannon"), names_to="metric", values_to="alpha") %>%
  filter(metric %in% "Shannon") %>%
  ggplot(aes(x=WOL_label, y=alpha, fill = WOL_label)) +
  #geom_boxplot(outlier.alpha=0) +
  geom_boxplot() + #outlier.alpha=0
  scale_fill_viridis_d(alpha = 0.8) +
  # facet_grid(metric~., scales = "free") +
  theme_bw() +
  theme(
    strip.background = element_blank(),
    panel.grid=element_blank(),
    legend.position = "none",
    axis.text.x = element_text(angle = 90, vjust = 0.5, hjust = 1)
  ) +
  labs(
    x="Week of life", color="",
    y="Shannon diversity"
  )

ggsave(here("Output", "FIG1C_alphadiv_no_T0.pdf"), width = 2.5, height = 2.5, useDingbats = F)

```

\newpage
## Linear model with just time

```{r, results='asis'}
my_formula <- "alpha ~ WOL"

summaries_df <- s2 %>%
  filter(!TimePoint %in% "T0") %>%
  pivot_longer(c("Richness", "Shannon"), names_to="metric", values_to="alpha") %>%
  group_by(metric) %>%
  do(tidy_lmer(nlme::lme(as.formula(my_formula), random=~1|SubjectID, data=., na.action=na.omit))) %>%
  ungroup() %>%
  filter(!grepl("Intercept", term)) %>%
  mutate(term = str_replace(term, "timepoint_numeric", "Time")) %>%
  mutate(term = if_else(str_detect(term, "growth_cat"), str_remove(term, "growth_cat"), term))

summaries_df %>%
  pander(digits = 2, split.table = Inf, split.cells = 20, caption = glue("Linear model of the alpha diversity measurement with numeric time as the predictive variable. SubjectID added as a random effect. Full formula: {my_formula}"))
```

## Linear model with log(npo_days_prop) and bwt_z

```{r, results='asis'}
my_formula <- "alpha ~ bwt_z + WOL + log_npo_days_prop"

summaries_df <- s2 %>%
  filter(!TimePoint %in% "T0") %>%
  pivot_longer(c("Richness", "Shannon"), names_to="metric", values_to="alpha") %>%
  group_by(metric) %>%
  do(tidy_lmer(nlme::lme(as.formula(my_formula), random=~1|SubjectID, data=., na.action=na.omit))) %>%
  ungroup() %>%
  filter(!grepl("Intercept", term)) %>%
  mutate(term = str_replace(term, "timepoint_numeric", "Time")) %>%
  mutate(term = if_else(str_detect(term, "growth_cat"), str_remove(term, "growth_cat"), term))

summaries_df %>%
  pander(digits = 2, split.table = Inf, split.cells = 20, caption = glue("Linear model of the alpha diversity measurement with numeric time and log_npo_days_prop as the predictive variables. SubjectID added as a random effect. Full formula: {my_formula}"))
```


## Linear model with log(AbxHoursPerHOL)

```{r, results='asis'}
my_formula <- "alpha ~ bwt_z + WOL + log_AbxHoursPerHOL"

summaries_df <- s2 %>%
  filter(!TimePoint %in% "T0") %>%
  pivot_longer(c("Richness", "Shannon"), names_to="metric", values_to="alpha") %>%
  group_by(metric) %>%
  do(tidy_lmer(nlme::lme(as.formula(my_formula), random=~1|SubjectID, data=., na.action=na.omit))) %>%
  ungroup() %>%
  filter(!grepl("Intercept", term)) %>%
  mutate(term = str_replace(term, "timepoint_numeric", "Time")) %>%
  mutate(term = if_else(str_detect(term, "growth_cat"), str_remove(term, "growth_cat"), term))

summaries_df %>%
  pander(digits = 2, split.table = Inf, split.cells = 20, caption = glue("Linear model of the alpha diversity measurement with numeric time and log_AbxHoursPerHOL as the predictive variables. SubjectID added as a random effect. Full formula: {my_formula}"))
```

# FIGURE 1D - Beta-div. over time


```{r fig.height=3, fig.width=5}
s2 %>%
  pcoaplus(bc) %>%
  plot(color=WOL) +
    scale_color_viridis_c() +
    theme_bw() +
    theme(
      axis.text=element_blank(),
      axis.ticks=element_blank(),
      panel.grid = element_blank(), 
      aspect.ratio = 1,
      strip.background = element_blank()
    ) + 
    labs(color="Week of life", title="Bray-Curtis")

ggsave(here("Output", "FIG1D_betadiv.pdf"), width = 5, height = 3, useDingbats = F)

```

## Permanova with just time

```{r, results='asis'}

# summaries_df <- 
#   s2 %>%
#     do(adonisplus(
#       ., distmat = bc, formula = distmat ~ WOL,
#       sample_id_var = SampleID, rep_meas_var = SubjectID,
#       shuffle = c(WOL = "within"), permutations=perm)) %>%
#   mutate(metric = "Bray-Curtis") %>%
#   filter(!term %in% c("Residuals", "Total")) %>%
#   select(-one_of(c("sumsq", "meansq"))) %>%
#   select(metric, everything())
# 
# write_rds(summaries_df, here("FigureData", "permanova_results_01.rds"))

summaries_df <- read_rds(here("FigureData", "permanova_results_01.rds"))

summaries_df %>%
  pander()

```

## Permanova with log_npo_days_prop

```{r, results='asis'}

# summaries_df <- 
#   s2 %>%
#     do(adonisplus(
#       ., distmat = bc, formula = distmat ~ bwt_z + log_npo_days_prop + WOL,
#       sample_id_var = SampleID, rep_meas_var = SubjectID,
#       shuffle = c(WOL = "within", log_npo_days_prop = "within"), permutations=perm)) %>%
#   mutate(metric = "Bray-Curtis") %>%
#   filter(!term %in% c("Residuals", "Total")) %>%
#   select(-one_of(c("sumsq", "meansq"))) %>%
#   select(metric, everything())
# 
# write_rds(summaries_df, here("FigureData", "permanova_results_02.rds"))

summaries_df <- read_rds(here("FigureData", "permanova_results_02.rds"))

summaries_df %>%
  pander()

```

## Permanova with log_AbxHoursPerHOL

```{r, results='asis'}

# summaries_df <- 
#   s2 %>%
#     do(adonisplus(
#       ., distmat = bc, formula = distmat ~ bwt_z + log_AbxHoursPerHOL + WOL,
#       sample_id_var = SampleID, rep_meas_var = SubjectID,
#       shuffle = c(WOL = "within", log_AbxHoursPerHOL = "within"), permutations=perm)) %>%
#   mutate(metric = "Bray-Curtis") %>%
#   filter(!term %in% c("Residuals", "Total")) %>%
#   select(-one_of(c("sumsq", "meansq"))) %>%
#   select(metric, everything())
# 
# write_rds(summaries_df, here("FigureData", "permanova_results_03.rds"))

summaries_df <- read_rds(here("FigureData", "permanova_results_03.rds"))

summaries_df %>%
  pander()

```

# FIGURE 1E

Taxa abundance panel - Naomi is doing this

# SUPP FIG 1A - alpha diversity growth faltering

```{r fig.width=6}

  # mutate(growth_label = case_when(growth_cat %in% "normal" ~ "Normal",
  #                                 growth_cat %in% "mild retardation" ~ "Mild",
  #                                 growth_cat %in% "moderate to severe retardation" ~ "Moderate to severe")) %>%

s2 %>%
  pivot_longer(c("Richness", "Shannon"), names_to="metric", values_to="alpha") %>%
  filter(metric %in% "Shannon") %>%
  mutate(growth_label = case_when(growth_cat %in% "normal" ~ "Normal",
                                  growth_cat %in% "mild retardation" ~ "Mild",
                                  growth_cat %in% "moderate to severe retardation" ~ "Moderate\nto severe")) %>%
  mutate(growth_label = fct_relevel(growth_label, "Normal", "Mild")) %>%
  ggplot(aes(x=growth_label, y=alpha, fill=WOL_label)) +
  geom_boxplot() + #outlier.alpha=0
  scale_fill_viridis_d(alpha = 0.8) +
  # facet_grid(metric~., scales = "free") +
  theme_bw() +
  theme(
    strip.background = element_blank(),
    panel.grid=element_blank(),
    legend.position = "none",
    axis.text.x = element_text(angle = 45, hjust = 1, vjust = 1)
  ) +
  labs(
    x="Growth faltering", color="Week of life",
    y="Shannon diversity"
  )

ggsave(here("Output", "SUPP_FIG1A_alphadiv_gfalt.pdf"), width = 3.5, height = 2.5, useDingbats = F)

```

## Linear model with just growth faltering

```{r, results='asis'}
my_formula <- "alpha ~ growth_cat"

summaries_df <- s2 %>%
  filter(!TimePoint %in% "T0") %>%
  pivot_longer(c("Richness", "Shannon"), names_to="metric", values_to="alpha") %>%
  group_by(metric) %>%
  do(tidy_lmer(nlme::lme(as.formula(my_formula), random=~1|SubjectID, data=., na.action=na.omit))) %>%
  ungroup() %>%
  filter(!grepl("Intercept", term)) %>%
  mutate(term = str_replace(term, "timepoint_numeric", "Time")) %>%
  mutate(term = if_else(str_detect(term, "growth_cat"), str_remove(term, "growth_cat"), term))

summaries_df %>%
  pander(digits = 2, split.table = Inf, split.cells = 20, caption = glue("Linear model of the alpha diversity measurement with growth_cat as the predictive variable. SubjectID added as a random effect. Full formula: {my_formula}"))
```

## Linear model with log_npo_days_prop

```{r, results='asis'}
my_formula <- "alpha ~ bwt_z + WOL * growth_cat + log_npo_days_prop"

summaries_df <- s2 %>%
  filter(!TimePoint %in% "T0") %>%
  pivot_longer(c("Richness", "Shannon"), names_to="metric", values_to="alpha") %>%
  group_by(metric) %>%
  do(tidy_lmer(nlme::lme(as.formula(my_formula), random=~1|SubjectID, data=., na.action=na.omit))) %>%
  ungroup() %>%
  filter(!grepl("Intercept", term)) %>%
  mutate(term = str_replace(term, "timepoint_numeric", "Time")) %>%
  mutate(term = if_else(str_detect(term, "growth_cat"), str_remove(term, "growth_cat"), term))

summaries_df %>%
  pander(digits = 2, split.table = Inf, split.cells = 20, caption = glue("Linear model of the alpha diversity measurement with numeric time and growth categories as the predictive variables. SubjectID added as a random effect. Full formula: {my_formula}"))
```


## Linear model with log_AbxHoursPerHOL

```{r, results='asis'}
my_formula <- "alpha ~ bwt_z + WOL * growth_cat + log_AbxHoursPerHOL"

summaries_df <- s2 %>%
  filter(!TimePoint %in% "T0") %>%
  pivot_longer(c("Richness", "Shannon"), names_to="metric", values_to="alpha") %>%
  group_by(metric) %>%
  do(tidy_lmer(nlme::lme(as.formula(my_formula), random=~1|SubjectID, data=., na.action=na.omit))) %>%
  ungroup() %>%
  filter(!grepl("Intercept", term)) %>%
  mutate(term = str_replace(term, "timepoint_numeric", "Time")) %>%
  mutate(term = if_else(str_detect(term, "growth_cat"), str_remove(term, "growth_cat"), term))

summaries_df %>%
  pander(digits = 2, split.table = Inf, split.cells = 20, caption = glue("Linear model of the alpha diversity measurement with numeric time and growth categories as the predictive variables. SubjectID added as a random effect. Full formula: {my_formula}"))
```


# SUPP FIG 1B - beta diversity growth faltering

```{r fig.height=3, fig.width=5}
s2 %>%
  mutate(growth_label = case_when(growth_cat %in% "normal" ~ "None",
                                  growth_cat %in% "mild retardation" ~ "Mild",
                                  growth_cat %in% "moderate to severe retardation" ~ "Moderate\nto severe")) %>%
  mutate(growth_label = fct_relevel(growth_label, "None", "Mild")) %>%
  pcoaplus(bc) %>%
  plot(color=growth_label) +
    scale_color_manual(values=ann_colors$growth_label) +
    theme_bw() +
    theme(
      axis.text=element_blank(),
      axis.ticks=element_blank(),
      panel.grid = element_blank(), 
      aspect.ratio = 1,
      strip.background = element_blank()
    ) + 
    labs(color="", title="")

ggsave(here("Output", "SUPP_FIG1B_betadiv.pdf"), width = 3.5, height = 2.5, useDingbats = F)

```

## Permanova with growth_cat and time

```{r, results='asis'}

# summaries_df <- 
#   s2 %>%
#     do(adonisplus(
#       ., distmat = bc, formula = distmat ~ growth_cat * WOL,
#       sample_id_var = SampleID, rep_meas_var = SubjectID,
#       shuffle = c(WOL = "within", growth_cat = "between"), permutations=perm)) %>%
#   mutate(metric = "Bray-Curtis") %>%
#   filter(!term %in% c("Residuals", "Total")) %>%
#   select(-one_of(c("sumsq", "meansq"))) %>%
#   select(metric, everything())
# 
# write_rds(summaries_df, here("FigureData", "permanova_results_04.rds"))

summaries_df <- read_rds(here("FigureData", "permanova_results_04.rds"))

summaries_df %>%
  pander()

```

## Permanova with log_npo_days_prop

```{r, results='asis'}

# summaries_df <- 
#   s2 %>%
#     do(adonisplus(
#       ., distmat = bc, formula = distmat ~ bwt_z + log_npo_days_prop + growth_cat * WOL,
#       sample_id_var = SampleID, rep_meas_var = SubjectID,
#       shuffle = c(WOL = "within", growth_cat = "between", log_npo_days_prop = "within"), permutations=perm)) %>%
#   mutate(metric = "Bray-Curtis") %>%
#   filter(!term %in% c("Residuals", "Total")) %>%
#   select(-one_of(c("sumsq", "meansq"))) %>%
#   select(metric, everything())
# 
# write_rds(summaries_df, here("FigureData", "permanova_results_05.rds"))

summaries_df <- read_rds(here("FigureData", "permanova_results_05.rds"))

summaries_df %>%
  pander()

```

## Permanova with log_AbxHoursPerHOL

```{r, results='asis'}

# summaries_df <- 
#   s2 %>%
#     do(adonisplus(
#       ., distmat = bc, formula = distmat ~ bwt_z + log_AbxHoursPerHOL + growth_cat * WOL,
#       sample_id_var = SampleID, rep_meas_var = SubjectID,
#       shuffle = c(WOL = "within", growth_cat = "between", log_AbxHoursPerHOL = "within"), permutations=perm)) %>%
#   mutate(metric = "Bray-Curtis") %>%
#   filter(!term %in% c("Residuals", "Total")) %>%
#   select(-one_of(c("sumsq", "meansq"))) %>%
#   select(metric, everything())
# 
# write_rds(summaries_df, here("FigureData", "permanova_results_06.rds"))

summaries_df <- read_rds(here("FigureData", "permanova_results_06.rds"))

summaries_df %>%
  pander()

```


# SUPP FIG 1C - alpha diversity growth failure

```{r fig.width=8}

s2 %>%
  pivot_longer(c("Richness", "Shannon"), names_to="metric", values_to="alpha") %>%
  filter(metric %in% "Shannon") %>%
  ggplot(aes(x=growth_failure, y=alpha, fill=WOL_label)) +
  geom_boxplot() + #outlier.alpha=0
  scale_fill_viridis_d(alpha = 0.8) +
  # facet_grid(metric~., scales = "free") +
  theme_bw() +
  theme(
    strip.background = element_blank(),
    panel.grid=element_blank(),
    legend.position = "none",
    axis.text.x = element_text(angle = 45, hjust = 1, vjust = 1)
  ) +
  labs(
    x="Growth failure", color="Week of life",
    y="Shannon diversity"
  )

ggsave(here("Output", "SUPP_FIG1C_alphadiv_gfail.pdf"), width = 3.5, height = 2.5, useDingbats = F)

```

## Linear model with just growth_failure

```{r, results='asis'}
my_formula <- "alpha ~ growth_failure"

summaries_df <- s2 %>%
  filter(!TimePoint %in% "T0") %>%
  pivot_longer(c("Richness", "Shannon"), names_to="metric", values_to="alpha") %>%
  group_by(metric) %>%
  do(tidy_lmer(nlme::lme(as.formula(my_formula), random=~1|SubjectID, data=., na.action=na.omit))) %>%
  ungroup() %>%
  filter(!grepl("Intercept", term)) %>%
  mutate(term = str_replace(term, "timepoint_numeric", "Time")) %>%
  mutate(term = if_else(str_detect(term, "growth_cat"), str_remove(term, "growth_cat"), term))

summaries_df %>%
  pander(digits = 2, split.table = Inf, split.cells = 20, caption = glue("Linear model of the alpha diversity measurement with growth_failure as the predictive variable. SubjectID added as a random effect. Full formula: {my_formula}"))
```

## Linear model with log_npo_days_prop

```{r, results='asis'}
my_formula <- "alpha ~ bwt_z + WOL * growth_failure + log_npo_days_prop"

summaries_df <- s2 %>%
  filter(!TimePoint %in% "T0") %>%
  pivot_longer(c("Richness", "Shannon"), names_to="metric", values_to="alpha") %>%
  group_by(metric) %>%
  do(tidy_lmer(nlme::lme(as.formula(my_formula), random=~1|SubjectID, data=., na.action=na.omit))) %>%
  ungroup() %>%
  filter(!grepl("Intercept", term)) %>%
  mutate(term = str_replace(term, "timepoint_numeric", "Time")) %>%
  mutate(term = if_else(str_detect(term, "growth_cat"), str_remove(term, "growth_cat"), term))

summaries_df %>%
  pander(digits = 2, split.table = Inf, split.cells = 20, caption = glue("Linear model of the alpha diversity measurement with numeric time and growth_failure as the predictive variables. SubjectID added as a random effect. Full formula: {my_formula}"))
```


## Linear model with log_AbxHoursPerHOL

```{r, results='asis'}
my_formula <- "alpha ~ bwt_z + WOL * growth_failure + log_AbxHoursPerHOL"

summaries_df <- s2 %>%
  filter(!TimePoint %in% "T0") %>%
  pivot_longer(c("Richness", "Shannon"), names_to="metric", values_to="alpha") %>%
  group_by(metric) %>%
  do(tidy_lmer(nlme::lme(as.formula(my_formula), random=~1|SubjectID, data=., na.action=na.omit))) %>%
  ungroup() %>%
  filter(!grepl("Intercept", term)) %>%
  mutate(term = str_replace(term, "timepoint_numeric", "Time")) %>%
  mutate(term = if_else(str_detect(term, "growth_cat"), str_remove(term, "growth_cat"), term))

summaries_df %>%
  pander(digits = 2, split.table = Inf, split.cells = 20, caption = glue("Linear model of the alpha diversity measurement with numeric time and growth_failure as the predictive variables. SubjectID added as a random effect. Full formula: {my_formula}"))
```



# SUPP FIG 1D - beta diversity growth failure

```{r fig.height=3, fig.width=5}
s2 %>%
  pcoaplus(bc) %>%
  plot(color=growth_failure) +
    scale_color_manual(values=ann_colors$growth_fail) +
    theme_bw() +
    theme(
      axis.text=element_blank(),
      axis.ticks=element_blank(),
      panel.grid = element_blank(), 
      aspect.ratio = 1,
      strip.background = element_blank()
    ) + 
    labs(color="", title="")

ggsave(here("Output", "SUPP_FIG1D_betadiv.pdf"), width = 3, height = 2.5, useDingbats = F)

```

## Permanova with growth_failure and time

```{r, results='asis'}

# summaries_df <- 
#   s2 %>%
#     do(adonisplus(
#       ., distmat = bc, formula = distmat ~ growth_failure * WOL,
#       sample_id_var = SampleID, rep_meas_var = SubjectID,
#       shuffle = c(WOL = "within", growth_failure = "between"), permutations=perm)) %>%
#   mutate(metric = "Bray-Curtis") %>%
#   filter(!term %in% c("Residuals", "Total")) %>%
#   select(-one_of(c("sumsq", "meansq"))) %>%
#   select(metric, everything())
# 
# write_rds(summaries_df, here("FigureData", "permanova_results_07.rds"))

summaries_df <- read_rds(here("FigureData", "permanova_results_07.rds"))

summaries_df %>%
  pander()

```

## Permanova with log_npo_days_prop

```{r, results='asis'}

# summaries_df <- 
#   s2 %>%
#     do(adonisplus(
#       ., distmat = bc, formula = distmat ~ bwt_z + log_npo_days_prop + growth_failure * WOL,
#       sample_id_var = SampleID, rep_meas_var = SubjectID,
#       shuffle = c(WOL = "within", growth_failure = "between", log_npo_days_prop = "within"), permutations=perm)) %>%
#   mutate(metric = "Bray-Curtis") %>%
#   filter(!term %in% c("Residuals", "Total")) %>%
#   select(-one_of(c("sumsq", "meansq"))) %>%
#   select(metric, everything())
# 
# write_rds(summaries_df, here("FigureData", "permanova_results_08.rds"))

summaries_df <- read_rds(here("FigureData", "permanova_results_08.rds"))

summaries_df %>%
  pander()

```

## Permanova with log_AbxHoursPerHOL

```{r, results='asis'}

# summaries_df <- 
#   s2 %>%
#     do(adonisplus(
#       ., distmat = bc, formula = distmat ~ bwt_z + log_AbxHoursPerHOL + growth_failure * WOL,
#       sample_id_var = SampleID, rep_meas_var = SubjectID,
#       shuffle = c(WOL = "within", growth_failure = "between", log_AbxHoursPerHOL = "within"), permutations=perm)) %>%
#   mutate(metric = "Bray-Curtis") %>%
#   filter(!term %in% c("Residuals", "Total")) %>%
#   select(-one_of(c("sumsq", "meansq"))) %>%
#   select(metric, everything())
# 
# write_rds(summaries_df, here("FigureData", "permanova_results_09.rds"))

summaries_df <- read_rds(here("FigureData", "permanova_results_09.rds"))

summaries_df %>%
  pander()

```

# SUPP FIGURE 4E - gene pathways abundance - growth faltering

### Linear models of pathways

```{r, results='asis'}

number_of_top_pathways = 25

props_toTest <- kegg_path_props %>%
  mutate(props_log10 = log10(props)) %>%
  mutate(PathwayName = fct_lump(PathwayName, n = number_of_top_pathways, w = .$props, other_level = "Other")) %>%
  filter(!PathwayName %in% c("Metabolic pathways", "Other"))

form1 = "props_log10 ~ bwt_z + WOL * growth_cat + log_npo_days_prop"
random1 = "SubjectID"

summaries_df <- run_lmer_on_props(factor_to_test = "PathwayName", props_toTest = props_toTest, s_toTest = s2, form1 = form1, rep_mes_label = random1, p_cutoff =  0.05) 

summaries_df %>%
  mutate(term = str_remove(term, "growth_cat")) %>%
  pander(split.table=Inf, digits=2, caption = glue("Linear models of log10 transformed abundances of KEGG pathways"), emphasize.strong.rows = which(.$fdr<=0.05))

```

```{r}

pathways_to_select <- summaries_df %>%
  mutate(term = str_remove(term, "growth_cat")) %>%
  filter(term %in% c("moderate to severe retardation", "mild retardation")) %>%
  pull(PathwayName) %>%
  unique() %>%
  as.character()

plot_df <- props_toTest %>%
  filter(PathwayName %in% pathways_to_select) %>%
  left_join(s2, join_by(SampleID))

g <- ggplot(plot_df, aes(x = props, y = PathwayName, color = growth_cat)) +
  theme_bw() +
  theme(legend.position = "bottom") +
  geom_boxplot(outlier.shape = 1, outlier.alpha = 0.5) +
  scale_color_manual(values=ann_colors$growth_cat) +
  labs(y = "KEGG pathway",
       x = "Relative abundance",
       color = "")

plot(g)

ggsave(here("Output", glue("SUPP_FIGURE_4E_growth_falt_kegg_pathway_abundance.pdf")), height = 2.75, width = 5)

```

# FIGURE 2B - gene pathways linear estimate of differences

```{r}

summaries_df <- run_lmer_on_props(factor_to_test = "PathwayName", props_toTest = props_toTest, s_toTest = s2, form1 = form1,rep_mes_label = random1, p_cutoff =  1)

plot_df <- summaries_df %>%
  mutate(term = str_remove(term, "growth_cat")) %>%
  filter(term %in% c("moderate to severe retardation", "mild retardation"), PathwayName %in% pathways_to_select) %>%
  mutate(CI95_lwr = round(Value - 1.96*Std.Error, 2)) %>%
  mutate(CI95_upr = round(Value + 1.96*Std.Error, 2)) %>%
  mutate(Significant = case_when(fdr < 0.05 ~ "p.adj < 0.05",
                                 fdr > 0.05 ~ "Not significant")) %>%
  arrange(Value) %>%
  mutate(PathwayName = fct_reorder(PathwayName, -Value))

g <- ggplot(plot_df, aes(Value, PathwayName, shape = Significant, color = term)) +
  theme_bw() +
  theme(legend.position = "bottom", legend.justification.bottom = "left", legend.direction = "vertical", legend.byrow = T) +
  geom_pointrange(aes(xmin=Value-Std.Error, xmax=Value+Std.Error), position = position_dodge(width = 0.4)) +
  geom_vline(xintercept = 0, linetype = "longdash") +
  scale_shape_manual(values = c(1,19)) +
  scale_color_manual(values=ann_colors$growth_cat[2:3]) +
  labs(y = "KEGG Pathway",
       x = "",
       shape = "Significance",
       color = "Growth faltering")
# <-- Decreased vs. normal growth    Increased vs. normal growth -->
plot(g)

ggsave(here("Output", glue("FIGURE_2B_growth_falt_kegg_pathway_difference.pdf")), height = 2.5, width = 5)

pathways_from_fig_2B <- as.character(unique(plot_df$PathwayName))

```

# SUPP FIGURE 4F - gene pathways abundance - growth failure

### Linear models of pathways

```{r, results='asis'}

number_of_top_pathways = 25

props_toTest <- kegg_path_props %>%
  mutate(props_log10 = log10(props)) %>%
  mutate(PathwayName = fct_lump(PathwayName, n = number_of_top_pathways, w = .$props, other_level = "Other")) %>%
  filter(!PathwayName %in% c("Metabolic pathways", "Other"))

form1 = "props_log10 ~ bwt_z + WOL * growth_failure + log_npo_days_prop"
random1 = "SubjectID"

summaries_df <- run_lmer_on_props(factor_to_test = "PathwayName", props_toTest = props_toTest, s_toTest = s2, form1 = form1, rep_mes_label = random1, p_cutoff =  0.05) 

summaries_df %>%
  mutate(term = str_remove(term, "growth_failure")) %>%
  pander(split.table=Inf, digits=2, caption = glue("Linear models of log10 transformed abundances of KEGG pathways"), emphasize.strong.rows = which(.$fdr<=0.05))

```

```{r}

pathways_to_select <- pathways_from_fig_2B

plot_df <- props_toTest %>%
  filter(PathwayName %in% pathways_to_select) %>%
  left_join(s2, join_by(SampleID))

g <- ggplot(plot_df, aes(x = props, y = PathwayName, color = growth_failure)) +
  theme_bw() +
  theme(legend.position = "bottom") +
  geom_boxplot(outlier.shape = 1, outlier.alpha = 0.5) +
  scale_color_manual(values=ann_colors$growth_fail) +
  labs(y = "KEGG pathway",
       x = "Relative abundance",
       color = "")

plot(g)

ggsave(here("Output", glue("SUPP_FIGURE_4F_growth_failure_kegg_pathway_abundance.pdf")), height = 2.75, width = 5)

```

# FIGURE 2C - gene pathways linear estimate of differences

```{r}

summaries_df <- run_lmer_on_props(factor_to_test = "PathwayName", props_toTest = props_toTest, s_toTest = s2, form1 = form1,rep_mes_label = random1, p_cutoff =  1)

plot_df <- summaries_df %>%
  mutate(term = str_remove(term, "growth_failure")) %>%
  filter(term %in% c("Yes")) %>%
  filter(PathwayName %in% pathways_to_select) %>%
  mutate(CI95_lwr = round(Value - 1.96*Std.Error, 2)) %>%
  mutate(CI95_upr = round(Value + 1.96*Std.Error, 2)) %>%
  mutate(Significant = case_when(fdr < 0.05 ~ "p.adj < 0.05",
                                 fdr > 0.05 ~ "Not significant")) %>%
  arrange(Value) %>%
  mutate(PathwayName = fct_reorder(PathwayName, -Value))

g <- ggplot(plot_df, aes(Value, PathwayName, shape = Significant, color = term)) +
  theme_bw() +
  theme(legend.position = "bottom", legend.justification.bottom = "left", legend.direction = "vertical", legend.byrow = T) +
  geom_pointrange(aes(xmin=Value-Std.Error, xmax=Value+Std.Error), position = position_dodge(width = 0.4)) +
  geom_vline(xintercept = 0, linetype = "longdash") +
  scale_shape_manual(values = c(1,19)) +
  scale_color_manual(values=ann_colors$growth_fail) +
  labs(y = "KEGG Pathway",
       x = "Change vs. normal growth",
       shape = "Significance",
       color = "Growth failure")
# <-- Decreased vs. normal growth    Increased vs. normal growth -->
plot(g)

ggsave(here("Output", glue("FIGURE_2C_growth_failure_kegg_pathway_difference.pdf")), height = 2.5, width = 5)

```

# FIGURE 2A - just time and top 500 KEGG

Just with time variable

Taking the top 500 KEGG orthologs and testing them

```{r message=FALSE, warning=FALSE, paged.print=TRUE, results='asis'}

kegg_defnitions <- read_tsv(here("Data", "ko_definition.list"), col_names = c("Ortholog", "Definition")) %>%
  mutate(Ortholog = str_remove(Ortholog, "ko:"))

path_annotation <- prokarotic_paths %>%
  select(Ortholog, PathwayName) %>%
  group_by(Ortholog) %>%
  summarise(Pathways = paste0(PathwayName, collapse = ", ")) %>%
  ungroup()

top500 <- kegg_props %>%
  filter(SampleID %in% s2$SampleID) %>%
  group_by(geneID) %>%
  summarise(mean_props = mean(props)) %>%
  arrange(desc(mean_props)) %>%
  slice(1:500)

kegg_props_subset <- kegg_props %>%
  select(SampleID, Ortholog = geneID, count, props) %>%
  filter(Ortholog %in% top500$geneID) %>%
  mutate(props_log10 = log10(props))

form1 = "props_log10 ~ WOL"
random1 = "SubjectID"

# summaries_df <- run_lmer_on_props(factor_to_test = "Ortholog", props_toTest = kegg_props_subset, s_toTest = s2, form1 = form1, p_cutoff = 1, rep_mes_label = random1)
# 
# save(summaries_df, file = here("Data", glue("just_time_kegg500_changes.RDS")))

load(file = here("Data", glue("just_time_kegg500_changes.RDS")))
  
plot_df <- summaries_df %>%
  mutate(Significant = case_when(fdr < 0.05 ~ "p.adj < 0.05",
                                 fdr > 0.05 ~ "Not significant")) %>%
  mutate(direction = case_when(Value > 0 ~ "Increased",
                               Value < 0 ~ "Decreased")) %>%
  left_join(kegg_defnitions, by = "Ortholog") %>%
  tidyr::separate_wider_delim(cols = Definition, delim = ";", names = c("gene", "description")) %>%
  mutate(ortho_lable = paste(Ortholog, gene))

library(ggrepel)

g <- ggplot(data=plot_df, aes(x=Value, y=-log10(fdr), col=direction, label = ortho_lable)) + 
  geom_point() +
  # geom_text_repel() +
  theme_bw() +
  geom_hline(yintercept=-log10(0.05), linetype = "longdash") +
  scale_shape_manual(values = c(1,19)) +
  # theme(legend.position = "none") +
  labs(y = "-log10(fdr value)",
       x = "Estimate of linear model",
       color = "Change over time",
       subtitle = "")
#  (formula: log10(props) ~ WOL)
#Dashed line represents fdr = 0.05

plot(g)

ggsave(here("Output", glue("FIGURE_2A_top_500_kegg_orthologs_just_time.pdf")), height = 2.25, width = 4)

table_name <- "FIGURE_2A_top_500_kegg_orthologs_just_time.csv"

my_table <- summaries_df %>%
  left_join(kegg_defnitions, by = "Ortholog") %>%
  left_join(path_annotation, by = "Ortholog")

write_csv(my_table, here("Output", table_name))

print(glue("All orthologs saved together as {table_name}"))

```

# SUPP FIGURE 4C - Gfalt and Kegg terms

FDR plot for growth faltering

## With log_npo_days_prop

```{r message=FALSE, warning=FALSE, paged.print=TRUE, results='asis'}

form1 = "props_log10 ~ bwt_z + WOL * growth_cat + log_npo_days_prop"
random1 = "SubjectID"

summaries_df <- run_lmer_on_props(factor_to_test = "Ortholog", props_toTest = kegg_props_subset, s_toTest = s2, form1 = form1, p_cutoff = 1, rep_mes_label = random1)

to_select <- summaries_df %>%
  mutate(term = str_remove(term, "growth_cat")) %>%
  filter(term %in% c("moderate to severe retardation", "mild retardation"), fdr < 0.05) %>%
  pull(Ortholog) %>%
  as.character() %>%
  unique()

changed_in_gfalt <- to_select
  
plot_df <- summaries_df %>%
  mutate(term = str_remove(term, "growth_cat")) %>%
  filter(term %in% c("moderate to severe retardation", "mild retardation"), Ortholog %in% to_select) %>%
  left_join(kegg_defnitions, by = "Ortholog") %>%
  mutate(ortho_lable = paste(Ortholog, str_trunc(Definition, 30, "right"))) %>%
  mutate(CI95_lwr = round(Value - 1.96*Std.Error, 2)) %>%
  mutate(CI95_upr = round(Value + 1.96*Std.Error, 2)) %>%
  mutate(Significant = case_when(fdr < 0.05 ~ "p.adj < 0.05",
                                 fdr > 0.05 ~ "Not significant")) %>%
  arrange(Value) %>%
  mutate(ortho_lable = fct_reorder(ortho_lable, Value))

g <- ggplot(plot_df, aes(Value, ortho_lable, color = term, shape = Significant)) +
  theme_bw() +
  geom_pointrange(aes(xmin=Value-Std.Error, xmax=Value+Std.Error), position = position_dodge(width = 0.4)) +
  geom_vline(xintercept = 0, linetype = "longdash") +
  scale_shape_manual(values = c(1,19)) +
  scale_color_manual(values=ann_colors$growth_cat[2:3]) +
    labs(y = "Ortholog / Description",
       x = "<-- Decreased vs. normal growth    Increased vs. normal growth -->",
       shape = "Significance",
       color = "")

plot(g)

ggsave(here("Output", glue("SUPP_FIGURE_4C_growth_faltering_kegg500.pdf")), height = 2.5, width = 6.5)

table_name <- glue("SUPP_FIGURE_4C_growth_faltering_kegg500.csv")

my_table <- summaries_df %>%
  mutate(term = str_remove(term, "growth_cat")) %>%
  filter(term %in% c("moderate to severe retardation", "mild retardation"), fdr < 0.05) %>%
  left_join(kegg_defnitions, by = "Ortholog") %>%
  left_join(path_annotation, by = "Ortholog")

write_csv(my_table, here("Output", table_name))

print(glue("All orthologs saved together as {table_name}"))

```




# SUPP FIGURE 4D - Gfailure and Kegg terms

FDR plot for growth failure

## With log_npo_days_prop

```{r message=FALSE, warning=FALSE, paged.print=TRUE, results='asis'}

form1 = "props_log10 ~ bwt_z + WOL * growth_failure + log_npo_days_prop"
random1 = "SubjectID"

summaries_df <- run_lmer_on_props(factor_to_test = "Ortholog", props_toTest = kegg_props_subset, s_toTest = s2, form1 = form1, p_cutoff = 1, rep_mes_label = random1)
  
plot_df <- summaries_df %>%
  mutate(term = str_remove(term, "growth_failure")) %>%
  filter(term %in% c("Yes"), Ortholog %in% c(changed_in_gfalt, "K00868")) %>%
  left_join(kegg_defnitions, by = "Ortholog") %>%
  mutate(ortho_lable = paste(Ortholog, str_trunc(Definition, 30, "right"))) %>%
  mutate(CI95_lwr = round(Value - 1.96*Std.Error, 2)) %>%
  mutate(CI95_upr = round(Value + 1.96*Std.Error, 2)) %>%
  mutate(Significant = case_when(fdr < 0.05 ~ "p.adj < 0.05",
                                 fdr > 0.05 ~ "Not significant")) %>%
  arrange(Value) %>%
  mutate(ortho_lable = fct_reorder(ortho_lable, Value))

g <- ggplot(plot_df, aes(Value, ortho_lable, color = term, shape = Significant)) +
  theme_bw() +
  geom_pointrange(aes(xmin=Value-Std.Error, xmax=Value+Std.Error), position = position_dodge(width = 0.4)) +
  geom_vline(xintercept = 0, linetype = "longdash") +
  scale_shape_manual(values = c(1,19)) +
  scale_color_manual(values=ann_colors$growth_fail) +
    labs(y = "Ortholog / Description",
       x = "<-- Decreased vs. normal growth    Increased vs. normal growth -->",
       shape = "Significance",
       color = "")

plot(g)

ggsave(here("Output", glue("SUPP_FIGURE_4D_growth_failure_kegg500.pdf")), height = 2.5, width = 6.5)

table_name <- glue("SUPP_FIGURE_4D_growth_failure_kegg500.csv")

my_table <- summaries_df %>%
  mutate(term = str_remove(term, "growth_failure")) %>%
  filter(term %in% c("Yes"), Ortholog %in% changed_in_gfalt) %>%
  left_join(kegg_defnitions, by = "Ortholog") %>%
  left_join(path_annotation, by = "Ortholog")

write_csv(my_table, here("Output", table_name))

print(glue("All orthologs saved together as {table_name}"))

```

```{r}

#all the orthologs
actually_changed <- summaries_df %>%
  filter(fdr < 0.05, term %in% "growth_failureYes")
#K00868

```

# SUPP FIGURE 4A - venn diagram of kegg genes in growth faltering

```{r}

library(ggVennDiagram)

kegg_pres_abs <- kegg_props_raw %>%
  left_join(s2 %>% select(SampleID, growth_cat), join_by(SampleID)) %>%
  group_by(growth_cat, geneID) %>%
  summarise(present = ifelse(sum(count) > 0, "Present", "Absent"))

venn_list <- list("Normal" = kegg_pres_abs %>% filter(growth_cat %in% "normal", present %in% "Present") %>% pull(geneID),
                "Mild retardation" = kegg_pres_abs %>% filter(growth_cat %in% "mild retardation", present %in% "Present") %>% pull(geneID),
                "Moderate to severe retardation" = kegg_pres_abs %>% filter(growth_cat %in% "moderate to severe retardation", present %in% "Present") %>% pull(geneID))

ggVennDiagram(venn_list) +
  scale_x_continuous(expand = expansion(mult = .2)) +
  scale_fill_distiller(palette = "RdBu")

ggsave(here("Output", "SUPP FIGURE 4A kegg genes shared among growth faltering.pdf"), height = 5, width = 6)

```

# SUPP FIGURE 4B - venn diagram of kegg genes in growth failure

```{r}

kegg_pres_abs <- kegg_props_raw %>%
  left_join(s2 %>% select(SampleID, growth_failure), join_by(SampleID)) %>%
  group_by(growth_failure, geneID) %>%
  summarise(present = ifelse(sum(count) > 0, "Present", "Absent"))

venn_list <- list("Yes" = kegg_pres_abs %>% filter(growth_failure %in% "Yes", present %in% "Present") %>% pull(geneID),
                "No" = kegg_pres_abs %>% filter(growth_failure %in% "No", present %in% "Present") %>% pull(geneID))

ggVennDiagram(venn_list) +
  scale_x_continuous(expand = expansion(mult = .2)) +
  scale_fill_distiller(palette = "RdBu")

ggsave(here("Output", "SUPP FIGURE 4B kegg genes shared among growth failure.pdf"), height = 5, width = 6)

```

# Random Forests

```{r load stuff}
library(Boruta)
library(mlbench)
library(caret)
library(pROC)

load(file = here("Data", "random_forest_models.RData"))

```

```{r}

top25 <- summed_props_long %>%
  group_by(Taxa) %>%
  summarise(mean_props = mean(props)) %>%
  ungroup() %>%
  arrange(desc(mean_props)) %>%
  slice(1:29) %>%
  filter(!str_detect(Taxa, "Ralstonia|Cutibacterium|Vibrio"))

training_data <- s2 %>%
  filter(growth_failure %in% "No") %>%
  left_join(summed_props_long, by = "SampleID") %>%
  filter(Taxa %in% top25$Taxa) %>%
  select(SampleID, DOL, Taxa, props) %>% #SubjectID, 
  mutate(Taxa = str_replace_all(Taxa, " ", "_")) %>%
  pivot_wider(names_from = "Taxa", values_from = "props", values_fill = 0) %>%
  column_to_rownames(var = "SampleID")

set.seed(43)
validation_babies <- sample(s2$SubjectID, 15)
validation_samples <- s2 %>%
  filter(growth_failure %in% "No") %>%
  filter(SubjectID %in% validation_babies) %>%
  pull(SampleID)

validation_data <- training_data[validation_samples,]
ind2 <- setdiff(rownames(training_data), rownames(validation_data))
training_data <- training_data[ind2,]

testing_data <- s2 %>%
  filter(growth_failure %in% "Yes") %>%
  left_join(summed_props_long, by = "SampleID") %>%
  filter(Taxa %in% top25$Taxa) %>%
  select(SampleID, DOL, Taxa, props) %>% #SubjectID, 
  mutate(Taxa = str_replace_all(Taxa, " ", "_")) %>%
  pivot_wider(names_from = "Taxa", values_from = "props", values_fill = 0) %>%
  column_to_rownames(var = "SampleID")

#sanity check, make sure same baby is not in both validation and training
# samples_in_training <- rownames(training_data)
# training_babies <- s2 %>%
#   filter(SampleID %in% samples_in_training) %>%
#   pull(SubjectID)
# intersect(training_babies, validation_babies) #should be "character(0)"

```

# FIGURE 3A - rf model with 25 taxa

```{r}

rf_model_improved <- rf_model_25taxa

p2 <- predict(rf_model_improved, testing_data)

plot_df <- tibble(class = "Regular growth", MMI = as.vector(rf_model_improved$predicted), actual_DOL = as.vector(rf_model_improved$y), SampleID = rownames(training_data)) %>%
  bind_rows(tibble(class = "Growth failure", MMI = as.vector(p2), actual_DOL = testing_data$DOL, SampleID = rownames(testing_data))) %>%
  mutate(class = fct_relevel(class, "Regular growth"))

growth_class_colors <- c("Regular growth" = "#9FCB6A", "Growth failure" = "#8367CC")

# mod2 = nls(MMI ~ SSlogis(actual_DOL, Asym, xmid, scal), data=plot_df %>% filter(class %in% "Growth failure"))
# mod2

# mod2 = nls(MMI ~ SSfpl(actual_DOL, A, B, xmid, scal), data=plot_df %>% filter(class %in% "Growth failure"))
# weird error
# Error in nls(y ~ cbind(1, 1/(1 + exp((xmid - x)/exp(lscal)))), data = xy,  : 
#   step factor 0.000488281 reduced below 'minFactor' of 0.000976562
# mod2

for_save <- plot_df %>%
  ggplot(mapping = aes(x=actual_DOL, y=MMI, color=class)) +
  theme_bw() +
  geom_point() +
  # annotate(geom="text", x=10, y=70, label=glue("Logistic curve for growth failure:\n Horiz. asymptote={round(mod2$m$getPars()['Asym'],2)}\n  xmid={round(mod2$m$getPars()['xmid'],2)}\n  scale={round(mod2$m$getPars()['scal'],2)}"), color = "#AD8152FF", hjust = 0) +
  geom_smooth(
    method = "nls", formula = y ~ SSlogis(x, Asym, xmid, scal),
    se = FALSE
  ) +
  scale_color_manual(values = growth_class_colors) +
  labs(
    y = "Predicted DOL / microbiome maturity index (MMI)",
    x = "Actual DOL",
    color = "Class",
    title = "",
    subtitle = ""
  )

ggsave(plot = for_save, here("Output", "FIGURE 3A random forest prediction for both growth legend.pdf"), height = 3.5, width = 5)

for_save + theme(legend.position = "none")

ggsave(here("Output", "FIGURE 3A random forest prediction for both growth no legend.pdf"), height = 5.5, width = 10)

```

# FIGURE 3B - RMMI reg. growth vs. GF vs. validation

```{r, fig.width=5, fig.height=3}

# mod1 is the SSlogis model for class %in% "Regular growth"
# mod2 is the SSlogis model for class %in% "Growth failure"

reg_growth_df <- tibble(class = "Regular growth", MMI = as.vector(rf_model_improved$predicted), actual_DOL = as.vector(rf_model_improved$y))

mod1 = nls(MMI ~ SSlogis(actual_DOL, Asym, xmid, scal), data=reg_growth_df)

p3 <- predict(rf_model_improved, validation_data)

plot_df <- plot_df %>%
  bind_rows(tibble(class = "Validation", MMI = as.vector(p3), actual_DOL = validation_data$DOL, SampleID = rownames(validation_data))) %>%
  mutate(class = fct_relevel(class, "Training"))

plot_df$curve_value <- predict(mod1, newdata = plot_df)

plot_df %<>%
  mutate(RMMI = MMI - curve_value) %>%
  mutate(class = fct_relevel(class, "Regular growth", "Growth failure"))

plot_df %>%
  ggplot(aes(x = class, y = RMMI, color = class)) +
  geom_boxplot(outlier.alpha = 0) +
  geom_quasirandom(dodge.width = 0.75, alpha = 0.25) +
  theme_bw() +
  scale_color_manual(values = growth_class_colors) +
  theme(legend.position = "") +
  labs(x = "Class", y = "Relative microbiota maturity index")

ggsave(here("Output", "FIGURE 3B RMMI comparison.pdf"), height = 3, width = 4)

```

# TABLE 3C - Mixed-effects model for RMMI

```{r, results='asis'}

test_df <- plot_df %>%
  left_join(s2 %>% select(SampleID, SubjectID), join_by(SampleID))

summary_df <- tidy_lmer(nlme::lme(fixed = RMMI ~ class, random = ~ 1 | SubjectID, data = test_df)) 

summary_df %>% pander(digits=2, split.cells=15)

summary_df %>% write_csv(here("Output", "FIGURE 3B summary table.csv"))

```

# FIGURE 3C Top five features for each model

```{r, fig.height=4, fig.width=8}

imp <- importance(rf_model_25taxa, type = 1)
imp_df_25taxa <- tibble(Feature = dimnames(imp)[[1]], MeanDecAccu = as.vector(imp)) %>%
  mutate(model = "25 taxa")
imp <- importance(rf_model_improved_78taxa, type = 1)
imp_df_improved_78taxa <- tibble(Feature = dimnames(imp)[[1]], MeanDecAccu = as.vector(imp)) %>%
  mutate(model = "71 taxa, boruta optimized")
imp <- importance(rf_model_just_kegg, type = 1)
imp_df_just_kegg <- tibble(Feature = dimnames(imp)[[1]], MeanDecAccu = as.vector(imp)) %>%
  mutate(model = "165 KEGG genes, boruta optimized") %>%
  left_join(kegg_defnitions, by = c("Feature"="Ortholog")) %>%
  mutate(Feature = paste(Feature, str_trunc(Definition, 20, "right")))

imp_df <- bind_rows(imp_df_25taxa, imp_df_improved_78taxa, imp_df_just_kegg) %>%
  group_by(model) %>%
  slice_max(n = 5, order_by = MeanDecAccu) %>%
  ungroup() %>%
  droplevels() %>%
  mutate(Feature = str_remove(Feature, "[a-zA-Z]+?_")) %>%
  mutate(Feature = str_replace(Feature, "_", " ")) %>%
  mutate(Feature = fct_reorder(Feature, MeanDecAccu)) %>%
  mutate(model = fct_relevel(model, "25 taxa", "71 taxa, boruta optimized"))

imp_df %>%
  ggplot(aes(x = MeanDecAccu, y = Feature)) +
  facet_wrap(~model, ncol = 3, scales = "free_x", strip.position = "right") +
  # facet_wrap(~model, ncol = 3, scales = "free_x", labeller = labeller(group = label_wrap_gen(width = 25))) +
  geom_point() +
  theme_bw() +
  theme(axis.text.x = element_text(angle = 45, hjust = 1, vjust = 1)) +
  labs(x = "Mean decrease in accuracy",
       y = "Top 5 features in models")

ggsave(here("Output", "FIGURE 3C Top 5 features.pdf"), height = 3, width = 5.5)

```

# FIGURE 3D

```{r}

roc25taxa <- plot_df_25taxa %>%
  mutate(class = fct_relevel(class, "Regular growth")) %>%
  roc(class ~ RMMI, data = .)

roc78taxa <- plot_df_78taxa %>%
  mutate(class = fct_relevel(class, "Regular growth")) %>%
  roc(class ~ RMMI, data = .)

rockegg <- plot_df_kegg %>%
  mutate(class = fct_relevel(class, "Regular growth")) %>%
  roc(class ~ RMMI, data = .)

ggroc(list(`25 taxa`=roc25taxa, `71 taxa`=roc78taxa, `165 KEGG genes`=rockegg)) +
  theme_bw() +
  labs(title = "", color = "Model", y = "Sensitivity", x = "Specificity") +
  geom_abline(slope = 1, intercept = c(1,0), lty = 2) +
  scale_colour_aaas()

ggsave(filename = here::here("Output","FIGURE 3D ROC curves.pdf"), device = "pdf", height = 3, width = 4)

```

# Supp Figure 3A - alpha diversity 0-21 day gfalt

```{r fig.height=6, fig.width=8}

s_toPlot <- s_toTest <- s2 %>%
  filter(DOL < 22) %>%
  droplevels()

# WOL_colors <- setNames((viridis(length(levels(s2$WOL_label)), alpha = 0.8)), levels(s2$WOL_label))

WOL_colors <- c(`1` = "#440154CC", `2` = "#481A6CCC", `3` = "#472F7DCC", `4` = "#414487CC")

s_toPlot %>%
  pivot_longer(c("Richness", "Shannon"), names_to="metric", values_to="alpha") %>%
  filter(metric %in% "Shannon") %>%
  mutate(growth_label = case_when(growth_cat %in% "normal" ~ "None",
                                  growth_cat %in% "mild retardation" ~ "Mild",
                                  growth_cat %in% "moderate to severe retardation" ~ "Moderate\nto severe")) %>%
  mutate(growth_label = fct_relevel(growth_label, "None", "Mild")) %>%
  ggplot(aes(x=growth_label, y=alpha, fill=WOL_label)) +
  geom_boxplot() + #outlier.alpha=0
  scale_fill_manual(values = WOL_colors) +
  # facet_grid(metric~., scales = "free") +
  theme_bw() +
  theme(
    strip.background = element_blank(),
    panel.grid= element_blank(),
    # legend.position = "none",
    axis.text.x = element_text(angle = 45, hjust = 1, vjust = 1)
  ) +
  labs(
    x="Growth faltering", fill="Week of life",
    y="Shannon diversity"
  )

ggsave(here("Output", "SUPP_FIG3A_alphadiv_gfalt.pdf"), width = 3.5, height = 2.5, useDingbats = F)

```

## Linear model with log_npo_days_prop

```{r, results='asis'}
my_formula <- "alpha ~ bwt_z + WOL * growth_cat + log_npo_days_prop"

summaries_df <- s_toTest %>%
  pivot_longer(c("Richness", "Shannon"), names_to="metric", values_to="alpha") %>%
  group_by(metric) %>%
  do(tidy_lmer(nlme::lme(as.formula(my_formula), random=~1|SubjectID, data=., na.action=na.omit))) %>%
  ungroup() %>%
  filter(!grepl("Intercept", term)) %>%
  mutate(term = str_replace(term, "timepoint_numeric", "Time")) %>%
  mutate(term = if_else(str_detect(term, "growth_cat"), str_remove(term, "growth_cat"), term))

summaries_df %>%
  pander(digits = 2, split.table = Inf, split.cells = 20, caption = glue("Linear model of the alpha diversity measurement with numeric time and growth categories as the predictive variables. SubjectID added as a random effect. Full formula: {my_formula}"))
```

# Supp Figure 3B - beta diversity 0-21 day gfalt

```{r fig.height=3, fig.width=5}

s_toPlot %>%
  mutate(growth_label = case_when(growth_cat %in% "normal" ~ "None",
                                  growth_cat %in% "mild retardation" ~ "Mild",
                                  growth_cat %in% "moderate to severe retardation" ~ "Moderate\nto severe")) %>%
  mutate(growth_label = fct_relevel(growth_label, "None", "Mild")) %>%
  pcoaplus(bc) %>%
  plot(color=growth_label) +
    scale_color_manual(values=ann_colors$growth_label) +
    theme_bw() +
    theme(
      axis.text=element_blank(),
      axis.ticks=element_blank(),
      panel.grid = element_blank(), 
      aspect.ratio = 1,
      strip.background = element_blank()
    ) + 
    labs(color="Growth faltering", title="")

ggsave(here("Output", "SUPP_FIG3B_betadiv.pdf"), width = 3.5, height = 2.5, useDingbats = F)

```

## Permanova with log_npo_days_prop

```{r, results='asis'}

# summaries_df <-
#   s_toTest %>%
#     do(adonisplus(
#       ., distmat = bc, formula = distmat ~ bwt_z + log_npo_days_prop + growth_cat * WOL,
#       sample_id_var = SampleID, rep_meas_var = SubjectID,
#       shuffle = c(WOL = "within", growth_cat = "between", log_npo_days_prop = "within"), permutations=perm)) %>%
#   mutate(metric = "Bray-Curtis") %>%
#   filter(!term %in% c("Residuals", "Total")) %>%
#   select(-one_of(c("sumsq", "meansq"))) %>%
#   select(metric, everything())
# 
# write_rds(summaries_df, here("FigureData", "permanova_results_10.rds"))

summaries_df <- read_rds(here("FigureData", "permanova_results_10.rds"))

summaries_df %>%
  pander()

```

# Supp Figure 3C - alpha diversity 0-21 days gfail

```{r}

s_toPlot %>%
  pivot_longer(c("Richness", "Shannon"), names_to="metric", values_to="alpha") %>%
  filter(metric %in% "Shannon") %>%
  ggplot(aes(x=growth_failure, y=alpha, fill=WOL_label)) +
  geom_boxplot() + #outlier.alpha=0
  scale_fill_manual(values = WOL_colors) +
  # facet_grid(metric~., scales = "free") +
  theme_bw() +
  theme(
    strip.background = element_blank(),
    panel.grid=element_blank(),
    # legend.position = "none",
    axis.text.x = element_text(angle = 45, hjust = 1, vjust = 1)
  ) +
  labs(
    x="Growth failure", fill="Week of life",
    y="Shannon diversity"
  )

ggsave(here("Output", "SUPP_FIG3C_alphadiv_gfail.pdf"), width = 3.5, height = 2.5, useDingbats = F)

```

## Linear model with log_npo_days_prop

```{r, results='asis'}
my_formula <- "alpha ~ bwt_z + WOL * growth_failure + log_npo_days_prop"

summaries_df <- s_toTest %>%
  pivot_longer(c("Richness", "Shannon"), names_to="metric", values_to="alpha") %>%
  group_by(metric) %>%
  do(tidy_lmer(nlme::lme(as.formula(my_formula), random=~1|SubjectID, data=., na.action=na.omit))) %>%
  ungroup() %>%
  filter(!grepl("Intercept", term)) %>%
  mutate(term = str_replace(term, "timepoint_numeric", "Time")) %>%
  mutate(term = if_else(str_detect(term, "growth_cat"), str_remove(term, "growth_cat"), term))

summaries_df %>%
  pander(digits = 2, split.table = Inf, split.cells = 20, caption = glue("Linear model of the alpha diversity measurement with numeric time and growth_failure as the predictive variables. SubjectID added as a random effect. Full formula: {my_formula}"))
```

# Supp Figure 3D - beta diversity 0-21 days gfail

```{r}
s_toPlot %>%
  pcoaplus(bc) %>%
  plot(color=growth_failure) +
    scale_color_manual(values=ann_colors$growth_fail) +
    theme_bw() +
    theme(
      axis.text=element_blank(),
      axis.ticks=element_blank(),
      panel.grid = element_blank(), 
      aspect.ratio = 1,
      strip.background = element_blank()
    ) + 
    labs(color="Growth failure", title="")

ggsave(here("Output", "SUPP_FIG3D_betadiv.pdf"), width = 3, height = 2.5, useDingbats = F)
```

## Permanova with log_npo_days_prop

```{r, results='asis'}

# summaries_df <-
#   s_toTest %>%
#     do(adonisplus(
#       ., distmat = bc, formula = distmat ~ bwt_z + log_npo_days_prop + growth_failure * WOL,
#       sample_id_var = SampleID, rep_meas_var = SubjectID,
#       shuffle = c(WOL = "within", growth_failure = "between", log_npo_days_prop = "within"), permutations=perm)) %>%
#   mutate(metric = "Bray-Curtis") %>%
#   filter(!term %in% c("Residuals", "Total")) %>%
#   select(-one_of(c("sumsq", "meansq"))) %>%
#   select(metric, everything())
# 
# write_rds(summaries_df, here("FigureData", "permanova_results_11.rds"))

summaries_df <- read_rds(here("FigureData", "permanova_results_11.rds"))

summaries_df %>%
  pander()

```

# Supp Figure 3E - alpha diversity 34-37 weeks PMA gfalt

```{r fig.height=6, fig.width=8}

s_subset2 <- read_rds(here("Data", "subset_of_babies_great34wk_less37wk.rds"))

s_toPlot <- s_toTest <- s2 %>%
  filter(SampleID %in% s_subset2$SampleID) %>%
  droplevels()

# WOL_colors <- setNames((viridis(length(levels(s2$WOL_label)), alpha = 0.8)), levels(s2$WOL_label))

WOL_colors <- c(`3` = "#472F7DCC", `4` = "#414487CC", 
                `5` = "#39568CCC", `6` = "#31688ECC", `7` = "#2A788ECC", `8` = "#23888ECC", 
                `9` = "#1F988BCC", `10` = "#22A884CC", `11` = "#35B779CC", `12` = "#54C568CC", 
                `13` = "#7AD151CC"
)

s_toPlot %>%
  pivot_longer(c("Richness", "Shannon"), names_to="metric", values_to="alpha") %>%
  filter(metric %in% "Shannon") %>%
  mutate(growth_label = case_when(growth_cat %in% "normal" ~ "None",
                                  growth_cat %in% "mild retardation" ~ "Mild",
                                  growth_cat %in% "moderate to severe retardation" ~ "Moderate\nto severe")) %>%
  mutate(growth_label = fct_relevel(growth_label, "None", "Mild")) %>%
  ggplot(aes(x=growth_label, y=alpha, fill=WOL_label)) +
  geom_boxplot() + #outlier.alpha=0
  scale_fill_manual(values = WOL_colors) +
  # facet_grid(metric~., scales = "free") +
  theme_bw() +
  theme(
    strip.background = element_blank(),
    panel.grid= element_blank(),
    # legend.position = "none",
    axis.text.x = element_text(angle = 45, hjust = 1, vjust = 1)
  ) +
  labs(
    x="Growth faltering", fill="Week of life",
    y="Shannon diversity"
  )

ggsave(here("Output", "SUPP_FIG3E_alphadiv_gfalt.pdf"), width = 3.5, height = 2.5, useDingbats = F)

```

## Linear model with log_npo_days_prop

```{r, results='asis'}
my_formula <- "alpha ~ bwt_z + WOL * growth_cat + log_npo_days_prop"

summaries_df <- s_toTest %>%
  pivot_longer(c("Richness", "Shannon"), names_to="metric", values_to="alpha") %>%
  group_by(metric) %>%
  do(tidy_lmer(nlme::lme(as.formula(my_formula), random=~1|SubjectID, data=., na.action=na.omit))) %>%
  ungroup() %>%
  filter(!grepl("Intercept", term)) %>%
  mutate(term = str_replace(term, "timepoint_numeric", "Time")) %>%
  mutate(term = if_else(str_detect(term, "growth_cat"), str_remove(term, "growth_cat"), term))

summaries_df %>%
  pander(digits = 2, split.table = Inf, split.cells = 20, caption = glue("Linear model of the alpha diversity measurement with numeric time and growth categories as the predictive variables. SubjectID added as a random effect. Full formula: {my_formula}"))
```

# Supp Figure 3F - beta diversity 34-37 weeks PMA gfalt

```{r fig.height=3, fig.width=5}

s_toPlot %>%
  mutate(growth_label = case_when(growth_cat %in% "normal" ~ "None",
                                  growth_cat %in% "mild retardation" ~ "Mild",
                                  growth_cat %in% "moderate to severe retardation" ~ "Moderate\nto severe")) %>%
  mutate(growth_label = fct_relevel(growth_label, "None", "Mild")) %>%
  pcoaplus(bc) %>%
  plot(color=growth_label) +
    scale_color_manual(values=ann_colors$growth_label) +
    theme_bw() +
    theme(
      axis.text=element_blank(),
      axis.ticks=element_blank(),
      panel.grid = element_blank(), 
      aspect.ratio = 1,
      strip.background = element_blank()
    ) + 
    labs(color="Growth faltering", title="")

ggsave(here("Output", "SUPP_FIG3F_betadiv.pdf"), width = 3.5, height = 2.5, useDingbats = F)

```

## Permanova with log_npo_days_prop

```{r, results='asis'}

# summaries_df <-
#   s_toTest %>%
#     do(adonisplus(
#       ., distmat = bc, formula = distmat ~ bwt_z + log_npo_days_prop + growth_cat * WOL,
#       sample_id_var = SampleID, rep_meas_var = SubjectID,
#       shuffle = c(WOL = "within", growth_cat = "between", log_npo_days_prop = "within"), permutations=perm)) %>%
#   mutate(metric = "Bray-Curtis") %>%
#   filter(!term %in% c("Residuals", "Total")) %>%
#   select(-one_of(c("sumsq", "meansq"))) %>%
#   select(metric, everything())
# 
# write_rds(summaries_df, here("FigureData", "permanova_results_12.rds"))

summaries_df <- read_rds(here("FigureData", "permanova_results_12.rds"))

summaries_df %>%
  pander()

```

# Supp Figure 3G - alpha diversity 34-37 weeks PMA gfail

```{r}

s_toPlot %>%
  pivot_longer(c("Richness", "Shannon"), names_to="metric", values_to="alpha") %>%
  filter(metric %in% "Shannon") %>%
  ggplot(aes(x=growth_failure, y=alpha, fill=WOL_label)) +
  geom_boxplot() + #outlier.alpha=0
  scale_fill_manual(values = WOL_colors) +
  # facet_grid(metric~., scales = "free") +
  theme_bw() +
  theme(
    strip.background = element_blank(),
    panel.grid=element_blank(),
    # legend.position = "none",
    axis.text.x = element_text(angle = 45, hjust = 1, vjust = 1)
  ) +
  labs(
    x="Growth failure", fill="Week of life",
    y="Shannon diversity"
  )

ggsave(here("Output", "SUPP_FIG3G_alphadiv_gfail.pdf"), width = 3.5, height = 2.5, useDingbats = F)

```

## Linear model with log_npo_days_prop

```{r, results='asis'}
my_formula <- "alpha ~ bwt_z + WOL * growth_failure + log_npo_days_prop"

summaries_df <- s_toTest %>%
  pivot_longer(c("Richness", "Shannon"), names_to="metric", values_to="alpha") %>%
  group_by(metric) %>%
  do(tidy_lmer(nlme::lme(as.formula(my_formula), random=~1|SubjectID, data=., na.action=na.omit))) %>%
  ungroup() %>%
  filter(!grepl("Intercept", term)) %>%
  mutate(term = str_replace(term, "timepoint_numeric", "Time")) %>%
  mutate(term = if_else(str_detect(term, "growth_cat"), str_remove(term, "growth_cat"), term))

summaries_df %>%
  pander(digits = 2, split.table = Inf, split.cells = 20, caption = glue("Linear model of the alpha diversity measurement with numeric time and growth_failure as the predictive variables. SubjectID added as a random effect. Full formula: {my_formula}"))
```

# Supp Figure 3H - beta diversity 34-37 weeks PMA gfail

```{r}
s_toPlot %>%
  pcoaplus(bc) %>%
  plot(color=growth_failure) +
    scale_color_manual(values=ann_colors$growth_fail) +
    theme_bw() +
    theme(
      axis.text=element_blank(),
      axis.ticks=element_blank(),
      panel.grid = element_blank(), 
      aspect.ratio = 1,
      strip.background = element_blank()
    ) + 
    labs(color="Growth failure", title="")

ggsave(here("Output", "SUPP_FIG3H_betadiv.pdf"), width = 3, height = 2.5, useDingbats = F)
```

## Permanova with log_npo_days_prop

```{r, results='asis'}

# summaries_df <-
#   s_toTest %>%
#     do(adonisplus(
#       ., distmat = bc, formula = distmat ~ bwt_z + log_npo_days_prop + growth_failure * WOL,
#       sample_id_var = SampleID, rep_meas_var = SubjectID,
#       shuffle = c(WOL = "within", growth_failure = "between", log_npo_days_prop = "within"), permutations=perm)) %>%
#   mutate(metric = "Bray-Curtis") %>%
#   filter(!term %in% c("Residuals", "Total")) %>%
#   select(-one_of(c("sumsq", "meansq"))) %>%
#   select(metric, everything())
# 
# write_rds(summaries_df, here("FigureData", "permanova_results_13.rds"))

summaries_df <- read_rds(here("FigureData", "permanova_results_13.rds"))

summaries_df %>%
  pander()

```

# End of report

```{r Generate time stamped report, eval=FALSE, include=FALSE}
#notes
#neat: you can run this following command in the console to give your reports custom names (or date-stamp them)
#rmarkdown::render('ronen_stein_Report_shotgun.Rmd',output_file = paste('CEASE.report.', Sys.Date(), '.pdf', sep=''))

rstudioapi::documentSaveAll()

script_dir <- file.path("Scripts","Rmds")
output_dir <- file.path("Output")

report_fp <- here::here(script_dir, "FIGURESANDTABLES.Rmd")
output_fp <- here::here(output_dir, paste('sagori_microbiome_figures_and_tables_', Sys.Date(), '.pdf', sep=''))

rmarkdown::render(report_fp, output_file = output_fp)

```
