---
title: |
  ![](CHOPMC_logo.pdf){width=5in}
  \break
  Sagori Mukhopadhyay - Runs 1 to 5 - Figures and tables
author: "Scott G. Daniel & Naomi G. Wilson; PennCHOP Microbiome Program"
date: "`r Sys.Date()`"
geometry: margin=3cm
output: 
    pdf_document:
        keep_tex: false
        toc: true
        toc_depth: 4
---

```{r setup, echo=FALSE}
### ================
###   knitr setup
### ================
library(knitr)
opts_chunk$set(
  tidy=FALSE,
  cache=FALSE,
  echo=FALSE,
  warning=FALSE,
  message=FALSE,
  dpi=100,
  fig.width=4,
  fig.height=4,
  fig.align = "center"
  )
```

```{r libs and funcs, message=FALSE, warning=FALSE}

library(here)
library(tidyverse)
library(qiimer)
library(vegan)
library(ape)
library(usedist)

## Visualization packages
library(pander)
library(kableExtra)

library(pheatbuilder)
library(ggbeeswarm)
library(ggsci)
library(viridis)
library(RColorBrewer)

# stats packages
library(adonisplus)
library(nlme)
library(emmeans) # for lmer post-hoc tests
library(broom)
library(ghibli)
library(glue)
library(magrittr)
select <- dplyr::select
library(patchwork)
library(qualpalr)
library(randomForest)
library(microbenchmark)
library(Boruta)
library(mlbench)
library(caret)
library(pROC)

cut_MONTH_timepoints <- function(DOL) {
  breakpoints <- c(0,27,55,83,111)
  my_labels <- c(1,2,3,4)
  cut(DOL, breaks = breakpoints, labels = my_labels, include.lowest = T, right = T)
}

cut_timepoints <- function(DOL) {
  breakpoints <- c(0,6,13,20,27,34,41,48,55,62,69,76,83,90,97,104,111)
  my_labels <- c(1,2,3,4,5,6,7,8,9,10,11,12,13,14,15,16)
  cut(DOL, breaks = breakpoints, labels = my_labels, include.lowest = T, right = T)
}
```

```{r load data}

load(file = here("Data","shotgun_data_2026-04-28.RData")) #source for raw data
load(file = here("Data","sagori_taxonomic_data.RData")) #source for fixed s2 table
load(file = here("Data","sagori_kegg_data.RData")) #source for gene abundance data
table_export <- table_iterator()

```

```{r log variables}

s2 %<>%
  mutate(npo_days_prop_min = min(npo_days_prop[npo_days_prop > 0])) %>%
  mutate(npo_days_prop = ifelse(npo_days_prop == 0, npo_days_prop_min/2, npo_days_prop)) %>%
  mutate(log_npo_days_prop = log(npo_days_prop)) %>%
  mutate(AbxHoursPerHOL_min = min(AbxHoursPerHOL[AbxHoursPerHOL > 0])) %>%
  mutate(AbxHoursPerHOL = ifelse(AbxHoursPerHOL == 0, AbxHoursPerHOL_min/2, AbxHoursPerHOL)) %>%
  mutate(log_AbxHoursPerHOL = log(AbxHoursPerHOL))

```

```{r moar colors}

ann_colors <- list(category = c(`1` = "#66C2A5", `2` = "#FC8D62", `3` = "#8DA0CB"
),
growth_cat = c(normal = "#92BBD9FF", `mild retardation` = "#DCCA2CFF", 
`moderate to severe retardation` = "#6FB382FF"),
TimePoint = c(T0 = "#440154FF", 
T1 = "#48186AFF", T2 = "#472D7BFF", T3 = "#424086FF", T4 = "#3B528BFF", 
T5 = "#33638DFF", T6 = "#2C728EFF", T7 = "#26828EFF", T8 = "#21908CFF", 
T9 = "#1F9F88FF", T10 = "#27AD81FF", T11 = "#3EBC74FF", T12 = "#5DC863FF", 
T13 = "#82D34DFF", T14 = "#AADC32FF", T15 = "#D5E21AFF", T16 = "#FDE725FF"
),
growth_fail = c(No = "#9FCB6A", Yes = "#8367CC"),
growth_label = c(None = "#92BBD9FF", Mild = "#DCCA2CFF", 
`Moderate\nto severe` = "#6FB382FF"))

```

```{r sra upload}

s_sra <- s2 %>%
  filter(which_run %in% c("run4", "run5")) %>%
  rename(sample_name = SampleID, collection_date = date_collected) %>%
  mutate(collection_date = if_else(is.na(collection_date), "2019", collection_date)) %>%
  mutate(collection_date = str_remove(collection_date, ".+/")) %>%
  mutate(collection_date = if_else(str_length(collection_date) == 2, paste0("20", collection_date), collection_date)) %>%
  mutate(host = "Homo sapiens") %>%
  mutate(organism = "human gut metagenome") %>%
  mutate(env_broad_scale = "dense settlement biome", env_local_scale = "human-associated environment", env_medium = "fecal material", geo_loc_name = "USA", lat_lon = "39.9473 N 75.1958 W") %>%
  select(sample_name, organism, collection_date, env_broad_scale, env_local_scale, env_medium, geo_loc_name, host, lat_lon, WOL, host_sex = Sex, growth_failure, description = SampleType, log_npo_days_prop, log_AbxHoursPerHOL, growth_cat, bwt_z, PMA, DOL) %>%
  group_by(growth_cat) %>%
  mutate(replicate = paste0(growth_cat, row_number())) %>%
  ungroup()

s_sra %>%
  write_tsv(file = here("Data","sample_table_for_sra.tsv"))

sra_metadata <- read_csv(here("Data","sagori_run4_and_5_samples.csv"), col_names = c("sample_name","filename","filename2")) %>%
  mutate(filename = str_replace(filename, "/scr1/users/danielsg/sagori_run_[45]/data_files/", "")) %>%
  mutate(filename = str_replace(filename, "_R1", "_1")) %>%
  mutate(filename2 = str_replace(filename2, "/scr1/users/danielsg/sagori_run_[45]/data_files/", "")) %>%
  mutate(filename2 = str_replace(filename2, "_R2", "_2")) %>%
  mutate(library_ID = str_replace(filename, "_1.fastq.gz", "")) %>%
  right_join(s_sra, by = "sample_name") %>%
  mutate(title = paste("DNA sequencing of gut bacteria in babies: ",replicate)) %>%
  mutate(library_strategy = "WGS") %>%
  mutate(library_source = "METAGENOMIC") %>%
  mutate(library_selection = "RANDOM") %>%
  mutate(library_layout = "paired") %>%
  mutate(platform = "ILLUMINA") %>%
  mutate(instrument_model = "Illumina HiSeq 2500") %>%
  mutate(design_description = "Shotgun libraries were generated using the Nextera Flex Library Prep kit and sequenced on the Illumina HiSeq 2500 using 2x125 bp chemistry") %>%
  mutate(filetype = "fastq")

sra_metadata %>% select(sample_name, library_ID, title, library_strategy, library_source, library_selection, library_layout, platform, instrument_model, design_description, filetype, filename, filename2) %>%
  write_tsv(file = here("Data","sra_metadata.tsv"))

for_copying <- c(sra_metadata$filename, sra_metadata$filename2)
write_lines(for_copying, file = here("Data","file_list_for_sra_upload.txt"))

```

# FIGURE 1B - Growth velocity and birthweight

```{r}

# s2 %>%
#   write_csv(here("Output", "metadata_that_scott_has.csv"))

#fixing npo_days_prop for s2
# s2 %<>%
#     mutate(npo_days_n = ifelse(SampleID %in% "SRC.738128", 1, npo_days_n)) %>%
#   mutate(npo_days_n = ifelse(SampleID %in% "SRC.738559", 1, npo_days_n)) %>%
#   mutate(npo_days_prop = npo_days_n / DOL) %>%
#   mutate(npo_days_prop = ifelse(npo_days_prop == Inf, 1, npo_days_prop))

fig1b_data <- s2 %>%
  select(redcap_id, bwt_pct, GV_36wk, growth_cat, growth_failure) %>%
  unique() %>%
  mutate(growth_label = case_when(growth_cat %in% "normal" & growth_failure %in% "No" ~ "No failure, no faltering",
                                  growth_cat %in% "mild retardation" & growth_failure %in% "No" ~ "No failure, mild faltering",
                                  growth_cat %in% "moderate to severe retardation" & growth_failure %in% "No" ~ "No failure, sev faltering",
                                  growth_cat %in% "normal" & growth_failure %in% "Yes" ~ "Failure, no faltering",
                                  growth_cat %in% "mild retardation" & growth_failure %in% "Yes" ~ "Failure, mild faltering",
                                  growth_cat %in% "moderate to severe retardation" & growth_failure %in% "Yes" ~ "Failure, sev faltering")) %>%
  mutate(growth_label = fct_relevel(growth_label, "Failure, no faltering",
                                    "Failure, mild faltering",
                                    "Failure, sev faltering",
                                    "No failure, no faltering",
                                    "No failure, mild faltering",
                                    "No failure, sev faltering"))

growth_label_colors <- setNames(ghibli_palette(name = "PonyoMedium", n = 6, type = "discrete"), unique(fig1b_data$growth_label))

growth_label_colors <- c(`No failure, mild faltering` = "#4C413FFF", `Failure, sev faltering` = "#EEBCB1FF", 
`No failure, no faltering` = "#278B9AFF", `No failure, sev faltering` = "#B50A2AFF", 
`Failure, no faltering` = "#DE7862FF", `Failure, mild faltering` = "#D8AF39FF"
)

fig1b_data %>%
  ggplot(aes(x=bwt_pct, y=GV_36wk, colour = growth_label)) +
  geom_point() +
  scale_color_manual(values = growth_label_colors) +
  theme_bw() +
  labs(y = "Growth velocity at 36 weeks",
       x = "Birthweight percentile",
       color = "Growth failure and faltering")

ggsave(here("Output", "FIG1B_GVandbwtpct.pdf"), width = 5, height = 3, useDingbats = F)

```

# FIGURE 1C - Alpha diversity

```{r fig.width=8}

s2 %>%
  filter(!TimePoint %in% "T0") %>%
  pivot_longer(c("Richness", "Shannon"), names_to="metric", values_to="alpha") %>%
  filter(metric %in% "Shannon") %>%
  ggplot(aes(x=WOL_label, y=alpha, fill = WOL_label)) +
  #geom_boxplot(outlier.alpha=0) +
  geom_boxplot() + #outlier.alpha=0
  scale_fill_viridis_d(alpha = 0.8) +
  # facet_grid(metric~., scales = "free") +
  theme_bw() +
  theme(
    strip.background = element_blank(),
    panel.grid=element_blank(),
    legend.position = "none",
    axis.text.x = element_text(angle = 90, vjust = 0.5, hjust = 1)
  ) +
  labs(
    x="Week of life", color="",
    y="Shannon diversity"
  )

ggsave(here("Output", "FIG1C_alphadiv_no_T0.pdf"), width = 2.5, height = 2.5, useDingbats = F)

```

\newpage
## Linear model with just time

```{r, results='asis'}
my_formula <- "alpha ~ WOL"

summaries_df <- s2 %>%
  filter(!TimePoint %in% "T0") %>%
  pivot_longer(c("Richness", "Shannon"), names_to="metric", values_to="alpha") %>%
  group_by(metric) %>%
  do(tidy_lmer(nlme::lme(as.formula(my_formula), random=~1|SubjectID, data=., na.action=na.omit))) %>%
  ungroup() %>%
  filter(!grepl("Intercept", term)) %>%
  mutate(term = str_replace(term, "timepoint_numeric", "Time")) %>%
  mutate(term = if_else(str_detect(term, "growth_cat"), str_remove(term, "growth_cat"), term))

summaries_df %>%
  pander(digits = 2, split.table = Inf, split.cells = 20, caption = glue("Linear model of the alpha diversity measurement with numeric time as the predictive variable. SubjectID added as a random effect. Full formula: {my_formula}"))
```

## Linear model with log(npo_days_prop) and bwt_z

```{r, results='asis'}
my_formula <- "alpha ~ bwt_z + WOL + log_npo_days_prop"

summaries_df <- s2 %>%
  filter(!TimePoint %in% "T0") %>%
  pivot_longer(c("Richness", "Shannon"), names_to="metric", values_to="alpha") %>%
  group_by(metric) %>%
  do(tidy_lmer(nlme::lme(as.formula(my_formula), random=~1|SubjectID, data=., na.action=na.omit))) %>%
  ungroup() %>%
  filter(!grepl("Intercept", term)) %>%
  mutate(term = str_replace(term, "timepoint_numeric", "Time")) %>%
  mutate(term = if_else(str_detect(term, "growth_cat"), str_remove(term, "growth_cat"), term))

summaries_df %>%
  pander(digits = 2, split.table = Inf, split.cells = 20, caption = glue("Linear model of the alpha diversity measurement with numeric time and log_npo_days_prop as the predictive variables. SubjectID added as a random effect. Full formula: {my_formula}"))
```


## Linear model with log(AbxHoursPerHOL)

```{r, results='asis'}
my_formula <- "alpha ~ bwt_z + WOL + log_AbxHoursPerHOL"

summaries_df <- s2 %>%
  filter(!TimePoint %in% "T0") %>%
  pivot_longer(c("Richness", "Shannon"), names_to="metric", values_to="alpha") %>%
  group_by(metric) %>%
  do(tidy_lmer(nlme::lme(as.formula(my_formula), random=~1|SubjectID, data=., na.action=na.omit))) %>%
  ungroup() %>%
  filter(!grepl("Intercept", term)) %>%
  mutate(term = str_replace(term, "timepoint_numeric", "Time")) %>%
  mutate(term = if_else(str_detect(term, "growth_cat"), str_remove(term, "growth_cat"), term))

summaries_df %>%
  pander(digits = 2, split.table = Inf, split.cells = 20, caption = glue("Linear model of the alpha diversity measurement with numeric time and log_AbxHoursPerHOL as the predictive variables. SubjectID added as a random effect. Full formula: {my_formula}"))
```

# FIGURE 1D - Beta-div. over time


```{r fig.height=3, fig.width=5}
s2 %>%
  pcoaplus(bc) %>%
  plot(color=WOL) +
    scale_color_viridis_c() +
    theme_bw() +
    theme(
      axis.text=element_blank(),
      axis.ticks=element_blank(),
      panel.grid = element_blank(), 
      aspect.ratio = 1,
      strip.background = element_blank()
    ) + 
    labs(color="Week of life", title="Bray-Curtis")

ggsave(here("Output", "FIG1D_betadiv.pdf"), width = 5, height = 3, useDingbats = F)

```

## Permanova with just time

```{r, results='asis'}

# summaries_df <- 
#   s2 %>%
#     do(adonisplus(
#       ., distmat = bc, formula = distmat ~ WOL,
#       sample_id_var = SampleID, rep_meas_var = SubjectID,
#       shuffle = c(WOL = "within"), permutations=perm)) %>%
#   mutate(metric = "Bray-Curtis") %>%
#   filter(!term %in% c("Residuals", "Total")) %>%
#   select(-one_of(c("sumsq", "meansq"))) %>%
#   select(metric, everything())
# 
# write_rds(summaries_df, here("FigureData", "permanova_results_01.rds"))

summaries_df <- read_rds(here("FigureData", "permanova_results_01.rds"))

summaries_df %>%
  pander()

```

## Permanova with log_npo_days_prop

```{r, results='asis'}

# summaries_df <- 
#   s2 %>%
#     do(adonisplus(
#       ., distmat = bc, formula = distmat ~ bwt_z + log_npo_days_prop + WOL,
#       sample_id_var = SampleID, rep_meas_var = SubjectID,
#       shuffle = c(WOL = "within", log_npo_days_prop = "within"), permutations=perm)) %>%
#   mutate(metric = "Bray-Curtis") %>%
#   filter(!term %in% c("Residuals", "Total")) %>%
#   select(-one_of(c("sumsq", "meansq"))) %>%
#   select(metric, everything())
# 
# write_rds(summaries_df, here("FigureData", "permanova_results_02.rds"))

summaries_df <- read_rds(here("FigureData", "permanova_results_02.rds"))

summaries_df %>%
  pander()

```

## Permanova with log_AbxHoursPerHOL

```{r, results='asis'}

# summaries_df <- 
#   s2 %>%
#     do(adonisplus(
#       ., distmat = bc, formula = distmat ~ bwt_z + log_AbxHoursPerHOL + WOL,
#       sample_id_var = SampleID, rep_meas_var = SubjectID,
#       shuffle = c(WOL = "within", log_AbxHoursPerHOL = "within"), permutations=perm)) %>%
#   mutate(metric = "Bray-Curtis") %>%
#   filter(!term %in% c("Residuals", "Total")) %>%
#   select(-one_of(c("sumsq", "meansq"))) %>%
#   select(metric, everything())
# 
# write_rds(summaries_df, here("FigureData", "permanova_results_03.rds"))

summaries_df <- read_rds(here("FigureData", "permanova_results_03.rds"))

summaries_df %>%
  pander()

```

# FIGURE 1E

Taxa abundance panel - Naomi is doing this

# SUPP FIG 1A - alpha diversity growth faltering

```{r fig.width=6}

  # mutate(growth_label = case_when(growth_cat %in% "normal" ~ "Normal",
  #                                 growth_cat %in% "mild retardation" ~ "Mild",
  #                                 growth_cat %in% "moderate to severe retardation" ~ "Moderate to severe")) %>%

s2 %>%
  pivot_longer(c("Richness", "Shannon"), names_to="metric", values_to="alpha") %>%
  filter(metric %in% "Shannon") %>%
  mutate(growth_label = case_when(growth_cat %in% "normal" ~ "Normal",
                                  growth_cat %in% "mild retardation" ~ "Mild",
                                  growth_cat %in% "moderate to severe retardation" ~ "Moderate\nto severe")) %>%
  mutate(growth_label = fct_relevel(growth_label, "Normal", "Mild")) %>%
  ggplot(aes(x=growth_label, y=alpha, fill=WOL_label)) +
  geom_boxplot() + #outlier.alpha=0
  scale_fill_viridis_d(alpha = 0.8) +
  # facet_grid(metric~., scales = "free") +
  theme_bw() +
  theme(
    strip.background = element_blank(),
    panel.grid=element_blank(),
    legend.position = "none",
    axis.text.x = element_text(angle = 45, hjust = 1, vjust = 1)
  ) +
  labs(
    x="Growth faltering", color="Week of life",
    y="Shannon diversity"
  )

ggsave(here("Output", "SUPP_FIG1A_alphadiv_gfalt.pdf"), width = 3.5, height = 2.5, useDingbats = F)

```

## Linear model with just growth faltering

```{r, results='asis'}
my_formula <- "alpha ~ growth_cat"

summaries_df <- s2 %>%
  filter(!TimePoint %in% "T0") %>%
  pivot_longer(c("Richness", "Shannon"), names_to="metric", values_to="alpha") %>%
  group_by(metric) %>%
  do(tidy_lmer(nlme::lme(as.formula(my_formula), random=~1|SubjectID, data=., na.action=na.omit))) %>%
  ungroup() %>%
  filter(!grepl("Intercept", term)) %>%
  mutate(term = str_replace(term, "timepoint_numeric", "Time")) %>%
  mutate(term = if_else(str_detect(term, "growth_cat"), str_remove(term, "growth_cat"), term))

summaries_df %>%
  pander(digits = 2, split.table = Inf, split.cells = 20, caption = glue("Linear model of the alpha diversity measurement with growth_cat as the predictive variable. SubjectID added as a random effect. Full formula: {my_formula}"))
```

## Linear model with log_npo_days_prop

```{r, results='asis'}
my_formula <- "alpha ~ bwt_z + WOL * growth_cat + log_npo_days_prop"

summaries_df <- s2 %>%
  filter(!TimePoint %in% "T0") %>%
  pivot_longer(c("Richness", "Shannon"), names_to="metric", values_to="alpha") %>%
  group_by(metric) %>%
  do(tidy_lmer(nlme::lme(as.formula(my_formula), random=~1|SubjectID, data=., na.action=na.omit))) %>%
  ungroup() %>%
  filter(!grepl("Intercept", term)) %>%
  mutate(term = str_replace(term, "timepoint_numeric", "Time")) %>%
  mutate(term = if_else(str_detect(term, "growth_cat"), str_remove(term, "growth_cat"), term))

summaries_df %>%
  pander(digits = 2, split.table = Inf, split.cells = 20, caption = glue("Linear model of the alpha diversity measurement with numeric time and growth categories as the predictive variables. SubjectID added as a random effect. Full formula: {my_formula}"))
```


## Linear model with log_AbxHoursPerHOL

```{r, results='asis'}
my_formula <- "alpha ~ bwt_z + WOL * growth_cat + log_AbxHoursPerHOL"

summaries_df <- s2 %>%
  filter(!TimePoint %in% "T0") %>%
  pivot_longer(c("Richness", "Shannon"), names_to="metric", values_to="alpha") %>%
  group_by(metric) %>%
  do(tidy_lmer(nlme::lme(as.formula(my_formula), random=~1|SubjectID, data=., na.action=na.omit))) %>%
  ungroup() %>%
  filter(!grepl("Intercept", term)) %>%
  mutate(term = str_replace(term, "timepoint_numeric", "Time")) %>%
  mutate(term = if_else(str_detect(term, "growth_cat"), str_remove(term, "growth_cat"), term))

summaries_df %>%
  pander(digits = 2, split.table = Inf, split.cells = 20, caption = glue("Linear model of the alpha diversity measurement with numeric time and growth categories as the predictive variables. SubjectID added as a random effect. Full formula: {my_formula}"))
```


# SUPP FIG 1B - beta diversity growth faltering

```{r fig.height=3, fig.width=5}
s2 %>%
  mutate(growth_label = case_when(growth_cat %in% "normal" ~ "None",
                                  growth_cat %in% "mild retardation" ~ "Mild",
                                  growth_cat %in% "moderate to severe retardation" ~ "Moderate\nto severe")) %>%
  mutate(growth_label = fct_relevel(growth_label, "None", "Mild")) %>%
  pcoaplus(bc) %>%
  plot(color=growth_label) +
    scale_color_manual(values=ann_colors$growth_label) +
    theme_bw() +
    theme(
      axis.text=element_blank(),
      axis.ticks=element_blank(),
      panel.grid = element_blank(), 
      aspect.ratio = 1,
      strip.background = element_blank()
    ) + 
    labs(color="", title="")

ggsave(here("Output", "SUPP_FIG1B_betadiv.pdf"), width = 3.5, height = 2.5, useDingbats = F)

```

## Permanova with growth_cat and time

```{r, results='asis'}

# summaries_df <- 
#   s2 %>%
#     do(adonisplus(
#       ., distmat = bc, formula = distmat ~ growth_cat * WOL,
#       sample_id_var = SampleID, rep_meas_var = SubjectID,
#       shuffle = c(WOL = "within", growth_cat = "between"), permutations=perm)) %>%
#   mutate(metric = "Bray-Curtis") %>%
#   filter(!term %in% c("Residuals", "Total")) %>%
#   select(-one_of(c("sumsq", "meansq"))) %>%
#   select(metric, everything())
# 
# write_rds(summaries_df, here("FigureData", "permanova_results_04.rds"))

summaries_df <- read_rds(here("FigureData", "permanova_results_04.rds"))

summaries_df %>%
  pander()

```

## Permanova with log_npo_days_prop

```{r, results='asis'}

# summaries_df <- 
#   s2 %>%
#     do(adonisplus(
#       ., distmat = bc, formula = distmat ~ bwt_z + log_npo_days_prop + growth_cat * WOL,
#       sample_id_var = SampleID, rep_meas_var = SubjectID,
#       shuffle = c(WOL = "within", growth_cat = "between", log_npo_days_prop = "within"), permutations=perm)) %>%
#   mutate(metric = "Bray-Curtis") %>%
#   filter(!term %in% c("Residuals", "Total")) %>%
#   select(-one_of(c("sumsq", "meansq"))) %>%
#   select(metric, everything())
# 
# write_rds(summaries_df, here("FigureData", "permanova_results_05.rds"))

summaries_df <- read_rds(here("FigureData", "permanova_results_05.rds"))

summaries_df %>%
  pander()

```

## Permanova with log_AbxHoursPerHOL

```{r, results='asis'}

# summaries_df <- 
#   s2 %>%
#     do(adonisplus(
#       ., distmat = bc, formula = distmat ~ bwt_z + log_AbxHoursPerHOL + growth_cat * WOL,
#       sample_id_var = SampleID, rep_meas_var = SubjectID,
#       shuffle = c(WOL = "within", growth_cat = "between", log_AbxHoursPerHOL = "within"), permutations=perm)) %>%
#   mutate(metric = "Bray-Curtis") %>%
#   filter(!term %in% c("Residuals", "Total")) %>%
#   select(-one_of(c("sumsq", "meansq"))) %>%
#   select(metric, everything())
# 
# write_rds(summaries_df, here("FigureData", "permanova_results_06.rds"))

summaries_df <- read_rds(here("FigureData", "permanova_results_06.rds"))

summaries_df %>%
  pander()

```


# SUPP FIG 1C - alpha diversity growth failure

```{r fig.width=8}

s2 %>%
  pivot_longer(c("Richness", "Shannon"), names_to="metric", values_to="alpha") %>%
  filter(metric %in% "Shannon") %>%
  ggplot(aes(x=growth_failure, y=alpha, fill=WOL_label)) +
  geom_boxplot() + #outlier.alpha=0
  scale_fill_viridis_d(alpha = 0.8) +
  # facet_grid(metric~., scales = "free") +
  theme_bw() +
  theme(
    strip.background = element_blank(),
    panel.grid=element_blank(),
    legend.position = "none",
    axis.text.x = element_text(angle = 45, hjust = 1, vjust = 1)
  ) +
  labs(
    x="Growth failure", color="Week of life",
    y="Shannon diversity"
  )

ggsave(here("Output", "SUPP_FIG1C_alphadiv_gfail.pdf"), width = 3.5, height = 2.5, useDingbats = F)

```

## Linear model with just growth_failure

```{r, results='asis'}
my_formula <- "alpha ~ growth_failure"

summaries_df <- s2 %>%
  filter(!TimePoint %in% "T0") %>%
  pivot_longer(c("Richness", "Shannon"), names_to="metric", values_to="alpha") %>%
  group_by(metric) %>%
  do(tidy_lmer(nlme::lme(as.formula(my_formula), random=~1|SubjectID, data=., na.action=na.omit))) %>%
  ungroup() %>%
  filter(!grepl("Intercept", term)) %>%
  mutate(term = str_replace(term, "timepoint_numeric", "Time")) %>%
  mutate(term = if_else(str_detect(term, "growth_cat"), str_remove(term, "growth_cat"), term))

summaries_df %>%
  pander(digits = 2, split.table = Inf, split.cells = 20, caption = glue("Linear model of the alpha diversity measurement with growth_failure as the predictive variable. SubjectID added as a random effect. Full formula: {my_formula}"))
```

## Linear model with log_npo_days_prop

```{r, results='asis'}
my_formula <- "alpha ~ bwt_z + WOL * growth_failure + log_npo_days_prop"

summaries_df <- s2 %>%
  filter(!TimePoint %in% "T0") %>%
  pivot_longer(c("Richness", "Shannon"), names_to="metric", values_to="alpha") %>%
  group_by(metric) %>%
  do(tidy_lmer(nlme::lme(as.formula(my_formula), random=~1|SubjectID, data=., na.action=na.omit))) %>%
  ungroup() %>%
  filter(!grepl("Intercept", term)) %>%
  mutate(term = str_replace(term, "timepoint_numeric", "Time")) %>%
  mutate(term = if_else(str_detect(term, "growth_cat"), str_remove(term, "growth_cat"), term))

summaries_df %>%
  pander(digits = 2, split.table = Inf, split.cells = 20, caption = glue("Linear model of the alpha diversity measurement with numeric time and growth_failure as the predictive variables. SubjectID added as a random effect. Full formula: {my_formula}"))
```


## Linear model with log_AbxHoursPerHOL

```{r, results='asis'}
my_formula <- "alpha ~ bwt_z + WOL * growth_failure + log_AbxHoursPerHOL"

summaries_df <- s2 %>%
  filter(!TimePoint %in% "T0") %>%
  pivot_longer(c("Richness", "Shannon"), names_to="metric", values_to="alpha") %>%
  group_by(metric) %>%
  do(tidy_lmer(nlme::lme(as.formula(my_formula), random=~1|SubjectID, data=., na.action=na.omit))) %>%
  ungroup() %>%
  filter(!grepl("Intercept", term)) %>%
  mutate(term = str_replace(term, "timepoint_numeric", "Time")) %>%
  mutate(term = if_else(str_detect(term, "growth_cat"), str_remove(term, "growth_cat"), term))

summaries_df %>%
  pander(digits = 2, split.table = Inf, split.cells = 20, caption = glue("Linear model of the alpha diversity measurement with numeric time and growth_failure as the predictive variables. SubjectID added as a random effect. Full formula: {my_formula}"))
```



# SUPP FIG 1D - beta diversity growth failure

```{r fig.height=3, fig.width=5}
s2 %>%
  pcoaplus(bc) %>%
  plot(color=growth_failure) +
    scale_color_manual(values=ann_colors$growth_fail) +
    theme_bw() +
    theme(
      axis.text=element_blank(),
      axis.ticks=element_blank(),
      panel.grid = element_blank(), 
      aspect.ratio = 1,
      strip.background = element_blank()
    ) + 
    labs(color="", title="")

ggsave(here("Output", "SUPP_FIG1D_betadiv.pdf"), width = 3, height = 2.5, useDingbats = F)

```

## Permanova with growth_failure and time

```{r, results='asis'}

# summaries_df <- 
#   s2 %>%
#     do(adonisplus(
#       ., distmat = bc, formula = distmat ~ growth_failure * WOL,
#       sample_id_var = SampleID, rep_meas_var = SubjectID,
#       shuffle = c(WOL = "within", growth_failure = "between"), permutations=perm)) %>%
#   mutate(metric = "Bray-Curtis") %>%
#   filter(!term %in% c("Residuals", "Total")) %>%
#   select(-one_of(c("sumsq", "meansq"))) %>%
#   select(metric, everything())
# 
# write_rds(summaries_df, here("FigureData", "permanova_results_07.rds"))

summaries_df <- read_rds(here("FigureData", "permanova_results_07.rds"))

summaries_df %>%
  pander()

```

## Permanova with log_npo_days_prop

```{r, results='asis'}

# summaries_df <- 
#   s2 %>%
#     do(adonisplus(
#       ., distmat = bc, formula = distmat ~ bwt_z + log_npo_days_prop + growth_failure * WOL,
#       sample_id_var = SampleID, rep_meas_var = SubjectID,
#       shuffle = c(WOL = "within", growth_failure = "between", log_npo_days_prop = "within"), permutations=perm)) %>%
#   mutate(metric = "Bray-Curtis") %>%
#   filter(!term %in% c("Residuals", "Total")) %>%
#   select(-one_of(c("sumsq", "meansq"))) %>%
#   select(metric, everything())
# 
# write_rds(summaries_df, here("FigureData", "permanova_results_08.rds"))

summaries_df <- read_rds(here("FigureData", "permanova_results_08.rds"))

summaries_df %>%
  pander()

```

## Permanova with log_AbxHoursPerHOL

```{r, results='asis'}

# summaries_df <- 
#   s2 %>%
#     do(adonisplus(
#       ., distmat = bc, formula = distmat ~ bwt_z + log_AbxHoursPerHOL + growth_failure * WOL,
#       sample_id_var = SampleID, rep_meas_var = SubjectID,
#       shuffle = c(WOL = "within", growth_failure = "between", log_AbxHoursPerHOL = "within"), permutations=perm)) %>%
#   mutate(metric = "Bray-Curtis") %>%
#   filter(!term %in% c("Residuals", "Total")) %>%
#   select(-one_of(c("sumsq", "meansq"))) %>%
#   select(metric, everything())
# 
# write_rds(summaries_df, here("FigureData", "permanova_results_09.rds"))

summaries_df <- read_rds(here("FigureData", "permanova_results_09.rds"))

summaries_df %>%
  pander()

```

# SUPP FIGURE 4E - gene pathways abundance - growth faltering

### Linear models of pathways

```{r, results='asis'}

number_of_top_pathways = 25

props_toTest <- kegg_path_props %>%
  mutate(props_log10 = log10(props)) %>%
  mutate(PathwayName = fct_lump(PathwayName, n = number_of_top_pathways, w = .$props, other_level = "Other")) %>%
  filter(!PathwayName %in% c("Metabolic pathways", "Other"))

form1 = "props_log10 ~ bwt_z + WOL * growth_cat + log_npo_days_prop"
random1 = "SubjectID"

summaries_df <- run_lmer_on_props(factor_to_test = "PathwayName", props_toTest = props_toTest, s_toTest = s2, form1 = form1, rep_mes_label = random1, p_cutoff =  0.05) 

summaries_df %>%
  mutate(term = str_remove(term, "growth_cat")) %>%
  pander(split.table=Inf, digits=2, caption = glue("Linear models of log10 transformed abundances of KEGG pathways"), emphasize.strong.rows = which(.$fdr<=0.05))

```

```{r}

pathways_to_select <- summaries_df %>%
  mutate(term = str_remove(term, "growth_cat")) %>%
  filter(term %in% c("moderate to severe retardation", "mild retardation")) %>%
  pull(PathwayName) %>%
  unique() %>%
  as.character()

plot_df <- props_toTest %>%
  filter(PathwayName %in% pathways_to_select) %>%
  left_join(s2, join_by(SampleID))

g <- ggplot(plot_df, aes(x = props, y = PathwayName, color = growth_cat)) +
  theme_bw() +
  theme(legend.position = "bottom") +
  geom_boxplot(outlier.shape = 1, outlier.alpha = 0.5) +
  scale_color_manual(values=ann_colors$growth_cat) +
  labs(y = "KEGG pathway",
       x = "Relative abundance",
       color = "")

plot(g)

ggsave(here("Output", glue("SUPP_FIGURE_4E_growth_falt_kegg_pathway_abundance.pdf")), height = 2.75, width = 5)

```

# FIGURE 2B - gene pathways linear estimate of differences

```{r}

summaries_df <- run_lmer_on_props(factor_to_test = "PathwayName", props_toTest = props_toTest, s_toTest = s2, form1 = form1,rep_mes_label = random1, p_cutoff =  1)

plot_df <- summaries_df %>%
  mutate(term = str_remove(term, "growth_cat")) %>%
  filter(term %in% c("moderate to severe retardation", "mild retardation"), PathwayName %in% pathways_to_select) %>%
  mutate(CI95_lwr = round(Value - 1.96*Std.Error, 2)) %>%
  mutate(CI95_upr = round(Value + 1.96*Std.Error, 2)) %>%
  mutate(Significant = case_when(fdr < 0.05 ~ "p.adj < 0.05",
                                 fdr > 0.05 ~ "Not significant")) %>%
  arrange(Value) %>%
  mutate(PathwayName = fct_reorder(PathwayName, -Value))

g <- ggplot(plot_df, aes(Value, PathwayName, shape = Significant, color = term)) +
  theme_bw() +
  theme(legend.position = "bottom", legend.justification.bottom = "left", legend.direction = "vertical", legend.byrow = T) +
  geom_pointrange(aes(xmin=Value-Std.Error, xmax=Value+Std.Error), position = position_dodge(width = 0.4)) +
  geom_vline(xintercept = 0, linetype = "longdash") +
  scale_shape_manual(values = c(1,19)) +
  scale_color_manual(values=ann_colors$growth_cat[2:3]) +
  labs(y = "KEGG Pathway",
       x = "",
       shape = "Significance",
       color = "Growth faltering")
# <-- Decreased vs. normal growth    Increased vs. normal growth -->
plot(g)

ggsave(here("Output", glue("FIGURE_2B_growth_falt_kegg_pathway_difference.pdf")), height = 2.5, width = 5)

pathways_from_fig_2B <- as.character(unique(plot_df$PathwayName))

```

# SUPP FIGURE 4F - gene pathways abundance - growth failure

### Linear models of pathways

```{r, results='asis'}

number_of_top_pathways = 25

props_toTest <- kegg_path_props %>%
  mutate(props_log10 = log10(props)) %>%
  mutate(PathwayName = fct_lump(PathwayName, n = number_of_top_pathways, w = .$props, other_level = "Other")) %>%
  filter(!PathwayName %in% c("Metabolic pathways", "Other"))

form1 = "props_log10 ~ bwt_z + WOL * growth_failure + log_npo_days_prop"
random1 = "SubjectID"

summaries_df <- run_lmer_on_props(factor_to_test = "PathwayName", props_toTest = props_toTest, s_toTest = s2, form1 = form1, rep_mes_label = random1, p_cutoff =  0.05) 

summaries_df %>%
  mutate(term = str_remove(term, "growth_failure")) %>%
  pander(split.table=Inf, digits=2, caption = glue("Linear models of log10 transformed abundances of KEGG pathways"), emphasize.strong.rows = which(.$fdr<=0.05))

```

```{r}

pathways_to_select <- pathways_from_fig_2B

plot_df <- props_toTest %>%
  filter(PathwayName %in% pathways_to_select) %>%
  left_join(s2, join_by(SampleID))

g <- ggplot(plot_df, aes(x = props, y = PathwayName, color = growth_failure)) +
  theme_bw() +
  theme(legend.position = "bottom") +
  geom_boxplot(outlier.shape = 1, outlier.alpha = 0.5) +
  scale_color_manual(values=ann_colors$growth_fail) +
  labs(y = "KEGG pathway",
       x = "Relative abundance",
       color = "")

plot(g)

ggsave(here("Output", glue("SUPP_FIGURE_4F_growth_failure_kegg_pathway_abundance.pdf")), height = 2.75, width = 5)

```

# FIGURE 2C - gene pathways linear estimate of differences

```{r}

summaries_df <- run_lmer_on_props(factor_to_test = "PathwayName", props_toTest = props_toTest, s_toTest = s2, form1 = form1,rep_mes_label = random1, p_cutoff =  1)

plot_df <- summaries_df %>%
  mutate(term = str_remove(term, "growth_failure")) %>%
  filter(term %in% c("Yes")) %>%
  filter(PathwayName %in% pathways_to_select) %>%
  mutate(CI95_lwr = round(Value - 1.96*Std.Error, 2)) %>%
  mutate(CI95_upr = round(Value + 1.96*Std.Error, 2)) %>%
  mutate(Significant = case_when(fdr < 0.05 ~ "p.adj < 0.05",
                                 fdr > 0.05 ~ "Not significant")) %>%
  arrange(Value) %>%
  mutate(PathwayName = fct_reorder(PathwayName, -Value))

g <- ggplot(plot_df, aes(Value, PathwayName, shape = Significant, color = term)) +
  theme_bw() +
  theme(legend.position = "bottom", legend.justification.bottom = "left", legend.direction = "vertical", legend.byrow = T) +
  geom_pointrange(aes(xmin=Value-Std.Error, xmax=Value+Std.Error), position = position_dodge(width = 0.4)) +
  geom_vline(xintercept = 0, linetype = "longdash") +
  scale_shape_manual(values = c(1,19)) +
  scale_color_manual(values=ann_colors$growth_fail) +
  labs(y = "KEGG Pathway",
       x = "Change vs. normal growth",
       shape = "Significance",
       color = "Growth failure")
# <-- Decreased vs. normal growth    Increased vs. normal growth -->
plot(g)

ggsave(here("Output", glue("FIGURE_2C_growth_failure_kegg_pathway_difference.pdf")), height = 2.5, width = 5)

```

# FIGURE 2A - just time and top 500 KEGG

Just with time variable

Taking the top 500 KEGG orthologs and testing them

```{r message=FALSE, warning=FALSE, paged.print=TRUE, results='asis'}

kegg_defnitions <- read_tsv(here("Data", "ko_definition.list"), col_names = c("Ortholog", "Definition")) %>%
  mutate(Ortholog = str_remove(Ortholog, "ko:"))

path_annotation <- prokarotic_paths %>%
  select(Ortholog, PathwayName) %>%
  group_by(Ortholog) %>%
  summarise(Pathways = paste0(PathwayName, collapse = ", ")) %>%
  ungroup()

top500 <- kegg_props %>%
  filter(SampleID %in% s2$SampleID) %>%
  group_by(geneID) %>%
  summarise(mean_props = mean(props)) %>%
  arrange(desc(mean_props)) %>%
  slice(1:500)

kegg_props_subset <- kegg_props %>%
  select(SampleID, Ortholog = geneID, count, props) %>%
  filter(Ortholog %in% top500$geneID) %>%
  mutate(props_log10 = log10(props))

form1 = "props_log10 ~ WOL"
random1 = "SubjectID"

# summaries_df <- run_lmer_on_props(factor_to_test = "Ortholog", props_toTest = kegg_props_subset, s_toTest = s2, form1 = form1, p_cutoff = 1, rep_mes_label = random1)
# 
# save(summaries_df, file = here("Data", glue("just_time_kegg500_changes.RDS")))

load(file = here("Data", glue("just_time_kegg500_changes.RDS")))
  
plot_df <- summaries_df %>%
  mutate(Significant = case_when(fdr < 0.05 ~ "p.adj < 0.05",
                                 fdr > 0.05 ~ "Not significant")) %>%
  mutate(direction = case_when(Value > 0 ~ "Increased",
                               Value < 0 ~ "Decreased")) %>%
  left_join(kegg_defnitions, by = "Ortholog") %>%
  tidyr::separate_wider_delim(cols = Definition, delim = ";", names = c("gene", "description")) %>%
  mutate(ortho_lable = paste(Ortholog, gene))

library(ggrepel)

g <- ggplot(data=plot_df, aes(x=Value, y=-log10(fdr), col=direction, label = ortho_lable)) + 
  geom_point() +
  # geom_text_repel() +
  theme_bw() +
  geom_hline(yintercept=-log10(0.05), linetype = "longdash") +
  scale_shape_manual(values = c(1,19)) +
  # theme(legend.position = "none") +
  labs(y = "-log10(fdr value)",
       x = "Estimate of linear model",
       color = "Change over time",
       subtitle = "")
#  (formula: log10(props) ~ WOL)
#Dashed line represents fdr = 0.05

plot(g)

ggsave(here("Output", glue("FIGURE_2A_top_500_kegg_orthologs_just_time.pdf")), height = 2.25, width = 4)

table_name <- "FIGURE_2A_top_500_kegg_orthologs_just_time.csv"

my_table <- summaries_df %>%
  left_join(kegg_defnitions, by = "Ortholog") %>%
  left_join(path_annotation, by = "Ortholog")

write_csv(my_table, here("Output", table_name))

print(glue("All orthologs saved together as {table_name}"))

```

# SUPP FIGURE 4C - Gfalt and Kegg terms

FDR plot for growth faltering

## With log_npo_days_prop

```{r message=FALSE, warning=FALSE, paged.print=TRUE, results='asis'}

form1 = "props_log10 ~ bwt_z + WOL * growth_cat + log_npo_days_prop"
random1 = "SubjectID"

summaries_df <- run_lmer_on_props(factor_to_test = "Ortholog", props_toTest = kegg_props_subset, s_toTest = s2, form1 = form1, p_cutoff = 1, rep_mes_label = random1)

to_select <- summaries_df %>%
  mutate(term = str_remove(term, "growth_cat")) %>%
  filter(term %in% c("moderate to severe retardation", "mild retardation"), fdr < 0.05) %>%
  pull(Ortholog) %>%
  as.character() %>%
  unique()

changed_in_gfalt <- to_select
  
plot_df <- summaries_df %>%
  mutate(term = str_remove(term, "growth_cat")) %>%
  filter(term %in% c("moderate to severe retardation", "mild retardation"), Ortholog %in% to_select) %>%
  left_join(kegg_defnitions, by = "Ortholog") %>%
  mutate(ortho_lable = paste(Ortholog, str_trunc(Definition, 30, "right"))) %>%
  mutate(CI95_lwr = round(Value - 1.96*Std.Error, 2)) %>%
  mutate(CI95_upr = round(Value + 1.96*Std.Error, 2)) %>%
  mutate(Significant = case_when(fdr < 0.05 ~ "p.adj < 0.05",
                                 fdr > 0.05 ~ "Not significant")) %>%
  arrange(Value) %>%
  mutate(ortho_lable = fct_reorder(ortho_lable, Value))

g <- ggplot(plot_df, aes(Value, ortho_lable, color = term, shape = Significant)) +
  theme_bw() +
  geom_pointrange(aes(xmin=Value-Std.Error, xmax=Value+Std.Error), position = position_dodge(width = 0.4)) +
  geom_vline(xintercept = 0, linetype = "longdash") +
  scale_shape_manual(values = c(1,19)) +
  scale_color_manual(values=ann_colors$growth_cat[2:3]) +
    labs(y = "Ortholog / Description",
       x = "<-- Decreased vs. normal growth    Increased vs. normal growth -->",
       shape = "Significance",
       color = "")

plot(g)

ggsave(here("Output", glue("SUPP_FIGURE_4C_growth_faltering_kegg500.pdf")), height = 2.5, width = 6.5)

table_name <- glue("SUPP_FIGURE_4C_growth_faltering_kegg500.csv")

my_table <- summaries_df %>%
  mutate(term = str_remove(term, "growth_cat")) %>%
  filter(term %in% c("moderate to severe retardation", "mild retardation"), fdr < 0.05) %>%
  left_join(kegg_defnitions, by = "Ortholog") %>%
  left_join(path_annotation, by = "Ortholog")

write_csv(my_table, here("Output", table_name))

print(glue("All orthologs saved together as {table_name}"))

```




# SUPP FIGURE 4D - Gfailure and Kegg terms

FDR plot for growth failure

## With log_npo_days_prop

```{r message=FALSE, warning=FALSE, paged.print=TRUE, results='asis'}

form1 = "props_log10 ~ bwt_z + WOL * growth_failure + log_npo_days_prop"
random1 = "SubjectID"

summaries_df <- run_lmer_on_props(factor_to_test = "Ortholog", props_toTest = kegg_props_subset, s_toTest = s2, form1 = form1, p_cutoff = 1, rep_mes_label = random1)
  
plot_df <- summaries_df %>%
  mutate(term = str_remove(term, "growth_failure")) %>%
  filter(term %in% c("Yes"), Ortholog %in% c(changed_in_gfalt, "K00868")) %>%
  left_join(kegg_defnitions, by = "Ortholog") %>%
  mutate(ortho_lable = paste(Ortholog, str_trunc(Definition, 30, "right"))) %>%
  mutate(CI95_lwr = round(Value - 1.96*Std.Error, 2)) %>%
  mutate(CI95_upr = round(Value + 1.96*Std.Error, 2)) %>%
  mutate(Significant = case_when(fdr < 0.05 ~ "p.adj < 0.05",
                                 fdr > 0.05 ~ "Not significant")) %>%
  arrange(Value) %>%
  mutate(ortho_lable = fct_reorder(ortho_lable, Value))

g <- ggplot(plot_df, aes(Value, ortho_lable, color = term, shape = Significant)) +
  theme_bw() +
  geom_pointrange(aes(xmin=Value-Std.Error, xmax=Value+Std.Error), position = position_dodge(width = 0.4)) +
  geom_vline(xintercept = 0, linetype = "longdash") +
  scale_shape_manual(values = c(1,19)) +
  scale_color_manual(values=ann_colors$growth_fail) +
    labs(y = "Ortholog / Description",
       x = "<-- Decreased vs. normal growth    Increased vs. normal growth -->",
       shape = "Significance",
       color = "")

plot(g)

ggsave(here("Output", glue("SUPP_FIGURE_4D_growth_failure_kegg500.pdf")), height = 2.5, width = 6.5)

table_name <- glue("SUPP_FIGURE_4D_growth_failure_kegg500.csv")

my_table <- summaries_df %>%
  mutate(term = str_remove(term, "growth_failure")) %>%
  filter(term %in% c("Yes"), Ortholog %in% changed_in_gfalt) %>%
  left_join(kegg_defnitions, by = "Ortholog") %>%
  left_join(path_annotation, by = "Ortholog")

write_csv(my_table, here("Output", table_name))

print(glue("All orthologs saved together as {table_name}"))

```

```{r}

#all the orthologs
actually_changed <- summaries_df %>%
  filter(fdr < 0.05, term %in% "growth_failureYes")
#K00868

```

# SUPP FIGURE 4A - venn diagram of kegg genes in growth faltering

```{r}

library(ggVennDiagram)

kegg_pres_abs <- kegg_props_raw %>%
  left_join(s2 %>% select(SampleID, growth_cat), join_by(SampleID)) %>%
  group_by(growth_cat, geneID) %>%
  summarise(present = ifelse(sum(count) > 0, "Present", "Absent"))

venn_list <- list("Normal" = kegg_pres_abs %>% filter(growth_cat %in% "normal", present %in% "Present") %>% pull(geneID),
                "Mild retardation" = kegg_pres_abs %>% filter(growth_cat %in% "mild retardation", present %in% "Present") %>% pull(geneID),
                "Moderate to severe retardation" = kegg_pres_abs %>% filter(growth_cat %in% "moderate to severe retardation", present %in% "Present") %>% pull(geneID))

ggVennDiagram(venn_list) +
  scale_x_continuous(expand = expansion(mult = .2)) +
  scale_fill_distiller(palette = "RdBu")

ggsave(here("Output", "SUPP FIGURE 4A kegg genes shared among growth faltering.pdf"), height = 5, width = 6)

```

# SUPP FIGURE 4B - venn diagram of kegg genes in growth failure

```{r}

kegg_pres_abs <- kegg_props_raw %>%
  left_join(s2 %>% select(SampleID, growth_failure), join_by(SampleID)) %>%
  group_by(growth_failure, geneID) %>%
  summarise(present = ifelse(sum(count) > 0, "Present", "Absent"))

venn_list <- list("Yes" = kegg_pres_abs %>% filter(growth_failure %in% "Yes", present %in% "Present") %>% pull(geneID),
                "No" = kegg_pres_abs %>% filter(growth_failure %in% "No", present %in% "Present") %>% pull(geneID))

ggVennDiagram(venn_list) +
  scale_x_continuous(expand = expansion(mult = .2)) +
  scale_fill_distiller(palette = "RdBu")

ggsave(here("Output", "SUPP FIGURE 4B kegg genes shared among growth failure.pdf"), height = 5, width = 6)

```

# Random Forests

```{r load stuff}
library(Boruta)
library(mlbench)
library(caret)
library(pROC)

load(file = here("Data", "random_forest_models.RData"))

```

```{r}

top25 <- summed_props_long %>%
  group_by(Taxa) %>%
  summarise(mean_props = mean(props)) %>%
  ungroup() %>%
  arrange(desc(mean_props)) %>%
  slice(1:29) %>%
  filter(!str_detect(Taxa, "Ralstonia|Cutibacterium|Vibrio"))

training_data <- s2 %>%
  filter(growth_failure %in% "No") %>%
  left_join(summed_props_long, by = "SampleID") %>%
  filter(Taxa %in% top25$Taxa) %>%
  select(SampleID, DOL, Taxa, props) %>% #SubjectID, 
  mutate(Taxa = str_replace_all(Taxa, " ", "_")) %>%
  pivot_wider(names_from = "Taxa", values_from = "props", values_fill = 0) %>%
  column_to_rownames(var = "SampleID")

set.seed(43)
validation_babies <- sample(s2$SubjectID, 15)
validation_samples <- s2 %>%
  filter(growth_failure %in% "No") %>%
  filter(SubjectID %in% validation_babies) %>%
  pull(SampleID)

validation_data <- training_data[validation_samples,]
ind2 <- setdiff(rownames(training_data), rownames(validation_data))
training_data <- training_data[ind2,]

testing_data <- s2 %>%
  filter(growth_failure %in% "Yes") %>%
  left_join(summed_props_long, by = "SampleID") %>%
  filter(Taxa %in% top25$Taxa) %>%
  select(SampleID, DOL, Taxa, props) %>% #SubjectID, 
  mutate(Taxa = str_replace_all(Taxa, " ", "_")) %>%
  pivot_wider(names_from = "Taxa", values_from = "props", values_fill = 0) %>%
  column_to_rownames(var = "SampleID")

#sanity check, make sure same baby is not in both validation and training
# samples_in_training <- rownames(training_data)
# training_babies <- s2 %>%
#   filter(SampleID %in% samples_in_training) %>%
#   pull(SubjectID)
# intersect(training_babies, validation_babies) #should be "character(0)"

```

# FIGURE 3A - rf model with 25 taxa

```{r}

rf_model_improved <- rf_model_25taxa

p2 <- predict(rf_model_improved, testing_data)

plot_df <- tibble(class = "Regular growth", MMI = as.vector(rf_model_improved$predicted), actual_DOL = as.vector(rf_model_improved$y), SampleID = rownames(training_data)) %>%
  bind_rows(tibble(class = "Growth failure", MMI = as.vector(p2), actual_DOL = testing_data$DOL, SampleID = rownames(testing_data))) %>%
  mutate(class = fct_relevel(class, "Regular growth"))

growth_class_colors <- c("Regular growth" = "#9FCB6A", "Growth failure" = "#8367CC")

# mod2 = nls(MMI ~ SSlogis(actual_DOL, Asym, xmid, scal), data=plot_df %>% filter(class %in% "Growth failure"))
# mod2

# mod2 = nls(MMI ~ SSfpl(actual_DOL, A, B, xmid, scal), data=plot_df %>% filter(class %in% "Growth failure"))
# weird error
# Error in nls(y ~ cbind(1, 1/(1 + exp((xmid - x)/exp(lscal)))), data = xy,  : 
#   step factor 0.000488281 reduced below 'minFactor' of 0.000976562
# mod2

for_save <- plot_df %>%
  ggplot(mapping = aes(x=actual_DOL, y=MMI, color=class)) +
  theme_bw() +
  geom_point() +
  # annotate(geom="text", x=10, y=70, label=glue("Logistic curve for growth failure:\n Horiz. asymptote={round(mod2$m$getPars()['Asym'],2)}\n  xmid={round(mod2$m$getPars()['xmid'],2)}\n  scale={round(mod2$m$getPars()['scal'],2)}"), color = "#AD8152FF", hjust = 0) +
  geom_smooth(
    method = "nls", formula = y ~ SSlogis(x, Asym, xmid, scal),
    se = FALSE
  ) +
  scale_color_manual(values = growth_class_colors) +
  labs(
    y = "Predicted DOL / microbiome maturity index (MMI)",
    x = "Actual DOL",
    color = "Class",
    title = "",
    subtitle = ""
  )

ggsave(plot = for_save, here("Output", "FIGURE 3A random forest prediction for both growth legend.pdf"), height = 3.5, width = 5)

for_save + theme(legend.position = "none")

ggsave(here("Output", "FIGURE 3A random forest prediction for both growth no legend.pdf"), height = 5.5, width = 10)

```

# FIGURE 3B - RMMI reg. growth vs. GF vs. validation

```{r, fig.width=5, fig.height=3}

# mod1 is the SSlogis model for class %in% "Regular growth"
# mod2 is the SSlogis model for class %in% "Growth failure"

reg_growth_df <- tibble(class = "Regular growth", MMI = as.vector(rf_model_improved$predicted), actual_DOL = as.vector(rf_model_improved$y))

mod1 = nls(MMI ~ SSlogis(actual_DOL, Asym, xmid, scal), data=reg_growth_df)

p3 <- predict(rf_model_improved, validation_data)

plot_df <- plot_df %>%
  bind_rows(tibble(class = "Validation", MMI = as.vector(p3), actual_DOL = validation_data$DOL, SampleID = rownames(validation_data))) %>%
  mutate(class = fct_relevel(class, "Training"))

plot_df$curve_value <- predict(mod1, newdata = plot_df)

plot_df %<>%
  mutate(RMMI = MMI - curve_value) %>%
  mutate(class = fct_relevel(class, "Regular growth", "Growth failure"))

plot_df %>%
  ggplot(aes(x = class, y = RMMI, color = class)) +
  geom_boxplot(outlier.alpha = 0) +
  geom_quasirandom(dodge.width = 0.75, alpha = 0.25) +
  theme_bw() +
  scale_color_manual(values = growth_class_colors) +
  theme(legend.position = "") +
  labs(x = "Class", y = "Relative microbiota maturity index")

ggsave(here("Output", "FIGURE 3B RMMI comparison.pdf"), height = 3, width = 4)

```

# TABLE 3C - Mixed-effects model for RMMI

```{r, results='asis'}

test_df <- plot_df %>%
  left_join(s2 %>% select(SampleID, SubjectID), join_by(SampleID))

summary_df <- tidy_lmer(nlme::lme(fixed = RMMI ~ class, random = ~ 1 | SubjectID, data = test_df)) 

summary_df %>% pander(digits=2, split.cells=15)

summary_df %>% write_csv(here("Output", "FIGURE 3B summary table.csv"))

```

# FIGURE 3C Top five features for each model

```{r, fig.height=4, fig.width=8}

imp <- importance(rf_model_25taxa, type = 1)
imp_df_25taxa <- tibble(Feature = dimnames(imp)[[1]], MeanDecAccu = as.vector(imp)) %>%
  mutate(model = "25 taxa")
imp <- importance(rf_model_improved_78taxa, type = 1)
imp_df_improved_78taxa <- tibble(Feature = dimnames(imp)[[1]], MeanDecAccu = as.vector(imp)) %>%
  mutate(model = "71 taxa, boruta optimized")
imp <- importance(rf_model_just_kegg, type = 1)
imp_df_just_kegg <- tibble(Feature = dimnames(imp)[[1]], MeanDecAccu = as.vector(imp)) %>%
  mutate(model = "165 KEGG genes, boruta optimized") %>%
  left_join(kegg_defnitions, by = c("Feature"="Ortholog")) %>%
  mutate(Feature = paste(Feature, str_trunc(Definition, 20, "right")))

imp_df <- bind_rows(imp_df_25taxa, imp_df_improved_78taxa, imp_df_just_kegg) %>%
  group_by(model) %>%
  slice_max(n = 5, order_by = MeanDecAccu) %>%
  ungroup() %>%
  droplevels() %>%
  mutate(Feature = str_remove(Feature, "[a-zA-Z]+?_")) %>%
  mutate(Feature = str_replace(Feature, "_", " ")) %>%
  mutate(Feature = fct_reorder(Feature, MeanDecAccu)) %>%
  mutate(model = fct_relevel(model, "25 taxa", "71 taxa, boruta optimized"))

imp_df %>%
  ggplot(aes(x = MeanDecAccu, y = Feature)) +
  facet_wrap(~model, ncol = 3, scales = "free_x", strip.position = "right") +
  # facet_wrap(~model, ncol = 3, scales = "free_x", labeller = labeller(group = label_wrap_gen(width = 25))) +
  geom_point() +
  theme_bw() +
  theme(axis.text.x = element_text(angle = 45, hjust = 1, vjust = 1)) +
  labs(x = "Mean decrease in accuracy",
       y = "Top 5 features in models")

ggsave(here("Output", "FIGURE 3C Top 5 features.pdf"), height = 3, width = 5.5)

```

# FIGURE 3D

```{r}

roc25taxa <- plot_df_25taxa %>%
  mutate(class = fct_relevel(class, "Regular growth")) %>%
  roc(class ~ RMMI, data = .)

roc78taxa <- plot_df_78taxa %>%
  mutate(class = fct_relevel(class, "Regular growth")) %>%
  roc(class ~ RMMI, data = .)

rockegg <- plot_df_kegg %>%
  mutate(class = fct_relevel(class, "Regular growth")) %>%
  roc(class ~ RMMI, data = .)

ggroc(list(`25 taxa`=roc25taxa, `71 taxa`=roc78taxa, `165 KEGG genes`=rockegg)) +
  theme_bw() +
  labs(title = "", color = "Model", y = "Sensitivity", x = "Specificity") +
  geom_abline(slope = 1, intercept = c(1,0), lty = 2) +
  scale_colour_aaas()

ggsave(filename = here::here("Output","FIGURE 3D ROC curves.pdf"), device = "pdf", height = 3, width = 4)

```

# Supp Figure 3A - alpha diversity 0-21 day gfalt

```{r fig.height=6, fig.width=8}

s_toPlot <- s_toTest <- s2 %>%
  filter(DOL < 22) %>%
  droplevels()

# WOL_colors <- setNames((viridis(length(levels(s2$WOL_label)), alpha = 0.8)), levels(s2$WOL_label))

WOL_colors <- c(`1` = "#440154CC", `2` = "#481A6CCC", `3` = "#472F7DCC", `4` = "#414487CC")

s_toPlot %>%
  pivot_longer(c("Richness", "Shannon"), names_to="metric", values_to="alpha") %>%
  filter(metric %in% "Shannon") %>%
  mutate(growth_label = case_when(growth_cat %in% "normal" ~ "None",
                                  growth_cat %in% "mild retardation" ~ "Mild",
                                  growth_cat %in% "moderate to severe retardation" ~ "Moderate\nto severe")) %>%
  mutate(growth_label = fct_relevel(growth_label, "None", "Mild")) %>%
  ggplot(aes(x=growth_label, y=alpha, fill=WOL_label)) +
  geom_boxplot() + #outlier.alpha=0
  scale_fill_manual(values = WOL_colors) +
  # facet_grid(metric~., scales = "free") +
  theme_bw() +
  theme(
    strip.background = element_blank(),
    panel.grid= element_blank(),
    # legend.position = "none",
    axis.text.x = element_text(angle = 45, hjust = 1, vjust = 1)
  ) +
  labs(
    x="Growth faltering", fill="Week of life",
    y="Shannon diversity"
  )

ggsave(here("Output", "SUPP_FIG3A_alphadiv_gfalt.pdf"), width = 3.5, height = 2.5, useDingbats = F)

```

## Linear model with log_npo_days_prop

```{r, results='asis'}
my_formula <- "alpha ~ bwt_z + WOL * growth_cat + log_npo_days_prop"

summaries_df <- s_toTest %>%
  pivot_longer(c("Richness", "Shannon"), names_to="metric", values_to="alpha") %>%
  group_by(metric) %>%
  do(tidy_lmer(nlme::lme(as.formula(my_formula), random=~1|SubjectID, data=., na.action=na.omit))) %>%
  ungroup() %>%
  filter(!grepl("Intercept", term)) %>%
  mutate(term = str_replace(term, "timepoint_numeric", "Time")) %>%
  mutate(term = if_else(str_detect(term, "growth_cat"), str_remove(term, "growth_cat"), term))

summaries_df %>%
  pander(digits = 2, split.table = Inf, split.cells = 20, caption = glue("Linear model of the alpha diversity measurement with numeric time and growth categories as the predictive variables. SubjectID added as a random effect. Full formula: {my_formula}"))
```

# Supp Figure 3B - beta diversity 0-21 day gfalt

```{r fig.height=3, fig.width=5}

s_toPlot %>%
  mutate(growth_label = case_when(growth_cat %in% "normal" ~ "None",
                                  growth_cat %in% "mild retardation" ~ "Mild",
                                  growth_cat %in% "moderate to severe retardation" ~ "Moderate\nto severe")) %>%
  mutate(growth_label = fct_relevel(growth_label, "None", "Mild")) %>%
  pcoaplus(bc) %>%
  plot(color=growth_label) +
    scale_color_manual(values=ann_colors$growth_label) +
    theme_bw() +
    theme(
      axis.text=element_blank(),
      axis.ticks=element_blank(),
      panel.grid = element_blank(), 
      aspect.ratio = 1,
      strip.background = element_blank()
    ) + 
    labs(color="Growth faltering", title="")

ggsave(here("Output", "SUPP_FIG3B_betadiv.pdf"), width = 3.5, height = 2.5, useDingbats = F)

```

## Permanova with log_npo_days_prop

```{r, results='asis'}

# summaries_df <-
#   s_toTest %>%
#     do(adonisplus(
#       ., distmat = bc, formula = distmat ~ bwt_z + log_npo_days_prop + growth_cat * WOL,
#       sample_id_var = SampleID, rep_meas_var = SubjectID,
#       shuffle = c(WOL = "within", growth_cat = "between", log_npo_days_prop = "within"), permutations=perm)) %>%
#   mutate(metric = "Bray-Curtis") %>%
#   filter(!term %in% c("Residuals", "Total")) %>%
#   select(-one_of(c("sumsq", "meansq"))) %>%
#   select(metric, everything())
# 
# write_rds(summaries_df, here("FigureData", "permanova_results_10.rds"))

summaries_df <- read_rds(here("FigureData", "permanova_results_10.rds"))

summaries_df %>%
  pander()

```

# Supp Figure 3C - alpha diversity 0-21 days gfail

```{r}

s_toPlot %>%
  pivot_longer(c("Richness", "Shannon"), names_to="metric", values_to="alpha") %>%
  filter(metric %in% "Shannon") %>%
  ggplot(aes(x=growth_failure, y=alpha, fill=WOL_label)) +
  geom_boxplot() + #outlier.alpha=0
  scale_fill_manual(values = WOL_colors) +
  # facet_grid(metric~., scales = "free") +
  theme_bw() +
  theme(
    strip.background = element_blank(),
    panel.grid=element_blank(),
    # legend.position = "none",
    axis.text.x = element_text(angle = 45, hjust = 1, vjust = 1)
  ) +
  labs(
    x="Growth failure", fill="Week of life",
    y="Shannon diversity"
  )

ggsave(here("Output", "SUPP_FIG3C_alphadiv_gfail.pdf"), width = 3.5, height = 2.5, useDingbats = F)

```

## Linear model with log_npo_days_prop

```{r, results='asis'}
my_formula <- "alpha ~ bwt_z + WOL * growth_failure + log_npo_days_prop"

summaries_df <- s_toTest %>%
  pivot_longer(c("Richness", "Shannon"), names_to="metric", values_to="alpha") %>%
  group_by(metric) %>%
  do(tidy_lmer(nlme::lme(as.formula(my_formula), random=~1|SubjectID, data=., na.action=na.omit))) %>%
  ungroup() %>%
  filter(!grepl("Intercept", term)) %>%
  mutate(term = str_replace(term, "timepoint_numeric", "Time")) %>%
  mutate(term = if_else(str_detect(term, "growth_cat"), str_remove(term, "growth_cat"), term))

summaries_df %>%
  pander(digits = 2, split.table = Inf, split.cells = 20, caption = glue("Linear model of the alpha diversity measurement with numeric time and growth_failure as the predictive variables. SubjectID added as a random effect. Full formula: {my_formula}"))
```

# Supp Figure 3D - beta diversity 0-21 days gfail

```{r}
s_toPlot %>%
  pcoaplus(bc) %>%
  plot(color=growth_failure) +
    scale_color_manual(values=ann_colors$growth_fail) +
    theme_bw() +
    theme(
      axis.text=element_blank(),
      axis.ticks=element_blank(),
      panel.grid = element_blank(), 
      aspect.ratio = 1,
      strip.background = element_blank()
    ) + 
    labs(color="Growth failure", title="")

ggsave(here("Output", "SUPP_FIG3D_betadiv.pdf"), width = 3, height = 2.5, useDingbats = F)
```

## Permanova with log_npo_days_prop

```{r, results='asis'}

# summaries_df <-
#   s_toTest %>%
#     do(adonisplus(
#       ., distmat = bc, formula = distmat ~ bwt_z + log_npo_days_prop + growth_failure * WOL,
#       sample_id_var = SampleID, rep_meas_var = SubjectID,
#       shuffle = c(WOL = "within", growth_failure = "between", log_npo_days_prop = "within"), permutations=perm)) %>%
#   mutate(metric = "Bray-Curtis") %>%
#   filter(!term %in% c("Residuals", "Total")) %>%
#   select(-one_of(c("sumsq", "meansq"))) %>%
#   select(metric, everything())
# 
# write_rds(summaries_df, here("FigureData", "permanova_results_11.rds"))

summaries_df <- read_rds(here("FigureData", "permanova_results_11.rds"))

summaries_df %>%
  pander()

```

# Supp Figure 3E - alpha diversity 34-37 weeks PMA gfalt

```{r fig.height=6, fig.width=8}

s_subset2 <- read_rds(here("Data", "subset_of_babies_great34wk_less37wk.rds"))

s_toPlot <- s_toTest <- s2 %>%
  filter(SampleID %in% s_subset2$SampleID) %>%
  droplevels()

# WOL_colors <- setNames((viridis(length(levels(s2$WOL_label)), alpha = 0.8)), levels(s2$WOL_label))

WOL_colors <- c(`3` = "#472F7DCC", `4` = "#414487CC", 
                `5` = "#39568CCC", `6` = "#31688ECC", `7` = "#2A788ECC", `8` = "#23888ECC", 
                `9` = "#1F988BCC", `10` = "#22A884CC", `11` = "#35B779CC", `12` = "#54C568CC", 
                `13` = "#7AD151CC"
)

s_toPlot %>%
  pivot_longer(c("Richness", "Shannon"), names_to="metric", values_to="alpha") %>%
  filter(metric %in% "Shannon") %>%
  mutate(growth_label = case_when(growth_cat %in% "normal" ~ "None",
                                  growth_cat %in% "mild retardation" ~ "Mild",
                                  growth_cat %in% "moderate to severe retardation" ~ "Moderate\nto severe")) %>%
  mutate(growth_label = fct_relevel(growth_label, "None", "Mild")) %>%
  ggplot(aes(x=growth_label, y=alpha, fill=WOL_label)) +
  geom_boxplot() + #outlier.alpha=0
  scale_fill_manual(values = WOL_colors) +
  # facet_grid(metric~., scales = "free") +
  theme_bw() +
  theme(
    strip.background = element_blank(),
    panel.grid= element_blank(),
    # legend.position = "none",
    axis.text.x = element_text(angle = 45, hjust = 1, vjust = 1)
  ) +
  labs(
    x="Growth faltering", fill="Week of life",
    y="Shannon diversity"
  )

ggsave(here("Output", "SUPP_FIG3E_alphadiv_gfalt.pdf"), width = 3.5, height = 2.5, useDingbats = F)

```

## Linear model with log_npo_days_prop

```{r, results='asis'}
my_formula <- "alpha ~ bwt_z + WOL * growth_cat + log_npo_days_prop"

summaries_df <- s_toTest %>%
  pivot_longer(c("Richness", "Shannon"), names_to="metric", values_to="alpha") %>%
  group_by(metric) %>%
  do(tidy_lmer(nlme::lme(as.formula(my_formula), random=~1|SubjectID, data=., na.action=na.omit))) %>%
  ungroup() %>%
  filter(!grepl("Intercept", term)) %>%
  mutate(term = str_replace(term, "timepoint_numeric", "Time")) %>%
  mutate(term = if_else(str_detect(term, "growth_cat"), str_remove(term, "growth_cat"), term))

summaries_df %>%
  pander(digits = 2, split.table = Inf, split.cells = 20, caption = glue("Linear model of the alpha diversity measurement with numeric time and growth categories as the predictive variables. SubjectID added as a random effect. Full formula: {my_formula}"))
```

# Supp Figure 3F - beta diversity 34-37 weeks PMA gfalt

```{r fig.height=3, fig.width=5}

s_toPlot %>%
  mutate(growth_label = case_when(growth_cat %in% "normal" ~ "None",
                                  growth_cat %in% "mild retardation" ~ "Mild",
                                  growth_cat %in% "moderate to severe retardation" ~ "Moderate\nto severe")) %>%
  mutate(growth_label = fct_relevel(growth_label, "None", "Mild")) %>%
  pcoaplus(bc) %>%
  plot(color=growth_label) +
    scale_color_manual(values=ann_colors$growth_label) +
    theme_bw() +
    theme(
      axis.text=element_blank(),
      axis.ticks=element_blank(),
      panel.grid = element_blank(), 
      aspect.ratio = 1,
      strip.background = element_blank()
    ) + 
    labs(color="Growth faltering", title="")

ggsave(here("Output", "SUPP_FIG3F_betadiv.pdf"), width = 3.5, height = 2.5, useDingbats = F)

```

## Permanova with log_npo_days_prop

```{r, results='asis'}

# summaries_df <-
#   s_toTest %>%
#     do(adonisplus(
#       ., distmat = bc, formula = distmat ~ bwt_z + log_npo_days_prop + growth_cat * WOL,
#       sample_id_var = SampleID, rep_meas_var = SubjectID,
#       shuffle = c(WOL = "within", growth_cat = "between", log_npo_days_prop = "within"), permutations=perm)) %>%
#   mutate(metric = "Bray-Curtis") %>%
#   filter(!term %in% c("Residuals", "Total")) %>%
#   select(-one_of(c("sumsq", "meansq"))) %>%
#   select(metric, everything())
# 
# write_rds(summaries_df, here("FigureData", "permanova_results_12.rds"))

summaries_df <- read_rds(here("FigureData", "permanova_results_12.rds"))

summaries_df %>%
  pander()

```

# Supp Figure 3G - alpha diversity 34-37 weeks PMA gfail

```{r}

s_toPlot %>%
  pivot_longer(c("Richness", "Shannon"), names_to="metric", values_to="alpha") %>%
  filter(metric %in% "Shannon") %>%
  ggplot(aes(x=growth_failure, y=alpha, fill=WOL_label)) +
  geom_boxplot() + #outlier.alpha=0
  scale_fill_manual(values = WOL_colors) +
  # facet_grid(metric~., scales = "free") +
  theme_bw() +
  theme(
    strip.background = element_blank(),
    panel.grid=element_blank(),
    # legend.position = "none",
    axis.text.x = element_text(angle = 45, hjust = 1, vjust = 1)
  ) +
  labs(
    x="Growth failure", fill="Week of life",
    y="Shannon diversity"
  )

ggsave(here("Output", "SUPP_FIG3G_alphadiv_gfail.pdf"), width = 3.5, height = 2.5, useDingbats = F)

```

## Linear model with log_npo_days_prop

```{r, results='asis'}
my_formula <- "alpha ~ bwt_z + WOL * growth_failure + log_npo_days_prop"

summaries_df <- s_toTest %>%
  pivot_longer(c("Richness", "Shannon"), names_to="metric", values_to="alpha") %>%
  group_by(metric) %>%
  do(tidy_lmer(nlme::lme(as.formula(my_formula), random=~1|SubjectID, data=., na.action=na.omit))) %>%
  ungroup() %>%
  filter(!grepl("Intercept", term)) %>%
  mutate(term = str_replace(term, "timepoint_numeric", "Time")) %>%
  mutate(term = if_else(str_detect(term, "growth_cat"), str_remove(term, "growth_cat"), term))

summaries_df %>%
  pander(digits = 2, split.table = Inf, split.cells = 20, caption = glue("Linear model of the alpha diversity measurement with numeric time and growth_failure as the predictive variables. SubjectID added as a random effect. Full formula: {my_formula}"))
```

# Supp Figure 3H - beta diversity 34-37 weeks PMA gfail

```{r}
s_toPlot %>%
  pcoaplus(bc) %>%
  plot(color=growth_failure) +
    scale_color_manual(values=ann_colors$growth_fail) +
    theme_bw() +
    theme(
      axis.text=element_blank(),
      axis.ticks=element_blank(),
      panel.grid = element_blank(), 
      aspect.ratio = 1,
      strip.background = element_blank()
    ) + 
    labs(color="Growth failure", title="")

ggsave(here("Output", "SUPP_FIG3H_betadiv.pdf"), width = 3, height = 2.5, useDingbats = F)
```

## Permanova with log_npo_days_prop

```{r, results='asis'}

# summaries_df <-
#   s_toTest %>%
#     do(adonisplus(
#       ., distmat = bc, formula = distmat ~ bwt_z + log_npo_days_prop + growth_failure * WOL,
#       sample_id_var = SampleID, rep_meas_var = SubjectID,
#       shuffle = c(WOL = "within", growth_failure = "between", log_npo_days_prop = "within"), permutations=perm)) %>%
#   mutate(metric = "Bray-Curtis") %>%
#   filter(!term %in% c("Residuals", "Total")) %>%
#   select(-one_of(c("sumsq", "meansq"))) %>%
#   select(metric, everything())
# 
# write_rds(summaries_df, here("FigureData", "permanova_results_13.rds"))

summaries_df <- read_rds(here("FigureData", "permanova_results_13.rds"))

summaries_df %>%
  pander()

```



```{r load data2}

load(file = here("Data", "sagori_metabolomics_data.RData"))
table_export <- table_iterator()
holupdates <- read.delim(file = here("Data", "stool_sequenced_update_20250829.txt"), sep = "\t")

s2 <- s2 %>%
  dplyr::rename(HOL_old=HOL) %>%
  left_join(holupdates %>% select(!HOL) %>% dplyr::rename(HOL=HOL_New) %>% mutate(redcap_id=as.character(redcap_id)), by = c("redcap_id", "SampleID")) %>%
  mutate(DOL=HOL/24) %>%
  mutate(WOL=as.numeric(cut_timepoints(DOL)),
         MOL=as.numeric(cut_MONTH_timepoints(DOL))) %>%
  mutate(growth_cat_old=growth_cat) #%>% 
  # mutate(growth_cat = factor(x = growth_cat_old, levels = c("normal", "mild retardation", "moderate to severe retardation"), labels = c("normal", "faltering", "faltering"))) # testing with faltering collapsed into normal vs. any faltering (mild or mod2sev)

s_toTest <- s2
ann_colors = list(
  category = setNames(brewer.pal(length(levels(s_toTest$category)), "Set2"), levels(s_toTest$category)),
  growth_cat = setNames(as.character(ghibli_palette("YesterdayMedium", direction = -1)[1:length(levels(s_toTest$growth_cat))]), levels(s_toTest$growth_cat)),
  TimePoint = setNames(viridis(length(levels(s_toTest$TimePoint))), levels(s_toTest$TimePoint)),
  growth_failure = setNames(qualpal(2)$hex, levels(s_toTest$growth_failure))
)
```

\newpage
# ~ TAXA ~
# FIGURE 1E  

Taxa abundance panel  
```{r fig1d, fig.height = 2, fig.width=4, dpi=300}
study_group_var <- "WOL"
is_virus <- grepl("Viruses", adf$Kingdom)
is_archaea <- grepl("Archaea", adf$Kingdom)
is_euk <- grepl("Eukaryota", adf$Kingdom)
deleteme = is_archaea | is_virus | is_euk
adf_special <- adf[!deleteme,]
counts_no_viruses <- counts[!deleteme,]
# order matters here. do more specific first:
adf_special <- adf_special %>%
  mutate(RANK = ifelse(Genus=="Staphylococcus", "Staphylococcus", ifelse(Genus=="Bifidobacterium", "Bifidobacterium", ifelse(Family=="Enterobacteriaceae", "Enterobacteriaceae", ifelse(Family == "Enterococcaceae", "Enterococcaceae", ifelse(Family=="Clostridiaceae", "Clostridiaceae", Phylum)))))) %>%
    mutate(RANK=ifelse(is.na(RANK), Phylum, RANK))

a_special <- simplify_assignments(adf_special, rank1="Phylum", rank2="RANK")
summed_cts_special <- rowsum(counts_no_viruses, a_special)
summed_props_special <- sweep(summed_cts_special, 2, colSums(summed_cts_special), "/")

s_subset <- s2 %>%
  # filter(!TimePoint %in% "T0") %>%
  select(SampleID, DOL, WOL, PMA, sample_type=SampleType, subject_id=SubjectID, TimePoint, study_day)
# all(colnames(summed_props_g) %in% unique(s2$SampleID, s$SampleID))

# Top 4 Phyla plus highlighted families or genera or classes within:
plotme2 = summed_props_special %>% 
  as.data.frame() %>% 
  rownames_to_column("Taxa") %>% 
  mutate(Taxa=ifelse(Taxa=="Proteobacteria Proteobacteria", "Other Proteobacteria", ifelse(Taxa=="Firmicutes Firmicutes", "Other Firmicutes", ifelse(Taxa=="Actinobacteria Actinobacteria", "Other Actinobacteria", ifelse(Taxa=="Bacteroidetes Bacteroidetes", "Bacteroidetes", ifelse(Taxa %in% c("Firmicutes Staphylococcus", "Actinobacteria Bifidobacterium", "Proteobacteria Enterobacteriaceae", "Firmicutes Lachnospiraceae", "Firmicutes Ruminococcaceae", "Firmicutes Enterococcaceae", "Firmicutes Clostridiaceae", "Firmicutes Bacilli"), Taxa, "Other Phyla")))))) %>%
  # filter(Taxa != "Bacteria unclassified") %>%  #!grepl(pattern = "Other", x = Taxa), 
  pivot_longer(-Taxa, names_to="SampleID", values_to="props") %>%
  right_join(s_subset, by="SampleID") %>%

  group_by(SampleID, Taxa, sample_type, subject_id, !!sym(study_group_var)) %>%
  summarize(sumprops = sum(props, na.rm = T), meanprops=mean(props, na.rm=T), sdprops=sd(props, na.rm = T), n=n()) %>%
  ungroup() %>%
  # mutate(se=sdprops/sqrt(n)) %>%
  # mutate(Taxa2 = fct_lump(Taxa, 30, w=sumprops, other_level = "Other Bacteria")) 
  # mutate(Taxa = reorder(Taxa, sumprops))
  mutate(Taxa = factor(Taxa, levels = c("Proteobacteria Enterobacteriaceae", "Other Proteobacteria", "Firmicutes Staphylococcus", "Firmicutes Enterococcaceae", "Firmicutes Clostridiaceae", "Other Firmicutes", "Actinobacteria Bifidobacterium", "Other Actinobacteria", "Bacteroidetes", "Other Phyla"), labels = c("Enterobacteriaceae", "Other Proteobacteria", "Staphylococcus", "Enterococcaceae", "Clostridiaceae", "Other Firmicutes", "Bifidobacterium", "Other Actinobacteria", "Bacteroidetes", "Other Phyla")))

ann_colors2 = factor_palette(plotme2$Taxa, ggsci::pal_d3("category10")(10)[c(1,2,3,4,5,6,7,9,10,8)])

library(scales)

plotme2 %>% 
  ggplot(aes(x=get(study_group_var), y=sumprops)) +
    geom_smooth(aes(group=Taxa, color=Taxa), alpha=0.2, size=0.5) + 
    scale_color_manual(values = ann_colors2) +
    scale_x_continuous(expand=c(0,0)) +
    theme_bw() +
    scale_y_continuous(labels = scales::percent, limit=c(0,NA), oob=squish, expand = c(0.01,0)) +
      coord_cartesian(clip = "off") +
    theme(
      # smaller verion:
      # legend.position.inside = c(0.57,.5),
      # legend.position = "inside", 
      
      # bottom legend version:
      # legend.position = "bottom", 
      
      # right legend version:
      legend.position = "right", 
      
      legend.key.size = unit(c(0.13), units = "in"),
      legend.background = element_rect(fill = NA), 
      legend.text = element_text(size=8),
      legend.title=element_blank(),
      panel.grid = element_blank(),
      axis.ticks = element_line(linewidth = 0.3),
      panel.border = element_rect(linewidth = 0.3),
      axis.text = element_text(colour = "black", size=8)
      ) +
    guides(color=guide_legend(ncol=1, override.aes = list(lwd= 1, fill=NA))) +
    labs(x=study_group_var, y="Relative abundance", fill="")

ggsave(here("Output", "FIG1E_taxaRA_WOL_smooth_rightsidekey.pdf"), width = 4, height = 2, useDingbats = F, dpi = 400)
# ggsave(here("Output", "FIG1E_taxaRA_WOL_smooth_smaller.pdf"), width = 4, height = 3, useDingbats = F, dpi = 400)
```

\newpage
Taxa boxplots
```{r fig.width=10, fig.height=8}
minRAmean = 0.01
genus_plot <- data.frame(summed_props) %>%
  rownames_to_column("Taxa") %>%
  mutate(Taxa = gsub("  ", "", Taxa)) %>%
  pivot_longer(-Taxa, names_to="SampleID", values_to="props") %>%
  right_join(s_subset, by = "SampleID") %>%
  group_by(Taxa, sample_type) %>%
    mutate(sumprops=sum(props), meanprops=mean(props), maxprops=max(props)) %>%
  ungroup() %>%
  filter(sumprops>0, meanprops>minRAmean) 

# genus_plot %>%
#   mutate(Taxalabel = str_wrap(as.character(Taxa), width = 25)) %>%
#   ggplot(aes(x=WOL, y=props, fill=WOL, group = WOL)) +
#   geom_boxplot(outlier.size = 0.25, outlier.shape = 21) +
#   facet_wrap(~Taxalabel, ncol=4, scales="free_y") +
#   scale_fill_viridis_c(alpha=0.8) +
#   scale_y_continuous(transform = "log10") +
#   theme_bw() +
#   theme(panel.grid = element_blank(), strip.background = element_blank())
```

\newpage
# FIGURE 1F  
## Linear model  
Taxa change with age  
```{r}
ps=1e-6
props_toTest <- data.frame(summed_props) %>%
  rownames_to_column("Taxa") %>%
  mutate(Taxa = gsub("  ", "", Taxa)) %>%
  pivot_longer(-Taxa, names_to="SampleID", values_to="props") %>%
  filter(Taxa %in% unique(genus_plot$Taxa)) %>%
  mutate(log10_props = log10(props+ps))

props_toTest_g <- data.frame(summed_props_g) %>%
  rownames_to_column("Taxa") %>%
  mutate(Taxa = gsub("  ", "", Taxa)) %>%
  pivot_longer(-Taxa, names_to="SampleID", values_to="props") %>%
  left_join(s %>% select(SampleID, SampleType), by = "SampleID") %>%
  mutate(present=props>0) %>%
  filter(SampleType=="Feces") %>%
  group_by(Taxa, SampleType) %>%
    mutate(sumprops=sum(props), meanprops=mean(props), maxprops=max(props), presence=sum(present)/dim(summed_props_g)[2]) %>%
  ungroup() %>%
  filter(sumprops>0, meanprops>minRAmean) %>%
  mutate(log10_props = log10(props+ps)) %>%
  select(!SampleType)

form1 = "log10_props ~ WOL"
# form2 = "log10_props ~ WOL * growth_cat + npo_days_prop"
# form3 = "log10_props ~ WOL * growth_cat + AbxHoursPerHOL"
random1 = "subject_id"

summaries_df <- run_lmer_on_props(factor_to_test = "Taxa", props_toTest = props_toTest, s_toTest = s_subset, form1 = form1, rep_mes_label = random1, p_cutoff =  0.05) 

summaries_df %>%
  # mutate(term = str_remove(term, "growth_cat")) %>%
  pander(split.table=Inf, digits=2, caption = glue("Linear models of log10 transformed relative abundance of genera. Filtered with p.value < 0.05. SubjectID added as a random effect. Full formula: {form1}"), emphasize.strong.rows = which(.$fdr<=0.05))

table_export(summaries_df, "taxonomic_analysis_table_")

```

```{r fig1f, fig.width=8, fig.height=4}
minRAmean = 0.01
sigtaxa = summaries_df %>% filter(fdr<0.05, term=="WOL") %>% pull(Taxa)
genus_plot %>%
  filter(Taxa %in% sigtaxa) %>%
  filter(!(Taxa %in% c("Proteobacteria Klebsiella", "Proteobacteria Enterobacter", "Proteobacteria Enterobacterales", "Proteobacteria Enterobacteriaceae"))) %>% # show only taxa that were classified to the species level for main text figure
  # filter(!(Taxa %in% c("Proteobacteria Klebsiella", "Proteobacteria Enterobacter", "Proteobacteria Enterobacterales", "Proteobacteria Enterobacteriaceae", "Proteobacteria Citrobacter freundii", "Proteobacteria Klebsiella aerogenes"))) %>% # show only taxa that were classified to the species level for main text figure, and Scott asked me to drop the last 2 for final figure edits
   mutate(Taxalabel = str_wrap(paste0(word(as.character(Taxa)), ": ", word(as.character(Taxa), start = 2, end = -1)), width = 15)) %>%
 
  left_join(summaries_df %>% 
              filter(term=="WOL") %>% 
              mutate(q_label = ifelse(fdr>=0.05, "n.s.", ifelse(fdr<0.0001, 
                                      paste0("q=", scientific(fdr, digits = 2)), paste0("q=", as.character(signif(fdr, digits = 2)))))), by="Taxa") %>%
  group_by(Taxa) %>% mutate(max=max(props), min=min(props)) %>% ungroup() %>%
  ggplot(aes(x=WOL, y=props, fill=WOL, group = WOL)) +
    geom_boxplot(outlier.size = 1, outlier.shape = 21,
                 position=position_dodge(0.25), lwd=0.25, outlier.color = NA) +
    geom_text(mapping = aes(label = q_label, x=12, y=10), size = 3, check_overlap = T) +
    facet_wrap(~Taxalabel, ncol=6) +
    scale_fill_viridis_c(alpha=0.8) +
    # scale_fill_manual(values = ann_colors$growth_cat) +
    scale_y_continuous(expand = c(0, 0.5), transform = "log10") +
    theme_bw() +
    labs(y="Relative abundance (log scale)", fill="WOL", x="WOL") +
    theme(panel.grid = element_blank(), strip.background = element_blank())
  
ggsave(here("Output", "FIG1F_taxaRA_fdr05_WOL_boxplots.pdf"), width=8, height = 4, useDingbats = F, dpi = 600)
```

# SUPP FIGURE 1E  

\newpage
## Linear model  
Taxa change with age  
```{r}
minRAmean = 0.01
s_subset <- s2 %>%
  # filter(!TimePoint %in% "T0") %>%
  select(SampleID, WOL, PMA, growth_cat, growth_failure, sample_type=SampleType, subject_id=SubjectID, TimePoint, study_day)

# sigtaxa = summaries_df %>% filter(fdr<0.05, term=="WOL") %>% pull(Taxa)
plotme <- data.frame(summed_props) %>%
    rownames_to_column("Taxa") %>%
    mutate(Taxa = gsub("  ", "", Taxa)) %>%
    pivot_longer(-Taxa, names_to="SampleID", values_to="props") %>%
    right_join(s_subset, by = "SampleID") %>%
    group_by(Taxa, sample_type) %>%
      mutate(sumprops=sum(props), meanprops=mean(props), maxprops=max(props)) %>%
    ungroup() %>%
    filter(sumprops>0, meanprops>minRAmean) %>%
    mutate(Taxalabel = paste0(word(as.character(Taxa)), " -\n", 
                              word(as.character(Taxa), start = 2, end = -1))) 
ps=1e-6
props_toTest <- data.frame(summed_props) %>%
  rownames_to_column("Taxa") %>%
  mutate(Taxa = gsub("  ", "", Taxa)) %>%
  pivot_longer(-Taxa, names_to="SampleID", values_to="props") %>%
  filter(Taxa %in% unique(plotme$Taxa)) %>%
  mutate(log10_props = log10(props+ps))

form1 = "log10_props ~ WOL"
# form2 = "log10_props ~ WOL * growth_cat + npo_days_prop"
# form3 = "log10_props ~ WOL * growth_cat + AbxHoursPerHOL"
random1 = "subject_id"

# summaries_df <- run_lmer_on_props(factor_to_test = "Taxa", props_toTest = props_toTest, s_toTest = s_subset, form1 = form1, rep_mes_label = random1, p_cutoff =  0.05) 

summaries_df <- props_toTest %>%
  right_join(s_subset, by = "SampleID") %>%
  group_by(Taxa, growth_cat) %>%
  do(tidy_lmer(nlme::lme(as.formula(form1), random = as.formula(paste("~ 1 |", random1)), data=., na.action=na.omit))) %>%
    ungroup() %>%
    filter(term != '(Intercept)') %>%
    group_by(term) %>%
    mutate(fdr = p.adjust(p.value, method="BH")) %>%
    ungroup() 

summaries_df %>%
  filter(p.value<0.05) %>%
  # mutate(term = str_remove(term, "growth_cat")) %>%
  pander(split.table=Inf, digits=2, caption = glue("Linear models of log10 transformed relative abundance of genera. Filtered with p.value < 0.05. SubjectID added as a random effect. Full formula: {form1}"), emphasize.strong.rows = which(.$fdr<=0.05))

table_export(summaries_df, "taxonomic_analysis_table_")
``` 

```{r sfig1e, fig.width=8, fig.height=16}
sigtaxa = summaries_df %>% filter(fdr<0.05, term=="WOL") %>% pull(Taxa)
plotme %>% filter(Taxa %in% sigtaxa) %>%
  # filter(!(Taxa %in% c("Proteobacteria Klebsiella", "Proteobacteria Enterobacter", "Proteobacteria Enterobacterales", "Proteobacteria Enterobacteriaceae"))) %>% # show only taxa that were classified to the species level for main text figure
  left_join(summaries_df %>% 
              filter(term=="WOL") %>% 
              mutate(q_label = ifelse(fdr>=0.05, "n.s.", ifelse(fdr<0.0001, 
                                      paste0("q=", scientific(fdr, digits = 2)), paste0("q=", as.character(signif(fdr, digits = 2)))))), by = c("Taxa", "growth_cat") ) %>%
  group_by(Taxa) %>% mutate(max=max(props), min=min(props)) %>% ungroup() %>%
  mutate(growth_cat = factor(growth_cat,levels = c("normal", "mild retardation", "moderate to severe retardation"), labels = c("None", "Mild", "Moderate to\nsevere"))) %>%
  ggplot(aes(x=growth_cat, y=props+ps)) +
    geom_boxplot(aes(fill=factor(WOL)), outlier.size = 0.5, outlier.shape = 21,
                 position=position_dodge(0.9), lwd=0.25, outlier.color = NA) +
    geom_text(mapping = aes(label = q_label, x=growth_cat, y=10), size = 3, check_overlap = T) +
    facet_wrap(~Taxalabel, ncol=2) +
    scale_fill_viridis_d(alpha=0.8, guide="none") +
    # scale_fill_manual(values = ann_colors$growth_cat) +
    scale_y_continuous(expand = c(0, 0.5), transform = "log10") +
    theme_bw() +
    labs(y="Relative abundance (log scale)", fill="WOL", x="Growth category") +
    theme(panel.grid = element_blank(), strip.background = element_blank())

ggsave(here("Output", "SUPP_FIG1E_taxaRA_growthcat_WOL_boxplots.pdf"), width =8, height = 16, useDingbats = F, dpi = 600)
```


\newpage
# SUPP FIGURE 1F  
## Linear model  
Taxa change with age  
```{r}
form1 = "log10_props ~ WOL"
# form2 = "log10_props ~ WOL * growth_cat + npo_days_prop"
# form3 = "log10_props ~ WOL * growth_cat + AbxHoursPerHOL"
random1 = "subject_id"
# summaries_df <- run_lmer_on_props(factor_to_test = "Taxa", props_toTest = props_toTest, s_toTest = s_subset, form1 = form1, rep_mes_label = random1, p_cutoff =  0.05) 

summaries_df <- props_toTest %>%
  right_join(s_subset, by = "SampleID") %>%
  group_by(Taxa, growth_failure) %>%
  do(tidy_lmer(nlme::lme(as.formula(form1), random = as.formula(paste("~ 1 |", random1)), data=., na.action=na.omit))) %>%
    ungroup() %>%
    filter(term != '(Intercept)') %>%
    group_by(term) %>%
    mutate(fdr = p.adjust(p.value, method="BH")) %>%
    ungroup() 

summaries_df %>%
  filter(p.value<0.05) %>%
  # mutate(term = str_remove(term, "growth_cat")) %>%
  pander(split.table=Inf, digits=2, caption = glue("Linear models of log10 transformed relative abundance of genera. Filtered with p.value < 0.05. SubjectID added as a random effect. Full formula: {form1}"), emphasize.strong.rows = which(.$fdr<=0.05))

table_export(summaries_df, "taxonomic_analysis_table_")
``` 

```{r fig.width=8, fig.height=16}
sigtaxa = summaries_df %>% filter(fdr<0.05, term=="WOL") %>% pull(Taxa)
plotme %>% filter(Taxa %in% sigtaxa) %>%
  # filter(!(Taxa %in% c("Proteobacteria Klebsiella", "Proteobacteria Enterobacter", "Proteobacteria Enterobacterales", "Proteobacteria Enterobacteriaceae"))) %>% # show only taxa that were classified to the species level for main text figure
  left_join(summaries_df %>% 
              filter(term=="WOL") %>% 
              mutate(q_label = ifelse(fdr>=0.05, "n.s.", ifelse(fdr<0.0001, 
                                      paste0("q=", scientific(fdr, digits = 2)), paste0("q=", as.character(signif(fdr, digits = 2)))))), by = c("Taxa", "growth_failure") ) %>%
  group_by(Taxa) %>% mutate(max=max(props), min=min(props)) %>% ungroup() %>%
  ggplot(aes(x=growth_failure, y=props)) +
    geom_boxplot(aes(fill=factor(WOL)), outlier.size = 0.5, outlier.shape = 21,
                 position=position_dodge(0.9), lwd=0.25, outlier.color = NA) +
    geom_text(mapping = aes(label = q_label, x=growth_failure, y=10), size = 3, check_overlap = T) +
  # plotme %>%
  # ggplot(aes(x=growth_failure, y=props)) +
  #   geom_boxplot(aes(fill=factor(WOL)), outlier.size = 0.5, outlier.shape = 21) +
    # geom_text(mapping = aes(label = paste0("q=", scientific(fdr, digits = 2)), x=12, y=max*0.96), size = 3, check_overlap = T) +
    facet_wrap(~Taxalabel, ncol=2) +
    scale_fill_viridis_d(alpha=0.8) +
    # scale_fill_manual(values = ann_colors$growth_cat) +
    scale_y_continuous(expand = c(0, 0.5), transform = "log10") +
    theme_bw() +
    labs(y="Relative abundance (log scale)", fill="WOL", x="Growth failure") +
    theme(panel.grid = element_blank(), strip.background = element_blank())

ggsave(here("Output", "SUPP_FIG1F_taxaRA_growthfail_WOL_boxplots.pdf"), width =8, height = 16, useDingbats = F, dpi = 600)
```

## Week by week  
### SPECIES
Is there a week where the groups diverge?  
Asterisk/star for fdr-adjusted p-value<0.05. Diamond for p-unadjusted<0.05  
Linear models were used to estimate the differences between study groups at each week of life. The relative abundances were log10 transformed. Multiple tests were adjusted for false discovery rate using Benjamini-Hochberg method. Only the terms with p<0.1 are shown in the table below.

Growth Faltering:  
```{r fig.height=9, fig.width=8}
myform="log10_props ~ study_group"

summaries_df <- props_toTest %>%
  inner_join(s2 %>% select(SampleID, study_group=growth_cat, sample_type=SampleType, TimePoint, SubjectID, WOL), by = "SampleID") %>%
  filter(!is.na(study_group)) %>%
  group_by(WOL, Taxa, study_group, sample_type) %>%
    mutate(mean=mean(log10_props)) %>%
  ungroup() %>%
  
  group_by(Taxa) %>% mutate(max=max(mean)) %>% ungroup() %>%
  
  group_by(Taxa, WOL) %>%
    mutate(ngroups=n_distinct(study_group)) %>%
  ungroup() %>%
  
  filter(ngroups>=2) %>% # Only test PMA_rounded where you have both study groups
  
  group_by(Taxa, WOL, max) %>%
    do(tidy(lm(as.formula(myform), data=.))) %>%
  ungroup() %>%
  
  filter(!grepl("Intercept|Residuals", term)) %>%
  mutate(term = sub("study_group", "", term)) %>%
  
  group_by(term, WOL) %>%
    mutate(fdr = p.adjust(p.value, method="BH")) %>%
  ungroup()

props_toTest %>%
  inner_join(s2 %>% select(SampleID, study_group=growth_cat, sample_type=SampleType, TimePoint, SubjectID, WOL), by = "SampleID") %>%
  filter(!is.na(study_group)) %>%
  # mutate(Taxa = reorder(Taxa, -log10_props)) %>%
  # mutate(Taxa = fct_relabel(Taxa, function(x) gsub(" ", "\n", x))) %>%
  group_by(WOL, Taxa, study_group, sample_type) %>%
    summarise(mean=mean(log10_props), sd=sd(log10_props), n=n(), max=max(log10_props)) %>%
  ungroup() %>%
  mutate(se=sd/sqrt(n)) %>%
  ggplot(aes(y=10^(mean), x=WOL)) +
    # geom_vline(aes(xintercept = PMA_rounded), lwd=0.3, color="grey") + # guide lines
    geom_line(aes(color=study_group, group = study_group)) +  
    geom_point(aes(color=study_group)) +
    geom_point(data = summaries_df %>% filter(fdr<0.1, term==levels(s2$growth_cat)[2]), mapping = aes(x=WOL, y=10^max +.2, color=term), shape=8) +
    geom_point(data = summaries_df %>% filter(fdr>=0.1, p.value<0.05, term==levels(s2$growth_cat)[2]), mapping = aes(x=WOL, y=10^max +.2, color=term), shape=23) +

    geom_point(data = summaries_df %>% filter(fdr<0.1, term==levels(s2$growth_cat)[3]), mapping = aes(color=term, x=WOL, y=10^max +.4), shape=8) +
    geom_point(data = summaries_df %>% filter(p.value<0.05, fdr>=0.1, term==levels(s2$growth_cat)[3]), mapping = aes(color=term, x=WOL, y=10^max +.4), shape=23) +
    scale_color_manual(values = ann_colors$growth_cat) +
    scale_y_continuous(transform = "log10", labels = scales::percent) +
    theme_bw() +
  theme(legend.position = "bottom") +
    theme(panel.grid = element_blank()) +
    geom_errorbar(aes(ymin = 10^(mean-se), ymax=10^(mean+se), color=study_group), alpha=0.5, width = 0.3) +
    facet_wrap(~Taxa, ncol=3, scales="free") +
  labs(y="Concentration (umol/g log-scale)")

# table_export(summaries_df, "Taxonomic_table_export")
summaries_df %>%
  filter(p.value < 0.05) %>%
  pander(split.tab=Inf, emphasize.strong.rows = which(.$fdr<=0.05)) #%>%
  #column_spec(c(2:5), width = "3em") ## if the table is running off the page, you can manually specify the width of the columns

rm(myform)
```

Growth Failure:  
Asterisk/star for fdr-adjusted p-value<0.05. Diamond for p-unadjusted<0.05  
```{r fig.height=9, fig.width=8}
myform="log10_props ~ study_group"

summaries_df <- props_toTest %>%
  inner_join(s2 %>% select(SampleID, study_group=growth_failure, sample_type=SampleType, TimePoint, SubjectID, WOL), by = "SampleID") %>%
  filter(!is.na(study_group)) %>%
  group_by(WOL, Taxa, study_group, sample_type) %>%
    mutate(mean=mean(log10_props)) %>%
  ungroup() %>%
  
  group_by(Taxa) %>% mutate(max=max(mean)) %>% ungroup() %>%
  
  group_by(Taxa, WOL) %>%
    mutate(ngroups=n_distinct(study_group)) %>%
  ungroup() %>%
  
  filter(ngroups>=2) %>% # Only test PMA_rounded where you have both study groups
  
  group_by(Taxa, WOL, max) %>%
    do(tidy(lm(as.formula(myform), data=.))) %>%
  ungroup() %>%
  
  filter(!grepl("Intercept|Residuals", term)) %>%
  mutate(term = sub("study_group", "", term)) %>%
  
  group_by(term, WOL) %>%
    mutate(fdr = p.adjust(p.value, method="BH")) %>%
  ungroup()

props_toTest %>%
  inner_join(s2 %>% select(SampleID, study_group=growth_failure, sample_type=SampleType, TimePoint, SubjectID, WOL), by = "SampleID") %>%
  filter(!is.na(study_group)) %>%
  # mutate(Taxa = reorder(Taxa, -log10_props)) %>%
  # mutate(Taxa = fct_relabel(Taxa, function(x) gsub(" ", "\n", x))) %>%
  group_by(WOL, Taxa, study_group, sample_type) %>%
    summarise(mean=mean(log10_props), sd=sd(log10_props), n=n(), max=max(log10_props)) %>%
  ungroup() %>%
  mutate(se=sd/sqrt(n)) %>%
  ggplot(aes(y=10^(mean), x=WOL)) +
    # geom_vline(aes(xintercept = PMA_rounded), lwd=0.3, color="grey") + # guide lines
    geom_line(aes(color=study_group, group = study_group)) +  
    geom_point(aes(color=study_group)) +
    geom_point(data = summaries_df %>% filter(fdr<0.05), mapping = aes(x=WOL, y=10^max +.2, color=term), shape=8) +
    geom_point(data = summaries_df %>% filter(fdr>0.05, p.value<0.05), mapping = aes(x=WOL, y=10^max +.2, color=term), shape=23) +
    # geom_point(data = summaries_df %>% filter(p.value<0.05, term=="moderate to severe retardation"), mapping = aes(color=term, x=WOL, y=10^max +.4), shape=8) +
    scale_color_manual(values = ann_colors$growth_failure) +
    scale_y_continuous(transform = "log10", labels = scales::percent) +
    theme_bw() +
  theme(legend.position = "bottom") +
    theme(panel.grid = element_blank()) +
    geom_errorbar(aes(ymin = 10^(mean-se), ymax=10^(mean+se), color=study_group), alpha=0.5, width = 0.3) +
    facet_wrap(~Taxa, ncol=3, scales="free") +
  labs(y="Concentration (umol/g log-scale)")

# table_export(summaries_df, "Taxonomic_table_export")

summaries_df %>%
  filter(p.value < 0.05) %>%
  pander(split.tab=Inf, emphasize.strong.rows = which(.$fdr<=0.05)) #%>%
  #column_spec(c(2:5), width = "3em") ## if the table is running off the page, you can manually specify the width of the columns

rm(myform)
```


### GENUS  
Growth Faltering:  
```{r fig.height=9, fig.width=8}
myform="log10_props ~ study_group"

summaries_df <- props_toTest_g %>%
  inner_join(s2 %>% select(SampleID, study_group=growth_cat, sample_type=SampleType, TimePoint, SubjectID, WOL), by = "SampleID") %>%
  filter(!is.na(study_group)) %>%
  group_by(WOL, Taxa, study_group, sample_type) %>%
    mutate(mean=mean(log10_props)) %>%
  ungroup() %>%
  
  group_by(Taxa) %>% mutate(max=max(mean)) %>% ungroup() %>%
  
  group_by(Taxa, WOL) %>%
    mutate(ngroups=n_distinct(study_group)) %>%
  ungroup() %>%
  
  filter(ngroups>=2) %>% # Only test PMA_rounded where you have both study groups
  
  group_by(Taxa, WOL, max) %>%
    do(tidy(lm(as.formula(myform), data=.))) %>%
  ungroup() %>%
  
  filter(!grepl("Intercept|Residuals", term)) %>%
  mutate(term = sub("study_group", "", term)) %>%
  
  group_by(term, WOL) %>%
    mutate(fdr = p.adjust(p.value, method="BH")) %>%
  ungroup()

props_toTest_g %>%
  inner_join(s2 %>% select(SampleID, study_group=growth_cat, sample_type=SampleType, TimePoint, SubjectID, WOL), by = "SampleID") %>%
  filter(!is.na(study_group)) %>%
  # mutate(Taxa = reorder(Taxa, -log10_props)) %>%
  # mutate(Taxa = fct_relabel(Taxa, function(x) gsub(" ", "\n", x))) %>%
  group_by(WOL, Taxa, study_group, sample_type) %>%
    summarise(mean=mean(log10_props), sd=sd(log10_props), n=n(), max=max(log10_props)) %>%
  ungroup() %>%
  mutate(se=sd/sqrt(n)) %>%
  ggplot(aes(y=10^(mean), x=WOL)) +
    # geom_vline(aes(xintercept = PMA_rounded), lwd=0.3, color="grey") + # guide lines
    geom_line(aes(color=study_group, group = study_group)) +  
    geom_point(aes(color=study_group)) +
    geom_point(data = summaries_df %>% filter(fdr<0.1, term==levels(s2$growth_cat)[2]), mapping = aes(x=WOL, y=10^max +.2, color=term), shape=8) +
    geom_point(data = summaries_df %>% filter(fdr>=0.1, p.value<0.05, term==levels(s2$growth_cat)[2]), mapping = aes(x=WOL, y=10^max +.2, color=term), shape=23) +

    geom_point(data = summaries_df %>% filter(fdr<0.1, term==levels(s2$growth_cat)[3]), mapping = aes(color=term, x=WOL, y=10^max +.4), shape=8) +
    geom_point(data = summaries_df %>% filter(p.value<0.05, fdr>=0.1, term==levels(s2$growth_cat)[3]), mapping = aes(color=term, x=WOL, y=10^max +.4), shape=23) +
    scale_color_manual(values = ann_colors$growth_cat) +
    scale_y_continuous(transform = "log10", labels = scales::percent) +
    theme_bw() +
  theme(legend.position = "bottom") +
    theme(panel.grid = element_blank()) +
    geom_errorbar(aes(ymin = 10^(mean-se), ymax=10^(mean+se), color=study_group), alpha=0.5, width = 0.3) +
    facet_wrap(~Taxa, ncol=3, scales="free") +
  labs(y="Concentration (umol/g log-scale)")

# table_export(summaries_df, "Taxonomic_table_export")
summaries_df %>%
  filter(p.value < 0.05) %>%
  pander(split.tab=Inf, emphasize.strong.rows = which(.$fdr<=0.05)) #%>%
  #column_spec(c(2:5), width = "3em") ## if the table is running off the page, you can manually specify the width of the columns

rm(myform)
```

Growth Failure:  
Asterisk/star for fdr-adjusted p-value<0.05. Diamond for p-unadjusted<0.05  
```{r fig.height=9, fig.width=8}
myform="log10_props ~ study_group"

summaries_df <- props_toTest_g %>%
  inner_join(s2 %>% select(SampleID, study_group=growth_failure, sample_type=SampleType, TimePoint, SubjectID, WOL), by = "SampleID") %>%
  filter(!is.na(study_group)) %>%
  group_by(WOL, Taxa, study_group, sample_type) %>%
    mutate(mean=mean(log10_props)) %>%
  ungroup() %>%
  
  group_by(Taxa) %>% mutate(max=max(mean)) %>% ungroup() %>%
  
  group_by(Taxa, WOL) %>%
    mutate(ngroups=n_distinct(study_group)) %>%
  ungroup() %>%
  
  filter(ngroups>=2) %>% # Only test PMA_rounded where you have both study groups
  
  group_by(Taxa, WOL, max) %>%
    do(tidy(lm(as.formula(myform), data=.))) %>%
  ungroup() %>%
  
  filter(!grepl("Intercept|Residuals", term)) %>%
  mutate(term = sub("study_group", "", term)) %>%
  
  group_by(term, WOL) %>%
    mutate(fdr = p.adjust(p.value, method="BH")) %>%
  ungroup()

props_toTest_g %>%
  inner_join(s2 %>% select(SampleID, study_group=growth_failure, sample_type=SampleType, TimePoint, SubjectID, WOL), by = "SampleID") %>%
  filter(!is.na(study_group)) %>%
  # mutate(Taxa = reorder(Taxa, -log10_props)) %>%
  # mutate(Taxa = fct_relabel(Taxa, function(x) gsub(" ", "\n", x))) %>%
  group_by(WOL, Taxa, study_group, sample_type) %>%
    summarise(mean=mean(log10_props), sd=sd(log10_props), n=n(), max=max(log10_props)) %>%
  ungroup() %>%
  mutate(se=sd/sqrt(n)) %>%
  ggplot(aes(y=10^(mean), x=WOL)) +
    # geom_vline(aes(xintercept = PMA_rounded), lwd=0.3, color="grey") + # guide lines
    geom_line(aes(color=study_group, group = study_group)) +  
    geom_point(aes(color=study_group)) +
    geom_point(data = summaries_df %>% filter(fdr<0.05), mapping = aes(x=WOL, y=10^max +.2, color=term), shape=8) +
    geom_point(data = summaries_df %>% filter(fdr>0.05, p.value<0.05), mapping = aes(x=WOL, y=10^max +.2, color=term), shape=23) +
    # geom_point(data = summaries_df %>% filter(p.value<0.05, term=="moderate to severe retardation"), mapping = aes(color=term, x=WOL, y=10^max +.4), shape=8) +
    scale_color_manual(values = ann_colors$growth_failure) +
    scale_y_continuous(transform = "log10", labels = scales::percent) +
    theme_bw() +
  theme(legend.position = "bottom") +
    theme(panel.grid = element_blank()) +
    geom_errorbar(aes(ymin = 10^(mean-se), ymax=10^(mean+se), color=study_group), alpha=0.5, width = 0.3) +
    facet_wrap(~Taxa, ncol=3, scales="free") +
  labs(y="Concentration (umol/g log-scale)")

# table_export(summaries_df, "Taxonomic_table_export")

summaries_df %>%
  filter(p.value < 0.05) %>%
  pander(split.tab=Inf, emphasize.strong.rows = which(.$fdr<=0.05)) #%>%
  #column_spec(c(2:5), width = "3em") ## if the table is running off the page, you can manually specify the width of the columns

rm(myform)
```



## Monthly 
### SPECIES
Is there a week where the groups diverge?  
Asterisk/star for fdr-adjusted p-value<0.05. Diamond for p-unadjusted<0.05  
Linear models were used to estimate the differences between study groups at each week of life. The relative abundances were log10 transformed. Multiple tests were adjusted for false discovery rate using Benjamini-Hochberg method. Only the terms with p<0.1 are shown in the table below.

Growth Faltering:  
```{r fig.height=9, fig.width=8}
myform="log10_props ~ study_group"

summaries_df <- props_toTest %>%
  inner_join(s2 %>% select(SampleID, study_group=growth_cat, sample_type=SampleType, TimePoint, SubjectID, MOL), by = "SampleID") %>%
  filter(!is.na(study_group)) %>%
  group_by(MOL, Taxa, study_group, sample_type) %>%
    mutate(mean=mean(log10_props)) %>%
  ungroup() %>%
  
  group_by(Taxa) %>% mutate(max=max(mean)) %>% ungroup() %>%
  
  group_by(Taxa, MOL) %>%
    mutate(ngroups=n_distinct(study_group)) %>%
  ungroup() %>%
  
  filter(ngroups>=2) %>% # Only test PMA_rounded where you have both study groups
  
  group_by(Taxa, MOL, max) %>%
    do(tidy(lm(as.formula(myform), data=.))) %>%
  ungroup() %>%
  
  filter(!grepl("Intercept|Residuals", term)) %>%
  mutate(term = sub("study_group", "", term)) %>%
  
  group_by(term, MOL) %>%
    mutate(fdr = p.adjust(p.value, method="BH")) %>%
  ungroup()

props_toTest %>%
  inner_join(s2 %>% select(SampleID, study_group=growth_cat, sample_type=SampleType, TimePoint, SubjectID, MOL), by = "SampleID") %>%
  filter(!is.na(study_group)) %>%
  # mutate(Taxa = reorder(Taxa, -log10_props)) %>%
  # mutate(Taxa = fct_relabel(Taxa, function(x) gsub(" ", "\n", x))) %>%
  group_by(MOL, Taxa, study_group, sample_type) %>%
    summarise(mean=mean(log10_props), sd=sd(log10_props), n=n(), max=max(log10_props)) %>%
  ungroup() %>%
  mutate(se=sd/sqrt(n)) %>%
  ggplot(aes(y=10^(mean), x=MOL)) +
    # geom_vline(aes(xintercept = PMA_rounded), lwd=0.3, color="grey") + # guide lines
    geom_line(aes(color=study_group, group = study_group)) +  
    geom_point(aes(color=study_group)) +
    geom_point(data = summaries_df %>% filter(fdr<0.1, term==levels(s2$growth_cat)[2]), mapping = aes(x=MOL, y=10^max +.2, color=term), shape=8) +
    geom_point(data = summaries_df %>% filter(fdr>=0.1, p.value<0.05, term==levels(s2$growth_cat)[2]), mapping = aes(x=MOL, y=10^max +.2, color=term), shape=23) +

    geom_point(data = summaries_df %>% filter(fdr<0.1, term==levels(s2$growth_cat)[3]), mapping = aes(color=term, x=MOL, y=10^max +.4), shape=8) +
    geom_point(data = summaries_df %>% filter(p.value<0.05, fdr>=0.1, term==levels(s2$growth_cat)[3]), mapping = aes(color=term, x=MOL, y=10^max +.4), shape=23) +
    scale_color_manual(values = ann_colors$growth_cat) +
    scale_y_continuous(transform = "log10", labels = scales::percent) +
    theme_bw() +
  theme(legend.position = "bottom") +
    theme(panel.grid = element_blank()) +
    geom_errorbar(aes(ymin = 10^(mean-se), ymax=10^(mean+se), color=study_group), alpha=0.5, width = 0.3) +
    facet_wrap(~Taxa, ncol=3, scales="free") +
  labs(y="Concentration (umol/g log-scale)")

# table_export(summaries_df, "Taxonomic_table_export")
summaries_df %>%
  filter(p.value < 0.05) %>%
  pander(split.tab=Inf, emphasize.strong.rows = which(.$fdr<=0.05)) #%>%
  #column_spec(c(2:5), width = "3em") ## if the table is running off the page, you can manually specify the width of the columns

rm(myform)
```

Growth Failure:  
Asterisk/star for fdr-adjusted p-value<0.05. Diamond for p-unadjusted<0.05  
```{r fig.height=9, fig.width=8}
myform="log10_props ~ study_group"

summaries_df <- props_toTest %>%
  inner_join(s2 %>% select(SampleID, study_group=growth_failure, sample_type=SampleType, TimePoint, SubjectID, MOL), by = "SampleID") %>%
  filter(!is.na(study_group)) %>%
  group_by(MOL, Taxa, study_group, sample_type) %>%
    mutate(mean=mean(log10_props)) %>%
  ungroup() %>%
  
  group_by(Taxa) %>% mutate(max=max(mean)) %>% ungroup() %>%
  
  group_by(Taxa, MOL) %>%
    mutate(ngroups=n_distinct(study_group)) %>%
  ungroup() %>%
  
  filter(ngroups>=2) %>% # Only test PMA_rounded where you have both study groups
  
  group_by(Taxa, MOL, max) %>%
    do(tidy(lm(as.formula(myform), data=.))) %>%
  ungroup() %>%
  
  filter(!grepl("Intercept|Residuals", term)) %>%
  mutate(term = sub("study_group", "", term)) %>%
  
  group_by(term, MOL) %>%
    mutate(fdr = p.adjust(p.value, method="BH")) %>%
  ungroup()

props_toTest %>%
  inner_join(s2 %>% select(SampleID, study_group=growth_failure, sample_type=SampleType, TimePoint, SubjectID, MOL), by = "SampleID") %>%
  filter(!is.na(study_group)) %>%
  # mutate(Taxa = reorder(Taxa, -log10_props)) %>%
  # mutate(Taxa = fct_relabel(Taxa, function(x) gsub(" ", "\n", x))) %>%
  group_by(MOL, Taxa, study_group, sample_type) %>%
    summarise(mean=mean(log10_props), sd=sd(log10_props), n=n(), max=max(log10_props)) %>%
  ungroup() %>%
  mutate(se=sd/sqrt(n)) %>%
  ggplot(aes(y=10^(mean), x=MOL)) +
    # geom_vline(aes(xintercept = PMA_rounded), lwd=0.3, color="grey") + # guide lines
    geom_line(aes(color=study_group, group = study_group)) +  
    geom_point(aes(color=study_group)) +
    geom_point(data = summaries_df %>% filter(fdr<0.05), mapping = aes(x=MOL, y=10^max +.2, color=term), shape=8) +
    geom_point(data = summaries_df %>% filter(fdr>0.05, p.value<0.05), mapping = aes(x=MOL, y=10^max +.2, color=term), shape=23) +
    # geom_point(data = summaries_df %>% filter(p.value<0.05, term=="moderate to severe retardation"), mapping = aes(color=term, x=MOL, y=10^max +.4), shape=8) +
    scale_color_manual(values = ann_colors$growth_failure) +
    scale_y_continuous(transform = "log10", labels = scales::percent) +
    theme_bw() +
  theme(legend.position = "bottom") +
    theme(panel.grid = element_blank()) +
    geom_errorbar(aes(ymin = 10^(mean-se), ymax=10^(mean+se), color=study_group), alpha=0.5, width = 0.3) +
    facet_wrap(~Taxa, ncol=3, scales="free") +
  labs(y="Concentration (umol/g log-scale)")

# table_export(summaries_df, "Taxonomic_table_export")

summaries_df %>%
  filter(p.value < 0.05) %>%
  pander(split.tab=Inf, emphasize.strong.rows = which(.$fdr<=0.05)) #%>%
  #column_spec(c(2:5), width = "3em") ## if the table is running off the page, you can manually specify the width of the columns

rm(myform)
```


### GENUS  
Growth Faltering:  
```{r fig.height=9, fig.width=8}
myform="log10_props ~ study_group"

summaries_df <- props_toTest_g %>%
  inner_join(s2 %>% select(SampleID, study_group=growth_cat, sample_type=SampleType, TimePoint, SubjectID, WOL), by = "SampleID") %>%
  filter(!is.na(study_group)) %>%
  group_by(WOL, Taxa, study_group, sample_type) %>%
    mutate(mean=mean(log10_props)) %>%
  ungroup() %>%
  
  group_by(Taxa) %>% mutate(max=max(mean)) %>% ungroup() %>%
  
  group_by(Taxa, WOL) %>%
    mutate(ngroups=n_distinct(study_group)) %>%
  ungroup() %>%
  
  filter(ngroups>=2) %>% # Only test PMA_rounded where you have both study groups
  
  group_by(Taxa, WOL, max) %>%
    do(tidy(lm(as.formula(myform), data=.))) %>%
  ungroup() %>%
  
  filter(!grepl("Intercept|Residuals", term)) %>%
  mutate(term = sub("study_group", "", term)) %>%
  
  group_by(term, WOL) %>%
    mutate(fdr = p.adjust(p.value, method="BH")) %>%
  ungroup()

props_toTest_g %>%
  inner_join(s2 %>% select(SampleID, study_group=growth_cat, sample_type=SampleType, TimePoint, SubjectID, WOL), by = "SampleID") %>%
  filter(!is.na(study_group)) %>%
  # mutate(Taxa = reorder(Taxa, -log10_props)) %>%
  # mutate(Taxa = fct_relabel(Taxa, function(x) gsub(" ", "\n", x))) %>%
  group_by(WOL, Taxa, study_group, sample_type) %>%
    summarise(mean=mean(log10_props), sd=sd(log10_props), n=n(), max=max(log10_props)) %>%
  ungroup() %>%
  mutate(se=sd/sqrt(n)) %>%
  ggplot(aes(y=10^(mean), x=WOL)) +
    # geom_vline(aes(xintercept = PMA_rounded), lwd=0.3, color="grey") + # guide lines
    geom_line(aes(color=study_group, group = study_group)) +  
    geom_point(aes(color=study_group)) +
    geom_point(data = summaries_df %>% filter(fdr<0.1, term==levels(s2$growth_cat)[2]), mapping = aes(x=WOL, y=10^max +.2, color=term), shape=8) +
    geom_point(data = summaries_df %>% filter(fdr>=0.1, p.value<0.05, term==levels(s2$growth_cat)[2]), mapping = aes(x=WOL, y=10^max +.2, color=term), shape=23) +

    geom_point(data = summaries_df %>% filter(fdr<0.1, term==levels(s2$growth_cat)[3]), mapping = aes(color=term, x=WOL, y=10^max +.4), shape=8) +
    geom_point(data = summaries_df %>% filter(p.value<0.05, fdr>=0.1, term==levels(s2$growth_cat)[3]), mapping = aes(color=term, x=WOL, y=10^max +.4), shape=23) +
    scale_color_manual(values = ann_colors$growth_cat) +
    scale_y_continuous(transform = "log10", labels = scales::percent) +
    theme_bw() +
  theme(legend.position = "bottom") +
    theme(panel.grid = element_blank()) +
    geom_errorbar(aes(ymin = 10^(mean-se), ymax=10^(mean+se), color=study_group), alpha=0.5, width = 0.3) +
    facet_wrap(~Taxa, ncol=3, scales="free") +
  labs(y="Concentration (umol/g log-scale)")

# table_export(summaries_df, "Taxonomic_table_export")
summaries_df %>%
  filter(p.value < 0.05) %>%
  pander(split.tab=Inf, emphasize.strong.rows = which(.$fdr<=0.05)) #%>%
  #column_spec(c(2:5), width = "3em") ## if the table is running off the page, you can manually specify the width of the columns

rm(myform)
```

Growth Failure:  
Asterisk/star for fdr-adjusted p-value<0.05. Diamond for p-unadjusted<0.05  
```{r fig.height=9, fig.width=8}
myform="log10_props ~ study_group"

summaries_df <- props_toTest_g %>%
  inner_join(s2 %>% select(SampleID, study_group=growth_failure, sample_type=SampleType, TimePoint, SubjectID, WOL), by = "SampleID") %>%
  filter(!is.na(study_group)) %>%
  group_by(WOL, Taxa, study_group, sample_type) %>%
    mutate(mean=mean(log10_props)) %>%
  ungroup() %>%
  
  group_by(Taxa) %>% mutate(max=max(mean)) %>% ungroup() %>%
  
  group_by(Taxa, WOL) %>%
    mutate(ngroups=n_distinct(study_group)) %>%
  ungroup() %>%
  
  filter(ngroups>=2) %>% # Only test PMA_rounded where you have both study groups
  
  group_by(Taxa, WOL, max) %>%
    do(tidy(lm(as.formula(myform), data=.))) %>%
  ungroup() %>%
  
  filter(!grepl("Intercept|Residuals", term)) %>%
  mutate(term = sub("study_group", "", term)) %>%
  
  group_by(term, WOL) %>%
    mutate(fdr = p.adjust(p.value, method="BH")) %>%
  ungroup()

props_toTest_g %>%
  inner_join(s2 %>% select(SampleID, study_group=growth_failure, sample_type=SampleType, TimePoint, SubjectID, WOL), by = "SampleID") %>%
  filter(!is.na(study_group)) %>%
  # mutate(Taxa = reorder(Taxa, -log10_props)) %>%
  # mutate(Taxa = fct_relabel(Taxa, function(x) gsub(" ", "\n", x))) %>%
  group_by(WOL, Taxa, study_group, sample_type) %>%
    summarise(mean=mean(log10_props), sd=sd(log10_props), n=n(), max=max(log10_props)) %>%
  ungroup() %>%
  mutate(se=sd/sqrt(n)) %>%
  ggplot(aes(y=10^(mean), x=WOL)) +
    # geom_vline(aes(xintercept = PMA_rounded), lwd=0.3, color="grey") + # guide lines
    geom_line(aes(color=study_group, group = study_group)) +  
    geom_point(aes(color=study_group)) +
    geom_point(data = summaries_df %>% filter(fdr<0.05), mapping = aes(x=WOL, y=10^max +.2, color=term), shape=8) +
    geom_point(data = summaries_df %>% filter(fdr>0.05, p.value<0.05), mapping = aes(x=WOL, y=10^max +.2, color=term), shape=23) +
    # geom_point(data = summaries_df %>% filter(p.value<0.05, term=="moderate to severe retardation"), mapping = aes(color=term, x=WOL, y=10^max +.4), shape=8) +
    scale_color_manual(values = ann_colors$growth_failure) +
    scale_y_continuous(transform = "log10", labels = scales::percent) +
    theme_bw() +
  theme(legend.position = "bottom") +
    theme(panel.grid = element_blank()) +
    geom_errorbar(aes(ymin = 10^(mean-se), ymax=10^(mean+se), color=study_group), alpha=0.5, width = 0.3) +
    facet_wrap(~Taxa, ncol=3, scales="free") +
  labs(y="Concentration (umol/g log-scale)")

# table_export(summaries_df, "Taxonomic_table_export")

summaries_df %>%
  filter(p.value < 0.05) %>%
  pander(split.tab=Inf, emphasize.strong.rows = which(.$fdr<=0.05)) #%>%
  #column_spec(c(2:5), width = "3em") ## if the table is running off the page, you can manually specify the width of the columns

rm(myform)
```




## Continuous trajectories over age  
### SPECIES
```{r fig.height=9, fig.width=8}
myform="log10_props ~ study_group * WOL"
myform_covariates="log10_props ~ log_npo_days_prop + bwt_z + study_group * WOL"

summaries_df <- run_lmer_on_props(factor_to_test = "Taxa", props_toTest = props_toTest, s_toTest = s2 %>% select(SampleID, study_group=growth_cat, sample_type=SampleType, TimePoint, SubjectID, WOL, npo_days_prop, bwt_z) %>% mutate(log_npo_days_prop = log10(npo_days_prop + min(s2$npo_days_prop[s2$npo_days_prop>0])/2)), form1 = myform, rep_mes_label = "SubjectID", p_cutoff =  0.05) 

summaries_df %>%
  filter(p.value < 0.05) %>%
  pander(split.table=Inf, digits=2, caption = glue("Linear mixed effects models of log10 transformed relative abundance with subject id as random effect. Filtered with p.value < 0.05. Full formula: {form1}"), emphasize.strong.rows = which(.$fdr<=0.05))

# summaries_df %>% filter(grepl(":", term), p.value<0.05)
props_toTest %>%
  inner_join(s2 %>% select(SampleID, study_group=growth_cat, sample_type=SampleType, TimePoint, SubjectID, HOL), by = "SampleID") %>%
  filter(!is.na(study_group)) %>%
  ggplot(aes(y=props, x=HOL)) +
    geom_point(aes(color=study_group), alpha=0.1) +
    geom_smooth(aes(color=study_group, group = study_group), se = F) +  
    scale_color_manual(values = ann_colors$growth_cat) +
    scale_y_continuous(transform = "log10", labels = scales::percent) +
    theme_bw() +
  theme(legend.position = "bottom", panel.grid = element_blank()) +
    facet_wrap(~Taxa, ncol=3, scales="free") +
  labs(y="Concentration (umol/g log-scale)")
```


```{r fig.height=3, fig.width=6}
sigtaxa <- summaries_df %>% filter(grepl(":", term), fdr<0.1) %>% pull(Taxa)
tmp=summaries_df %>% filter(grepl(":", term), fdr<0.1) %>% mutate(siggroup = gsub("study_group|:WOL", "", term)) %>% mutate(`Growth Faltering` = factor(x = siggroup, levels = c("none", "mild retardation", "moderate to severe retardation"), labels = c("none", "mild faltering", "moderate to severe faltering"))) %>% mutate(ypos=if_else(Taxa == "Firmicutes Clostridium perfringens", 1e-5, 1e-5))

props_toTest %>%
  inner_join(s2 %>% select(SampleID, study_group=growth_cat, sample_type=SampleType, TimePoint, SubjectID, HOL), by = "SampleID") %>%
  filter(!is.na(study_group), Taxa %in% sigtaxa) %>%
  # ggplot(aes(y=props, x=HOL)) +
  #   geom_point(aes(color=study_group), alpha=0.1) +
  #   geom_smooth(aes(color=study_group, group = study_group), se = F) +  
    mutate(`Growth Faltering` = factor(x = study_group, labels = c("none", "mild faltering", "moderate to severe faltering"))) %>%
  filter(!is.na(study_group)) %>%
  ggplot(aes(y=props, x=HOL)) +
    geom_point(aes(color=`Growth Faltering`), alpha=0.1) +
    geom_smooth(aes(color=`Growth Faltering`, group = `Growth Faltering`), se = F) +  
    geom_label(data = tmp, aes(x=2200, y=ypos, color=`Growth Faltering`, label = paste("q=", signif(fdr, 2))), fill=NA, show.legend = FALSE) + 
    scale_color_manual(values = as.character(ann_colors$growth_cat)) +
    scale_y_continuous(transform = "log10", labels = scales::percent) +
    theme_bw() +
  theme(legend.position = "bottom", panel.grid = element_blank(), text = element_text(size=10, color="black"), axis.text = element_text(size=8, color="black"), strip.text = element_text(size=10, color="black", face="italic"), legend.text = element_text(size=10, color="black")) +    
  facet_wrap(~gsub("Firmicutes |Proteobacteria ", "", Taxa), ncol=3, scales="free") +
  labs(y="Relative Abundance", x="Hours of Life")

ggsave(here("Output", "SUPPFIG4B_speciesRA_GrFaltering.pdf"), width=6, height = 3, useDingbats = F, dpi = 600)
```

WITH COVARIATES: `r myform_covariates`   
```{r}
myform_covariates="log10_props ~ log_npo_days_prop + bwt_z + study_group * WOL"

summaries_df <- run_lmer_on_props(factor_to_test = "Taxa", props_toTest = props_toTest, s_toTest = s2 %>% select(SampleID, study_group=growth_cat, sample_type=SampleType, TimePoint, SubjectID, WOL, npo_days_prop, bwt_z) %>% mutate(log_npo_days_prop = log10(npo_days_prop + min(s2$npo_days_prop[s2$npo_days_prop>0])/2)), form1 = myform_covariates, rep_mes_label = "SubjectID", p_cutoff =  0.05)

summaries_df %>%
  filter(fdr < 0.1, grepl(":", term)) %>%
  pander(split.table=Inf, digits=2, caption = glue("Linear mixed effects models of log10 transformed relative abundance with subject id as random effect. Filtered with fdr < 0.1 and only interaction terms. Full formula: {myform_covariates}"), emphasize.strong.rows = which(.$fdr<=0.05))
```


```{r fig.height=9, fig.width=8}
# GROWTH FAILURE
summaries_df <- run_lmer_on_props(factor_to_test = "Taxa", props_toTest = props_toTest, s_toTest = s2 %>% select(SampleID, study_group=growth_failure, sample_type=SampleType, TimePoint, SubjectID, WOL), form1 = myform, rep_mes_label = "SubjectID", p_cutoff =  0.05) 

summaries_df %>%
  filter(p.value < 0.05) %>%
  pander(split.table=Inf, digits=2, caption = glue("Linear mixed effects models of log10 transformed relative abundance with subject id as random effect. Filtered with p.value < 0.05. Full formula: {form1}"), emphasize.strong.rows = which(.$fdr<=0.05))
rm(myform)

props_toTest %>%
  inner_join(s2 %>% select(SampleID, study_group=growth_failure, sample_type=SampleType, TimePoint, SubjectID, HOL), by = "SampleID") %>%
  filter(!is.na(study_group)) %>%
  ggplot(aes(y=props, x=HOL)) +
    geom_point(aes(color=study_group), alpha=0.1) +
    geom_smooth(aes(color=study_group, group = study_group), se = F) +  
    scale_color_manual(values = ann_colors$growth_failure) +
    scale_y_continuous(transform = "log10", labels = scales::percent) +
    theme_bw() +
  theme(legend.position = "bottom", panel.grid = element_blank()) +
    facet_wrap(~Taxa, ncol=3, scales="free") +
  labs(y="Concentration (umol/g log-scale)")

```

```{r fig.width=8, fig.height=5}
sigtaxa <- summaries_df %>% filter(grepl(":", term), fdr<0.1) %>% pull(Taxa)
tmp=summaries_df %>% filter(grepl(":", term), fdr<0.1) %>% mutate(siggroup = gsub("study_group|:WOL", "", term)) %>% mutate(ypos=if_else(Taxa=="Proteobacteria Enterobacteriaceae", 1e-5, if_else(Taxa=="Firmicutes Staphylococcus epidermidis", 0.2, 1e-6)))

props_toTest %>%
  inner_join(s2 %>% select(SampleID, study_group=growth_failure, sample_type=SampleType, TimePoint, SubjectID, HOL), by = "SampleID") %>%
  filter(!is.na(study_group), Taxa %in% sigtaxa) %>%
  ggplot(aes(y=props, x=HOL)) +
    geom_point(aes(color=study_group), alpha=0.1) +
    geom_smooth(aes(color=study_group, group = study_group), se = F) +  
    geom_label(data = tmp, aes(x=2100, y=ypos, color=siggroup, label = paste("q=", signif(fdr, 2))), fill=NA, show.legend = F) + 
    scale_color_manual(values = ann_colors$growth_failure) +
    scale_y_continuous(transform = "log10", labels = scales::percent) +
    theme_bw() +
  theme(legend.position = "bottom", panel.grid = element_blank()) +
    facet_wrap(~gsub("Firmicutes |Proteobacteria ", "", Taxa), ncol=3, scales="free") +
    theme(legend.position = "bottom", panel.grid = element_blank(), text = element_text(size=10, color="black"), axis.text = element_text(size=8, color="black"), strip.text = element_text(size=10, color="black", face="italic"), legend.text = element_text(size=10, color="black")) +    
  labs(y="Relative Abundance", color="Growth Failure")

ggsave(here("Output", "SUPPFIG2B_speciesRA_GrFailure.pdf"), width=6, height = 3, useDingbats = F, dpi = 600)
```

WITH COVARIATES: `r myform_covariates`   
```{r}
myform_covariates="log10_props ~ study_group * WOL + log_npo_days_prop + bwt_z"

summaries_df <- run_lmer_on_props(factor_to_test = "Taxa", props_toTest = props_toTest, s_toTest = s2 %>% select(SampleID, study_group=growth_failure, sample_type=SampleType, TimePoint, SubjectID, WOL, npo_days_n, npo_days_prop, bwt_z) %>% mutate(log_npo_days_prop = log10(npo_days_prop + min(s2$npo_days_prop[s2$npo_days_prop>0])/2)), form1 = myform_covariates, rep_mes_label = "SubjectID", p_cutoff =  0.05) 

summaries_df %>%
  filter(fdr < 0.1, grepl(":", term)) %>%
  pander(split.table=Inf, digits=2, caption = glue("Linear mixed effects models of log10 transformed relative abundance with subject id as random effect. Filtered with fdr < 0.1 and only interaction terms. Full formula: {myform_covariates}"), emphasize.strong.rows = which(.$fdr<=0.05))
```

### GENUS  
#### NEW Supp Fig 2 (12/18/25)   
```{r fig.height=3, fig.width=8}
myform="log10_props ~ study_group * WOL"

summaries_df <- run_lmer_on_props(factor_to_test = "Taxa", props_toTest = props_toTest_g, s_toTest = s2 %>% select(SampleID, study_group=growth_cat, sample_type=SampleType, TimePoint, SubjectID, WOL), form1 = myform, rep_mes_label = "SubjectID", p_cutoff =  0.05) 

summaries_df %>%
  filter(p.value < 0.05) %>%
  pander(split.table=Inf, digits=2, caption = glue("Linear mixed effects models of log10 transformed relative abundance with subject id as random effect. Filtered with p.value < 0.05. Full formula: {form1}"), emphasize.strong.rows = which(.$fdr<=0.05))

sigtaxa <- summaries_df %>% filter(grepl(":", term), fdr<0.1) %>% pull(Taxa)
tmp=summaries_df %>% filter(grepl(":", term), fdr<0.1) %>% mutate(siggroup = gsub("study_group|:WOL", "", term)) %>% mutate(`Growth Faltering` = factor(x = siggroup, levels = c("none", "mild retardation", "moderate to severe retardation"), labels = c("none", "mild faltering", "moderate to severe faltering"))) %>% mutate(ypos=if_else(Taxa == "Proteobacteria Escherichia", 1e-5, 1e-6))

props_toTest_g %>%
  inner_join(s2 %>% select(SampleID, study_group=growth_cat, sample_type=SampleType, TimePoint, SubjectID, HOL), by = "SampleID") %>%
  filter(!is.na(study_group), Taxa %in% sigtaxa) %>%
  # ggplot(aes(y=props, x=HOL)) +
  #   geom_point(aes(color=study_group), alpha=0.1) +
  #   geom_smooth(aes(color=study_group, group = study_group), se = F) +  
    mutate(`Growth Faltering` = factor(x = study_group, labels = c("none", "mild faltering", "moderate to severe faltering"))) %>%
  filter(!is.na(study_group)) %>%
  ggplot(aes(y=props, x=HOL)) +
    geom_point(aes(color=`Growth Faltering`), alpha=0.1) +
    geom_smooth(aes(color=`Growth Faltering`, group = `Growth Faltering`), se = F) +  
    geom_label(data = tmp, aes(x=1900, y=ypos, color=`Growth Faltering`, label = paste("q=", signif(fdr, 2))), fill=NA, show.legend = FALSE) + 
    scale_color_manual(values = as.character(ann_colors$growth_cat)) +
    scale_y_continuous(transform = "log10", labels = scales::percent) +
    theme_bw() +
  theme(legend.position = "bottom", panel.grid = element_blank(), text = element_text(size=10, color="black"), axis.text = element_text(size=8, color="black"), strip.text = element_text(size=10, color="black", face="italic"), legend.text = element_text(size=10, color="black")) +    
  facet_wrap(~gsub("Firmicutes |Proteobacteria ", "", Taxa), ncol=3, scales="free") +
  labs(y="Relative Abundance", x="Hours of Life")

ggsave(here("Output", "SUPPFIG2A_genusRA_GrFaltering.pdf"), width=8, height = 3, useDingbats = F, dpi = 600)
```

WITH COVARIATES: `r myform_covariates`   
```{r}
myform_covariates="log10_props ~ log_npo_days_prop + bwt_z + study_group * WOL"

summaries_df <- run_lmer_on_props(factor_to_test = "Taxa", props_toTest = props_toTest_g, s_toTest = s2 %>% select(SampleID, study_group=growth_cat, sample_type=SampleType, TimePoint, SubjectID, WOL, npo_days_prop, bwt_z) %>% mutate(log_npo_days_prop = log10(npo_days_prop + min(s2$npo_days_prop[s2$npo_days_prop>0])/2)), form1 = myform_covariates, rep_mes_label = "SubjectID", p_cutoff =  0.05)

summaries_df %>%
  filter(fdr < 0.1, grepl(":", term)) %>%
  pander(split.table=Inf, digits=2, caption = glue("Linear mixed effects models of log10 transformed relative abundance with subject id as random effect. Filtered with fdr < 0.1 and only interaction terms. Full formula: {myform_covariates}"), emphasize.strong.rows = which(.$fdr<=0.05))
```


```{r fig.height=3, fig.width=6}
# GROWTH FAILURE
summaries_df <- run_lmer_on_props(factor_to_test = "Taxa", props_toTest = props_toTest_g, s_toTest = s2 %>% select(SampleID, study_group=growth_failure, sample_type=SampleType, TimePoint, SubjectID, WOL), form1 = myform, rep_mes_label = "SubjectID", p_cutoff =  0.05) 

summaries_df %>%
  filter(p.value < 0.05) %>%
  pander(split.table=Inf, digits=2, caption = glue("Linear mixed effects models of log10 transformed relative abundance with subject id as random effect. Filtered with p.value < 0.05. Full formula: {form1}"), emphasize.strong.rows = which(.$fdr<=0.05))
rm(myform)

sigtaxa <- summaries_df %>% filter(grepl(":", term), fdr<0.1) %>% pull(Taxa)
tmp=summaries_df %>% filter(grepl(":", term), fdr<0.1) %>% mutate(siggroup = gsub("study_group|:WOL", "", term)) %>% mutate(ypos=if_else(Taxa=="Firmicutes Staphylococcus", 0.5, 1e-6))

props_toTest_g %>%
  inner_join(s2 %>% select(SampleID, study_group=growth_failure, sample_type=SampleType, TimePoint, SubjectID, HOL), by = "SampleID") %>%
  filter(!is.na(study_group), Taxa %in% sigtaxa) %>%
  ggplot(aes(y=props, x=HOL)) +
    geom_point(aes(color=study_group), alpha=0.1) +
    geom_smooth(aes(color=study_group, group = study_group), se = F) +  
    geom_label(data = tmp, aes(x=2100, y=ypos, color=siggroup, label = paste("q=", signif(fdr, 2))), fill=NA, show.legend = F) + 
    scale_color_manual(values = ann_colors$growth_failure) +
    scale_y_continuous(transform = "log10", labels = scales::percent) +
    theme_bw() +
  theme(legend.position = "bottom", panel.grid = element_blank()) +
    facet_wrap(~gsub("Firmicutes |Proteobacteria ", "", Taxa), ncol=3, scales="free") +
    theme(legend.position = "bottom", panel.grid = element_blank(), text = element_text(size=10, color="black"), axis.text = element_text(size=8, color="black"), strip.text = element_text(size=10, color="black", face="italic"), legend.text = element_text(size=10, color="black")) +    
  labs(y="Relative Abundance", color="Growth Failure")

ggsave(here("Output", "SUPPFIG2A_genusRA_GrFailure.pdf"), width=6, height = 3, useDingbats = F, dpi = 600)
```

WITH COVARIATES: `r myform_covariates`   
```{r}
myform_covariates="log10_props ~ log_npo_days_prop + bwt_z + study_group * WOL"

summaries_df <- run_lmer_on_props(factor_to_test = "Taxa", props_toTest = props_toTest_g, s_toTest = s2 %>% select(SampleID, study_group=growth_failure, sample_type=SampleType, TimePoint, SubjectID, WOL, npo_days_prop, bwt_z) %>% mutate(log_npo_days_prop = log10(npo_days_prop + min(s2$npo_days_prop[s2$npo_days_prop>0])/2)), form1 = myform_covariates, rep_mes_label = "SubjectID", p_cutoff = 0.05)

summaries_df %>%
  filter(fdr < 0.1, grepl(":", term)) %>%
  pander(split.table=Inf, digits=2, caption = glue("Linear mixed effects models of log10 transformed relative abundance with subject id as random effect. Filtered with fdr < 0.1 and only interaction terms. Full formula: {myform_covariates}"), emphasize.strong.rows = which(.$fdr<=0.05))
```

# OXYGEN USAGE  
```{r}
library(whatbacteria)
dbedit <- bind_rows(c(taxon="Klebsiella aerogenes", rank="Species", aerobic_status="facultative anaerobe", gram_stain="Gram-negative", doi="Google"), 
          c(taxon="Citrobacter freundii", rank="Species", aerobic_status="facultative anaerobe", gram_stain="Gram-negative", doi="Google"),
          c(taxon="Enterobacterales", rank="Order", aerobic_status="facultative anaerobe", gram_stain="Gram-negative", doi="Google"),
          c(taxon="Klebsiella pneumoniae", rank="Species", aerobic_status="facultative anaerobe", gram_stain="Gram-negative", doi="Google"),
          c(taxon="Proteus mirabilis", rank="Species", aerobic_status="facultative anaerobe", gram_stain="Gram-negative", doi="Google"),
          c(taxon="Klebsiella oxytoca", rank="Species", aerobic_status="facultative anaerobe", gram_stain="Gram-negative", doi="Google"),
          c(taxon="Klebsiella", rank="Genus", aerobic_status="facultative anaerobe", gram_stain="Gram-negative", doi="Google"),
           whatbacteria::taxon_phenotypes)
# bind_cols(unique(props_toTest$Taxa), what_phenotype(db = dbedit, lineage = unique(props_toTest$Taxa), phenotype ="aerobic_status")) # %>%

props_toTest <- props_toTest %>%
  mutate(aerobic_status = what_phenotype(db = dbedit, lineage = Taxa, phenotype = "aerobic_status"),
         gram_stain = what_phenotype(db = dbedit, lineage = Taxa, phenotype = "gram_stain"))
```

```{r fig.width=6.5}
props_toTest %>%
  group_by(SampleID, aerobic_status) %>%
    summarise(sumprops = sum(props), sumpropslog=sum(log10_props)) %>%
  ungroup() %>%
  inner_join(s2) %>%
  ggplot(aes(y=sumprops, x=growth_failure)) +
    geom_boxplot(aes(group = interaction(growth_failure,aerobic_status), color=aerobic_status)) +
  # ggpubr::stat_compare_means(mapping = aes(group = growth_failure)) + 
  theme_bw() 

props_toTest %>%
  group_by(SampleID, aerobic_status) %>%
    summarise(sumprops = sum(props), sumpropslog=sum(log10_props)) %>%
  ungroup() %>%
  inner_join(s2) %>%
  ggplot(aes(y=sumprops, x=WOL)) +
    geom_boxplot(aes(group = interaction(growth_failure,aerobic_status,WOL), fill=growth_failure, color=aerobic_status), outlier.shape=NA) +
  scale_fill_manual(values = c("white", "lightgrey")) +
  # ggpubr::stat_compare_means(mapping = aes(group = growth_failure)) + 
  theme_bw() 

# RATIO
props_aerobic_ratio <- props_toTest %>%
  group_by(SampleID, aerobic_status) %>%
    summarise(sumprops = sum(props)) %>%
  ungroup() %>%
  mutate(sumpropslog = log10(sumprops)) %>%
  group_by(SampleID) %>%
    arrange(aerobic_status) %>%
    summarise(propsratio=last(sumprops)/first(sumprops), logratio=last(sumpropslog)-first(sumpropslog), numerator=last(aerobic_status), denominator=first(aerobic_status)) %>%
  inner_join(s2) 

props_aerobic_ratio %>%
  ggplot(aes(y=propsratio, x=growth_failure)) +
  geom_boxplot(aes(group = growth_failure)) +
  ggpubr::stat_compare_means(mapping = aes(group = growth_failure)) + 
  scale_y_log10() +
  theme_bw() +
  labs(y="Obligate : Facultative Anaerobes ratio")

# props_aerobic_ratio %>%
#   ggplot(aes(y=propsratio, x=WOL)) +
#   geom_boxplot(aes(group = interaction(WOL, growth_failure), fill=growth_failure), outlier.shape=21) +
  # ggpubr::stat_compare_means(mapping = aes(group = growth_failure), hide.ns = F, label = "p.signif", method = "wilcox.test") + 
  # scale_y_log10() +
  # scale_fill_manual(values = ann_colors$growth_failure) +
  # theme_bw() 
```

```{r fig.width=6.5}
props_toTest %>%
  group_by(SampleID, aerobic_status) %>%
    summarise(sumprops = sum(props), sumpropslog=sum(log10_props)) %>%
  ungroup() %>%
  inner_join(s2) %>%
  ggplot(aes(y=sumprops, x=growth_cat)) +
    geom_boxplot(aes(group = interaction(growth_cat,aerobic_status), color=aerobic_status)) +
  # ggpubr::stat_compare_means(mapping = aes(group = growth_cat)) + 
  theme_bw() 

props_toTest %>%
  group_by(SampleID, aerobic_status) %>%
    summarise(sumprops = sum(props), sumpropslog=sum(log10_props)) %>%
  ungroup() %>%
  inner_join(s2) %>%
  ggplot(aes(y=sumprops, x=WOL)) +
    geom_boxplot(aes(group = interaction(growth_cat,aerobic_status,WOL), fill=growth_cat, color=aerobic_status), outlier.shape=NA) +
  scale_fill_manual(values = c("white", "lightgrey", "darkgrey")) +
  # ggpubr::stat_compare_means(mapping = aes(group = growth_cat)) + 
  theme_bw() +
  theme(legend.position = "bottom")

props_toTest %>%
  inner_join(s2 %>% select(SampleID, WOL, growth_failure, growth_cat)) %>%
  group_by(aerobic_status, Taxa, WOL, growth_failure) %>%
    summarise(meanprop = mean(props), sumpropslog=sum(log10_props), n=n()) %>%
  ungroup() %>%
  # mutate(meanprop=sumprops/n) %>%
  ggplot(aes(y=meanprop, x=WOL)) +
    geom_col(aes(fill=Taxa)) +
  # facet_wrap(~paste(growth_failure, "Failure")+aerobic_status, ncol=2) +
  facet_grid(paste(growth_failure, "Failure")~aerobic_status) +
  scale_fill_viridis_d(option = "turbo") +
  theme_bw() +
  theme(legend.position = "right", legend.key.height = unit(0.5, units = "cm"))

props_toTest %>%
  inner_join(s2 %>% select(SampleID, WOL, growth_failure, growth_cat)) %>%
  group_by(aerobic_status, Taxa, WOL, growth_cat) %>%
    summarise(meanprop = mean(props), sumpropslog=sum(log10_props), n=n()) %>%
  ungroup() %>%
  # mutate(meanprop=sumprops/n) %>%
  ggplot(aes(y=meanprop, x=WOL)) +
    geom_col(aes(fill=Taxa)) +
  facet_grid(growth_cat~aerobic_status) +
  scale_fill_viridis_d(option = "turbo") +
  theme_bw() +
  theme(legend.position = "right", legend.key.height = unit(0.5, units = "cm"))

# RATIO
props_aerobic_ratio <- props_toTest %>%
  group_by(SampleID, aerobic_status) %>%
    summarise(sumprops = sum(props+1e-6)) %>%
  ungroup() %>%
  mutate(sumpropslog = log10(sumprops)) %>%
  group_by(SampleID) %>%
    arrange(aerobic_status) %>%
    summarise(propsratio=last(sumprops)/first(sumprops), logratio=last(sumpropslog)-first(sumpropslog), numerator=last(aerobic_status), denominator=first(aerobic_status)) %>%
  inner_join(s2) 

props_aerobic_ratio %>%
  ggplot(aes(y=propsratio, x=growth_cat)) +
  geom_boxplot(aes(group = growth_cat)) +
  ggpubr::stat_compare_means(mapping = aes(group = growth_cat)) + 
  scale_y_log10() +
  theme_bw() +
  labs(y="Obligate : Facultative Anaerobes ratio")

summaries_df <- props_aerobic_ratio %>% 
  filter(WOL<=14, !is.na(growth_cat), !is.na(logratio)) %>%
  group_by(WOL) %>%
    # tidy(lm(data = ., formula=logratio ~ growth_cat)
    do(tidy(lm(logratio ~ growth_cat, data=.))) %>%
  ungroup()  %>%
  filter(!grepl("Intercept|Residual", term)) %>%
  mutate(term=gsub("growth_cat", "", term))

props_aerobic_ratio %>%
  ggplot(aes(y=propsratio, x=WOL)) +
  geom_boxplot(aes(group = interaction(WOL, growth_cat), fill=growth_cat), outlier.shape=21) +
  # ggpubr::stat_compare_means(mapping = aes(group = growth_cat), hide.ns = F, label = "p.signif", show.legend = T, method = "wilcox.test", position = position_jitter(height =1)) + 
  geom_point(data=summaries_df %>% filter(term=="mild retardation", p.value<0.1), shape=8, mapping = aes(color=term, x=WOL, y=70)) +
  geom_point(data=summaries_df %>% filter(term=="moderate to severe retardation", p.value<0.1), shape=8, mapping = aes(color=term, x=WOL, y=100)) +
  scale_y_log10() +
  scale_fill_manual(values = ann_colors$growth_cat) +
  scale_color_manual(values = ann_colors$growth_cat, guide='none') +
  theme_bw() +
  theme(legend.position = "bottom")+
  labs(y="Obligate : Facultative Anaerobes ratio")

summaries_df %>% filter(p.value<0.1) %>% pander(split.tab=Inf)


summaries_df <- props_aerobic_ratio %>% 
  filter(WOL<=12, !is.na(growth_failure), !is.na(logratio)) %>%
  group_by(WOL) %>%
    # tidy(lm(data = ., formula=logratio ~ growth_cat)
    do(tidy(lm(logratio ~ growth_failure, data=.))) %>%
  ungroup()  %>%
  filter(!grepl("Intercept|Residual", term)) %>%
  mutate(term=gsub("growth_failure", "", term))

props_aerobic_ratio %>%
  ggplot(aes(y=propsratio, x=WOL)) +
  geom_boxplot(aes(group = interaction(WOL, growth_failure), fill=growth_failure), outlier.shape=21) +
  # ggpubr::stat_compare_means(mapping = aes(group = growth_cat), hide.ns = F, label = "p.signif", show.legend = T, method = "wilcox.test", position = position_jitter(height =1)) + 
  geom_point(data=summaries_df %>% filter(p.value<0.1), shape=8, mapping = aes(color=term, x=WOL, y=70)) +
  # geom_point(data=summaries_df %>% filter(term=="moderate to severe retardation", p.value<0.1), shape=8, mapping = aes(color=term, x=WOL, y=100)) +
  scale_y_log10() +
  scale_fill_manual(values = ann_colors$growth_failure) +
  scale_color_manual(values = ann_colors$growth_failure, guide='none') +
  theme_bw() +
  theme(legend.position = "bottom") +
  labs(y="Obligate : Facultative Anaerobes ratio")

summaries_df %>% filter(p.value<0.1) %>% pander(split.tab=Inf)
```

```{r fig.width=6.5}
summaries_df <- props_aerobic_ratio %>% 
  filter(WOL<=14, !is.na(growth_cat), !is.na(logratio)) %>%
  group_by(MOL) %>%
    # tidy(lm(data = ., formula=logratio ~ growth_cat)
    do(tidy(lm(logratio ~ growth_cat, data=.))) %>%
  ungroup()  %>%
  filter(!grepl("Intercept|Residual", term)) %>%
  mutate(term=gsub("growth_cat", "", term))

props_aerobic_ratio %>%
  ggplot(aes(y=propsratio, x=MOL)) +
  geom_boxplot(aes(group = interaction(MOL, growth_cat), fill=growth_cat), outlier.shape=21) +
  # ggpubr::stat_compare_means(mapping = aes(group = growth_cat), hide.ns = F, label = "p.signif", show.legend = T, method = "wilcox.test", position = position_jitter(height =1)) + 
  geom_point(data=summaries_df %>% filter(term=="mild retardation", p.value<0.1), shape=8, mapping = aes(color=term, x=MOL, y=70)) +
  geom_point(data=summaries_df %>% filter(term=="moderate to severe retardation", p.value<0.1), shape=8, mapping = aes(color=term, x=MOL, y=100)) +
  scale_y_log10() +
  scale_fill_manual(values = ann_colors$growth_cat) +
  scale_color_manual(values = ann_colors$growth_cat, guide='none') +
  theme_bw() +
  theme(legend.position = "bottom")+
  labs(y="Obligate : Facultative Anaerobes ratio")

summaries_df %>% filter(p.value<0.1) %>% pander(split.tab=Inf)


summaries_df <- props_aerobic_ratio %>% 
  filter(WOL<=12, !is.na(growth_failure), !is.na(logratio)) %>%
  group_by(MOL) %>%
    # tidy(lm(data = ., formula=logratio ~ growth_cat)
    do(tidy(lm(logratio ~ growth_failure, data=.))) %>%
  ungroup()  %>%
  filter(!grepl("Intercept|Residual", term)) %>%
  mutate(term=gsub("growth_failure", "", term))

props_aerobic_ratio %>%
  ggplot(aes(y=propsratio, x=MOL)) +
  geom_boxplot(aes(group = interaction(MOL, growth_failure), fill=growth_failure), outlier.shape=21) +
  # ggpubr::stat_compare_means(mapping = aes(group = growth_cat), hide.ns = F, label = "p.signif", show.legend = T, method = "wilcox.test", position = position_jitter(height =1)) + 
  geom_point(data=summaries_df %>% filter(p.value<0.1), shape=8, mapping = aes(color=term, x=MOL, y=70)) +
  # geom_point(data=summaries_df %>% filter(term=="moderate to severe retardation", p.value<0.1), shape=8, mapping = aes(color=term, x=WOL, y=100)) +
  scale_y_log10() +
  scale_fill_manual(values = ann_colors$growth_failure) +
  scale_color_manual(values = ann_colors$growth_failure, guide='none') +
  theme_bw() +
  theme(legend.position = "bottom") +
  labs(y="Obligate : Facultative Anaerobes ratio")

summaries_df %>% filter(p.value<0.1) %>% pander(split.tab=Inf)
```


\newpage
# ~ METABOLITES ~
# AGE  

# FIG 2D
Trajectory of metabolites over age (pretty summary)  
FIGURE 2 optional summary of metabolites over age  
```{r fig2eSUMMARY, fig.height = 2, fig.width =4, dpi=300}
study_group_var = "WOL"
# unique(stool_metab_toPlot$metab)
aa <- c("Tyrosine", "Alanine", "Glycine", "Isovalerate", "Phenylalanine", "Threonine", "Histidine", "Lysine", "Tryptophan", "Methionine", "Glutamine", "Glutamate", "Aspartate", "Pyroglutamate")
bcaa <- c("Isoleucine", "Leucine", "Valine")
krebs <- c("Citrate", "Succinate", "Pyruvate", "Fumarate", "Formate")
scfa <- c("Propionate", "Butyrate", "Acetate")
purines <- c("Purine", "Hypoxanthine")
sugar <- c("Galactose", "Glucose")
#Creatine
#Uracil
#Glycerol
voc <- c("Cadaverine", "Putrescine", "Dimethylamine")

# stool_metab_toPlot <- stool_metab_long %>%
#   inner_join(s2 %>% select(SampleID, growth_cat, growth_failure, TimePoint, SubjectID), by = "SampleID") %>% 
#   droplevels() %>%
#   mutate(metab_group = ifelse(metab %in% aa, "Amino acid", ifelse(metab %in% krebs, "Kreb's Cycle", ifelse(metab %in% scfa, "SCFA", ifelse(metab %in% purines, "Purines", ifelse(metab %in% sugar, "Sugars", ifelse(metab %in% voc, "VOC", metab)))))))

x=data.frame(t(stool_metab %>% select(!c(SampleID, redcap_id, nmr_no, `Stool weight`, Seq_run5_YN, Sequenced, in_36wklist, HOL, DOL,Inf_metBol_mlk))))
colnames(x) <- stool_metab$SampleID

# y <- x %>% rownames_to_column("metab") %>% mutate(metab_group = ifelse(metab %in% aa, "Amino acids", ifelse(metab %in% krebs, "Kreb's Cycle", ifelse(metab %in% scfa, "SCFAs", ifelse(metab %in% purines, "Purines", ifelse(metab %in% sugar, "Sugars", ifelse(metab %in% voc, "VOCs", metab)))))))

# Add number in each group:
y <- x %>% rownames_to_column("metab") %>% mutate(metab_group = ifelse(metab %in% bcaa, paste0("BCAAs (", length(bcaa), ")"), ifelse(metab %in% aa, paste0("Amino acids (", length(aa), ")"), ifelse(metab %in% krebs, paste0("Kreb's Cycle (", length(krebs), ")"), ifelse(metab %in% scfa, paste0("SCFAs (", length(scfa), ")"), ifelse(metab %in% purines, paste0("Purines (", length(purines), ")"), ifelse(metab %in% sugar, paste0("Sugars (", length(sugar), ")"), ifelse(metab %in% voc, paste0("VOCs (", length(voc), ")"), metab))))))))

metab_summarised <- rowsum(x, y$metab_group)
metab_summarised %>%
  rownames_to_column("metab") %>%
  pivot_longer(!metab, names_to = "SampleID", values_to = "umol_g") %>%
  inner_join(s2 %>% select(SampleID, growth_cat, growth_failure, TimePoint, SubjectID, WOL), by = "SampleID") %>%
  filter(metab != "Glycerol") %>%
  mutate(metab = reorder(metab, -umol_g)) %>%
  ggplot(aes(x=get(study_group_var), y=umol_g)) +
    geom_smooth(aes(group=metab, color=metab), alpha=0.2, size=0.5) + #, linetype = Taxa
    scale_color_d3(palette = "category20") +
    scale_x_continuous(expand=c(0,0)) +
    scale_y_continuous(limit=c(0,NA), oob=squish, expand = c(0.01,0)) +
    coord_cartesian(clip = "off") +
    theme_bw() +
    theme(
      legend.key.size = unit(c(0.13), units = "in"),
      legend.background = element_rect(fill = NA), 
      legend.text = element_text(size=8),
      legend.title=element_blank(),
      panel.grid = element_blank(),
      axis.ticks = element_line(linewidth = 0.3),
      panel.border = element_rect(linewidth = 0.3),
      axis.text = element_text(colour = "black", size=8)
      ) +
    guides(color=guide_legend(ncol=1, override.aes = list(lwd = 1, fill=NA))) +
    labs(x=study_group_var, fill="", y="umol/g")
  
ggsave(here("Output", "FIG2D_stoolmetabolites_WOL_smooth.pdf"), width = 4, height = 2, useDingbats = F, dpi = 400)
```

```{r fig.width=4}
s2 %>%
  ggplot(aes(y=AbxHoursPerHOL, x=WOL)) + 
  geom_smooth(aes(color=growth_failure))+
  theme_bw()

s2 %>%
  ggplot(aes(y=AbxHoursPerHOL, x=WOL)) + 
  geom_point(aes(color=SubjectID)) +
  geom_boxplot(aes(group=WOL), outlier.shape=NA, fill=NA) +
  facet_wrap(~growth_failure) +
  scale_color_discrete(guide='none') +
  theme_bw()

s2 %>%
  ggplot(aes(y=AbxHoursPerHOL, x=WOL)) + 
  geom_smooth(aes(color=growth_cat))+
  theme_bw()

s2 %>%
  ggplot(aes(y=AbxHoursPerHOL, x=WOL)) + 
  geom_point(aes(color=SubjectID)) +
  geom_boxplot(aes(group=WOL), outlier.shape=NA, fill=NA) +
  facet_wrap(~growth_cat) +
  scale_color_discrete(guide='none') +
  theme_bw()

pander(stool_metab_long %>% inner_join(s2 %>% select(SampleID, WOL,growth_cat, growth_failure)) %>% select(SampleID, WOL, growth_cat, growth_failure, metab, umol_g) %>% count(WOL, growth_failure, metab) %>% group_by(WOL, growth_failure) %>% summarise(`Average Sample Size`=mean(n)) %>% ungroup(), caption = "Sample sizes in growth failure groups (averaged over all metabolites)")

pander(stool_metab_long %>% inner_join(s2 %>% select(SampleID, WOL,growth_cat)) %>% select(SampleID, WOL, growth_cat, metab, umol_g) %>% count(WOL, growth_cat, metab) %>% group_by(WOL, growth_cat) %>% summarise(`Average Sample Size`=mean(n)) %>% ungroup(), caption = "Sample sizes in growth faltering groups (averaged over all metabolites)")

```



\newpage
SCFAs - only acetate, butyrate, propionate, lactate, formate, succinate
Trajectory of SCFAs
```{r fig2eSCFA, fig.width=7, fig.height=5}
stool_metab_long %>%
  filter(metab %in% c("Acetate", "Butyrate", "Propionate", "Lactate", "Succinate", "Formate")) %>%
  right_join(s2, by="SampleID") %>%
  filter(!is.na(metab)) %>%
  ggplot(aes(y=umol_g, x=WOL)) +
    geom_point(shape=21, color="darkgrey") +
    geom_boxplot(aes(group=WOL, fill=WOL), outlier.shape = NA) +
    facet_wrap(~metab, scales="free") +
    scale_fill_viridis_c() +
  theme_bw() +
  labs(y="Concentration (umol/g)") +
  theme(panel.grid = element_blank())

ggsave(here("Output", "FIG2E_stoolSCFAs_WOL_boxplots.pdf"), width = 7, height = 5, useDingbats = F, dpi = 400)

```

\newpage
## Linear model  
Metabolite change with age  
```{r}
props_toTest <- stool_metab_long %>%
  filter(metab != "Glycerol") %>%
  mutate(log10_umol_g = log10(umol_g))

form1 = "log10_umol_g ~ WOL"
# form2 = "log10_umol_g ~ WOL * growth_cat + npo_days_prop"
# form3 = "log10_umol_g ~ WOL * growth_cat + AbxHoursPerHOL"
random1 = "SubjectID"

summaries_df <- run_lmer_on_props(factor_to_test = "metab", props_toTest = props_toTest, s_toTest = s_toTest, form1 = form1,rep_mes_label = random1, p_cutoff =  0.05) 

summaries_df %>%
  mutate(term = str_remove(term, "growth_cat")) %>%
  arrange(fdr) %>%
  pander(split.table=Inf, digits=2, caption = glue("Linear models of log10 transformed concentrations of metabolites. Filtered with p.value < 0.05. SubjectID added as a random effect. Full formula: {form1}"), emphasize.strong.rows = which(.$fdr<=0.05))

table_export(summaries_df, "metabolites_analysis_table_")

```


\newpage
# All age significant metabolites  
```{r fig2eSIG, fig.width=15, fig.height=6}
sig_metab <- summaries_df %>% filter(fdr<0.05, term == "WOL") %>% pull(metab)
stool_metab_long %>%
  filter(metab %in% sig_metab) %>%
  right_join(s2, by="SampleID") %>%
  filter(!is.na(metab)) %>%
  left_join(summaries_df %>% filter(fdr<0.05, term == "WOL"), by = "metab") %>%
  group_by(metab) %>% mutate(median=median(umol_g), max=max(umol_g)) %>%
  ungroup() %>%
  mutate(metab = fct_reorder(metab, -umol_g)) %>%
  ggplot(aes(y=umol_g, x=WOL)) +
    geom_point(shape=21, color="darkgrey") +
    geom_boxplot(aes(group=WOL, fill=WOL), outlier.shape=NA) +
    geom_text(mapping = aes(label = paste0("q=", scientific(fdr, digits = 2)), x=8, y=max*.99), size = 3, check_overlap = T) +
    facet_wrap(~metab, scales="free", ncol=7) +
    scale_fill_viridis_c(alpha=0.8) +
  scale_y_continuous(transform="log10") +
  theme_bw() +
  theme(panel.grid = element_blank()) +
  labs(y="Concentration (umol/g log-scale)")

ggsave(here("Output", "FIG2E_stoolmetabolites_fdr05_WOL_boxplots.pdf"), width = 15, height = 6, useDingbats = F, dpi = 400)
```


# All metabolites squiggles  
```{r fig.width=10, fig.height=8}
lod=min(stool_metab_long %>% pull(umol_g), na.rm = T)
stool_metab_long %>% select(!HOL) %>%
  right_join(s2, by = "SampleID") %>%
  mutate(`Growth Faltering` = factor(x = growth_cat, labels = c("none", "mild faltering", "moderate to severe faltering"))) %>%
  filter(!is.na(growth_cat), !is.na(metab), metab != "Glycerol") %>%
  ggplot(aes(y=umol_g, x=HOL)) +
    # geom_hline(aes(yintercept = lod),  lty=2) +
    geom_point(aes(color=`Growth Faltering`), alpha=0.1) +
    geom_smooth(aes(color=`Growth Faltering`, group = `Growth Faltering`), se = F) +  
    scale_color_manual(values = as.character(ann_colors$growth_cat)) +
    scale_y_continuous(transform = "log10", limits = c(lod, NA), labels = signif) +
    theme_bw() +
  theme(legend.position = "bottom", panel.grid = element_blank(), text = element_text(size=10, color="black"), axis.text = element_text(size=8, color="black"), strip.text = element_text(size=8, color="black"), legend.text = element_text(size=10, color="black")) +    
  facet_wrap(~metab, ncol=7, scales="free_y") +
  # labs(y="Relative Abundance", x="Hours of Life")
 labs(y="Concentration (umol/g log-scale)")

ggsave(here("Output", "NEW_ALL_stoolmetabolites_GrFaltering.pdf"), width = 10, height = 8, useDingbats = F, dpi = 600)

stool_metab_long %>% select(!HOL) %>%
  right_join(s2, by = "SampleID") %>%
  # mutate(`Growth Faltering` = factor(x = growth_cat, labels = c("none", "mild faltering", "moderate to severe faltering"))) %>%
  filter(!is.na(growth_cat), !is.na(metab), metab != "Glycerol") %>%
  ggplot(aes(y=umol_g, x=HOL)) +
    geom_point(aes(color=growth_failure), alpha=0.1) +
    geom_smooth(aes(color=growth_failure, group=growth_failure), se = F) +  
    scale_color_manual(values = as.character(ann_colors$growth_failure)) +
    scale_y_continuous(transform = "log10", limits = c(lod, NA), labels = signif) +
    theme_bw() +
  theme(legend.position = "bottom", panel.grid = element_blank(), text = element_text(size=10, color="black"), axis.text = element_text(size=8, color="black"), strip.text = element_text(size=8, color="black"), legend.text = element_text(size=10, color="black")) +    
  facet_wrap(~metab, ncol=7, scales="free_y") +
  # labs(y="Relative Abundance", x="Hours of Life")
 labs(y="Concentration (umol/g log-scale)", color="Growth failure")

ggsave(here("Output", "NEW_ALL_stoolmetabolites_GrFailure.pdf"), width = 10, height = 8, useDingbats = F, dpi = 600)
```

\newpage
# SUPP FIGURE 3A - Metabolites over age by growth category   
## Linear models (grouped metabs)
```{r}
props_toTest <- metab_summarised %>%
  rownames_to_column("metab") %>%
  pivot_longer(!metab, names_to = "SampleID", values_to = "umol_g") %>%
  # inner_join(s2 %>% select(SampleID, growth_cat, growth_failure, TimePoint, SubjectID, WOL), by = "SampleID") %>%
  filter(!is.na(metab), metab!="Glycerol") %>%
  mutate(log10_umol_g = log10(umol_g + min(metab_summarised[metab_summarised>0])/10))

form1 = "log10_umol_g ~ WOL * growth_cat"
form2 = "log10_umol_g ~ WOL * growth_cat + npo_days_prop"
form3 = "log10_umol_g ~ WOL * growth_cat + AbxHoursPerHOL"
random1 = "SubjectID"

summaries_df <- run_lmer_on_props(factor_to_test = "metab", props_toTest = props_toTest, s_toTest = s2 %>% select(SampleID, growth_cat, growth_failure, TimePoint, SubjectID, WOL), form1 = form1, rep_mes_label = random1, p_cutoff = .05) 

summaries_df %>%
  mutate(term = str_remove(term, "growth_cat")) %>%
  pander(split.table=Inf, digits=2, caption = glue("Linear models of log10 transformed concentrations of groups of metabolites. Filtered with p.value < 0.05. SubjectID added as a random effect. Full formula: {form1}"), emphasize.strong.rows = which(.$fdr<=0.05))

table_export(summaries_df, "metabolites_analysis_table_")

```

```{r supp3a, fig.height=5, fig.width=7}
metab_summarised %>%
  rownames_to_column("metab") %>%
  pivot_longer(!metab, names_to = "SampleID", values_to = "umol_g") %>%
  inner_join(s2 %>% select(SampleID, growth_cat, growth_failure, TimePoint, SubjectID, WOL), by = "SampleID") %>%
  filter(!is.na(metab), metab!="Glycerol") %>%
  mutate(metab = reorder(metab, -umol_g)) %>%
  group_by(WOL, metab, growth_cat) %>%
  mutate(mean=mean(umol_g), sd=sd(umol_g), n=n()) %>%
  ungroup() %>%
  mutate(se = sd/sqrt(n)) %>%
  mutate(metab = fct_reorder(metab, -umol_g)) %>%
  arrange(growth_cat) %>%
  ggplot(aes(y=umol_g, x=WOL, group=growth_cat, color=growth_cat)) +
    geom_point(size=1.2, aes(shape=growth_cat), alpha=0.5) +
    geom_smooth(size=0.8, aes(fill=growth_cat), se=F) +
    # geom_point(size=1, aes(y=mean,shape=growth_cat)) +
    facet_wrap(~metab, scales="free") +
    scale_color_manual(values = ann_colors$growth_cat) +
    scale_fill_manual(values = ann_colors$growth_cat) +
    scale_shape_manual(values = c(21, 22, 23)) +
   # scale_y_continuous(limit=c(0,NA),oob=squish) +
  theme_bw() +
  theme(legend.position = "bottom", , panel.grid = element_blank()) +
  labs(fill="Growth", color="Growth", shape="Growth", y='Concentration (umol/g)')
ggsave(here("Output", "SUPP_FIG3A_stoolmetabolites_growthcat_grouped.pdf"), width = 7, height = 5, useDingbats = F, dpi = 400)
```

## Week by week  
```{r fig.height=9, fig.width=8}
myform="log10_umol_g ~ study_group"

summaries_df <- props_toTest %>%
  inner_join(s2 %>% select(SampleID, study_group=growth_cat, sample_type=SampleType, TimePoint, SubjectID, WOL), by = "SampleID") %>%
  filter(!is.na(study_group)) %>%
  group_by(WOL, metab, study_group, sample_type) %>%
    mutate(mean=mean(log10_umol_g)) %>%
  ungroup() %>%
  
  group_by(metab) %>% mutate(max=max(mean)) %>% ungroup() %>%
  
  group_by(metab, WOL) %>%
    mutate(ngroups=n_distinct(study_group)) %>%
  ungroup() %>%
  
  filter(ngroups>=2) %>% # Only test PMA_rounded where you have both study groups
  
  group_by(metab, WOL, max) %>%
    do(tidy(lm(as.formula(myform), data=.))) %>%
  ungroup() %>%
  
  filter(!grepl("Intercept|Residuals", term)) %>%
  mutate(term = sub("study_group", "", term)) %>%
  
  group_by(term, WOL) %>%
    mutate(fdr = p.adjust(p.value, method="BH")) %>%
  ungroup()

props_toTest %>%
  inner_join(s2 %>% select(SampleID, study_group=growth_cat, sample_type=SampleType, TimePoint, SubjectID, WOL), by = "SampleID") %>%
  filter(!is.na(study_group)) %>%
  group_by(WOL, metab, study_group, sample_type) %>%
    summarise(mean=mean(log10_umol_g), sd=sd(log10_umol_g), n=n(), max=max(log10_umol_g)) %>%
  ungroup() %>%
  mutate(se=sd/sqrt(n)) %>%
  ggplot(aes(y=10^(mean), x=WOL)) +
    # geom_vline(aes(xintercept = PMA_rounded), lwd=0.3, color="grey") + # guide lines
    geom_line(aes(color=study_group, group = study_group)) +  
    geom_point(aes(color=study_group)) +
    geom_point(data = summaries_df %>% filter(fdr<0.1, term==levels(s2$growth_cat)[2]), mapping = aes(color=term, x=WOL, y=10^(max*1.75)), shape=8) +
    geom_point(data = summaries_df %>% filter(fdr>=0.1, p.value<0.05, term==levels(s2$growth_cat)[2]), mapping = aes(color=term, x=WOL, y=10^(max*1.75)), shape=23) +
    geom_point(data = summaries_df %>% filter(fdr<0.1, term==levels(s2$growth_cat)[3]), mapping = aes(color=term, x=WOL, y=10^(max*2)), shape=8) +
    geom_point(data = summaries_df %>% filter(fdr>=0.1, p.value<0.05, term==levels(s2$growth_cat)[3]), mapping = aes(color=term, x=WOL, y=10^(max*2)), shape=23) +
    scale_color_manual(values = ann_colors$growth_cat) +
    scale_y_continuous(transform = "log10") +
    theme_bw() +
  theme(legend.position = "bottom") +
    theme(panel.grid = element_blank()) +
    geom_errorbar(aes(ymin = 10^(mean-se), ymax=10^(mean+se), color=study_group), alpha=0.5, width = 0.3) +
    facet_wrap(~metab, ncol=3, scales="free_y") +
  labs(y="Concentration (umol/g log-scale)")
```

Linear models were used to estimate the differences between study groups at each week of life. The relative abundances were log10 transformed. Multiple tests were adjusted for false discovery rate using Benjamini-Hochberg method. Only the terms with p<0.1 are shown in the table below.

For each week of life (WOL):  
lm(`r myform`)

```{r}
# table_export(summaries_df, "Taxonomic_table_export")

summaries_df %>%
  filter(p.value < 0.05) %>%
  pander(split.tab=Inf, emphasize.strong.rows = which(.$fdr<=0.05)) #%>%
  #column_spec(c(2:5), width = "3em") ## if the table is running off the page, you can manually specify the width of the columns

rm(myform)
```

## Monthly  
(every 4 weeks)   
```{r fig.height=9, fig.width=8}
myform="log10_umol_g ~ study_group"

summaries_df <- props_toTest %>%
  inner_join(s2 %>% select(SampleID, study_group=growth_cat, sample_type=SampleType, TimePoint, SubjectID, MOL), by = "SampleID") %>%
  filter(!is.na(study_group)) %>%
  group_by(MOL, metab, study_group, sample_type) %>%
    mutate(mean=mean(log10_umol_g)) %>%
  ungroup() %>%
  
  group_by(metab) %>% mutate(maxmean=max(mean), max=max(log10_umol_g)) %>% ungroup() %>%
  
  group_by(metab, MOL) %>%
    mutate(ngroups=n_distinct(study_group)) %>%
  ungroup() %>%
  
  filter(ngroups>=2) %>% # Only test PMA_rounded where you have both study groups
  
  group_by(metab, MOL, max) %>%
    do(tidy(lm(as.formula(myform), data=.))) %>%
  ungroup() %>%
  
  filter(!grepl("Intercept|Residuals", term)) %>%
  mutate(term = sub("study_group", "", term)) %>%
  
  group_by(term, MOL) %>%
    mutate(fdr = p.adjust(p.value, method="BH")) %>%
  ungroup()

props_toTest %>%
  inner_join(s2 %>% select(SampleID, study_group=growth_cat, sample_type=SampleType, TimePoint, SubjectID, MOL), by = "SampleID") %>%
  filter(!is.na(study_group)) %>%
  group_by(MOL, metab, study_group, sample_type) %>%
    summarise(mean=mean(log10_umol_g), sd=sd(log10_umol_g), n=n()) %>%
  ungroup() %>%
  mutate(se=sd/sqrt(n)) %>%
  ggplot(aes(y=10^(mean), x=MOL)) +
    # geom_vline(aes(xintercept = PMA_rounded), lwd=0.3, color="grey") + # guide lines
    geom_line(aes(color=study_group, group = study_group)) +  
    geom_point(aes(color=study_group)) +
    geom_point(data = summaries_df %>% filter(fdr<0.1, term==levels(s2$growth_cat)[2]), mapping = aes(color=term, x=MOL, y=10^(max*1.25)), shape=8) +
    geom_point(data = summaries_df %>% filter(fdr>=0.1, p.value<0.05, term==levels(s2$growth_cat)[2]), mapping = aes(color=term, x=MOL, y=10^(max*1.25)), shape=23) +
    geom_point(data = summaries_df %>% filter(fdr<0.1, term==levels(s2$growth_cat)[3]), mapping = aes(color=term, x=MOL, y=10^(max*1.3)), shape=8) +
    geom_point(data = summaries_df %>% filter(fdr>=0.1, p.value<0.05, term==levels(s2$growth_cat)[3]), mapping = aes(color=term, x=MOL, y=10^(max*1.3)), shape=23) +
    scale_color_manual(values = ann_colors$growth_cat) +
    scale_y_continuous(transform = "log10") +
    theme_bw() +
  theme(legend.position = "bottom") +
    theme(panel.grid = element_blank()) +
    geom_errorbar(aes(ymin = 10^(mean-se), ymax=10^(mean+se), color=study_group), alpha=0.5, width = 0.3) +
    facet_wrap(~metab, ncol=3, scales="free_y") +
  labs(y="Concentration (umol/g log-scale)", title = "")
```

Linear models were used to estimate the differences between study groups at each week of life. The relative abundances were log10 transformed. Multiple tests were adjusted for false discovery rate using Benjamini-Hochberg method. Only the terms with p<0.1 are shown in the table below.

For each week of life (MOL):  
lm(`r myform`)

```{r}
# table_export(summaries_df, "Taxonomic_table_export")

summaries_df %>%
  # filter(p.value < 0.1) %>%
  pander(split.tab=Inf, emphasize.strong.rows = which(.$fdr<=0.05)) #%>%
  #column_spec(c(2:5), width = "3em") ## if the table is running off the page, you can manually specify the width of the columns

rm(myform)
```

\newpage
# FIG 2E - select metabolites over time by growth category  
## Linear model - growth cateogry change with age
Growth category change with age  
```{r}
props_toTest <- stool_metab_long %>%
  filter(metab != "Glycerol") %>%
  mutate(log10_umol_g = log10(umol_g))

form1 = "log10_umol_g ~ WOL * growth_cat"
form1m = "log10_umol_g ~ MOL * growth_cat"
form2 = "log10_umol_g ~ WOL * growth_cat + npo_days_prop"
form3 = "log10_umol_g ~ WOL * growth_cat + AbxHoursPerHOL"
random1 = "SubjectID"

# Growth category WOL
summaries_df <- run_lmer_on_props(factor_to_test = "metab", props_toTest = props_toTest, s_toTest = s_toTest, form1 = form1,rep_mes_label = random1, p_cutoff = .05) 
summaries_df %>%
  mutate(term = str_remove(term, "growth_cat")) %>%
  pander(split.table=Inf, digits=2, caption = glue("Linear models of log10 transformed concentrations of metabolites. Filtered with p.value < 0.05. SubjectID added as a random effect. Full formula: {form1}"), emphasize.strong.rows = which(.$fdr<=0.05))
table_export(summaries_df, "metabolites_analysis_table_")

# Growth category MOL
summaries_df <- run_lmer_on_props(factor_to_test = "metab", props_toTest = props_toTest, s_toTest = s_toTest, form1 = form1m,rep_mes_label = random1, p_cutoff = .05) 
summaries_df %>%
  mutate(term = str_remove(term, "growth_cat")) %>%
  pander(split.table=Inf, digits=2, caption = glue("Linear models of log10 transformed concentrations of metabolites. Filtered with p.value < 0.05. SubjectID added as a random effect. Full formula: {form1m}"), emphasize.strong.rows = which(.$fdr<=0.05))

table_export(summaries_df, "metabolites_analysis_table_")

```

```{r fig.height=3, fig.width=8}
# sig_metab <- summaries_df %>% filter(fdr<0.2) %>% pull(metab)
stool_metab_long %>%
  # filter(metab %in% unique(summaries_df$metab)) %>%
  filter(metab %in% c("Butyrate", "Propionate", "Isovalerate", "Dimethylamine")) %>%
  right_join(s2, by="SampleID") %>%
  filter(!is.na(metab), metab!="Glycerol") %>%
  group_by(WOL, metab, growth_cat) %>%
  mutate(mean=mean(umol_g), sd=sd(umol_g), n=n()) %>%
  ungroup() %>%
  mutate(se = sd/sqrt(n)) %>%
  mutate(metab = fct_reorder(metab, -umol_g)) %>%
  arrange(growth_cat) %>%
  ggplot(aes(y=umol_g, x=WOL, group=growth_cat, color=growth_cat)) +
    geom_point(aes(shape=growth_cat), alpha=0.5) +
    geom_smooth(aes(fill=growth_cat), se=F) +
    # geom_point(aes(y=mean, shape=growth_failure)) +
    # geom_errorbar(aes(ymin=mean-se, ymax=mean+se)) +
    facet_wrap(~metab, scales="free_y", ncol=7) +
    scale_color_manual(values = ann_colors$growth_cat) +
    scale_fill_manual(values = ann_colors$growth_cat) +
    scale_shape_manual(values = c(21, 22, 24)) +
   # scale_y_continuous(limit=c(0,NA),oob=squish) +
  theme_bw() +
  theme(legend.position = "bottom", panel.grid = element_blank()) +
  labs(fill="Growth Category", color="Growth Category", shape="Growth Category", y='Concentration (umol/g)')
ggsave(here("Output", "FIG2E_metabolites_age_growthcategory_subset.pdf"), width = 8, height = 3, useDingbats = F, dpi = 400)
```

## Week by week  
\newpage
```{r fig.height=14, fig.width=9, out.width="5.5in"}
myform="log10_umol_g ~ study_group"

summaries_df <- props_toTest %>%
  inner_join(s2 %>% select(SampleID, study_group=growth_cat, sample_type=SampleType, TimePoint, SubjectID, WOL), by = "SampleID") %>%
  filter(!is.na(study_group)) %>%
  group_by(WOL, metab, study_group, sample_type) %>%
    mutate(mean=mean(log10_umol_g)) %>%
  ungroup() %>%
  
  group_by(metab) %>% mutate(max=max(mean)) %>% ungroup() %>%
  
  group_by(metab, WOL) %>%
    mutate(ngroups=n_distinct(study_group)) %>%
  ungroup() %>%
  
  filter(ngroups>=2) %>% # Only test PMA_rounded where you have both study groups
  
  group_by(metab, WOL, max) %>%
    do(tidy(lm(as.formula(myform), data=.))) %>%
  ungroup() %>%
  
  filter(!grepl("Intercept|Residuals", term)) %>%
  mutate(term = sub("study_group", "", term)) %>%
  
  group_by(term, WOL) %>%
    mutate(fdr = p.adjust(p.value, method="BH")) %>%
  ungroup()

props_toTest %>%
  inner_join(s2 %>% select(SampleID, study_group=growth_cat, sample_type=SampleType, TimePoint, SubjectID, WOL), by = "SampleID") %>%
  filter(!is.na(study_group)) %>%
  mutate(Taxa = reorder(metab, -log10_umol_g)) %>%
  mutate(Taxa = fct_relabel(Taxa, function(x) gsub(" ", "\n", x))) %>%
  group_by(WOL, metab, study_group, sample_type) %>%
    summarise(mean=mean(log10_umol_g), sd=sd(log10_umol_g), n=n(), max=max(log10_umol_g)) %>%
  ungroup() %>%
  mutate(se=sd/sqrt(n)) %>%
  ggplot(aes(y=10^(mean), x=WOL)) +
    # geom_vline(aes(xintercept = PMA_rounded), lwd=0.3, color="grey") + # guide lines
    geom_line(aes(color=study_group, group = study_group)) +  
    geom_point(aes(color=study_group)) +
    geom_point(data = summaries_df %>% filter(fdr<0.1, term==levels(s2$growth_cat)[2]), mapping = aes(color=term, x=WOL, y=10^(max*1.75)), shape=8) +
    geom_point(data = summaries_df %>% filter(fdr>=0.1, p.value<0.05, term==levels(s2$growth_cat)[2]), mapping = aes(color=term, x=WOL, y=10^(max*1.75)), shape=23) +
    geom_point(data = summaries_df %>% filter(fdr<0.1, term==levels(s2$growth_cat)[3]), mapping = aes(color=term, x=WOL, y=10^(max*2)), shape=8) +
    geom_point(data = summaries_df %>% filter(fdr>=0.1, p.value<0.05, term==levels(s2$growth_cat)[3]), mapping = aes(color=term, x=WOL, y=10^(max*2)), shape=23) +
    scale_color_manual(values = ann_colors$growth_cat) +
    scale_y_continuous(transform = "log10") +
    theme_bw() +
  theme(legend.position = "bottom") +
    theme(panel.grid = element_blank()) +
    geom_errorbar(aes(ymin = 10^(mean-se), ymax=10^(mean+se), color=study_group), alpha=0.5, width = 0.3) +
    facet_wrap(~metab, ncol=3, scales="free_y") +
  labs(y="Concentration (umol/g log-scale)")
```

Linear models were used to estimate the differences between study groups at each week of life. The relative abundances were log10 transformed. Multiple tests were adjusted for false discovery rate using Benjamini-Hochberg method. Only the terms with p<0.1 are shown in the table below.

For each week of life (WOL):  
lm(`r myform`)

```{r}
# table_export(summaries_df, "Taxonomic_table_export")

summaries_df %>%
  filter(p.value < 0.05) %>%
  pander(split.tab=Inf, emphasize.strong.rows = which(.$fdr<=0.05)) #%>%
  #column_spec(c(2:5), width = "3em") ## if the table is running off the page, you can manually specify the width of the columns

rm(myform)
```

## Monthly  
\newpage
```{r fig.height=14, fig.width=9, out.width="5.5in"}
myform="log10_umol_g ~ study_group"

summaries_df <- props_toTest %>%
  inner_join(s2 %>% select(SampleID, study_group=growth_cat, sample_type=SampleType, TimePoint, SubjectID, MOL), by = "SampleID") %>%
  filter(!is.na(study_group)) %>%
  group_by(MOL, metab, study_group, sample_type) %>%
    mutate(mean=mean(log10_umol_g)) %>%
  ungroup() %>%
  
  group_by(metab) %>% mutate(max=max(mean)) %>% ungroup() %>%
  
  group_by(metab, MOL) %>%
    mutate(ngroups=n_distinct(study_group)) %>%
  ungroup() %>%
  
  filter(ngroups>=2) %>% # Only test PMA_rounded where you have both study groups
  
  group_by(metab, MOL, max) %>%
    do(tidy(lm(as.formula(myform), data=.))) %>%
  ungroup() %>%
  
  filter(!grepl("Intercept|Residuals", term)) %>%
  mutate(term = sub("study_group", "", term)) %>%
  
  group_by(term, MOL) %>%
    mutate(fdr = p.adjust(p.value, method="BH")) %>%
  ungroup()

props_toTest %>%
  inner_join(s2 %>% select(SampleID, study_group=growth_cat, sample_type=SampleType, TimePoint, SubjectID, MOL), by = "SampleID") %>%
  filter(!is.na(study_group)) %>%
  mutate(Taxa = reorder(metab, -log10_umol_g)) %>%
  mutate(Taxa = fct_relabel(Taxa, function(x) gsub(" ", "\n", x))) %>%
  group_by(MOL, metab, study_group, sample_type) %>%
    summarise(mean=mean(log10_umol_g), sd=sd(log10_umol_g), n=n(), max=max(log10_umol_g)) %>%
  ungroup() %>%
  mutate(se=sd/sqrt(n)) %>%
  ggplot(aes(y=10^(mean), x=MOL)) +
    # geom_vline(aes(xintercept = PMA_rounded), lwd=0.3, color="grey") + # guide lines
    geom_line(aes(color=study_group, group = study_group)) +  
    geom_point(aes(color=study_group)) +
    geom_point(data = summaries_df %>% filter(fdr<0.1, term==levels(s2$growth_cat)[2]), mapping = aes(color=term, x=MOL, y=35), shape=8) +
    geom_point(data = summaries_df %>% filter(fdr>=0.1, p.value<0.05, term==levels(s2$growth_cat)[2]), mapping = aes(color=term, x=MOL, y=35), shape=23) +
    geom_point(data = summaries_df %>% filter(fdr<0.1, term==levels(s2$growth_cat)[3]), mapping = aes(color=term, x=MOL, y=20), shape=8) +
    geom_point(data = summaries_df %>% filter(fdr>=0.1, p.value<0.05, term==levels(s2$growth_cat)[3]), mapping = aes(color=term, x=MOL, y=20), shape=23) +
    scale_color_manual(values = ann_colors$growth_cat) +
    scale_y_continuous(transform = "log10") +
    theme_bw() +
  theme(legend.position = "bottom") +
    theme(panel.grid = element_blank()) +
    geom_errorbar(aes(ymin = 10^(mean-se), ymax=10^(mean+se), color=study_group), alpha=0.5, width = 0.3) +
    facet_wrap(~metab, ncol=3, scales="free_y") +
  labs(y="Concentration (umol/g log-scale)")
```

Linear models were used to estimate the differences between study groups at each week of life. The relative abundances were log10 transformed. Multiple tests were adjusted for false discovery rate using Benjamini-Hochberg method. Only the terms with p<0.1 are shown in the table below.

For each week of life (MOL):  
lm(`r myform`)

```{r}
# table_export(summaries_df, "Taxonomic_table_export")

summaries_df %>%
  filter(p.value < 0.05) %>%
  pander(split.tab=Inf, emphasize.strong.rows = which(.$fdr<=0.05)) #%>%
  #column_spec(c(2:5), width = "3em") ## if the table is running off the page, you can manually specify the width of the columns

rm(myform)
```

\newpage
```{r suppfig3b, fig.width=10, fig.height=8, eval=T}
# SUPP FIGURE 3B - Metabolites (all) over time by growth failure   
# Supp 3A but no grouping
stool_metab_long %>%
  right_join(s2, by="SampleID") %>%
  filter(!is.na(metab), metab!="Glycerol") %>%
  group_by(WOL, metab, growth_failure) %>%
  mutate(mean=mean(umol_g), sd=sd(umol_g), n=n()) %>%
  ungroup() %>%
  mutate(se = sd/sqrt(n)) %>%
  mutate(metab = fct_reorder(metab, -umol_g)) %>%
  arrange(growth_failure) %>%
  ggplot(aes(y=umol_g, x=WOL, group=growth_failure, color=growth_failure)) +
    geom_point(aes(shape=growth_failure), alpha=0.7) +
    geom_smooth(aes(fill=growth_failure), se=F) +
    # geom_point(aes(y=mean, shape=growth_failure)) +
    # geom_errorbar(aes(ymin=mean-se, ymax=mean+se)) +
    facet_wrap(~metab, scales="free_y", ncol=7) +
    scale_color_manual(values = ann_colors$growth_failure) +
    scale_fill_manual(values = ann_colors$growth_failure) +
    scale_shape_manual(values = c(21, 24)) +
   # scale_y_continuous(limit=c(0,NA),oob=squish) +
  theme_bw() +
  theme(legend.position = "bottom", panel.grid = element_blank()) +
  labs(fill="Growth Failure", color="Growth Failure", shape="Growth Failure", y='Concentration (umol/g)')
ggsave(here("Output", "SUPP_FIG3B_stoolmetabolites_growthfailure_ungrouped.pdf"), width = 10, height = 8, useDingbats = F, dpi = 400)
```
## Metabolites (all) over time by growth failure Linear models  
Growth failure change with age  
```{r}
props_toTest <- stool_metab_long %>%
  filter(metab != "Glycerol") %>%
  mutate(log10_umol_g = log10(umol_g))

form1 = "log10_umol_g ~ WOL * growth_failure"
form1m = "log10_umol_g ~ MOL * growth_failure"
form2 = "log10_umol_g ~ WOL * growth_failure + npo_days_prop"
form3 = "log10_umol_g ~ WOL * growth_failure + AbxHoursPerHOL"
random1 = "SubjectID"

# Growth failure WOL
summaries_df <- run_lmer_on_props(factor_to_test = "metab", props_toTest = props_toTest, s_toTest = s_toTest, form1 = form1,rep_mes_label = random1, p_cutoff = .05) 
summaries_df %>%
  mutate(term = str_remove(term, "growth_failure")) %>%
  pander(split.table=Inf, digits=2, caption = glue("Linear models of log10 transformed concentrations of metabolites. Filtered with p.value < 0.05. SubjectID added as a random effect. Full formula: {form1}"), emphasize.strong.rows = which(.$fdr<=0.05))
table_export(summaries_df, "metabolites_analysis_table_")

# Growth failure MOL
summaries_df <- run_lmer_on_props(factor_to_test = "metab", props_toTest = props_toTest, s_toTest = s_toTest, form1 = form1m,rep_mes_label = random1, p_cutoff = .05) 
summaries_df %>%
  mutate(term = str_remove(term, "growth_failure")) %>%
  pander(split.table=Inf, digits=2, caption = glue("Linear models of log10 transformed concentrations of metabolites. Filtered with p.value < 0.05. SubjectID added as a random effect. Full formula: {form1m}"), emphasize.strong.rows = which(.$fdr<=0.05))
table_export(summaries_df, "metabolites_analysis_table_")
```

# alt SUPP FIG 3b - grouped

```{r supp3bGROUPED, fig.height=5, fig.width=7, eval=T}
props_toTest <- metab_summarised %>%
  rownames_to_column("metab") %>%
  pivot_longer(!metab, names_to = "SampleID", values_to = "umol_g") %>%
  mutate(log10_umol_g = log10(umol_g + min(metab_summarised[metab_summarised>0])/10))

summaries_df <- run_lmer_on_props(factor_to_test = "metab", props_toTest = props_toTest %>% filter(!is.na(metab), metab!="Glycerol"), s_toTest = s2 %>% select(SampleID, growth_failure, sample_type=SampleType, TimePoint, SubjectID, WOL), form1 = "log10_umol_g ~ WOL * growth_failure",rep_mes_label = "SubjectID", p_cutoff = 1) 
table_export(summaries_df, "metabolites_analysis_table_SuppFig6C_")

summaries_df %>%
  mutate(term = str_remove(term, "growth_cat")) %>%
  pander(split.table=Inf, digits=2, caption = glue("Linear models of log10 transformed concentrations of metabolites. Filtered with p.value < 0.05. SubjectID added as a random effect. Full formula: log10_umol_g ~ WOL"), emphasize.strong.rows = which(.$fdr<=0.05))

props_toTest %>%
  inner_join(s2 %>% select(SampleID, growth_failure, sample_type=SampleType, TimePoint, SubjectID, WOL), by = "SampleID") %>%
  filter(!is.na(metab), metab!="Glycerol") %>%
  mutate(metab = reorder(metab, -umol_g)) %>%
  group_by(WOL, metab, growth_failure) %>%
  mutate(mean=mean(umol_g), sd=sd(umol_g), n=n()) %>%
  ungroup() %>%
  mutate(se = sd/sqrt(n)) %>%
  mutate(metab = fct_reorder(metab, -umol_g)) %>%
  arrange(growth_failure) %>%
  ggplot(aes(y=umol_g, x=WOL, group=growth_failure, color=growth_failure)) +
    geom_point(aes(shape=growth_failure), alpha=0.5) +
    geom_smooth(size=1, aes(fill=growth_failure), se=F) +
    facet_wrap(~metab, scales="free") +
    # scale_color_aaas() +
    scale_color_manual(values = ann_colors$growth_failure) +
    # scale_fill_aaas() +
    scale_fill_manual(values = ann_colors$growth_failure) +
    scale_shape_manual(values = c(21, 24)) +
   # scale_y_continuous(limit=c(0,NA),oob=squish) +
  theme_bw() +
  theme(legend.position = "bottom", panel.grid = element_blank()) +
  labs(fill="Growth Failure", color="Growth Failure", shape="Growth Failure", y='Concentration (umol/g)')
ggsave(here("Output", "SUPP_FIG3A_stoolmetabolites_growthfail_grouped.pdf"), width = 7, height = 5, useDingbats = F, dpi = 400)
```


## Week by week  
```{r fig.height=12, fig.width=8}
myform="log10_umol_g ~ study_group"

props_toTest <- stool_metab_long %>%
  filter(metab != "Glycerol") %>%
  mutate(log10_umol_g = log10(umol_g))

summaries_df <- props_toTest %>%
  inner_join(s2 %>% select(SampleID, study_group=growth_failure, sample_type=SampleType, TimePoint, SubjectID, WOL, AbxHoursPerHOL), by = "SampleID") %>%
  filter(!is.na(study_group), metab != "Glycerol") %>%
  group_by(WOL, metab, study_group, sample_type) %>%
    mutate(mean=mean(log10_umol_g)) %>%
  ungroup() %>%
  
  group_by(metab) %>% mutate(max=max(mean)) %>% ungroup() %>%
  
  group_by(metab, WOL) %>%
    mutate(ngroups=n_distinct(study_group)) %>%
  ungroup() %>%
  
  filter(ngroups==2) %>% # Only test PMA_rounded where you have both study groups
  
  group_by(metab, WOL, max) %>%
  # summarise(n=n())
    do(tidy(lm(as.formula(myform), data=.))) %>%
  ungroup() %>%
  
  filter(!grepl("Intercept|Residuals", term)) %>%
  mutate(term = sub("study_group", "", term)) %>%
  
  group_by(term, WOL) %>%
    mutate(fdr = p.adjust(p.value, method="BH")) %>%
  ungroup()

props_toTest %>%
  inner_join(s2 %>% select(SampleID, study_group=growth_failure, sample_type=SampleType, TimePoint, SubjectID, WOL, AbxHoursPerHOL), by = "SampleID") %>%
  filter(!is.na(study_group)) %>%
  mutate(Taxa = reorder(metab, -log10_umol_g)) %>%
  mutate(Taxa = fct_relabel(Taxa, function(x) gsub(" ", "\n", x))) %>%
  group_by(WOL, metab, study_group, sample_type) %>%
    summarise(mean=mean(log10_umol_g), sd=sd(log10_umol_g), n=n(), max=max(log10_umol_g)) %>%
  ungroup() %>%
  mutate(se=sd/sqrt(n)) %>%
  ggplot(aes(y=10^(mean), x=WOL)) +
    # geom_vline(aes(xintercept = PMA_rounded), lwd=0.3, color="grey") + # guide lines
    geom_line(aes(color=study_group, group = study_group)) +  
    geom_point(shape=21) +
    geom_point(data = summaries_df %>% filter(fdr<0.1), mapping = aes(x=WOL, y=10^(max*1.75)), shape=8, color="darkred") +
    geom_point(data = summaries_df %>% filter(p.value<0.05), mapping = aes(x=WOL, y=10^(max)+5), shape=23, color="red", fill="yellow") +
    # geom_point(data = summaries_df %>% filter(fdr<0.1, term=="moderate to severe retardation"), mapping = aes(color=term, x=WOL, y=10^(max*2)), shape=8) +
    scale_color_aaas() +
    # scale_color_manual(values = ann_colors$growth_failure) +
    scale_y_continuous(transform = "log10") +
    theme_bw() +
  theme(legend.position = "bottom") +
    theme(panel.grid = element_blank()) +
    geom_errorbar(aes(ymin = 10^(mean-se), ymax=10^(mean+se), color=study_group), alpha=0.5, width = 0.3) +
    facet_wrap(~metab, ncol=3, scales="free_y") +
  labs(y="Concentration (umol/g log-scale)")
```

Linear models were used to estimate the differences between study groups at each week of life. The relative abundances were log10 transformed. Multiple tests were adjusted for false discovery rate using Benjamini-Hochberg method. Only the terms with p<0.1 are shown in the table below.

For each week of life (WOL):  
lm(`r myform`)

```{r}
# table_export(summaries_df, "Taxonomic_table_export")

summaries_df %>%
  filter(p.value < 0.05) %>%
  pander(split.tab=Inf, emphasize.strong.rows = which(.$fdr<=0.05)) #%>%
  #column_spec(c(2:5), width = "3em") ## if the table is running off the page, you can manually specify the width of the columns

rm(myform)
```


## Monthly  
```{r fig.height=12, fig.width=8}
myform="log10_umol_g ~ study_group"

props_toTest <- stool_metab_long %>%
  filter(metab != "Glycerol") %>%
  mutate(log10_umol_g = log10(umol_g))

summaries_df <- props_toTest %>%
  inner_join(s2 %>% select(SampleID, study_group=growth_failure, sample_type=SampleType, TimePoint, SubjectID, MOL, AbxHoursPerHOL), by = "SampleID") %>%
  filter(!is.na(study_group), metab != "Glycerol") %>%
  group_by(MOL, metab, study_group, sample_type) %>%
    mutate(mean=mean(log10_umol_g)) %>%
  ungroup() %>%
  
  group_by(metab) %>% mutate(max=max(mean)) %>% ungroup() %>%
  
  group_by(metab, MOL) %>%
    mutate(ngroups=n_distinct(study_group)) %>%
  ungroup() %>%
  
  filter(ngroups==2) %>% # Only test PMA_rounded where you have both study groups
  
  group_by(metab, MOL, max) %>%
  # summarise(n=n())
    do(tidy(lm(as.formula(myform), data=.))) %>%
  ungroup() %>%
  
  filter(!grepl("Intercept|Residuals", term)) %>%
  mutate(term = sub("study_group", "", term)) %>%
  
  group_by(term, MOL) %>%
    mutate(fdr = p.adjust(p.value, method="BH")) %>%
  ungroup()

props_toTest %>%
  inner_join(s2 %>% select(SampleID, study_group=growth_failure, sample_type=SampleType, TimePoint, SubjectID, MOL, AbxHoursPerHOL), by = "SampleID") %>%
  filter(!is.na(study_group)) %>%
  mutate(Taxa = reorder(metab, -log10_umol_g)) %>%
  mutate(Taxa = fct_relabel(Taxa, function(x) gsub(" ", "\n", x))) %>%
  group_by(MOL, metab, study_group, sample_type) %>%
    summarise(mean=mean(log10_umol_g), sd=sd(log10_umol_g), n=n(), max=max(log10_umol_g)) %>%
  ungroup() %>%
  mutate(se=sd/sqrt(n)) %>%
  ggplot(aes(y=10^(mean), x=MOL)) +
    # geom_vline(aes(xintercept = PMA_rounded), lwd=0.3, color="grey") + # guide lines
    geom_line(aes(color=study_group, group = study_group)) +  
    geom_point(shape=21) +
    geom_point(data = summaries_df %>% filter(fdr<0.1), mapping = aes(x=MOL, y=10^(max*1.75)), shape=8, color="darkred") +
    geom_point(data = summaries_df %>% filter(p.value<0.05), mapping = aes(x=MOL, y=10^(max)+5), shape=23, color="red", fill="yellow") +
    # geom_point(data = summaries_df %>% filter(fdr<0.1, term=="moderate to severe retardation"), mapping = aes(color=term, x=MOL, y=10^(max*2)), shape=8) +
    scale_color_aaas() +
    # scale_color_manual(values = ann_colors$growth_failure) +
    scale_y_continuous(transform = "log10") +
    theme_bw() +
  theme(legend.position = "bottom") +
    theme(panel.grid = element_blank()) +
    geom_errorbar(aes(ymin = 10^(mean-se), ymax=10^(mean+se), color=study_group), alpha=0.5, width = 0.3) +
    facet_wrap(~metab, ncol=4, scales="free_y") +
  labs(y="Concentration (umol/g log-scale)")
```

Linear models were used to estimate the differences between study groups at each week of life. The relative abundances were log10 transformed. Multiple tests were adjusted for false discovery rate using Benjamini-Hochberg method. Only the terms with p<0.1 are shown in the table below.

For each week of life (MOL):  
lm(`r myform`)

```{r}
# table_export(summaries_df, "Taxonomic_table_export")

summaries_df %>%
  filter(p.value < 0.05) %>%
  pander(split.tab=Inf, emphasize.strong.rows = which(.$fdr<=0.05)) #%>%
  #column_spec(c(2:5), width = "3em") ## if the table is running off the page, you can manually specify the width of the columns

rm(myform)
```

## Week by week grouped metabolites   
```{r fig.height=9, fig.width=8}
myform="log10_umol_g ~ study_group"

props_toTest <- metab_summarised %>%
  rownames_to_column("metab") %>%
  pivot_longer(!metab, names_to = "SampleID", values_to = "umol_g") %>%
  mutate(log10_umol_g = log10(umol_g + min(metab_summarised[metab_summarised>0])/10))

summaries_df <- props_toTest %>%
  inner_join(s2 %>% select(SampleID, study_group=growth_failure, sample_type=SampleType, TimePoint, SubjectID, WOL, AbxHoursPerHOL), by = "SampleID") %>%
  filter(!is.na(study_group), metab != "Glycerol") %>%
  group_by(WOL, metab, study_group, sample_type) %>%
    mutate(mean=mean(log10_umol_g)) %>%
  ungroup() %>%
  
  group_by(metab) %>% mutate(max=max(mean)) %>% ungroup() %>%
  
  group_by(metab, WOL) %>%
    mutate(ngroups=n_distinct(study_group)) %>%
  ungroup() %>%
  
  filter(ngroups==2) %>% # Only test PMA_rounded where you have both study groups
  
  group_by(metab, WOL, max) %>%
  # summarise(n=n())
    do(tidy(lm(as.formula(myform), data=.))) %>%
  ungroup() %>%
  
  filter(!grepl("Intercept|Residuals", term)) %>%
  mutate(term = sub("study_group", "", term)) %>%
  
  group_by(term, WOL) %>%
    mutate(fdr = p.adjust(p.value, method="BH")) %>%
  ungroup()

props_toTest %>%
  inner_join(s2 %>% select(SampleID, study_group=growth_failure, sample_type=SampleType, TimePoint, SubjectID, WOL, AbxHoursPerHOL), by = "SampleID") %>%
  filter(!is.na(study_group)) %>%
  mutate(Taxa = reorder(metab, -log10_umol_g)) %>%
  mutate(Taxa = fct_relabel(Taxa, function(x) gsub(" ", "\n", x))) %>%
  group_by(WOL, metab, study_group, sample_type) %>%
    summarise(mean=mean(log10_umol_g), sd=sd(log10_umol_g), n=n(), max=max(log10_umol_g)) %>%
  ungroup() %>%
  mutate(se=sd/sqrt(n)) %>%
  ggplot(aes(y=10^(mean), x=WOL)) +
    # geom_vline(aes(xintercept = PMA_rounded), lwd=0.3, color="grey") + # guide lines
    geom_line(aes(color=study_group, group = study_group)) +  
    geom_point(shape=21) +
    geom_point(data = summaries_df %>% filter(fdr<0.1), mapping = aes(x=WOL, y=10^(max*1.75)), shape=8, color="darkred") +
    geom_point(data = summaries_df %>% filter(p.value<0.05), mapping = aes(x=WOL, y=10^(max*2)), shape=23, color="darkred") +
    # geom_point(data = summaries_df %>% filter(fdr<0.1, term=="moderate to severe retardation"), mapping = aes(color=term, x=WOL, y=10^(max*2)), shape=8) +
    scale_color_aaas() +
    # scale_color_manual(values = ann_colors$growth_failure) +
    scale_y_continuous(transform = "log10") +
    theme_bw() +
  theme(legend.position = "bottom") +
    theme(panel.grid = element_blank()) +
    geom_errorbar(aes(ymin = 10^(mean-se), ymax=10^(mean+se), color=study_group), alpha=0.5, width = 0.3) +
    facet_wrap(~metab, ncol=3, scales="free_y") +
  labs(y="Concentration (umol/g log-scale)")
```

Linear models were used to estimate the differences between study groups at each week of life. The relative abundances were log10 transformed. Multiple tests were adjusted for false discovery rate using Benjamini-Hochberg method. Only the terms with p<0.1 are shown in the table below.

For each week of life (WOL):  
lm(`r myform`)

```{r}
# table_export(summaries_df, "Taxonomic_table_export")

summaries_df %>%
  filter(p.value < 0.05) %>%
  pander(split.tab=Inf, emphasize.strong.rows = which(.$fdr<=0.05)) #%>%
  #column_spec(c(2:5), width = "3em") ## if the table is running off the page, you can manually specify the width of the columns

rm(myform)
```

## Monthly grouped metabolites   
```{r fig.height=9, fig.width=8}
myform="log10_umol_g ~ study_group"

props_toTest <- metab_summarised %>%
  rownames_to_column("metab") %>%
  pivot_longer(!metab, names_to = "SampleID", values_to = "umol_g") %>%
  mutate(log10_umol_g = log10(umol_g + min(metab_summarised[metab_summarised>0])/10))

summaries_df <- props_toTest %>%
  inner_join(s2 %>% select(SampleID, study_group=growth_failure, sample_type=SampleType, TimePoint, SubjectID, MOL, AbxHoursPerHOL), by = "SampleID") %>%
  filter(!is.na(study_group), metab != "Glycerol") %>%
  group_by(MOL, metab, study_group, sample_type) %>%
    mutate(mean=mean(log10_umol_g)) %>%
  ungroup() %>%
  
  group_by(metab) %>% mutate(max=max(mean)) %>% ungroup() %>%
  
  group_by(metab, MOL) %>%
    mutate(ngroups=n_distinct(study_group)) %>%
  ungroup() %>%
  
  filter(ngroups==2) %>% # Only test PMA_rounded where you have both study groups
  
  group_by(metab, MOL, max) %>%
  # summarise(n=n())
    do(tidy(lm(as.formula(myform), data=.))) %>%
  ungroup() %>%
  
  filter(!grepl("Intercept|Residuals", term)) %>%
  mutate(term = sub("study_group", "", term)) %>%
  
  group_by(term, MOL) %>%
    mutate(fdr = p.adjust(p.value, method="BH")) %>%
  ungroup()

props_toTest %>%
  inner_join(s2 %>% select(SampleID, study_group=growth_failure, sample_type=SampleType, TimePoint, SubjectID, MOL, AbxHoursPerHOL), by = "SampleID") %>%
  filter(!is.na(study_group)) %>%
  mutate(Taxa = reorder(metab, -log10_umol_g)) %>%
  mutate(Taxa = fct_relabel(Taxa, function(x) gsub(" ", "\n", x))) %>%
  group_by(MOL, metab, study_group, sample_type) %>%
    summarise(mean=mean(log10_umol_g), sd=sd(log10_umol_g), n=n(), max=max(log10_umol_g)) %>%
  ungroup() %>%
  mutate(se=sd/sqrt(n)) %>%
  ggplot(aes(y=10^(mean), x=MOL)) +
    # geom_vline(aes(xintercept = PMA_rounded), lwd=0.3, color="grey") + # guide lines
    geom_line(aes(color=study_group, group = study_group)) +  
    geom_point(shape=21) +
    geom_point(data = summaries_df %>% filter(fdr<0.1), mapping = aes(x=MOL, y=10), shape=8, color="darkred") +
    geom_point(data = summaries_df %>% filter(p.value<0.05), mapping = aes(x=MOL, y=10), shape=23, color="darkred") +
    # geom_point(data = summaries_df %>% filter(fdr<0.1, term=="moderate to severe retardation"), mapping = aes(color=term, x=MOL, y=10^(max*2)), shape=8) +
    scale_color_aaas() +
    # scale_color_manual(values = ann_colors$growth_failure) +
    scale_y_continuous(transform = "log10") +
    theme_bw() +
  theme(legend.position = "bottom") +
    theme(panel.grid = element_blank()) +
    geom_errorbar(aes(ymin = 10^(mean-se), ymax=10^(mean+se), color=study_group), alpha=0.5, width = 0.2) +
    facet_wrap(~metab, ncol=3, scales="free_y") +
  labs(y="Concentration (umol/g log-scale)")
```

Linear models were used to estimate the differences between study groups at each week of life. The relative abundances were log10 transformed. Multiple tests were adjusted for false discovery rate using Benjamini-Hochberg method. Only the terms with p<0.1 are shown in the table below.

For each week of life (MOL):  
lm(`r myform`)

```{r}
# table_export(summaries_df, "Taxonomic_table_export")

summaries_df %>%
  filter(p.value < 0.05) %>%
  pander(split.tab=Inf, emphasize.strong.rows = which(.$fdr<=0.05)) #%>%
  #column_spec(c(2:5), width = "3em") ## if the table is running off the page, you can manually specify the width of the columns

rm(myform)
```

\newpage
# FIG 2F - select metabolites over time by growth failure  
```{r fig.height=3, fig.width=8}
# sig_metab <- summaries_df %>% filter(fdr<0.2) %>% pull(metab)
stool_metab_long %>%
  # filter(metab %in% unique(summaries_df$metab)) %>%
  filter(metab %in% c("Butyrate", "Propionate", "Isovalerate", "Dimethylamine")) %>%
  right_join(s2, by="SampleID") %>%
  filter(!is.na(metab), metab!="Glycerol") %>%
  group_by(WOL, metab, growth_failure) %>%
  mutate(mean=mean(umol_g), sd=sd(umol_g), n=n()) %>%
  ungroup() %>%
  mutate(se = sd/sqrt(n)) %>%
  mutate(metab = fct_reorder(metab, -umol_g)) %>%
  arrange(growth_failure) %>%
  ggplot(aes(y=umol_g, x=WOL, group=growth_failure, color=growth_failure)) +
    geom_point(aes(shape=growth_failure), alpha=0.5) +
    geom_smooth(aes(fill=growth_failure), se=F) +
    # geom_point(aes(y=mean, shape=growth_failure)) +
    # geom_errorbar(aes(ymin=mean-se, ymax=mean+se)) +
    facet_wrap(~metab, scales="free_y", ncol=7) +
    scale_color_manual(values = ann_colors$growth_failure) +
    scale_fill_manual(values = ann_colors$growth_failure) +
    scale_shape_manual(values = c(21, 24)) +
   # scale_y_continuous(limit=c(0,NA),oob=squish) +
  theme_bw() +
  theme(legend.position = "bottom", panel.grid = element_blank()) +
  labs(fill="Growth Failure", color="Growth Failure", shape="Growth Failure", y='Concentration (umol/g)')
ggsave(here("Output", "FIG2F_metabolites_age_growthFail_subset.pdf"), width = 8, height = 3, useDingbats = F, dpi = 400)
```


\newpage
# FIG 2C Linear model estimate lineplot log10_umol_g ~ growth_cat  
```{r}
summaries_df <- run_lmer_on_props(factor_to_test = "metab", props_toTest = stool_metab_long %>% filter(metab != "Glycerol") %>% mutate(log10_umol_g = log10(umol_g)), s_toTest = s_toTest, form1 = "log10_umol_g ~ growth_cat", rep_mes_label = "SubjectID", p_cutoff =  1) %>% mutate(term = str_remove(term, "growth_cat")) #%>% filter(!grepl(":", term))

summaries_df %>%
  mutate(term = str_remove(term, "growth_cat")) %>%
  pander(split.table=Inf, digits=2, caption = glue("Linear models of log10 transformed concentrations of metabolites. Filtered with p.value < 0.05. SubjectID added as a random effect. Full formula: log10_umol_g ~ growth_cat"), emphasize.strong.rows = which(.$fdr<=0.05))

summaries_df %>%
  mutate(Significance = ifelse(fdr < 0.05, "q<0.05", "q>0.05")) %>%
  arrange(Value) %>%
  ggplot(aes(x=Value, y=fct_inorder(metab), color=term, group=term)) +
  geom_vline(aes(xintercept = 0), lty=2) +
  geom_linerange(aes(xmin = Value - Std.Error, xmax=Value + Std.Error), position=position_dodge(0.8)) +
  geom_point(aes(shape=Significance), size=2, position=position_dodge(0.8)) +
  scale_shape_manual(values = c(21, 16)) +
  scale_color_manual(values = ann_colors$growth_cat) +
  theme_bw() +
  theme(legend.position = "bottom", panel.grid = element_blank()) +
  labs(color="", y="Metabolite", x="Estimate", title="log10_umol_g ~ growth_cat")
  
ggsave(here("Output", "FIG2C_Metab_GrowthCat_LinearModEst.pdf"), width = 5, height = 8, useDingbats = F, dpi = 400)
```

\newpage
# FIG 2X Linear model estimate lineplot
```{r}
summaries_df <- run_lmer_on_props(factor_to_test = "metab", props_toTest = stool_metab_long %>% filter(metab != "Glycerol") %>% mutate(log10_umol_g = log10(umol_g)), s_toTest = s_toTest, form1 = "log10_umol_g ~ growth_failure", rep_mes_label = "SubjectID", p_cutoff =  1) %>% mutate(term = str_remove(term, "growth_failure")) #%>% filter(!grepl(":", term))

summaries_df %>%
  mutate(term = str_remove(term, "growth_failure")) %>%
  arrange(p.value) %>%
  pander(split.table=Inf, digits=2, caption = glue("Linear models of log10 transformed concentrations of metabolites. Filtered with p.value < 0.05. SubjectID added as a random effect. Full formula: log10_umol_g ~ growth_failure)"), emphasize.strong.rows = which(.$fdr<=0.05))

summaries_df %>%
  mutate(Significance = ifelse(fdr < 0.05, "q<0.05", ifelse(fdr<0.1, "q<0.1", "q>0.05"))) %>%
  arrange(Value) %>%
  ggplot(aes(x=Value, y=fct_inorder(metab), color=term, group=term)) +
  geom_vline(aes(xintercept = 0), lty=2) +
  geom_linerange(aes(xmin = Value - Std.Error, xmax=Value + Std.Error), position=position_dodge(0.8)) +
  geom_point(aes(shape=Significance), size=2, position=position_dodge(0.8)) +
  scale_shape_manual(values = c(21, 22, 16)) +
  scale_color_manual(values = ann_colors$growth_failure) +
  theme_bw() +
  theme(legend.position = "bottom", panel.grid = element_blank()) +
  labs(color="", y="Metabolite", x="Estimate", title="log10_umol_g ~ growth_failure")
  
ggsave(here("Output", "FIG2D_Metab_GrowthFail_LinearModEst.pdf"), width = 5, height = 8, useDingbats = F, dpi = 400)
```

\newpage
Growth category over time with NPO days  
```{r}
### Linear models of metabolites
# number_of_top_pathways = 25
props_toTest <- stool_metab_long %>%
  filter(metab != "Glycerol") %>%
  mutate(log10_umol_g = log10(umol_g))

form1 = "log10_umol_g ~ WOL * growth_cat"
form2 = "log10_umol_g ~ WOL * growth_cat + npo_days_prop"
form3 = "log10_umol_g ~ WOL * growth_cat + AbxHoursPerHOL"
random1 = "SubjectID"

summaries_df <- run_lmer_on_props(factor_to_test = "metab", props_toTest = props_toTest, s_toTest = s_toTest, form1 = form2,rep_mes_label = random1, p_cutoff =  0.05) 

summaries_df %>%
  mutate(term = str_remove(term, "growth_cat")) %>%
  pander(split.table=Inf, digits=2, caption = glue("Linear models of log10 transformed concentrations of metabolites. Filtered with p.value < 0.05. SubjectID added as a random effect. Full formula: {form2}"), emphasize.strong.rows = which(.$fdr<=0.05))

# table_export(summaries_df, "metabolites_analysis_table_")

```

\newpage
Growth category over time with Abx hours  
```{r}
### Linear models of metabolites
# number_of_top_pathways = 25
props_toTest <- stool_metab_long %>%
  filter(metab != "Glycerol") %>%
  mutate(log10_umol_g = log10(umol_g))

form1 = "log10_umol_g ~ WOL * growth_cat"
form2 = "log10_umol_g ~ WOL * growth_cat + npo_days_prop"
form3 = "log10_umol_g ~ WOL * growth_cat + AbxHoursPerHOL"
random1 = "SubjectID"

summaries_df <- run_lmer_on_props(factor_to_test = "metab", props_toTest = props_toTest, s_toTest = s_toTest, form1 = form3,rep_mes_label = random1, p_cutoff =  0.05) 

summaries_df %>%
  mutate(term = str_remove(term, "growth_cat")) %>%
  pander(split.table=Inf, digits=2, caption = glue("Linear models of log10 transformed concentrations of metabolites. Filtered with p.value < 0.05. SubjectID added as a random effect. Full formula: {form3}"), emphasize.strong.rows = which(.$fdr<=0.05))

# table_export(summaries_df, "metabolites_analysis_table_")

```



\newpage

# End of report

```{r Generate time stamped report, eval=FALSE, include=FALSE}
#notes
#neat: you can run this following command in the console to give your reports custom names (or date-stamp them)
#rmarkdown::render('ronen_stein_Report_shotgun.Rmd',output_file = paste('CEASE.report.', Sys.Date(), '.pdf', sep=''))

rstudioapi::documentSaveAll()

script_dir <- file.path("Scripts","Rmds")
output_dir <- file.path("Output")

report_fp <- here::here(script_dir, "FIGURESANDTABLES.Rmd")
output_fp <- here::here(output_dir, paste('sagori_microbiome_figures_and_tables_', Sys.Date(), '.pdf', sep=''))

rmarkdown::render(report_fp, output_file = output_fp)

```
