---
title: |
  ![](CHOPMC_logo.pdf){width=5in}  
  "Sagori - baby growth analysis - metabolomics"
author: "Scott G. Daniel, Ph.D."
date: \today
output: 
  pdf_document:
    toc: true
    toc_depth: 3
---

```{r knitr setup, echo=FALSE, message=FALSE}
library(knitr)
opts_chunk$set(
  tidy=FALSE,
  cache=FALSE,
  echo=FALSE,
  warning=FALSE,
  message=FALSE,
  dpi=100,
  fig.width=6,
  fig.height=6,
  fig.align = "center",
  fig_crop=FALSE
  )
```

```{r libraries, message=FALSE, warning=FALSE}
library(here)
library(ggbeeswarm)
library(grid)
library(magrittr)
library(pander)
library(pheatmap)
library(usedist)
library(qiimer)
library(reshape2)
library(jsonlite)
library(vegan)
library(ape)
library(tidyverse)
library(gghighlight)
library(RColorBrewer)
library(patchwork)
library(kylemisc)
library(ggpubr)
library(GGally)
library(lme4)
# library(tsne)
library(Rtsne)

library(glue)
library(kableExtra)

library(pheatbuilder)
library(ppcor)
library(scales)
library(ghibli)
library(viridis)
library(qualpalr)
library(ggVennDiagram)

panderOptions("knitr.auto.asis", FALSE)
#this also means that you have to put results='asis' in any block that has pander output
#this lets you loop through variables and produce multiple pander tables and ggplots in a
#single code block!

select <- dplyr::select

```

```{r functions}

source(here::here("Scripts","Functions","common_functions.R"))
source(here::here("Scripts","Functions","for_functional_analysis.R"))
source(here::here("Scripts","Functions","scott_daniel_useful_R_functions.R"))

```

```{r load data and set colors}

load(file = here("Data", "after_gene_abundance_analysis.RData"))

s_toPlot <- s_toTest <- s2

ann_colors = list(
  category = setNames(brewer.pal(length(levels(s_toTest$category)), "Set2"), levels(s_toTest$category)),
  growth_cat = setNames(as.character(ghibli_palette("YesterdayMedium", direction = -1)[1:length(levels(s_toTest$growth_cat))]), levels(s_toTest$growth_cat)),
  TimePoint = setNames(viridis(length(levels(s_toTest$TimePoint))), levels(s_toTest$TimePoint)),
  growth_fail = setNames(qualpal(2)$hex, levels(s_toTest$growth_failure))
)

table_iterator <- function() {
  table_count <- 0
  function(x, title = "table_export", ...) {
    table_count <<- table_count + 1
    table_name <- paste0(title, table_count, ".csv")
    table_path <- here("Output", table_name)
    write_csv(x = x, file = table_path, ...)
    print(paste0("Full table is saved as ", table_name, "."))
  }
}

table_export <- table_iterator()

```

# START

# Metabolomics - Stool

```{r message=FALSE, warning=FALSE}

#Stool metabolites in micromolars per gram (umol/g)

stool_metab_raw <- read_csv(here("Data","stool_metab.csv")) %>%
  rename(nmr_no = `NMR No.`) %>%
  select(-stool_date, -box, -location, -barcode_tube) %>%
  mutate(redcap_id = as.character(redcap_id))

stool_metab_mapping <- read_csv(here("Data", "stool_metab_mapping.csv")) %>%
  select(SampleID = SampleID_seq, nmr_no = `NMR No.`, Seq_run5_YN,	Inf_metBol_mlk,	DOL,	HOL,	in_36wklist)

stool_metab <- stool_metab_raw %>%
  inner_join(stool_metab_mapping, by = "nmr_no") %>%
  select(SampleID, everything()) %>%
  mutate(Sequenced = ifelse(SampleID %in% "Not seq", F, T)) %>%
  mutate(SampleID = ifelse(SampleID %in% "Not seq", paste0("inf", redcap_id, "nmr_no", nmr_no), SampleID))

```

## Total samples w/ microbiome and metabolite data

```{r}

x = list(
  Microbiota = unique(c(stool_metab %>% filter(Sequenced) %>% pull(SampleID), s_toTest$SampleID)),
  Metabolites = stool_metab$SampleID
)

ggVennDiagram(x) +
  scale_fill_gradient(low="blue",high = "orange") +
  scale_x_continuous(expand = expansion(mult = .2))

```

## Total babies w/ microbiome and/or metabolite data

```{r}

x = list(
  Microbiota = unique(c(stool_metab %>% filter(Sequenced) %>% pull(redcap_id), s_toTest$redcap_id)),
  Metabolites = unique(stool_metab$redcap_id)
)

ggVennDiagram(x) +
  scale_fill_gradient(low="blue",high = "orange") +
  scale_x_continuous(expand = expansion(mult = .2))

```


## Overview plots

### Histograms of metabolites

```{r, fig.width=10, fig.height=7}

metab_names <- stool_metab_raw %>%
  select(4:length(colnames(stool_metab_raw))) %>%
  colnames()

stool_metab_long <- stool_metab %>%
  pivot_longer(cols = all_of(metab_names), names_to = "metab", values_to = "umol_g", values_drop_na = T) %>%
  droplevels()

max_num <- max(stool_metab_long$umol_g)

stool_metab_long %<>%
  group_by(metab) %>%
  mutate(metab_min = ifelse(umol_g == 0, max_num, umol_g)) %>%
  mutate(umol_g = ifelse(umol_g == 0, min(metab_min)/2, umol_g)) %>%
  ungroup()

number_of_blocks <- ceiling(length(metab_names) / 9)
my_blocks <- chunk_it(metab_names, number_of_blocks)

for (i in 1:number_of_blocks) {
  subset <- my_blocks[[i]]
  
  g <- stool_metab_long %>%
    filter(metab %in% subset) %>%
    ggplot(aes(umol_g)) +
    geom_histogram() +
    facet_wrap(~metab, scales="free")
  
  plot(g)
  
}

```

### Samples by growth category

Zeros were replaced with 1/2 the minimum value per metabolite.

```{r, fig.width=10, fig.height=7}

metab_to_Plot <- stool_metab_long %>%
  inner_join(s_toPlot %>% select(SampleID, growth_cat, growth_failure, TimePoint, SubjectID), by = "SampleID") %>% 
  droplevels()

for (i in 1:number_of_blocks) {
  subset <- my_blocks[[i]]
  
  g <- metab_to_Plot %>%
    filter(!is.na(growth_cat), metab %in% subset) %>%
    ggplot(aes(x=growth_cat, y=umol_g, color=TimePoint)) +
    geom_quasirandom(dodge.width = 0.75, groupOnX = T) +
    theme_bw() +
    facet_wrap(~metab, scales="free_y") +
    theme(
      axis.text.x=element_text(angle=45, hjust=1, vjust=1),
      panel.grid = element_blank(),
      strip.background = element_blank(),
      legend.position = "bottom"
    ) +
    scale_color_manual(values = ann_colors$TimePoint) +
    scale_y_continuous(trans = "log10") +
    labs(
      x="Growth category", color="Timepoint",
      y="Concentration (umol/g)"
    )
  
  plot(g)
  
}
    
```

### Samples by growth failure

```{r, fig.width=10, fig.height=7}

for (i in 1:number_of_blocks) {
  subset <- my_blocks[[i]]
  
  g <- metab_to_Plot %>%
    filter(!is.na(growth_failure), metab %in% subset) %>%
    ggplot(aes(x=growth_failure, y=umol_g, color=TimePoint)) +
    geom_quasirandom(dodge.width = 0.75, groupOnX = T) +
    theme_bw() +
    facet_wrap(~metab, scales="free_y") +
    theme(
      axis.text.x=element_text(angle=45, hjust=1, vjust=1),
      panel.grid = element_blank(),
      strip.background = element_blank(),
      legend.position = "bottom"
    ) +
    scale_color_manual(values = ann_colors$TimePoint) +
    scale_y_continuous(trans = "log10") +
    labs(
      x="Growth failure", color="Timepoint",
      y="Concentration (umol/g)"
    )
  
  plot(g)
  
}

```

## Linear models for growth categories

Are there any significant differences between growth categories? We will log10 transform the values since they do not appear normally distributed.

```{r, results='asis'}

props_toTest <- stool_metab_long %>%
  mutate(log10_umol_g = log10(umol_g))

my_formula <- "log10_umol_g ~ timepoint_numeric * growth_cat + npo_days_prop + feeds_prop_hm + AbxHoursPerHOL + Sex"

summaries_df <- run_lmer_on_props(factor_to_test = "metab", props_toTest = props_toTest, s_toTest = s_toTest, form1 = my_formula, rep_mes_label = "SubjectID", p_cutoff = 1)

my_table <- summaries_df %>%
  mutate(term = str_replace(term, "growth_cat", "")) %>%
  mutate(term = str_replace(term, "SexM", "Male"))

my_table %>%
  filter(p.value < 0.05) %>%
  pander(digits = 2, split.table = Inf, caption = glue("Linear models of log10 transformed concentrations of metabolites. Filtered with p.value < 0.05. SubjectID added as a random effect. Full formula: {my_formula}"))

my_table %>% table_export(title = "Sagori_metabolites_table_")

```

A lot of them are strongly associated with the amount of Antibiotics given to the babies. Let's graph that.

### Samples by Abx use

```{r, fig.width=10, fig.height=7}

metabs_to_plot <- my_table %>%
  filter(fdr < 0.05, term %in% "AbxHoursPerHOL") %>%
  pull(metab) %>%
  unique()

number_of_blocks <- ceiling(length(metabs_to_plot) / 9)
my_blocks <- chunk_it(metabs_to_plot, number_of_blocks)

for (i in 1:number_of_blocks) {
  subset <- my_blocks[[i]]
  
  g <- metab_to_Plot %>%
    inner_join(s_toTest %>% select(SampleID, AbxHoursPerHOL), by = "SampleID") %>%
    filter(!is.na(AbxHoursPerHOL), metab %in% subset) %>%
    ggplot(aes(x=AbxHoursPerHOL, y=umol_g, color=TimePoint)) +
    geom_point() +
    theme_bw() +
    facet_wrap(~metab, scales="free_y") +
    theme(
      axis.text.x=element_text(angle=45, hjust=1, vjust=1),
      panel.grid = element_blank(),
      strip.background = element_blank(),
      legend.position = "bottom"
    ) +
    scale_color_manual(values = ann_colors$TimePoint) +
    scale_y_continuous(trans = "log10") +
    labs(
      x="AbxHoursPerHOL", color="Timepoint",
      y="Concentration (umol/g)"
    )
  
  plot(g)
  
}

```

## Linear models for growth failure

Are there any significant differences between growth regular and growth failure? We will log10 transform the values since they do not appear normally distributed.

```{r, results='asis'}

props_toTest <- stool_metab_long %>%
  mutate(log10_umol_g = log10(umol_g))

my_formula <- "log10_umol_g ~ timepoint_numeric * growth_failure + npo_days_prop + feeds_prop_hm + AbxHoursPerHOL + Sex"

summaries_df <- run_lmer_on_props(factor_to_test = "metab", props_toTest = props_toTest, s_toTest = s_toTest, form1 = my_formula, rep_mes_label = "SubjectID", p_cutoff = 1)

my_table <- summaries_df %>%
  mutate(term = str_replace(term, "growth_failure", "")) %>%
  mutate(term = str_replace(term, "SexM", "Male"))

my_table %>%
  filter(p.value < 0.05) %>%
  pander(digits = 2, split.table = Inf, caption = glue("Linear models of log10 transformed concentrations of metabolites. Filtered with p.value < 0.05. SubjectID added as a random effect. Full formula: {my_formula}"))

my_table %>% table_export(title = "Sagori_metabolites_table_")

```

# Metabolomics - Breast milk

```{r message=FALSE, warning=TRUE}

#Breast milk metabolites in millimolars (mM)

bm_metab_raw <- read_csv(here("Data","breast_milk_metab.csv")) %>%
  rename(nmr_no = `NMR No.`) %>%
  select(-box, -location, -bm_barcode) %>%
  mutate(redcap_id = as.character(redcap_id))

```


## Overview plots

### Histograms of metabolites

```{r, fig.width=10, fig.height=7}

metab_names <- bm_metab_raw %>%
  select(5:length(colnames(bm_metab_raw))) %>%
  colnames()

bm_metab_long <- bm_metab_raw %>%
  pivot_longer(cols = all_of(metab_names), names_to = "metab", values_to = "mM", values_drop_na = T) %>%
  droplevels()

# any(bm_metab_long$mM == 0)
# TRUE

max_num <- max(bm_metab_long$mM)

bm_metab_long %<>%
  group_by(metab) %>%
  mutate(metab_min = ifelse(mM == 0, max_num, mM)) %>%
  mutate(mM = ifelse(mM == 0, min(metab_min)/2, mM)) %>%
  ungroup()

number_of_blocks <- ceiling(length(metab_names) / 9)
my_blocks <- chunk_it(metab_names, number_of_blocks)

for (i in 1:number_of_blocks) {
  subset <- my_blocks[[i]]
  
  g <- bm_metab_long %>%
    filter(metab %in% subset) %>%
    ggplot(aes(mM)) +
    geom_histogram() +
    facet_wrap(~metab, scales="free")
  
  plot(g)
  
}

```


### Samples by growth category

Zeros were replaced with 1/2 the minimum value per metabolite.

```{r, fig.width=10, fig.height=7}

metab_to_Plot <- bm_metab_long %>%
  inner_join(s_toPlot %>% select(growth_cat, growth_failure, redcap_id) %>% unique(), by = "redcap_id") %>% 
  droplevels()

for (i in 1:number_of_blocks) {
  subset <- my_blocks[[i]]
  
  g <- metab_to_Plot %>%
    filter(!is.na(growth_cat), metab %in% subset) %>%
    ggplot(aes(x=growth_cat, y=mM, color=growth_cat)) +
    geom_boxplot(outlier.alpha = 0) +
    geom_quasirandom(dodge.width = 0.75, groupOnX = T) +
    theme_bw() +
    facet_wrap(~metab, scales="free_y") +
    theme(
      axis.text.x=element_text(angle=45, hjust=1, vjust=1),
      panel.grid = element_blank(),
      strip.background = element_blank(),
      legend.position = "none"
    ) +
    scale_y_continuous(trans = "log10") +
    scale_color_manual(values = ann_colors$growth_cat) +
    labs(
      x="Growth category",
      y="Concentration (mM)"
    )
  
  plot(g)
  
}
    
```

### Samples by growth failure

All breast milk samples were growth regular.

## Linear models

Are there any significant differences between growth categories? We will log10 transform the values since they do not appear normally distributed.

```{r, results='asis'}

props_toTest <- metab_to_Plot %>%
  mutate(log10_mM = log10(mM))

my_formula <- "log10_mM ~ growth_cat"
my_random_effect <- "~ 1 | redcap_id"
p_cutoff = 1

summaries_df <- props_toTest %>%
  group_by(metab) %>%
  do(tidy_lmer(nlme::lme(as.formula(my_formula), random = as.formula(my_random_effect), data=., na.action=na.omit))) %>%
  ungroup() %>%
  filter(term != '(Intercept)') %>%
  group_by(term) %>%
  mutate(fdr = p.adjust(p.value, method="BH")) %>%
  ungroup() %>%
  filter(p.value<p_cutoff)

my_table <- summaries_df %>%
  mutate(term = str_replace(term, "growth_cat", ""))

my_table %>%
  filter(p.value < 0.05) %>%
  pander(digits = 2, split.table = Inf, caption = glue("Linear models of log10 transformed concentrations of metabolites. Filtered with p.value < 0.05. SubjectID added as a random effect. Full formula: {my_formula}"))

my_table %>% table_export(title = "Sagori_metabolites_table_")

```

\newpage

# Correlating metabolites

In this section we will correlate metabolites to other variables and themselves. We will do this with the stool samples.

## Kegg pathways to metab

```{r}

metab_df <- stool_metab_long %>%
  mutate(log10_umol_g = log10(umol_g))

selected_pathways <- kegg_path_props %>%
  filter(!PathwayName %in% "Metabolic pathways") %>%
  mutate(PathwayName = fct_lump(PathwayName, n = number_of_top_pathways, w = .$props, other_level = "Other")) %>%
  filter(!PathwayName %in% "Other") %>%
  droplevels() %>%
  pull(PathwayName) %>%
  unique()

pathway_df <- kegg_path_props %>%
  filter(PathwayName %in% selected_pathways) %>%
  mutate(props_log10 = log10(props))

my_df <- inner_join(metab_df, pathway_df, by = c("SampleID")) %>%
  group_by(metab, PathwayName) %>%
  summarize(cor = cor(log10_umol_g, props_log10), p.value = broom::tidy(cor.test(log10_umol_g, props_log10, method = "pearson"))$p.value) %>%
  ungroup() %>%
  mutate(fdr = p.adjust(p.value, method = "BH"))

my_df_wide <- my_df %>%
  select(metab, PathwayName, cor) %>%
  pivot_wider(names_from = "PathwayName", values_from = cor)

my_matrix <- my_df_wide %>%
  column_to_rownames(var = "metab") %>%
  as.matrix()

significance_matrix <- my_df %>%
  mutate(starred = ifelse(fdr < 0.05, "*", "")) %>%
  pivot_wider(id_cols = metab, names_from = "PathwayName", values_from = starred) %>%
  column_to_rownames(var = "metab") %>%
  as.matrix()

my_matrix %>%
  pheat() %>%
  pheat_display_main(display_numbers = significance_matrix) %>%
  pheat_cluster_rows() %>%
  pheat_cluster_cols() %>%
  print()

```

```{r}
heatmap_name = "kegg_pathways_to_metabs.pdf"

my_matrix %>%
  pheat() %>%
  pheat_display_main(display_numbers = significance_matrix) %>%
  pheat_cluster_rows() %>%
  pheat_cluster_cols() %>%
  pheat_save(filename = here("Output", heatmap_name))
```

Heatmap saved as `r heatmap_name`.

\newpage

## KEGG orthologs to metab

Here we will correlate KEGG orthologs whose abundances changed with metabolites.

### All timepoints growth faltering_individual_gene_changes_growth_categories

```{r}

gene_list <- read_csv(here("Output", "All timepoints growth faltering_individual_gene_changes_growth_categories.csv")) %>%
  filter(fdr < 0.05) %>%
  mutate(geneShortDescr = paste(Ortholog, str_trunc(Definition, 30), sep = ": "))

ko_df <- kegg_props %>%
  right_join(gene_list, by = c("geneID"= "Ortholog")) %>%
  select(SampleID, geneShortDescr, count, props) %>%
  mutate(props_log10 = log10(props))

my_df <- inner_join(metab_df, ko_df, by = c("SampleID")) %>%
  group_by(metab, geneShortDescr) %>%
  summarize(cor = cor(log10_umol_g, props_log10), p.value = broom::tidy(cor.test(log10_umol_g, props_log10, method = "pearson"))$p.value) %>%
  ungroup() %>%
  mutate(fdr = p.adjust(p.value, method = "BH"))

my_df_wide <- my_df %>%
  select(metab, geneShortDescr, cor) %>%
  pivot_wider(names_from = "geneShortDescr", values_from = cor)

my_matrix <- my_df_wide %>%
  column_to_rownames(var = "metab") %>%
  as.matrix()

significance_matrix <- my_df %>%
  mutate(starred = ifelse(fdr < 0.05, "*", "")) %>%
  pivot_wider(id_cols = metab, names_from = "geneShortDescr", values_from = starred) %>%
  column_to_rownames(var = "metab") %>%
  as.matrix()

my_matrix %>%
  pheat() %>%
  pheat_display_main(display_numbers = significance_matrix) %>%
  pheat_cluster_rows() %>%
  pheat_cluster_cols() %>%
  print()

```

```{r}
heatmap_name = "KEGG_orthologs_to_metabs_all_timepoints.pdf"

my_matrix %>%
  pheat() %>%
  pheat_display_main(display_numbers = significance_matrix) %>%
  pheat_cluster_rows() %>%
  pheat_cluster_cols() %>%
  pheat_save(filename = here("Output", heatmap_name))
```

Heatmap saved as `r heatmap_name`.

\newpage

### 0to21day_individual_gene_changes_growth_categories

```{r}

gene_list1 <- read_csv(here("Output", "0to21day_individual_gene_changes_growth_categories.csv")) %>%
  filter(fdr < 0.05) %>%
  mutate(geneShortDescr = paste(Ortholog, str_trunc(Definition, 30), sep = ": ")) %>%
  select(Ortholog, geneShortDescr)

gene_list2 <- read_csv(here("Output", "0to21day_individual_gene_changes_growth_categories_gf.csv")) %>%
  filter(fdr < 0.05) %>%
  mutate(geneShortDescr = paste(Ortholog, str_trunc(Definition, 30), sep = ": ")) %>%
  select(Ortholog, geneShortDescr)

gene_list <- bind_rows(gene_list1, gene_list2) %>%
  unique()

ko_df <- kegg_props %>%
  right_join(gene_list, by = c("geneID"= "Ortholog")) %>%
  select(SampleID, geneShortDescr, count, props) %>%
  mutate(props_log10 = log10(props))

my_df <- inner_join(metab_df, ko_df, by = c("SampleID")) %>%
  group_by(metab, geneShortDescr) %>%
  summarize(cor = cor(log10_umol_g, props_log10), p.value = broom::tidy(cor.test(log10_umol_g, props_log10, method = "pearson"))$p.value) %>%
  ungroup() %>%
  mutate(fdr = p.adjust(p.value, method = "BH"))

my_df_wide <- my_df %>%
  select(metab, geneShortDescr, cor) %>%
  pivot_wider(names_from = "geneShortDescr", values_from = cor)

my_matrix <- my_df_wide %>%
  column_to_rownames(var = "metab") %>%
  as.matrix()

significance_matrix <- my_df %>%
  mutate(starred = ifelse(fdr < 0.05, "*", "")) %>%
  pivot_wider(id_cols = metab, names_from = "geneShortDescr", values_from = starred) %>%
  column_to_rownames(var = "metab") %>%
  as.matrix()

my_matrix %>%
  pheat() %>%
  pheat_display_main(display_numbers = significance_matrix) %>%
  pheat_cluster_rows() %>%
  pheat_cluster_cols() %>%
  print()

```

```{r}
heatmap_name = "KEGG_orthologs_to_metabs_0to21day.pdf"

my_matrix %>%
  pheat() %>%
  pheat_display_main(display_numbers = significance_matrix) %>%
  pheat_cluster_rows() %>%
  pheat_cluster_cols() %>%
  pheat_save(filename = here("Output", heatmap_name))
```

Heatmap saved as `r heatmap_name`.

\newpage

### 34to37weeks_individual_gene_changes_growth_categories

```{r}

gene_list1 <- read_csv(here("Output", "34to37weeks_individual_gene_changes_growth_categories.csv")) %>%
  filter(fdr < 0.05) %>%
  mutate(geneShortDescr = paste(Ortholog, str_trunc(Definition, 30), sep = ": ")) %>%
  select(Ortholog, geneShortDescr)

gene_list2 <- read_csv(here("Output", "34to37weeks_individual_gene_changes_growth_categories_gf.csv")) %>%
  filter(fdr < 0.05) %>%
  mutate(geneShortDescr = paste(Ortholog, str_trunc(Definition, 30), sep = ": ")) %>%
  select(Ortholog, geneShortDescr)

gene_list <- bind_rows(gene_list1, gene_list2) %>%
  unique()

ko_df <- kegg_props %>%
  right_join(gene_list, by = c("geneID"= "Ortholog")) %>%
  select(SampleID, geneShortDescr, count, props) %>%
  mutate(props_log10 = log10(props))

my_df <- inner_join(metab_df, ko_df, by = c("SampleID")) %>%
  group_by(metab, geneShortDescr) %>%
  summarize(cor = cor(log10_umol_g, props_log10), p.value = broom::tidy(cor.test(log10_umol_g, props_log10, method = "pearson"))$p.value) %>%
  ungroup() %>%
  mutate(fdr = p.adjust(p.value, method = "BH"))

my_df_wide <- my_df %>%
  select(metab, geneShortDescr, cor) %>%
  pivot_wider(names_from = "geneShortDescr", values_from = cor)

my_matrix <- my_df_wide %>%
  column_to_rownames(var = "metab") %>%
  as.matrix()

significance_matrix <- my_df %>%
  mutate(starred = ifelse(fdr < 0.05, "*", "")) %>%
  pivot_wider(id_cols = metab, names_from = "geneShortDescr", values_from = starred) %>%
  column_to_rownames(var = "metab") %>%
  as.matrix()

my_matrix %>%
  pheat() %>%
  pheat_display_main(display_numbers = significance_matrix) %>%
  pheat_cluster_rows() %>%
  pheat_cluster_cols() %>%
  print()

```

```{r}
heatmap_name = "KEGG_orthologs_to_metabs_34to37weeks.pdf"

my_matrix %>%
  pheat() %>%
  pheat_display_main(display_numbers = significance_matrix) %>%
  pheat_cluster_rows() %>%
  pheat_cluster_cols() %>%
  pheat_save(filename = here("Output", heatmap_name))
```

Heatmap saved as `r heatmap_name`.

\newpage

## Butyrate genes to the other metab

```{r}

but_df <- but_ipa %>%
  inner_join(s_toTest, by = "SampleID") %>%
  filter(SampleID %in% stool_metab$SampleID) %>%
  mutate(props_log10 = log10(props))

my_df <- inner_join(metab_df, but_df, by = c("SampleID")) %>%
  group_by(metab, geneID) %>%
  summarize(cor = cor(log10_umol_g, props_log10), p.value = broom::tidy(cor.test(log10_umol_g, props_log10, method = "pearson"))$p.value) %>%
  ungroup() %>%
  mutate(fdr = p.adjust(p.value, method = "BH"))

my_df_wide <- my_df %>%
  select(metab, geneID, cor) %>%
  pivot_wider(names_from = "geneID", values_from = cor)

my_matrix <- my_df_wide %>%
  column_to_rownames(var = "metab") %>%
  as.matrix()

significance_matrix <- my_df %>%
  mutate(starred = ifelse(fdr < 0.05, "*", "")) %>%
  pivot_wider(id_cols = metab, names_from = "geneID", values_from = starred) %>%
  column_to_rownames(var = "metab") %>%
  as.matrix()

my_matrix %>%
  pheat() %>%
  pheat_display_main(display_numbers = significance_matrix) %>%
  pheat_cluster_rows() %>%
  pheat_cluster_cols() %>%
  print()

```

```{r}
heatmap_name = "butyrate_genes_to_metabs.pdf"

my_matrix %>%
  pheat() %>%
  pheat_display_main(display_numbers = significance_matrix) %>%
  pheat_cluster_cols() %>%
  pheat_cluster_rows() %>%
  pheat_save(filename = here("Output", heatmap_name))
```

Heatmap saved as `r heatmap_name`.

\newpage

## Most abundant taxa to bile acids

We will filter out taxa that have less than 1% mean abundance across samples. We will also manually include butyrate producers (those are the ones with shortened names).

```{r}

but_producers <- read_csv(here("Data", "but_producers_bacdive_2024-03-19.csv"), skip = 1)

thing <- summed_props_long %>%
  mutate(Taxa = str_remove(Taxa, "\\w+ ")) %>%
  filter(Taxa %in% but_producers$species | str_detect(Taxa, "butyric"))

```

```{r}

taxa_df <- summed_props_long %>%
  group_by(Taxa) %>%
  mutate(mean_prop = mean(props)) %>%
  ungroup() %>%
  filter(mean_prop > 0.01) %>%
  bind_rows(thing) %>%
  mutate(props_log10 = log10(props))

my_df <- inner_join(metab_df, taxa_df, by = c("SampleID")) %>%
  group_by(metab, Taxa) %>%
  summarize(cor = cor(log10_umol_g, props_log10), p.value = broom::tidy(cor.test(log10_umol_g, props_log10, method = "pearson"))$p.value) %>%
  ungroup() %>%
  mutate(fdr = p.adjust(p.value, method = "BH"))

my_df_wide <- my_df %>%
  select(metab, Taxa, cor) %>%
  pivot_wider(names_from = "Taxa", values_from = cor)

my_matrix <- my_df_wide %>%
  column_to_rownames(var = "metab") %>%
  as.matrix()

significance_matrix <- my_df %>%
  mutate(starred = ifelse(fdr < 0.05, "*", "")) %>%
  pivot_wider(id_cols = metab, names_from = "Taxa", values_from = starred) %>%
  column_to_rownames(var = "metab") %>%
  as.matrix()

my_matrix %>%
  pheat() %>%
  pheat_display_main(display_numbers = significance_matrix) %>%
  pheat_cluster_rows() %>%
  pheat_cluster_cols() %>%
  print()

```

```{r}
heatmap_name = "most_abundant_taxa_to_metabs.pdf"

my_matrix %>%
  pheat() %>%
  pheat_display_main(display_numbers = significance_matrix) %>%
  pheat_cluster_rows() %>%
  pheat_cluster_cols() %>%
  pheat_save(filename = here("Output", heatmap_name))
```

Heatmap saved as `r heatmap_name`.

\newpage

## Metabolites to themselves

```{r}

my_df <- inner_join(metab_df, metab_df, by = c("SampleID")) %>%
  group_by(metab.x, metab.y) %>%
  summarize(cor = cor(log10_umol_g.x, log10_umol_g.y), p.value = broom::tidy(cor.test(log10_umol_g.x, log10_umol_g.y, method = "pearson"))$p.value) %>%
  ungroup() %>%
  mutate(fdr = p.adjust(p.value, method = "BH"))

my_df_wide <- my_df %>%
  select(metab.x, metab.y, cor) %>%
  pivot_wider(names_from = "metab.y", values_from = cor)

my_matrix <- my_df_wide %>%
  column_to_rownames(var = "metab.x") %>%
  as.matrix()

significance_matrix <- my_df %>%
  mutate(starred = ifelse(fdr < 0.05, "*", "")) %>%
  pivot_wider(id_cols = metab.x, names_from = "metab.y", values_from = starred) %>%
  column_to_rownames(var = "metab.x") %>%
  as.matrix()

my_matrix %>%
  pheat() %>%
  pheat_display_main(display_numbers = significance_matrix) %>%
  pheat_cluster_rows() %>%
  pheat_cluster_cols() %>%
  print()

```

```{r}
heatmap_name = "metabs_to_themselves.pdf"

my_matrix %>%
  pheat() %>%
  pheat_display_main(display_numbers = significance_matrix) %>%
  pheat_cluster_rows() %>%
  pheat_cluster_cols() %>%
  pheat_save(filename = here("Output", heatmap_name))
```

Heatmap saved as `r heatmap_name`.

\newpage



# END

```{r Generate time stamped report, eval=FALSE, include=FALSE}

rstudioapi::documentSaveAll()

script_dir <- file.path("Scripts","Rmds")
output_dir <- file.path("Output")

report_fp <- here::here(script_dir, "Shotgun_metabolomic_analysis.Rmd")
output_fp <- here::here(output_dir, paste('Sagori_growth_metabolomics_', Sys.Date(), '.pdf', sep=''))

rmarkdown::render(input = report_fp, output_file = output_fp)

```

```{r save data}

save.image(file = here("Data","sagori_metabolomics_data.RData"))

```