---
title: |
  ![](CHOPMC_logo.pdf){width=5in}
  \break
  Sagori Mukhopadhyay - Runs 1 to 5 - Growth Analysis
author: "Scott G. Daniel, PennCHOP Microbiome Program"
date: "`r Sys.Date()`"
geometry: margin=3cm
output: 
    pdf_document:
        keep_tex: false
        toc: true
        toc_depth: 4
---

```{r setup, echo=FALSE}
### ================
###   knitr setup
### ================
library(knitr)
opts_chunk$set(
  tidy=FALSE,
  cache=FALSE,
  echo=FALSE,
  warning=FALSE,
  message=FALSE,
  dpi=100,
  fig.width=8,
  fig.height=8,
  fig.align = "center"
  )
```

```{r libs and funcs, message=FALSE, warning=FALSE}

library(here)
library(tidyverse)
library(qiimer)
library(vegan)
library(ape)
library(usedist)

## Visualization packages
library(pander)
library(kableExtra)

library(pheatbuilder)
library(ggbeeswarm)
library(ggsci)
library(viridis)
library(RColorBrewer)

# stats packages
library(adonisplus)
library(nlme)
library(emmeans) # for lmer post-hoc tests
library(broom)
library(ghibli)
library(glue)
library(magrittr)
select <- dplyr::select
library(patchwork)
library(qualpalr)

```

```{r load data}

load(file = here("Data","shotgun_data_2026-04-28.RData"))

table_export <- table_iterator()

```

# Run overview

## Taxonomic heatmap

```{r}
prop_cut <- 0.001
```

Taxonomic assignments were performed using the Kraken program.

Heatmap charts were generated from the taxonomic assignments. Each column represents one sample and each row represents one taxon (typically a species). Ranks are included in the plot if the taxon is present in `r 100*prop_cut`% mean abundance in at least one sample type.

The chart is colored white if species were not observed in the sample, dark blue if species were observed at very low abundance.  This allows the reader to quickly survey species presence/absence.  Abundance values exceeding 40% are colored red, indicating an extremely dominant species.


```{r}
s_toPlot <- s %>%
  filter(kraken_counts > 0 | !is.na(kraken_counts)) %>%
  select(SampleID, which_run, category) %>%
  arrange(which_run, category) %>%
  droplevels()

# select taxa with mean relative abundance of 1% in at least one sample type
select_taxa <- summed_props %>%
  as.data.frame() %>% 
  rownames_to_column("Taxa") %>% 
  pivot_longer(-Taxa, names_to="SampleID", values_to="props") %>%
  right_join(s_toPlot, by="SampleID")  %>%
  group_by(Taxa) %>%
  mutate(mean_prop = mean(props)) %>%
  ungroup() %>%
  group_by(Taxa) %>%
  mutate(mean_prop = max(mean_prop)) %>%
  ungroup() %>%
  filter(mean_prop > prop_cut) %>%
  pull(Taxa) %>%
  as.character() %>%
  unique()

heatmap_fp <- "taxonomic_heatmap.pdf"

props_toPlot <- summed_props[select_taxa, s_toPlot$SampleID]
props_toPlot %>%
  pheat() %>%
  pheat_color_saturated() %>%
  pheat_cluster_rows() %>%
  pheat_annotate_cols(s_toPlot) %>%
  pheat_display_cols(gaps = factor_gaps(s_toPlot$category)) %>%
  pheat_save(filename = here("Output", "taxonomic_heatmap.pdf"))

```

Heatmap saved as `r heatmap_fp`.

# Week of life analysis

Table of samples across timepoints and birth categories:

```{r table and set colors, results='asis'}
## Subset the metadata file to only keep the samples you would like to test on
growth_failure <- read_csv(here("Data", "growth_failure.csv")) %>%
  select(REDCapID, GF36_10, GF36_3rd) %>%
  mutate(REDCapID = as.character(REDCapID)) %>%
  mutate(across(GF36_10:GF36_3rd, as.factor)) %>%
  unique()

abx_use <- read_tsv(here("Data", "AbxHoursPerHOL.tsv"))

diet_data <- read_csv(here("Data", "2023.12.20 Feed Data_per stool.csv")) %>%
  mutate(redcap_id = as.character(redcap_id))

s2 <- s %>%
  filter(Keep, !isControl, !is.na(TimePoint), !is.na(growth_cat)) %>%
  mutate(category = factor(category)) %>%
  mutate(growth_cat = case_when(Zdec36_gt0.8 == 1 & Zdec36_gt1.2 == 0 & Zdec36_gt2 == 0 ~ "mild retardation",
                                 Zdec36_gt0.8 == 0 & (Zdec36_gt1.2 == 1 | Zdec36_gt2 == 1) ~ "moderate to severe retardation",
                                 (Zdec36_gt0.8 + Zdec36_gt1.2 + Zdec36_gt2) == 0 ~ "normal")) %>%
  mutate(growth_cat = fct_relevel(growth_cat, "normal", "mild retardation")) %>%
  mutate(TimePoint = fct_relevel(TimePoint, "T0", "T1", "T2", "T3", "T4", "T5", "T6", "T7", "T8", "T9")) %>%
  left_join(growth_failure, by = c("redcap_id" = "REDCapID")) %>%
  left_join(diet_data, by = c("SampleID" = "sampleid", "redcap_id")) %>%
  left_join(abx_use, by = "SampleID") %>%
  mutate(AbxHoursPerHOL = ifelse(is.na(AbxHoursPerHOL), 0, AbxHoursPerHOL)) %>%
  mutate(growth_failure = fct_recode(GF36_3rd, Yes = "1", No = "0")) %>%
  mutate(npo_days_prop = npo_days_n / DOL) %>%
  mutate(npo_days_prop = ifelse(npo_days_prop == Inf, 1, npo_days_prop)) %>%
  droplevels()

## Set colors for each factor ahead of time so they are consistent through the report.
ann_colors = list(
  category = setNames(brewer.pal(length(levels(s2$category)), "Set2"), levels(s2$category)),
  growth_cat = setNames(as.character(ghibli_palette("YesterdayMedium", direction = -1)[1:length(levels(s2$growth_cat))]), levels(s2$growth_cat)),
  TimePoint = setNames(viridis(length(levels(s2$TimePoint))), levels(s2$TimePoint)),
  growth_fail = setNames(qualpal(2)$hex, levels(s2$growth_failure))
)

## Change this to be the breakdown of whatever variable the collaborator is interested in
addmargins(table(s2$category, s2$TimePoint, useNA = "ifany")) %>%
  pander(split.table=Inf)
```

```{r fixing stool HOL and npo_days_prop}

fixed_HOL <- read_csv(here("Data", "2025.2.7 Stool HOL and NPO Days Proportion.csv")) %>%
  rename(SampleID = sampleid, HOL = stool_hol) %>%
  mutate(redcap_id = as.character(redcap_id))

s2 %<>%
  select(-HOL, -npo_days_prop) %>%
  left_join(fixed_HOL, join_by(SampleID, redcap_id))

s2 %<>%
  mutate(DOL = HOL / 24)

s_toPlot <- s_toTest <- s2

```

```{r exploration, eval=FALSE, include=FALSE}
s_toPlot %>% select(SampleID, redcap_id, npo_days_n, DOL, npo_days_prop, feeds_prop_hm, AbxHoursPerHOL) %>% arrange(redcap_id, DOL) %>% view()

cor(s_toPlot[,c("npo_days_prop","AbxHoursPerHOL")], use="pairwise")

cor.test(formula = ~ npo_days_prop + AbxHoursPerHOL, data = s_toPlot)

summary(lm(formula = npo_days_prop ~ AbxHoursPerHOL, data = s_toPlot))

```

Samples per DOL, colored by TimePoint:

```{r changing to week of life}

s_toPlot %>%
  ggplot(aes(x = DOL, fill = TimePoint)) +
  geom_histogram(binwidth = 1, center = 0)

s_toPlot %>%
  group_by(TimePoint) %>%
  summarise(min_DOL = min(DOL), max_DOL = max(DOL)) %>%
  pander(caption = "Min and max DOL per TimePoint")

cut_timepoints <- function(DOL) {
  breakpoints <- c(0,6,13,20,27,34,41,48,55,62,69,76,83,90,97,104,111)
  my_labels <- c(1,2,3,4,5,6,7,8,9,10,11,12,13,14,15,16)
  cut(DOL, breaks = breakpoints, labels = my_labels, include.lowest = T, right = T)
}

s2 <- s_toPlot %>%
  mutate(WOL_label = cut_timepoints(DOL)) %>%
  mutate(WOL = as.numeric(WOL_label)) %>%
  mutate(timepoint_numeric = as.numeric(TimePoint))

s2 %>%
  group_by(WOL) %>%
  summarise(min_DOL = min(DOL), max_DOL = max(DOL)) %>%
  pander(caption = "Weeks of life")

```

Samples per week of life:

```{r}
s2 %>%
  ggplot(aes(x = DOL, fill = WOL_label)) +
  geom_histogram(binwidth = 1, center = 0)
```

## Differential abundance

```{r}
s_toPlot <- s_toTest <- s2

prop_cut = 0.01
```


We filtered down to bacteria that have a mean relative abundance of >`r 100*prop_cut`% across samples. We also filtered *Cutibacterium acnes* since it is a known contaminant. Only these bacteria have been tested in this report.

```{r replace zeros and filter}

# replace zeros per Taxa
max_num <- max(summed_props)
summed_props_long <- summed_props %>%
  as.data.frame() %>% 
  rownames_to_column("Taxa") %>% 
  pivot_longer(-Taxa, names_to="SampleID", values_to="props") %>%
  group_by(Taxa) %>%
  mutate(Taxa_min = ifelse(props == 0, max_num, props)) %>%
  mutate(props = ifelse(props == 0, min(Taxa_min)/2, props)) %>%
  ungroup() %>%
  select(-Taxa_min) %>%
  droplevels()

props_toTest <- summed_props_long %>%
  group_by(Taxa) %>%
  mutate(mean_prop = mean(props)) %>%
  ungroup() %>%
  filter(mean_prop > prop_cut) %>%
  filter(str_detect(Taxa, "Bacteria|acnes", negate = T)) %>%
  mutate(log10_props = log10(props))

```

The top taxa are:

```{r, results='asis'}

props_toTest %>%
  select(Taxa, mean_prop) %>%
  distinct() %>%
  arrange(desc(mean_prop)) %>%
  pander(digits = 2, split.table = Inf)

```



### Boxplots

#### By TimePoint

```{r fig.height=12, fig.width=8}
props_toTest %>%
  right_join(s_toTest, by="SampleID")  %>%
  mutate(Taxa = reorder(Taxa, -props)) %>%
  mutate(Taxa = fct_relabel(Taxa, function(x) gsub(" ", "\n", x))) %>%
  
  ggplot(aes(x=TimePoint, y=props, color=TimePoint)) +
    geom_boxplot(outlier.alpha=0) +
    #geom_quasirandom() +
    facet_wrap(~Taxa, scales="free", ncol = 4) +
    scale_color_manual(values=ann_colors$TimePoint) +
    scale_y_continuous(trans = "log10") +
    theme_bw() +
    theme(
      axis.text.x=element_text(angle=90, hjust=1, vjust=0.5),
      panel.grid = element_blank(),
      strip.background = element_blank(),
      legend.position = "bottom"
    ) +
    #guides(color=F) +
    labs(
      y="Relative abundance"
    )
```

#### By Growth category

```{r fig.height=12, fig.width=8}
props_toTest %>%
  right_join(s_toTest, by="SampleID")  %>%
  mutate(Taxa = reorder(Taxa, -props)) %>%
  mutate(Taxa = fct_relabel(Taxa, function(x) gsub(" ", "\n", x))) %>%
  mutate(growth_cat = fct_relabel(growth_cat, function(x) gsub(" ", "\n", x))) %>%
  
  ggplot(aes(x=growth_cat, y=props, color=TimePoint)) +
    # geom_boxplot(outlier.alpha=0) +#
    geom_quasirandom(dodge.width = 0.75) +
    facet_wrap(~Taxa, scales="free", ncol = 4) +
    scale_color_manual(values=ann_colors$TimePoint) +
    scale_y_continuous(trans = "log10") +
    theme_bw() +
    theme(
      axis.text.x=element_text(angle=45, hjust=1, vjust=1),
      panel.grid = element_blank(),
      strip.background = element_blank(),
      legend.position = "bottom"
    ) +
    #guides(color=F) +
    labs(
      y="Relative abundance"
    )
```


### Linear mixed-effects models

Linear models were used to estimate the change in relative abundance of select taxa between study groups. The relative abundances were log10 transformed. Multiple tests were adjusted for false discovery rate using Benjamini-Hochberg method. Only the terms with p<0.05 are shown in the table below. The value is the change in the log10-transformed relative abundance per unit time (or the change in category).

`AbxHoursPerHOL` is defined as the (Abx end time - Abx start time) / Hours of life for a given sample. Thus, samples immediately after a lengthy course of Abx will have a high `AbxHoursPerHOL` and this number will decrease the farther out a sample is from a course of Abx. Multiple courses of Abx are cumulative but the gaps between courses are not summed.

AbxHoursPerHOL has been taken out because it is collinear with npo_days_prop.

```{r, results='asis'}

s_toTest %<>%
  mutate(npo_days_prop_min = min(npo_days_prop[npo_days_prop > 0])) %>%
  mutate(npo_days_prop = ifelse(npo_days_prop == 0, npo_days_prop_min/2, npo_days_prop)) %>%
  mutate(log_npo_days_prop = log(npo_days_prop)) %>%
  mutate(AbxHoursPerHOL_min = min(AbxHoursPerHOL[AbxHoursPerHOL > 0])) %>%
  mutate(AbxHoursPerHOL = ifelse(AbxHoursPerHOL == 0, AbxHoursPerHOL_min/2, AbxHoursPerHOL)) %>%
  mutate(log_AbxHoursPerHOL = log(AbxHoursPerHOL))

my_formula <- "log10_props ~ bwt_z + log_npo_days_prop + WOL * growth_cat"

summaries_df <- run_lmer_on_props(factor_to_test = "Taxa", props_toTest = props_toTest, s_toTest = s_toTest, form1 = my_formula, rep_mes_label = "SubjectID", p_cutoff = 0.05)

table_export(summaries_df, "sagori_growth_analysis_table_")

summaries_df %>%
  mutate(term = str_replace(term, "WOL", "Time")) %>%
  mutate(term = if_else(str_detect(term, "growth_cat"), str_remove(term, "growth_cat"), term)) %>%
  pander(digits = 2, split.table = Inf, split.cells = 20, caption = glue("Linear model of the log10-transformed relative abundance with numeric time and growth categories as the predictive variables. SubjectID added as a random effect. Full formula: {my_formula}"), emphasize.strong.rows = which(.$fdr<=0.05))

```

#### Bonus plot

```{r}

props_toTest %>%
  right_join(s_toTest, by="SampleID")  %>%
  mutate(Taxa = reorder(Taxa, -props)) %>%
  mutate(Taxa = fct_relabel(Taxa, function(x) gsub(" ", "\n", x))) %>%
  
  ggplot(aes(x=AbxHoursPerHOL, y=props)) +
  geom_point(shape = 1, alpha = 0.5) +
  geom_smooth(method = "lm") +
  facet_wrap(~Taxa, scales="free", ncol = 4) +
  scale_y_continuous(trans = "log10") +
  theme_bw() +
  theme(
    axis.text.x=element_text(angle=90, hjust=1, vjust=0.5),
    panel.grid = element_blank(),
    strip.background = element_blank(),
    legend.position = "bottom"
  ) +
  labs(
    y="Relative abundance"
  )

```

## A. baumannii abundances

```{r}

a_baumannii <- summed_props_long %>%
  filter(str_detect(Taxa, "baumannii"))

a_baumannii %>%
  right_join(s_toPlot, by = "SampleID") %>%
  write_csv(here("Output", "a_baumannii_abundances.csv"))

```


### Boxplots

#### By TimePoint

```{r fig.height=12, fig.width=8}
a_baumannii %>%
  right_join(s_toTest, by="SampleID")  %>%
  mutate(Taxa = reorder(Taxa, -props)) %>%
  mutate(Taxa = fct_relabel(Taxa, function(x) gsub(" ", "\n", x))) %>%
  
  ggplot(aes(x=TimePoint, y=props, color=TimePoint)) +
  geom_boxplot(outlier.alpha=0) +
  geom_quasirandom() +
  geom_hline(yintercept = 4e-08, lty = "dashed") +
  scale_color_manual(values=ann_colors$TimePoint) +
  scale_y_continuous(trans = "log10") +
  theme_bw() +
  theme(
    axis.text.x=element_text(angle=90, hjust=1, vjust=0.5),
    panel.grid = element_blank(),
    strip.background = element_blank(),
    legend.position = "bottom"
  ) +
  labs(
    y="Relative abundance"
  )

ggsave(here("Output", "a_baumannii_abundance_by_timepoint.pdf"), width = 5, height = 5)
```

#### By Growth category

```{r fig.height=12, fig.width=8}
a_baumannii %>%
  right_join(s_toTest, by="SampleID")  %>%
  mutate(Taxa = reorder(Taxa, -props)) %>%
  mutate(Taxa = fct_relabel(Taxa, function(x) gsub(" ", "\n", x))) %>%
  mutate(growth_cat = fct_relabel(growth_cat, function(x) gsub(" ", "\n", x))) %>%
  
  ggplot(aes(x=growth_cat, y=props, color=TimePoint)) +
    geom_quasirandom(dodge.width = 0.75) +
    geom_hline(yintercept = 4e-08, lty = "dashed") +
    scale_color_manual(values=ann_colors$TimePoint) +
    scale_y_continuous(trans = "log10") +
    theme_bw() +
    theme(
      axis.text.x=element_text(angle=45, hjust=1, vjust=1),
      panel.grid = element_blank(),
      strip.background = element_blank(),
      legend.position = "bottom"
    ) +
    labs(
      y="Relative abundance"
    )

ggsave(here("Output", "a_baumannii_abundance_by_growth_category.pdf"), width = 5, height = 5)
```



## Beta diversity

Here, we use Bray-Curtis and Jaccard distances to compare the species composition of the samples to each other.

The plots below show the distance between each pair of samples in a single 2D plot.  It is not possible to plot the distances exactly on paper, so we have used a method of ordination called Principal Coordinates Analysis to select the best coordinate system for display.  The percentage of total variance captured along each axis is displayed on the chart.

### Plots

#### By TimePoint


```{r fig.height=3, fig.width=5}
s_toTest %>%
  pcoaplus(bc) %>%
  plot(color=TimePoint) +
    scale_color_manual(values=ann_colors$TimePoint) +
    theme_bw() +
    theme(
      axis.text=element_blank(),
      axis.ticks=element_blank(),
      panel.grid = element_blank(), 
      aspect.ratio = 1,
      strip.background = element_blank()
    ) + 
    labs(color="", title="Bray-Curtis")

s_toTest %>%
  pcoaplus(jd) %>%
  plot(color=TimePoint) +
    scale_color_manual(values=ann_colors$TimePoint) +
    theme_bw() +
    theme(
      axis.text=element_blank(),
      axis.ticks=element_blank(),
      panel.grid = element_blank(), 
      aspect.ratio = 1,
      strip.background = element_blank()
    ) + 
    labs(color="", title="Jaccard")
```

#### By Growth category

```{r fig.height=3, fig.width=5}
s_toTest %>%
  pcoaplus(bc) %>%
  plot(color=growth_cat) +
    scale_color_manual(values=ann_colors$growth_cat) +
    theme_bw() +
    theme(
      axis.text=element_blank(),
      axis.ticks=element_blank(),
      panel.grid = element_blank(), 
      aspect.ratio = 1,
      strip.background = element_blank()
    ) + 
    labs(color="", title="Bray-Curtis")

s_toTest %>%
  pcoaplus(jd) %>%
  plot(color=growth_cat) +
    scale_color_manual(values=ann_colors$growth_cat) +
    theme_bw() +
    theme(
      axis.text=element_blank(),
      axis.ticks=element_blank(),
      panel.grid = element_blank(), 
      aspect.ratio = 1,
      strip.background = element_blank()
    ) + 
    labs(color="", title="Jaccard")
```

### PERMANOVA

PERMANOVA test on Bray-Curtis and Jaccard distances to test if the centroids of the study groups can be distinguished from each other.

```{r, results='asis'}

s_permanova <- s_toTest %>%
  filter(!is.na(log_npo_days_prop))

summaries_df <- rbind(
  s_permanova %>%
    do(adonisplus(
      ., distmat = bc, formula = distmat ~ bwt_z + log_npo_days_prop + WOL * growth_cat,
      sample_id_var = SampleID, rep_meas_var = SubjectID,
      shuffle = c(growth_cat = "between", WOL = "within"), permutations=perm)) %>%
    ungroup() %>%
    mutate(metric = "Bray-Curtis"),

  s_permanova %>%
    do(adonisplus(
      ., distmat = jd, formula = distmat ~ bwt_z + log_npo_days_prop + WOL * growth_cat,
      sample_id_var = SampleID, rep_meas_var = SubjectID,
      shuffle = c(growth_cat = "between", WOL = "within"), permutations=perm)) %>%
    ungroup() %>%
    mutate(metric = "Jaccard")
) %>%
  filter(!term %in% c("Residuals", "Total")) %>%
  select(-one_of(c("sumsq", "meansq"))) %>%
  select(metric, everything())

write_rds(summaries_df, here("Data", "bcjd_permanova_results.rds"))

summaries_df <- read_rds(here("Data", "bcjd_permanova_results.rds"))

# If you don't need the shuffling functionality, you can use
# summaries_df <- adonisplus(s_toTest, wu, distmat ~ TimePoint)

summaries_df %>%
  mutate(term = str_replace(term, "timepoint_numeric", "Time")) %>%
  mutate(term = str_replace(term, "growth_cat", "Growth category")) %>%
  pander(emphasize.strong.rows = which(.$p.value<=0.05))

table_export(summaries_df, "sagori_growth_analysis_table_")

```

## Alpha diversity

Alpha diversity was assessed by the expected number of observed OTUs (out of rarefying sample size of `r format(richness_subsample_size, big.mark = ",", scientific = F)`) and Shannon index.

### Plots

#### By TimePoint

```{r fig.width=8}
s_toTest %>%
  pivot_longer(c("Richness", "Shannon"), names_to="metric", values_to="alpha") %>%

  ggplot(aes(x=TimePoint, y=alpha, color=TimePoint)) +
    geom_boxplot(outlier.alpha=0) +
    geom_quasirandom() +
    facet_grid(metric~., scales = "free") +
    scale_color_manual(values=ann_colors$TimePoint) +
    theme_bw() +
    theme(
      strip.background = element_blank(),
      panel.grid=element_blank(),
      legend.position = "none"
    ) +
    labs(
      x="", color="",
      y="Alpha diversity value"
    )
```

#### By Growth category

```{r fig.width=8}
s_toTest %>%
  pivot_longer(c("Richness", "Shannon"), names_to="metric", values_to="alpha") %>%
  ggplot(aes(x=growth_cat, y=alpha, color=TimePoint)) +
    geom_quasirandom(dodge.width = 0.75) +
    facet_grid(metric~., scales = "free") +
    scale_color_manual(values=ann_colors$TimePoint) +
    theme_bw() +
    theme(
      strip.background = element_blank(),
      panel.grid=element_blank(),
      legend.position = "bottom"
    ) +
    labs(
      x="Growth category", color="Timepoint",
      y="Alpha diversity value"
    )
```

### Linear mixed-effects model

Linear models were used to estimate the difference between timepoints.

```{r, results='asis'}

my_formula <- "alpha ~ bwt_z + log_npo_days_prop + WOL * growth_cat"

summaries_df <- s_toTest %>%
  pivot_longer(c("Richness", "Shannon"), names_to="metric", values_to="alpha") %>%
  group_by(metric) %>%
  do(tidy_lmer(nlme::lme(as.formula(my_formula), random=~1|SubjectID, data=., na.action=na.omit))) %>%
  ungroup() %>%
  filter(!grepl("Intercept", term)) %>%
  mutate(term = str_replace(term, "WOL", "Time")) %>%
  mutate(term = if_else(str_detect(term, "growth_cat"), str_remove(term, "growth_cat"), term))

summaries_df %>%
  pander(digits = 2, split.table = Inf, split.cells = 20, caption = glue("Linear model of the alpha diversity measurement with numeric time and growth categories as the predictive variables. SubjectID added as a random effect. Full formula: {my_formula}"), emphasize.strong.rows = which(.$p.value<=0.05))

```

```{r saving for random forest}

s2 <- s_toTest

save(s2, summed_props_long, summed_cts, summed_cts_g, summed_props, summed_props_g, growth_data, growth_failure, diet_data, abx_use, file = here("Data", "sagori_taxonomic_data.RData"))

```

# 0 to 21 day old babies

## Growth faltering

After filtering for > 0 days and < 22 days we are left with these samples per baby:

```{r fig.height=6, fig.width=8}

s_toPlot <- s_toTest <- s2 %>%
  filter(DOL < 22) %>%
  droplevels()

#subsetting timepoint colors for nicer legends
ann_colors$TimePoint <- ann_colors$TimePoint[names(ann_colors$TimePoint) %in% unique(s_toPlot$TimePoint)]

my_df <- s_toPlot %>%
  count(SubjectID) %>%
  arrange(desc(n)) %>%
  rename(`Number of samples` = n)

my_df %>%
  ggplot(aes(x = SubjectID, y = `Number of samples`, fill = `Number of samples`)) +
  geom_col() +
  theme_bw() +
  scale_fill_viridis_c() +
  theme(axis.text.x = element_text(angle = 90, hjust = 1, vjust = 1))

glue("With a total of {length(unique(s_toPlot$SubjectID))} babies.")

```

### Differential abundance

```{r}

prop_cut = 0.01

props_toTest <- summed_props_long %>%
  filter(SampleID %in% s_toPlot$SampleID) %>%
  group_by(Taxa) %>%
  mutate(mean_prop = mean(props)) %>%
  ungroup() %>%
  filter(mean_prop > prop_cut) %>%
  filter(str_detect(Taxa, "Bacteria|acnes", negate = T)) %>%
  mutate(log10_props = log10(props))

```

We filtered down to bacteria that have a mean relative abundance of >`r 100*prop_cut`% across these samples. We also filtered *Cutibacterium acnes* since it is a known contaminant. Only these bacteria have been tested in this subset.

The top taxa are:

```{r, results='asis'}

props_toTest %>%
  select(Taxa, mean_prop) %>%
  distinct() %>%
  arrange(desc(mean_prop)) %>%
  pander(digits = 2, split.table = Inf)

```

### Boxplots

#### By TimePoint

```{r fig.height=12, fig.width=8}
props_toTest %>%
  right_join(s_toTest, by="SampleID")  %>%
  mutate(Taxa = reorder(Taxa, -props)) %>%
  mutate(Taxa = fct_relabel(Taxa, function(x) gsub(" ", "\n", x))) %>%
  
  ggplot(aes(x=TimePoint, y=props, color=TimePoint)) +
    geom_boxplot(outlier.alpha=0) +
    #geom_quasirandom() +
    facet_wrap(~Taxa, scales="free", ncol = 4) +
    scale_color_manual(values=ann_colors$TimePoint) +
    scale_y_continuous(trans = "log10") +
    theme_bw() +
    theme(
      axis.text.x=element_text(angle=90, hjust=1, vjust=0.5),
      panel.grid = element_blank(),
      strip.background = element_blank(),
      legend.position = "bottom"
    ) +
    #guides(color=F) +
    labs(
      y="Relative abundance"
    )
```

#### By Growth category

```{r fig.height=12, fig.width=8}
props_toTest %>%
  right_join(s_toTest, by="SampleID")  %>%
  mutate(Taxa = reorder(Taxa, -props)) %>%
  mutate(Taxa = fct_relabel(Taxa, function(x) gsub(" ", "\n", x))) %>%
  mutate(growth_cat = fct_relabel(growth_cat, function(x) gsub(" ", "\n", x))) %>%
  
  ggplot(aes(x=growth_cat, y=props, color=TimePoint)) +
    # geom_boxplot(outlier.alpha=0) +#
    geom_quasirandom(dodge.width = 0.75) +
    facet_wrap(~Taxa, scales="free", ncol = 4) +
    scale_color_manual(values=ann_colors$TimePoint) +
    scale_y_continuous(trans = "log10") +
    theme_bw() +
    theme(
      axis.text.x=element_text(angle=45, hjust=1, vjust=1),
      panel.grid = element_blank(),
      strip.background = element_blank(),
      legend.position = "bottom"
    ) +
    #guides(color=F) +
    labs(
      y="Relative abundance"
    )

ggsave(filename = here("Output", "subset0to21daysold_taxa_abundance.pdf"), width = 10, height = 8, useDingbats = F)

```


#### Linear models

Linear models were used to estimate the change in relative abundance of select taxa between study groups. The relative abundances were log10 transformed. Multiple tests were adjusted for false discovery rate using Benjamini-Hochberg method. Only the terms with p<0.05 are shown in the table below. The value is the change in the log10-transformed relative abundance per unit time (or the change in category).


```{r, results='asis'}

my_formula <- "log10_props ~ bwt_z + log_npo_days_prop + WOL * growth_cat"

summaries_df <- run_lmer_on_props(factor_to_test = "Taxa", props_toTest = props_toTest, s_toTest = s_toTest, form1 = my_formula, p_cutoff = 0.05, rep_mes_label = "SubjectID")

table_export(summaries_df, "sagori_growth_analysis_table_")

summaries_df %>%
  mutate(term = str_replace(term, "timepoint_numeric", "Time")) %>%
  mutate(term = if_else(str_detect(term, "growth_cat"), str_remove(term, "growth_cat"), term)) %>%
  pander(digits = 2, split.table = Inf, split.cells = 20, caption = glue("Linear model of the log10-transformed relative abundance with numeric time and growth categories as the predictive variables. Full formula: {my_formula}"), emphasize.strong.rows = which(.$fdr<=0.05))

```

The ones that were closest to changing:

```{r}

to_select <- unique(summaries_df$Taxa)

props_toTest %>%
  filter(Taxa %in% to_select) %>%
  right_join(s_toTest, by="SampleID")  %>%
  mutate(Taxa = reorder(Taxa, -props)) %>%
  mutate(Taxa = fct_relabel(Taxa, function(x) gsub(" ", "\n", x))) %>%
  mutate(growth_cat = fct_relabel(growth_cat, function(x) gsub(" ", "\n", x))) %>%
  
  ggplot(aes(x=growth_cat, y=props, color=TimePoint)) +
    # geom_boxplot(outlier.alpha=0) +#
    geom_quasirandom(dodge.width = 0.75) +
    facet_wrap(~Taxa, scales="free", ncol = 4) +
    scale_color_manual(values=ann_colors$TimePoint) +
    scale_y_continuous(trans = "log10") +
    theme_bw() +
    theme(
      axis.text.x=element_text(angle=45, hjust=1, vjust=1),
      panel.grid = element_blank(),
      strip.background = element_blank(),
      legend.position = "bottom"
    ) +
    #guides(color=F) +
    labs(
      y="Relative abundance"
    )

ggsave(filename = here("Output", "subset0to21daysold_taxa_abundance_changed.pdf"), width = 7, height = 5, useDingbats = F)

```


### Beta diversity

Here, we use Bray-Curtis and Jaccard distances to compare the species composition of the samples to each other.

The plots below show the distance between each pair of samples in a single 2D plot.  It is not possible to plot the distances exactly on paper, so we have used a method of ordination called Principal Coordinates Analysis to select the best coordinate system for display.  The percentage of total variance captured along each axis is displayed on the chart.

#### Plots

#### By TimePoint


```{r fig.height=3, fig.width=5}
s_toTest %>%
  pcoaplus(bc) %>%
  plot(color=TimePoint) +
    scale_color_manual(values=ann_colors$TimePoint) +
    theme_bw() +
    theme(
      axis.text=element_blank(),
      axis.ticks=element_blank(),
      panel.grid = element_blank(), 
      aspect.ratio = 1,
      strip.background = element_blank()
    ) + 
    labs(color="", title="Bray-Curtis")

s_toTest %>%
  pcoaplus(jd) %>%
  plot(color=TimePoint) +
    scale_color_manual(values=ann_colors$TimePoint) +
    theme_bw() +
    theme(
      axis.text=element_blank(),
      axis.ticks=element_blank(),
      panel.grid = element_blank(), 
      aspect.ratio = 1,
      strip.background = element_blank()
    ) + 
    labs(color="", title="Jaccard")
```

#### By Growth category

```{r fig.height=3, fig.width=5}
s_toTest %>%
  pcoaplus(bc) %>%
  plot(color=growth_cat) +
    scale_color_manual(values=ann_colors$growth_cat) +
    theme_bw() +
    theme(
      axis.text=element_blank(),
      axis.ticks=element_blank(),
      panel.grid = element_blank(), 
      aspect.ratio = 1,
      strip.background = element_blank()
    ) + 
    labs(color="", title="Bray-Curtis")

ggsave(here("Output", "bray_curtis_growth_0to21daysold.pdf"), width = 5, height = 5)

s_toTest %>%
  pcoaplus(jd) %>%
  plot(color=growth_cat) +
    scale_color_manual(values=ann_colors$growth_cat) +
    theme_bw() +
    theme(
      axis.text=element_blank(),
      axis.ticks=element_blank(),
      panel.grid = element_blank(), 
      aspect.ratio = 1,
      strip.background = element_blank()
    ) + 
    labs(color="", title="Jaccard")

ggsave(here("Output", "jaccard_growth_0to21daysold.pdf"), width = 5, height = 5)

```

#### PERMANOVA

PERMANOVA test on Bray-Curtis and Jaccard distances to test if the centroids of the study groups can be distinguished from each other.

```{r, results='asis'}

s_permanova <- s_toTest %>%
  filter(!is.na(log_npo_days_prop))

summaries_df <- rbind(
  s_permanova %>%
    do(adonisplus(
      ., distmat = bc, formula = distmat ~ bwt_z + log_npo_days_prop + WOL * growth_cat,
      sample_id_var = SampleID, rep_meas_var = SubjectID,
      shuffle = c(growth_cat = "between", WOL = "within"), permutations=perm)) %>%
    ungroup() %>%
    mutate(metric = "Bray-Curtis"),
  
  s_permanova %>%
    do(adonisplus(
      ., distmat = jd, formula = distmat ~ bwt_z + log_npo_days_prop + WOL * growth_cat,
      sample_id_var = SampleID, rep_meas_var = SubjectID,
      shuffle = c(growth_cat = "between", WOL = "within"), permutations=perm)) %>%
    ungroup() %>%
    mutate(metric = "Jaccard")
) %>%
  filter(!term %in% c("Residuals", "Total")) %>%
  select(-one_of(c("sumsq", "meansq"))) %>%
  select(metric, everything())


# If you don't need the shuffling functionality, you can use
# summaries_df <- adonisplus(s_toTest, wu, distmat ~ TimePoint)

summaries_df %>%
  mutate(term = str_replace(term, "timepoint_numeric", "Time")) %>%
  mutate(term = str_replace(term, "growth_cat", "Growth category")) %>%
  pander(emphasize.strong.rows = which(.$p.value<=0.05))

table_export(summaries_df, "sagori_growth_analysis_table_")

```

### Alpha diversity

Alpha diversity was assessed by the expected number of observed OTUs (out of rarefying sample size of `r format(richness_subsample_size, big.mark = ",", scientific = F)`) and Shannon index.

#### Plots

#### By TimePoint

```{r fig.width=8}
s_toTest %>%
  pivot_longer(c("Richness", "Shannon"), names_to="metric", values_to="alpha") %>%

  ggplot(aes(x=TimePoint, y=alpha, color=TimePoint)) +
    geom_boxplot(outlier.alpha=0) +
    geom_quasirandom() +
    facet_grid(metric~., scales = "free") +
    scale_color_manual(values=ann_colors$TimePoint) +
    theme_bw() +
    theme(
      strip.background = element_blank(),
      panel.grid=element_blank(),
      legend.position = "none"
    ) +
    labs(
      x="", color="",
      y="Alpha diversity value"
    )
```

#### By Growth category

```{r fig.width=8}
s_toTest %>%
  pivot_longer(c("Richness", "Shannon"), names_to="metric", values_to="alpha") %>%
  mutate(growth_cat = fct_relabel(growth_cat, function(x) gsub(" ", "\n", x))) %>%
  ggplot(aes(x=growth_cat, y=alpha, color=TimePoint)) +
    geom_quasirandom(dodge.width = 0.75) +
    facet_grid(metric~., scales = "free") +
    scale_color_manual(values=ann_colors$TimePoint) +
    theme_bw() +
    theme(
      axis.text.x=element_text(angle=45, hjust=1, vjust=1),
      strip.background = element_blank(),
      panel.grid=element_blank(),
      legend.position = "bottom"
    ) +
    labs(
      x="Growth category", color="Timepoint",
      y="Alpha diversity value"
    )

ggsave(here("Output", "alpha_diversity_growth_0to21daysold.pdf"), width = 5, height = 5)

```

\newpage

#### Linear model

Linear models were used to estimate the difference between growth categories including time as a covariate.

```{r, results='asis'}

my_formula <- "alpha ~ bwt_z + log_npo_days_prop + WOL * growth_cat"
rep_mes_form <- "~ 1 | SubjectID"

summaries_df <- s_toTest %>%
  pivot_longer(c("Richness", "Shannon"), names_to="metric", values_to="alpha") %>%
  group_by(metric) %>%
  do(tidy_lmer(nlme::lme(as.formula(my_formula), random = as.formula(rep_mes_form), data=., na.action=na.omit))) %>%
  ungroup() %>%
  filter(!grepl("Intercept", term)) %>%
  mutate(term = str_replace(term, "timepoint_numeric", "Time")) %>%
  mutate(term = if_else(str_detect(term, "growth_cat"), str_remove(term, "growth_cat"), term))

summaries_df %>%
  pander(digits = 2, split.table = Inf, split.cells = 20, caption = glue("Linear model of the alpha diversity measurement with numeric time and growth categories as the predictive variables. Full formula: {my_formula}"), emphasize.strong.rows = which(.$p.value<=0.05))

```

## Growth failure

Growth failure is now those babies who are the bottom 3rd percentile in terms of normal growth.

### Differential abundance

#### Boxplots

#### By Growth failure

```{r fig.height=12, fig.width=8}
props_toTest %>%
  right_join(s_toTest, by="SampleID")  %>%
  mutate(Taxa = reorder(Taxa, -props)) %>%
  mutate(Taxa = fct_relabel(Taxa, function(x) gsub(" ", "\n", x))) %>%
  
  ggplot(aes(x=growth_failure, y=props, color=TimePoint)) +
    # geom_boxplot(outlier.alpha=0) +#
    geom_quasirandom(dodge.width = 0.75) +
    facet_wrap(~Taxa, scales="free", ncol = 4) +
    scale_color_manual(values=ann_colors$TimePoint) +
    scale_y_continuous(trans = "log10") +
    theme_bw() +
    theme(
      axis.text.x=element_text(angle=45, hjust=1, vjust=1),
      panel.grid = element_blank(),
      strip.background = element_blank(),
      legend.position = "bottom"
    ) +
    labs(
      y="Relative abundance",
      x="Growth failure at 36 weeks"
    )

ggsave(filename = here("Output", "0to21daysold_growth_failure_taxa_abundance.pdf"), width = 10, height = 8, useDingbats = F)

```


#### Linear models

Linear models were used to estimate the change in relative abundance of select taxa between study groups. The relative abundances were log10 transformed. Multiple tests were adjusted for false discovery rate using Benjamini-Hochberg method. Only the terms with p<0.05 are shown in the table below. The value is the change in the log10-transformed relative abundance per unit time (or the change in category).


```{r, results='asis'}

my_formula <- "log10_props ~ bwt_z + log_npo_days_prop + WOL * growth_failure"

summaries_df <- run_lmer_on_props(factor_to_test = "Taxa", props_toTest = props_toTest, s_toTest = s_toTest, form1 = my_formula, rep_mes_label = "SubjectID", p_cutoff = 0.05) %>%
  mutate(term = str_replace(term, "timepoint_numeric", "Time")) %>%
  mutate(term = str_replace(term, "growth_failureYes", "Had growth failure"))

table_export(summaries_df, "sagori_growth_analysis_table_")

summaries_df %>%
  pander(digits = 2, split.table = Inf, split.cells = 20, caption = glue("Linear model of the log10-transformed relative abundance with numeric time and growth failure as the predictive variables. Full formula: {my_formula}"), emphasize.strong.rows = which(.$fdr<=0.05))

```

The ones that were closest to changing:

```{r}

to_select <- unique(summaries_df$Taxa)

props_toTest %>%
  filter(Taxa %in% to_select) %>%
  right_join(s_toTest, by="SampleID")  %>%
  mutate(Taxa = reorder(Taxa, -props)) %>%
  mutate(Taxa = fct_relabel(Taxa, function(x) gsub(" ", "\n", x))) %>%
  
  ggplot(aes(x=growth_failure, y=props, color=TimePoint)) +
    # geom_boxplot(outlier.alpha=0) +#
    geom_quasirandom(dodge.width = 0.75) +
    facet_wrap(~Taxa, scales="free", ncol = 4) +
    scale_color_manual(values=ann_colors$TimePoint) +
    scale_y_continuous(trans = "log10") +
    theme_bw() +
    theme(
      axis.text.x=element_text(angle=45, hjust=1, vjust=1),
      panel.grid = element_blank(),
      strip.background = element_blank(),
      legend.position = "bottom"
    ) +
    labs(
      y="Relative abundance",
      x="Growth failure at 36 weeks"
    )

ggsave(filename = here("Output", "0to21daysold_growth_failure_taxa_abundance_changed.pdf"), width = 7, height = 5, useDingbats = F)

```

### Beta diversity

Here, we use Bray-Curtis and Jaccard distances to compare the species composition of the samples to each other.

The plots below show the distance between each pair of samples in a single 2D plot.  It is not possible to plot the distances exactly on paper, so we have used a method of ordination called Principal Coordinates Analysis to select the best coordinate system for display.  The percentage of total variance captured along each axis is displayed on the chart.

#### Plots

#### By Growth failure

```{r fig.height=3, fig.width=5}
s_toTest %>%
  pcoaplus(bc) %>%
  plot(color=growth_failure) +
    scale_color_manual(values=ann_colors$growth_fail) +
    theme_bw() +
    theme(
      axis.text=element_blank(),
      axis.ticks=element_blank(),
      panel.grid = element_blank(), 
      aspect.ratio = 1,
      strip.background = element_blank()
    ) + 
    labs(color="", title="Bray-Curtis")

ggsave(here("Output", "bray_curtis_growth_failure_0to21daysold.pdf"), width = 5, height = 5)

s_toTest %>%
  pcoaplus(jd) %>%
  plot(color=growth_failure) +
    scale_color_manual(values=ann_colors$growth_fail) +
    theme_bw() +
    theme(
      axis.text=element_blank(),
      axis.ticks=element_blank(),
      panel.grid = element_blank(), 
      aspect.ratio = 1,
      strip.background = element_blank()
    ) + 
    labs(color="", title="Jaccard")

ggsave(here("Output", "jaccard_growth_failure_0to21daysold.pdf"), width = 5, height = 5)

```

#### PERMANOVA

PERMANOVA test on Bray-Curtis and Jaccard distances to test if the centroids of the study groups can be distinguished from each other.

```{r, results='asis'}

s_permanova <- s_toTest %>%
  filter(!is.na(log_npo_days_prop))

summaries_df <- rbind(
  s_permanova %>%
    do(adonisplus(
      ., distmat = bc, formula = distmat ~ bwt_z + log_npo_days_prop + WOL * growth_cat,
      sample_id_var = SampleID, rep_meas_var = SubjectID,
      shuffle = c(growth_failure = "between", WOL = "within"), permutations=perm)) %>%
    ungroup() %>%
    mutate(metric = "Bray-Curtis"),
  
  s_permanova %>%
    do(adonisplus(
      ., distmat = jd, formula = distmat ~ bwt_z + log_npo_days_prop + WOL * growth_cat,
      sample_id_var = SampleID, rep_meas_var = SubjectID,
      shuffle = c(growth_failure = "between", WOL = "within"), permutations=perm)) %>%
    ungroup() %>%
    mutate(metric = "Jaccard")
) %>%
  filter(!term %in% c("Residuals", "Total")) %>%
  select(-one_of(c("sumsq", "meansq"))) %>%
  select(metric, everything())

summaries_df %>%
  mutate(term = str_replace(term, "timepoint_numeric", "Time")) %>%
  mutate(term = str_replace(term, "growth_failure", "Growth failure")) %>%
  pander(emphasize.strong.rows = which(.$p.value<=0.05))

table_export(summaries_df, "sagori_growth_analysis_table_")

```

### Alpha diversity

Alpha diversity was assessed by the expected number of observed OTUs (out of rarefying sample size of `r format(richness_subsample_size, big.mark = ",", scientific = F)`) and Shannon index.

#### Plots

#### By Growth failure

```{r fig.width=8}
s_toTest %>%
  pivot_longer(c("Richness", "Shannon"), names_to="metric", values_to="alpha") %>%
  ggplot(aes(x=growth_failure, y=alpha, color=TimePoint)) +
    geom_quasirandom(dodge.width = 0.75) +
    facet_grid(metric~., scales = "free") +
    scale_color_manual(values=ann_colors$TimePoint) +
    theme_bw() +
    theme(
      axis.text.x=element_text(angle=45, hjust=1, vjust=1),
      strip.background = element_blank(),
      panel.grid=element_blank(),
      legend.position = "bottom"
    ) +
    labs(
      x="Growth failure", color="Timepoint",
      y="Alpha diversity value"
    )

ggsave(here("Output", "alpha_diversity_growth_failure_0to21daysold.pdf"), width = 5, height = 5)

```

\newpage

#### Linear model

Linear models were used to estimate the difference between growth categories including time as a covariate.

```{r, results='asis'}

my_formula <- "alpha ~ bwt_z + log_npo_days_prop + WOL * growth_failure"
rep_mes_form <- "~ 1 | SubjectID"

summaries_df <- s_toTest %>%
  pivot_longer(c("Richness", "Shannon"), names_to="metric", values_to="alpha") %>%
  group_by(metric) %>%
  do(tidy_lmer(nlme::lme(as.formula(my_formula), random = as.formula(rep_mes_form), data=., na.action=na.omit))) %>%
  ungroup() %>%
  filter(!grepl("Intercept", term)) %>%
  mutate(term = str_replace(term, "timepoint_numeric", "Time")) %>%
  mutate(term = if_else(str_detect(term, "growth_cat"), str_remove(term, "growth_cat"), term))

table_export(summaries_df, "sagori_growth_analysis_table_")

summaries_df %>%
  pander(digits = 2, split.table = Inf, split.cells = 20, caption = glue("Linear model of the alpha diversity measurement with numeric time and growth failure as the predictive variables. Full formula: {my_formula}"), emphasize.strong.rows = which(.$p.value<=0.05))

```

# Post-menstrual age > 34 weeks and < 37 weeks

In this section, all of the statistics that include day-of-life (DOL) or hour-of-life (HOL) will instead be analyzed using post-menstrual age (PMA) which is equal to `Gestation in Weeks + (DOL/7)`. We are also going to restrict the analysis to less than 37 weeks and greater than 34 weeks with one sample per baby.

```{r}

s2 %<>%
  mutate(PMA = (DOL/7) + Gestation)

s2 %>%
  ggplot(aes(x = PMA, fill = WOL_label)) +
  geom_histogram(binwidth = 1, center = 0) +
  geom_vline(xintercept = 36, lty = 2) +
  labs(caption = "Dashed line represents 36 weeks PMA")

# check
# s2 %>%
#   filter(PMA > 36) %>%
#   select(SampleID, Zdec36_gt0.8, Zdec36_gt1.2, Zdec36_gt2, growth_cat) %>%
#   view()

```

Subset:

```{r}

s_subset2 <- read_rds(here("Data", "subset_of_babies_great34wk_less37wk.rds"))

```


## Growth faltering

After filtering for > 34 weeks and < 37 weeks PMA we are left with these samples per baby:

```{r fig.height=6, fig.width=8}

s_toPlot <- s_toTest <- s2 %>%
  filter(SampleID %in% s_subset2$SampleID) %>%
  droplevels()

#subsetting timepoint colors for nicer legends
ann_colors = list(
  category = setNames(brewer.pal(length(levels(s_toTest$category)), "Set2"), levels(s_toTest$category)),
  growth_cat = setNames(as.character(ghibli_palette("YesterdayMedium", direction = -1)[1:length(levels(s_toTest$growth_cat))]), levels(s_toTest$growth_cat)),
  TimePoint = setNames(viridis(length(levels(s_toTest$TimePoint))), levels(s_toTest$TimePoint)),
  growth_fail = setNames(qualpal(2)$hex, levels(s_toTest$growth_failure))
)
ann_colors$TimePoint <- ann_colors$TimePoint[names(ann_colors$TimePoint) %in% unique(s_toPlot$TimePoint)]

my_df <- s_toPlot %>%
  count(SubjectID) %>%
  arrange(desc(n)) %>%
  rename(`Number of samples` = n)

my_df %>%
  ggplot(aes(x = SubjectID, y = `Number of samples`, fill = `Number of samples`)) +
  geom_col() +
  theme_bw() +
  scale_fill_viridis_c() +
  theme(axis.text.x = element_text(angle = 90, hjust = 1, vjust = 1))

glue("With a total of {length(unique(s_toPlot$SubjectID))} babies.")

```

### Differential abundance

```{r}

prop_cut = 0.01

props_toTest <- summed_props_long %>%
  filter(SampleID %in% s_toPlot$SampleID) %>%
  group_by(Taxa) %>%
  mutate(mean_prop = mean(props)) %>%
  ungroup() %>%
  filter(mean_prop > prop_cut) %>%
  filter(str_detect(Taxa, "Bacteria|acnes", negate = T)) %>%
  mutate(log10_props = log10(props))

```

We filtered down to bacteria that have a mean relative abundance of >`r 100*prop_cut`% across these samples. We also filtered *Cutibacterium acnes* since it is a known contaminant. Only these bacteria have been tested in this subset.

The top taxa are:

```{r, results='asis'}

props_toTest %>%
  select(Taxa, mean_prop) %>%
  distinct() %>%
  arrange(desc(mean_prop)) %>%
  pander(digits = 2, split.table = Inf)

```

#### Boxplots

#### By TimePoint

```{r fig.height=12, fig.width=8}
props_toTest %>%
  right_join(s_toTest, by="SampleID")  %>%
  mutate(Taxa = reorder(Taxa, -props)) %>%
  mutate(Taxa = fct_relabel(Taxa, function(x) gsub(" ", "\n", x))) %>%
  
  ggplot(aes(x=TimePoint, y=props, color=TimePoint)) +
    geom_boxplot(outlier.alpha=0) +
    #geom_quasirandom() +
    facet_wrap(~Taxa, scales="free", ncol = 4) +
    scale_color_manual(values=ann_colors$TimePoint) +
    scale_y_continuous(trans = "log10") +
    theme_bw() +
    theme(
      axis.text.x=element_text(angle=90, hjust=1, vjust=0.5),
      panel.grid = element_blank(),
      strip.background = element_blank(),
      legend.position = "bottom"
    ) +
    #guides(color=F) +
    labs(
      y="Relative abundance"
    )
```

#### By Growth category

```{r fig.height=12, fig.width=8}
props_toTest %>%
  right_join(s_toTest, by="SampleID")  %>%
  mutate(Taxa = reorder(Taxa, -props)) %>%
  mutate(Taxa = fct_relabel(Taxa, function(x) gsub(" ", "\n", x))) %>%
  mutate(growth_cat = fct_relabel(growth_cat, function(x) gsub(" ", "\n", x))) %>%
  
  ggplot(aes(x=growth_cat, y=props, color=TimePoint)) +
    # geom_boxplot(outlier.alpha=0) +#
    geom_quasirandom(dodge.width = 0.75) +
    facet_wrap(~Taxa, scales="free", ncol = 4) +
    scale_color_manual(values=ann_colors$TimePoint) +
    scale_y_continuous(trans = "log10") +
    theme_bw() +
    theme(
      axis.text.x=element_text(angle=45, hjust=1, vjust=1),
      panel.grid = element_blank(),
      strip.background = element_blank(),
      legend.position = "bottom"
    ) +
    #guides(color=F) +
    labs(
      y="Relative abundance"
    )

ggsave(filename = here("Output", "subset34to37weeksPMA_taxa_abundance.pdf"), width = 10, height = 8, useDingbats = F)

```


#### Linear models

Linear models were used to estimate the change in relative abundance of select taxa between study groups. The relative abundances were log10 transformed. Multiple tests were adjusted for false discovery rate using Benjamini-Hochberg method. Only the terms with p<0.05 are shown in the table below. The value is the change in the log10-transformed relative abundance per unit time (or the change in category).


```{r, results='asis'}

my_formula <- "log10_props ~ bwt_z + log_npo_days_prop + WOL * growth_cat"

summaries_df <- run_lm_on_props(factor_to_test = "Taxa", props_toTest = props_toTest, s_toTest = s_toTest, form1 = my_formula, p_cutoff = 0.05)

table_export(summaries_df, "sagori_growth_analysis_table_")

summaries_df %>%
  mutate(term = str_replace(term, "timepoint_numeric", "Time")) %>%
  mutate(term = if_else(str_detect(term, "growth_cat"), str_remove(term, "growth_cat"), term)) %>%
  pander(digits = 2, split.table = Inf, split.cells = 20, caption = glue("Linear model of the log10-transformed relative abundance with numeric time and growth categories as the predictive variables. Full formula: {my_formula}"), emphasize.strong.rows = which(.$fdr<=0.05))

```

The ones that were closest to changing:

```{r}

to_select <- unique(summaries_df$Taxa)

props_toTest %>%
  filter(Taxa %in% to_select) %>%
  right_join(s_toTest, by="SampleID")  %>%
  mutate(Taxa = reorder(Taxa, -props)) %>%
  mutate(Taxa = fct_relabel(Taxa, function(x) gsub(" ", "\n", x))) %>%
  mutate(growth_cat = fct_relabel(growth_cat, function(x) gsub(" ", "\n", x))) %>%
  
  ggplot(aes(x=growth_cat, y=props, color=TimePoint)) +
    # geom_boxplot(outlier.alpha=0) +#
    geom_quasirandom(dodge.width = 0.75) +
    facet_wrap(~Taxa, scales="free", ncol = 4) +
    scale_color_manual(values=ann_colors$TimePoint) +
    scale_y_continuous(trans = "log10") +
    theme_bw() +
    theme(
      axis.text.x=element_text(angle=45, hjust=1, vjust=1),
      panel.grid = element_blank(),
      strip.background = element_blank(),
      legend.position = "bottom"
    ) +
    #guides(color=F) +
    labs(
      y="Relative abundance"
    )

ggsave(filename = here("Output", "subset34to37weeksPMA_taxa_abundance_changed.pdf"), width = 7, height = 5, useDingbats = F)

```


### Beta diversity

Here, we use Bray-Curtis and Jaccard distances to compare the species composition of the samples to each other.

The plots below show the distance between each pair of samples in a single 2D plot.  It is not possible to plot the distances exactly on paper, so we have used a method of ordination called Principal Coordinates Analysis to select the best coordinate system for display.  The percentage of total variance captured along each axis is displayed on the chart.

#### Plots

#### By TimePoint


```{r fig.height=3, fig.width=5}
s_toTest %>%
  pcoaplus(bc) %>%
  plot(color=TimePoint) +
    scale_color_manual(values=ann_colors$TimePoint) +
    theme_bw() +
    theme(
      axis.text=element_blank(),
      axis.ticks=element_blank(),
      panel.grid = element_blank(), 
      aspect.ratio = 1,
      strip.background = element_blank()
    ) + 
    labs(color="", title="Bray-Curtis")

s_toTest %>%
  pcoaplus(jd) %>%
  plot(color=TimePoint) +
    scale_color_manual(values=ann_colors$TimePoint) +
    theme_bw() +
    theme(
      axis.text=element_blank(),
      axis.ticks=element_blank(),
      panel.grid = element_blank(), 
      aspect.ratio = 1,
      strip.background = element_blank()
    ) + 
    labs(color="", title="Jaccard")
```

#### By Growth category

```{r fig.height=3, fig.width=5}
s_toTest %>%
  pcoaplus(bc) %>%
  plot(color=growth_cat) +
    scale_color_manual(values=ann_colors$growth_cat) +
    theme_bw() +
    theme(
      axis.text=element_blank(),
      axis.ticks=element_blank(),
      panel.grid = element_blank(), 
      aspect.ratio = 1,
      strip.background = element_blank()
    ) + 
    labs(color="", title="Bray-Curtis")

ggsave(here("Output", "bray_curtis_growth_34to37weeksPMA.pdf"), width = 5, height = 5)

s_toTest %>%
  pcoaplus(jd) %>%
  plot(color=growth_cat) +
    scale_color_manual(values=ann_colors$growth_cat) +
    theme_bw() +
    theme(
      axis.text=element_blank(),
      axis.ticks=element_blank(),
      panel.grid = element_blank(), 
      aspect.ratio = 1,
      strip.background = element_blank()
    ) + 
    labs(color="", title="Jaccard")

ggsave(here("Output", "jaccard_growth_34to37weeksPMA.pdf"), width = 5, height = 5)

```

#### PERMANOVA

PERMANOVA test on Bray-Curtis and Jaccard distances to test if the centroids of the study groups can be distinguished from each other.

```{r, results='asis'}

s_permanova <- s_toTest %>%
  filter(!is.na(log_npo_days_prop))

summaries_df <- rbind(
  s_permanova %>%
    do(adonisplus(
      ., distmat = bc, formula = distmat ~ bwt_z + log_npo_days_prop + WOL * growth_cat,
      sample_id_var = SampleID, rep_meas_var = NULL,
      shuffle = NULL, permutations=perm)) %>%
    ungroup() %>%
    mutate(metric = "Bray-Curtis"),
  
  s_permanova %>%
    do(adonisplus(
      ., distmat = jd, formula = distmat ~ bwt_z + log_npo_days_prop + WOL * growth_cat,
      sample_id_var = SampleID, rep_meas_var = NULL,
      shuffle = NULL, permutations=perm)) %>%
    ungroup() %>%
    mutate(metric = "Jaccard")
) %>%
  filter(!term %in% c("Residuals", "Total")) %>%
  select(-one_of(c("sumsq", "meansq"))) %>%
  select(metric, everything())


# If you don't need the shuffling functionality, you can use
# summaries_df <- adonisplus(s_toTest, wu, distmat ~ TimePoint)

summaries_df %>%
  mutate(term = str_replace(term, "timepoint_numeric", "Time")) %>%
  mutate(term = str_replace(term, "growth_cat", "Growth category")) %>%
  pander(emphasize.strong.rows = which(.$p.value<=0.05))

table_export(summaries_df, "sagori_growth_analysis_table_")

```

### Alpha diversity

Alpha diversity was assessed by the expected number of observed OTUs (out of rarefying sample size of `r format(richness_subsample_size, big.mark = ",", scientific = F)`) and Shannon index.

#### Plots

#### By TimePoint

```{r fig.width=8}
s_toTest %>%
  pivot_longer(c("Richness", "Shannon"), names_to="metric", values_to="alpha") %>%

  ggplot(aes(x=TimePoint, y=alpha, color=TimePoint)) +
    geom_boxplot(outlier.alpha=0) +
    geom_quasirandom() +
    facet_grid(metric~., scales = "free") +
    scale_color_manual(values=ann_colors$TimePoint) +
    theme_bw() +
    theme(
      strip.background = element_blank(),
      panel.grid=element_blank(),
      legend.position = "none"
    ) +
    labs(
      x="", color="",
      y="Alpha diversity value"
    )
```

#### By Growth category

```{r fig.width=8}
s_toTest %>%
  pivot_longer(c("Richness", "Shannon"), names_to="metric", values_to="alpha") %>%
  mutate(growth_cat = fct_relabel(growth_cat, function(x) gsub(" ", "\n", x))) %>%
  ggplot(aes(x=growth_cat, y=alpha, color=TimePoint)) +
    geom_quasirandom(dodge.width = 0.75) +
    facet_grid(metric~., scales = "free") +
    scale_color_manual(values=ann_colors$TimePoint) +
    theme_bw() +
    theme(
      axis.text.x=element_text(angle=45, hjust=1, vjust=1),
      strip.background = element_blank(),
      panel.grid=element_blank(),
      legend.position = "bottom"
    ) +
    labs(
      x="Growth category", color="Timepoint",
      y="Alpha diversity value"
    )

ggsave(here("Output", "alpha_diversity_growth_34to37weeksPMA.pdf"), width = 5, height = 5)

```

\newpage

#### Linear model

Linear models were used to estimate the difference between growth categories including time as a covariate.

```{r, results='asis'}

my_formula <- "alpha ~ bwt_z + log_npo_days_prop + WOL * growth_cat"

summaries_df <- s_toTest %>%
  pivot_longer(c("Richness", "Shannon"), names_to="metric", values_to="alpha") %>%
  group_by(metric) %>%
  do(tidy_lm(lm(as.formula(my_formula), data=., na.action=na.omit))) %>%
  ungroup() %>%
  filter(!grepl("Intercept", term)) %>%
  mutate(term = str_replace(term, "timepoint_numeric", "Time")) %>%
  mutate(term = if_else(str_detect(term, "growth_cat"), str_remove(term, "growth_cat"), term))

summaries_df %>%
  pander(digits = 2, split.table = Inf, split.cells = 20, caption = glue("Linear model of the alpha diversity measurement with numeric time and growth categories as the predictive variables. Full formula: {my_formula}"), emphasize.strong.rows = which(.$p.value<=0.05))

```

## Growth failure

### Differential abundance

#### Boxplots

#### By Growth failure

```{r fig.height=12, fig.width=8}
props_toTest %>%
  right_join(s_toTest, by="SampleID")  %>%
  mutate(Taxa = reorder(Taxa, -props)) %>%
  mutate(Taxa = fct_relabel(Taxa, function(x) gsub(" ", "\n", x))) %>%
  
  ggplot(aes(x=WOL, y=props, color=growth_failure)) +
    # geom_boxplot(outlier.alpha=0) +#
    # geom_quasirandom(dodge.width = 0.75) +
  geom_point() +
  geom_smooth(mapping = aes(group=growth_failure), method = "lm", se = F) +
    facet_wrap(~Taxa, scales="free", ncol = 4) +
  scale_color_manual(values=ann_colors$growth_fail) +
    #scale_color_manual(values=ann_colors$TimePoint) +
    scale_y_continuous(trans = "log10") +
    theme_bw() +
    theme(
      axis.text.x=element_text(angle=45, hjust=1, vjust=1),
      panel.grid = element_blank(),
      strip.background = element_blank(),
      legend.position = "bottom"
    ) +
    labs(
      y="Relative abundance",
      x="Chronological age in weeks"
    )

ggsave(filename = here("Output", "subset34to37weeks_growth_failure_taxa_abundance.pdf"), width = 10, height = 8, useDingbats = F)

```


#### Linear models

Linear models were used to estimate the change in relative abundance of select taxa between study groups. The relative abundances were log10 transformed. Multiple tests were adjusted for false discovery rate using Benjamini-Hochberg method. Only the terms with p<0.05 are shown in the table below. The value is the change in the log10-transformed relative abundance per unit time (or the change in category).


```{r, results='asis'}

my_formula <- "log10_props ~ bwt_z + log_npo_days_prop + WOL * growth_failure"

summaries_df <- run_lm_on_props(factor_to_test = "Taxa", props_toTest = props_toTest, s_toTest = s_toTest, form1 = my_formula, p_cutoff = 0.05) %>%
  mutate(term = str_replace(term, "timepoint_numeric", "Time")) %>%
  mutate(term = str_replace(term, "growth_failureYes", "Had growth failure"))

table_export(summaries_df, "sagori_growth_analysis_table_")

summaries_df %>%
  pander(digits = 2, split.table = Inf, split.cells = 20, caption = glue("Linear model of the log10-transformed relative abundance with numeric time and growth failure as the predictive variables. Full formula: {my_formula}"), emphasize.strong.rows = which(.$fdr<=0.05))

```

The ones that significantly changed after fdr correction:

```{r}

to_select <- summaries_df %>%
  filter(fdr < 0.05) %>%
  pull(Taxa) %>%
  unique()

props_toTest %>%
  filter(Taxa %in% to_select) %>%
  right_join(s_toTest, by="SampleID")  %>%
  mutate(Taxa = reorder(Taxa, -props)) %>%
  mutate(Taxa = fct_relabel(Taxa, function(x) gsub(" ", "\n", x))) %>%
  
  ggplot(aes(x=WOL, y=props, color=growth_failure)) +
    # geom_boxplot(outlier.alpha=0) +#
    # geom_quasirandom(dodge.width = 0.75) +
  geom_point() +
  geom_smooth(mapping = aes(group=growth_failure), method = "lm", se = F) +
    facet_wrap(~Taxa, scales="free", ncol = 4) +
  scale_color_manual(values=ann_colors$growth_fail) +
    #scale_color_manual(values=ann_colors$TimePoint) +
    scale_y_continuous(trans = "log10") +
    theme_bw() +
    theme(
      axis.text.x=element_text(angle=45, hjust=1, vjust=1),
      panel.grid = element_blank(),
      strip.background = element_blank(),
      legend.position = "bottom"
    ) +
    labs(
      y="Relative abundance",
      x="Chronological age in weeks"
    )

ggsave(filename = here("Output", "subset34to37weeks_growth_failure_taxa_abundance_changed.pdf"), width = 6, height = 4, useDingbats = F)

```

It looks like the changes are heavily influenced by outliers.

#### Let's do a sensitivity analysis:

inf185Cat1T3n1 is the outlier sample. It's data point -7.247 on a log10 scale is more than 3 standard deviations from the mean. To put it another way, it's smaller than Q1 - 1.5(IQR) where Q1 is the 1st quartile and IQR is the interquartile range (the range between the 1st and 3rd quartiles i.e. containing 75% of the data points).

```{r}

without_outlier <- props_toTest %>%
  filter(Taxa %in% to_select) %>%
  filter(!SampleID %in% "inf185Cat1T3n1")

my_formula <- "log10_props ~ bwt_z + log_npo_days_prop + WOL * growth_failure"

summaries_df <- run_lm_on_props(factor_to_test = "Taxa", props_toTest = without_outlier, s_toTest = s_toTest, form1 = my_formula, p_cutoff = 0.05) %>%
  mutate(term = str_replace(term, "timepoint_numeric", "Time")) %>%
  mutate(term = str_replace(term, "growth_failureYes", "Had growth failure"))

table_export(summaries_df, "sagori_growth_analysis_table_")

summaries_df %>%
  pander(digits = 2, split.table = Inf, split.cells = 20, caption = glue("Linear model of the log10-transformed relative abundance with numeric time and growth failure as the predictive variables. Full formula: {my_formula}"), emphasize.strong.rows = which(.$fdr<=0.05))

```

Graphing again:

```{r}

without_outlier %>%
  left_join(s_toTest, by="SampleID")  %>%
  mutate(Taxa = reorder(Taxa, -props)) %>%
  mutate(Taxa = fct_relabel(Taxa, function(x) gsub(" ", "\n", x))) %>%
  
  ggplot(aes(x=WOL, y=props, color=growth_failure)) +
    # geom_boxplot(outlier.alpha=0) +#
    # geom_quasirandom(dodge.width = 0.75) +
  geom_point() +
  geom_smooth(mapping = aes(group=growth_failure), method = "lm", se = F) +
    facet_wrap(~Taxa, scales="free", ncol = 4) +
  scale_color_manual(values=ann_colors$growth_fail) +
    #scale_color_manual(values=ann_colors$TimePoint) +
    scale_y_continuous(trans = "log10") +
    theme_bw() +
    theme(
      axis.text.x=element_text(angle=45, hjust=1, vjust=1),
      panel.grid = element_blank(),
      strip.background = element_blank(),
      legend.position = "bottom"
    ) +
    labs(
      y="Relative abundance",
      x="Chronological age in weeks"
    )

ggsave(filename = here("Output", "subset34to37weeks_growth_failure_taxa_abundance_changed_without_outlier.pdf"), width = 6, height = 4, useDingbats = F)

```

### Beta diversity

Here, we use Bray-Curtis and Jaccard distances to compare the species composition of the samples to each other.

The plots below show the distance between each pair of samples in a single 2D plot.  It is not possible to plot the distances exactly on paper, so we have used a method of ordination called Principal Coordinates Analysis to select the best coordinate system for display.  The percentage of total variance captured along each axis is displayed on the chart.

#### Plots

#### By Growth failure

```{r fig.height=3, fig.width=5}
s_toTest %>%
  pcoaplus(bc) %>%
  plot(color=growth_failure) +
    scale_color_manual(values=ann_colors$growth_fail) +
    theme_bw() +
    theme(
      axis.text=element_blank(),
      axis.ticks=element_blank(),
      panel.grid = element_blank(), 
      aspect.ratio = 1,
      strip.background = element_blank()
    ) + 
    labs(color="", title="Bray-Curtis")

ggsave(here("Output", "bray_curtis_growth_failure_34to37weeksPMA.pdf"), width = 5, height = 5)

s_toTest %>%
  pcoaplus(jd) %>%
  plot(color=growth_failure) +
    scale_color_manual(values=ann_colors$growth_fail) +
    theme_bw() +
    theme(
      axis.text=element_blank(),
      axis.ticks=element_blank(),
      panel.grid = element_blank(), 
      aspect.ratio = 1,
      strip.background = element_blank()
    ) + 
    labs(color="", title="Jaccard")

ggsave(here("Output", "jaccard_growth_failure_34to37weeksPMA.pdf"), width = 5, height = 5)

```

```{r fig.height=3, fig.width=5}
plot_1 <- s_toTest %>%
  pcoaplus(bc) %>%
  plot(color=growth_failure) +
    scale_color_manual(values=ann_colors$growth_fail) +
    theme_bw() +
    theme(
      axis.text=element_blank(),
      axis.ticks=element_blank(),
      panel.grid = element_blank(), 
      aspect.ratio = 1,
      strip.background = element_blank()
    ) + 
    labs(color="Growth failure", title="Bray-Curtis")

plot_2 <- s_toTest %>%
  pcoaplus(jd) %>%
  plot(color=growth_failure) +
    scale_color_manual(values=ann_colors$growth_fail) +
    theme_bw() +
    theme(
      axis.text=element_blank(),
      axis.ticks=element_blank(),
      panel.grid = element_blank(), 
      aspect.ratio = 1,
      strip.background = element_blank()
    ) + 
    labs(color="Growth failure", title="Jaccard")

my_plot <- plot_1 / plot_2 + plot_layout(guides = "collect")

ggsave(filename = here("Output", "beta_div_growth_failure_34to37weeksPMA.pdf"), plot = my_plot, width = 4, height = 5)

```

#### PERMANOVA

PERMANOVA test on Bray-Curtis and Jaccard distances to test if the centroids of the study groups can be distinguished from each other.

```{r, results='asis'}

s_permanova <- s_toTest %>%
  filter(!is.na(log_npo_days_prop))

summaries_df <- rbind(
  s_permanova %>%
    do(adonisplus(
      ., distmat = bc, formula = distmat ~ bwt_z + log_npo_days_prop + WOL * growth_cat,
      sample_id_var = SampleID, rep_meas_var = NULL,
      shuffle = NULL, permutations=perm)) %>%
    ungroup() %>%
    mutate(metric = "Bray-Curtis"),
  
  s_permanova %>%
    do(adonisplus(
      ., distmat = jd, formula = distmat ~ bwt_z + log_npo_days_prop + WOL * growth_cat,
      sample_id_var = SampleID, rep_meas_var = NULL,
      shuffle = NULL, permutations=perm)) %>%
    ungroup() %>%
    mutate(metric = "Jaccard")
) %>%
  filter(!term %in% c("Residuals", "Total")) %>%
  select(-one_of(c("sumsq", "meansq"))) %>%
  select(metric, everything())

summaries_df %>%
  mutate(term = str_replace(term, "timepoint_numeric", "Time")) %>%
  mutate(term = str_replace(term, "growth_failure", "Growth failure")) %>%
  pander(emphasize.strong.rows = which(.$p.value<=0.05))

table_export(summaries_df, "sagori_growth_analysis_table_")

```

### Alpha diversity

Alpha diversity was assessed by the expected number of observed OTUs (out of rarefying sample size of `r format(richness_subsample_size, big.mark = ",", scientific = F)`) and Shannon index.

#### Plots

#### By Growth failure

```{r fig.width=8}
s_toTest %>%
  pivot_longer(c("Richness", "Shannon"), names_to="metric", values_to="alpha") %>%
  ggplot(aes(x=WOL, y=alpha, color=growth_failure)) +
  geom_point() +
  geom_smooth(mapping = aes(group=growth_failure), method = "lm", se = F) +
  facet_grid(metric~., scales = "free") +
  scale_color_manual(values=ann_colors$growth_fail) +
  theme_bw() +
  theme(
    axis.text.x=element_text(angle=45, hjust=1, vjust=1),
    strip.background = element_blank(),
    panel.grid=element_blank(),
    legend.position = "bottom"
  ) +
  labs(
    x="Chronological age in weeks", color="Growth failure",
    y="Alpha diversity value"
  )

```

```{r}

my_plot <- s_toTest %>%
  pivot_longer(c("Richness", "Shannon"), names_to="metric", values_to="alpha") %>%
  ggplot(aes(x=WOL, y=alpha, color=growth_failure)) +
  geom_point() +
  geom_smooth(mapping = aes(group=growth_failure), method = "lm", se = F) +
  facet_grid(metric~., scales = "free") +
  scale_color_manual(values=ann_colors$growth_fail) +
    theme_bw() +
    theme(
      axis.text.x=element_text(angle=45, hjust=1, vjust=1),
      strip.background = element_blank(),
      panel.grid=element_blank(),
      legend.position = "none"
    ) +
    labs(
      x="Chronological age in weeks", color="Growth failure",
      y="Alpha diversity value"
    )

ggsave(plot = my_plot, filename = here("Output", "alpha_diversity_growth_failure_34to37weeksPMA.pdf"), width = 3, height = 3)

```

\newpage

#### Linear model

Linear models were used to estimate the difference between growth categories including time as a covariate.

```{r, results='asis'}

my_formula <- "alpha ~ bwt_z + log_npo_days_prop + WOL * growth_failure"

summaries_df <- s_toTest %>%
  pivot_longer(c("Richness", "Shannon"), names_to="metric", values_to="alpha") %>%
  group_by(metric) %>%
  do(tidy_lm(lm(as.formula(my_formula), data=., na.action=na.omit))) %>%
  ungroup() %>%
  filter(!grepl("Intercept", term)) %>%
  mutate(term = str_replace(term, "timepoint_numeric", "Time")) %>%
  mutate(term = str_replace(term, "growth_failureYes", "Had growth failure"))

table_export(summaries_df, "sagori_growth_analysis_table_")

summaries_df %>%
  pander(digits = 2, split.table = Inf, split.cells = 20, caption = glue("Linear model of the alpha diversity measurement with numeric time and growth failure as the predictive variables. Full formula: {my_formula}"), emphasize.strong.rows = which(.$p.value<=0.05))

```

#### Let's do a sensitivity analysis:

Again there seem to be outliers, this time in the normal growth group.

inf179Cat1T4n1 is more than 3 standard deviations away from the mean and is greater than Q3 + IQR*1.5. 

```{r}

summary_shannon <-summary(s_toTest$Shannon)

upper_limit_shannon <- summary_shannon[[5]] + IQR(s_toTest$Shannon) * 1.5
lower_limit_shannon <- summary_shannon[[2]] - IQR(s_toTest$Shannon) * 1.5

outlier_sample <- s_toTest %>%
  mutate(scale_shannon = scale(Shannon)) %>%
  filter(Shannon > upper_limit_shannon | Shannon < lower_limit_shannon) %>%
  pull(SampleID)

without_outlier <- s_toTest %>%
  filter(!SampleID %in% outlier_sample)

my_formula <- "alpha ~ bwt_z + log_npo_days_prop + WOL * growth_failure"

summaries_df <- without_outlier %>%
  pivot_longer(c("Richness", "Shannon"), names_to="metric", values_to="alpha") %>%
  group_by(metric) %>%
  do(tidy_lm(lm(as.formula(my_formula), data=., na.action=na.omit))) %>%
  ungroup() %>%
  filter(!grepl("Intercept", term)) %>%
  mutate(term = str_replace(term, "timepoint_numeric", "Time")) %>%
  mutate(term = str_replace(term, "growth_failureYes", "Had growth failure"))

table_export(summaries_df, "sagori_growth_analysis_table_")

summaries_df %>%
  pander(digits = 2, split.table = Inf, split.cells = 20, caption = glue("Linear model of the alpha diversity measurement with numeric time and growth failure as the predictive variables. Full formula: {my_formula}"), emphasize.strong.rows = which(.$p.value<=0.05))

```

Graph without outlier:

```{r fig.width=8}

my_plot <- without_outlier %>%
  pivot_longer(c("Richness", "Shannon"), names_to="metric", values_to="alpha") %>%
  ggplot(aes(x=WOL, y=alpha, color=growth_failure)) +
  geom_point() +
  geom_smooth(mapping = aes(group=growth_failure), method = "lm", se = F) +
  facet_grid(metric~., scales = "free") +
  scale_color_manual(values=ann_colors$growth_fail) +
    theme_bw() +
    theme(
      axis.text.x=element_text(angle=45, hjust=1, vjust=1),
      strip.background = element_blank(),
      panel.grid=element_blank(),
      legend.position = "none"
    ) +
    labs(
      x="Chronological age in weeks", color="Growth failure",
      y="Alpha diversity value"
    )

ggsave(plot = my_plot, filename = here("Output", "alpha_diversity_growth_failure_34to37weeksPMA_outlier_removed.pdf"), width = 3, height = 3)

```

# End of report

```{r savepoint, include=FALSE}

save.image(file = here("Data", "after_taxa_analysis.RData"))

```


```{r Generate time stamped report, eval=FALSE, include=FALSE}
#notes
#neat: you can run this following command in the console to give your reports custom names (or date-stamp them)
#rmarkdown::render('ronen_stein_Report_shotgun.Rmd',output_file = paste('CEASE.report.', Sys.Date(), '.pdf', sep=''))

rstudioapi::documentSaveAll()

script_dir <- file.path("Scripts","Rmds")
output_dir <- file.path("Output")

report_fp <- here::here(script_dir, "Shotgun_taxonomic_growth_analysis.Rmd")
output_fp <- here::here(output_dir, paste('sagori_microbiome_and_growth_', Sys.Date(), '.pdf', sep=''))

rmarkdown::render(report_fp, output_file = output_fp)

```
