---
title: |
  ![](CHOPMC_logo.pdf){width=5in}
  \break
  Sagori Mukhopadhyay - Runs 1 to 5 - Growth Analysis - Gene Abundances
author: "Scott G. Daniel, PennCHOP Microbiome Program"
date: \today
output: 
  pdf_document:
    toc: true
    toc_depth: 3
---

```{r knitr setup, echo=FALSE, message=FALSE}
library(knitr)
opts_chunk$set(
  tidy=FALSE,
  cache=FALSE,
  echo=FALSE,
  warning=FALSE,
  message=FALSE,
  dpi=100,
  fig.width=5,
  fig.height=3,
  fig.align = "center"
  )
```

```{r libraries, message=FALSE, warning=FALSE}
library(here)
library(ggbeeswarm)
library(grid)
library(magrittr)
library(pander)
library(pheatmap)
library(usedist)
library(qiimer)
library(reshape2)
library(jsonlite)
library(vegan)
library(ape)
library(tidyverse)
library(gghighlight)
library(RColorBrewer)
library(patchwork)
library(kylemisc)
library(ggpubr)
library(GGally)
library(glue)
library(emmeans)
library(adonisplus)

#MOAR cOLORS!!!
library(ggsci)
library(ghibli)
library(qualpalr)

panderOptions("knitr.auto.asis", FALSE)
#this also means that you have to put results='asis' in any block that has pander output
#this lets you loop through variables and produce multiple pander tables and ggplots in a
#single code block!

```

```{r functions}

source(here("Scripts","Functions","common_functions.R"))
source(here("Scripts","Functions","scott_daniel_useful_R_functions.R"))
source(here("Scripts","Functions","for_functional_analysis.R"))

```

```{r load savepoint, include=FALSE}
#you can load data here if you have a savepoint from your main R markdown (but remember that libraries do not get saved)
load(here("Data","after_taxa_analysis.RData"))

table_export <- table_iterator()
```

```{r constants}

ec_fp <- here("Data","ec.list")
ko_ec_fp <- here("Data", "ko_enzyme.list")
ko_definitions_fp <- here("Data", "ko_definition.list")
ko_mo_fp <- here("Data", "ko_module.list")
ko_mo_info_fp <- here("Data", "KEGG_modules_info.tsv")
mo_pathway_fp <- here("Data", "module_pathway.list")

```

# KEGG

Reads were aligned to the Kyoto Encyclopedia of Genes and Genomes (downloaded on February 28, 2019) using diamond sequence aligner (PMID:33828273). The resulting protein hits were mapped to KEGG ortholog database. Below is a PCoA plot of Bray-Curtis distances using relative abundances of KEGG orthologs.

```{r loading kegg data}

### KEGG orthology assignment

s_seq <- s2
library(viridis)
## Set colors for each factor ahead of time so they are consistent through the report.
ann_colors = list(
  category = setNames(brewer.pal(length(levels(s_seq$category)), "Set2"), levels(s_seq$category)),
  growth_cat = setNames(as.character(ghibli_palette("YesterdayMedium", direction = -1)[1:length(levels(s_seq$growth_cat))]), levels(s_seq$growth_cat)),
  TimePoint = setNames(viridis(length(levels(s_seq$TimePoint))), levels(s_seq$TimePoint)),
  growth_fail = setNames(qualpal(2)$hex, levels(s_seq$growth_failure))
)

kegg_1to3 <- read_table_if_no_rds(base_dir = here("Data","gene_hits","run1to3","20190228_kegg_species_prokaryotes"), name_of_table = kegg, rds_file_name = "kegg1to3.rds") %>%
  filter(SampleID %in% s_seq$SampleID)
kegg_4 <- read_table_if_no_rds(base_dir = here("Data","gene_hits","run4","20190228_kegg_species_prokaryotes"), name_of_table = kegg, rds_file_name = "kegg4.rds") %>%
  filter(SampleID %in% s_seq$SampleID)
kegg_5 <- read_table_if_no_rds(base_dir = here("Data","gene_hits","run5","20190228_kegg_species_prokaryotes"), name_of_table = kegg, rds_file_name = "kegg5.rds") %>%
  filter(SampleID %in% s_seq$SampleID)

kegg <- bind_rows(kegg_1to3, kegg_4, kegg_5)

kegg_props_raw <- kegg %>%
  group_by(SampleID) %>%
  mutate(sample_sum = sum(count)) %>%
  mutate(props = count / sample_sum) %>%
  ungroup()

# replace zeros per gene
kegg_props <- kegg_props_raw %>%
  group_by(geneID) %>%
  filter(sum(count) > 0) %>%
  mutate(geneID_min = min(props[props > 0])) %>%
  mutate(props = ifelse(props == 0, geneID_min/2, props)) %>%
  ungroup() %>%
  select(-geneID_min, -sample_sum, -database)

rm(kegg_1to3, kegg_4, kegg_5)

kegg_defnitions <- read_tsv(here("Data", "ko_definition.list"), col_names = c("Ortholog", "Definition")) %>%
  mutate(Ortholog = str_remove(Ortholog, "ko:"))

prokarotic_paths <- read_tsv(here("Data","prokaryotic_pathway.list"), col_names = c("PathwayID", "Description")) %>%
  mutate(PathwayID = paste0("path:map", PathwayID)) %>%
  left_join(read_rds(here("Data","kegg_pathways.RDS")), by = "PathwayID") %>%
  mutate(Ortholog = str_remove(Ortholog, "ko:")) %>%
  select(-Description)

path_annotation <- prokarotic_paths %>%
  select(Ortholog, PathwayName) %>%
  group_by(Ortholog) %>%
  summarise(Pathways = paste0(PathwayName, collapse = ", ")) %>%
  ungroup()

```

```{r glycoside hydrolases}
ec_names <- read_delim(ec_fp, delim='\t', col_names = F) %>%
  setNames(c("EC_id", "EC_desc")) %>%
  separate(EC_desc, into=c("EC_desc", "rest"), extra = "merge", sep = ';') %>%
  select(-rest)

kegg_info <- read.delim(ko_ec_fp, stringsAsFactors = F, sep='\t', header = F) %>%
  setNames(c("geneID", "EC_id")) %>%
  mutate(geneID = sub("ko:", "", geneID)) %>%
  group_by(geneID) %>%
  mutate(weight = 1/n()) %>%
  ungroup()

kegg_ec <- kegg_props %>%
  left_join(kegg_info, by="geneID") %>%
  mutate(count = count*weight, props = props*weight) %>%
  group_by(SampleID, EC_id) %>%
  summarize(count = sum(count), props = sum(props), .groups = "drop") %>%
  ungroup() %>%
  filter(!is.na(EC_id))

kegg_gh <- kegg_ec %>%
  filter(grepl("ec:3\\.2\\.1\\.", EC_id)) %>%
  droplevels()

```

```{r}
current_sub_study <- "All timepoints growth faltering"
```

# All timepoints

## Growth faltering

### Bray-curtis PCoA

First, we will examine overall differences using PCoA of the Bray-curtis' distances of the gene abundances between samples.

#### PCoA's

```{r}

if (current_sub_study %in% s_seq$sub_study) {
  s_toTest <- s_toPlot <- s_seq %>%
    filter(str_detect(sub_study, current_sub_study))
} else {
  s_toTest <- s_toPlot <- s_seq
}

ko <- kegg_props_raw %>%
  filter(SampleID %in% s_toPlot$SampleID) %>%
  select(SampleID, geneID, count) %>%
  spread_to_numeric_matrix(., "geneID", "SampleID", "count")

bc_kegg <- vegdist(t(ko))

dist_in <- usedist::dist_subset(bc_kegg, s_toPlot$SampleID)

ann_colors = list(
  category = setNames(brewer.pal(length(levels(s_toTest$category)), "Set2"), levels(s_toTest$category)),
  growth_cat = setNames(as.character(ghibli_palette("YesterdayMedium", direction = -1)[1:length(levels(s_toTest$growth_cat))]), levels(s_toTest$growth_cat)),
  TimePoint = setNames(viridis(length(levels(s_toTest$TimePoint))), levels(s_toTest$TimePoint)),
  growth_fail = setNames(qualpal(2)$hex, levels(s_toTest$growth_failure))
)

```

```{r fig.height=5, fig.width=5}
s_toPlot %>%
  pcoaplus(bc_kegg) %>%
  plot(color=TimePoint) +
    scale_color_manual(values=ann_colors$TimePoint) +
    theme_bw() +
    theme(
      axis.text=element_blank(),
      axis.ticks=element_blank(),
      panel.grid = element_blank(), 
      aspect.ratio = 1,
      strip.background = element_blank()
    ) + 
    labs(color="", title="Bray-Curtis")
```

```{r fig.height=5, fig.width=5}
s_toPlot %>%
  pcoaplus(bc_kegg) %>%
  plot(color=growth_cat) +
    scale_color_manual(values=ann_colors$growth_cat) +
    theme_bw() +
    theme(
      axis.text=element_blank(),
      axis.ticks=element_blank(),
      panel.grid = element_blank(), 
      aspect.ratio = 1,
      strip.background = element_blank()
    ) + 
    labs(color="", title="Bray-Curtis")
```

#### Permanova

AbxHoursPerHOL has been taken out because it is collinear with npo_days_prop.

Ok, now I know why I didn't redo this report because all of the gene abundance stuff we actually need is in FiguresANDTables.Rmd.

```{r, results='asis'}
s_permanova <- s_toTest %>%
  filter(!is.na(log_npo_days_prop))

summaries_df <-
  s_permanova %>%
    do(adonisplus(
      ., distmat = bc_kegg, formula = distmat ~ log_npo_days_prop + bwt_z + WOL * growth_cat,
      sample_id_var = SampleID, rep_meas_var = SubjectID,
      shuffle = c(growth_cat = "between", WOL = "within"), permutations=perm)) %>%
    ungroup() %>%
    mutate(metric = "Bray-Curtis") %>%
  filter(!term %in% c("Residual", "Total")) %>%
  select(metric, everything())

write_rds(summaries_df, here("Data", "kegg_bc_permanova_results.rds"))

summaries_df <- read_rds(here("Data", "kegg_bc_permanova_results.rds"))

table_export(summaries_df, "gene_analysis_table_")

summaries_df %>%
  pander(split.cells = Inf, digits = 2, emphasize.strong.rows = which(.$p.value<=0.05))

```

```{r}

sum_of_counts <- kegg %>%
  select(SampleID, Ortholog = geneID, count) %>%
  inner_join(prokarotic_paths, by = "Ortholog") %>%
  group_by(SampleID) %>%
  summarise(sample_sums = sum(count))

kegg_path_props <- kegg %>%
  select(SampleID, Ortholog = geneID, count) %>%
  inner_join(prokarotic_paths, by = "Ortholog") %>%
  group_by(PathwayName, SampleID) %>%
  summarise(count = sum(count)) %>%
  ungroup() %>%
  inner_join(sum_of_counts, by = "SampleID") %>%
  mutate(props = count / sample_sums)

kegg_path_props %<>%
  group_by(PathwayName) %>%
  filter(sum(count) > 0) %>%
  mutate(pathway_min = min(props[props > 0])) %>%
  mutate(props = ifelse(props == 0, pathway_min/2, props)) %>%
  ungroup() %>%
  select(-pathway_min, -sample_sums, -count) %>%
  droplevels()

```


```{r saving for random forest analysis}

save(ann_colors, ec_names, kegg, kegg_ec, kegg_gh, kegg_info, kegg_path_props, kegg_props, kegg_props_raw, ko, prokarotic_paths, file = here("Data", "sagori_kegg_data.RData"))

```

### Linear models of pathways

```{r, results='asis'}

number_of_top_pathways = 25

props_toTest <- kegg_path_props %>%
  mutate(props_log2 = log2(props)) %>%
  mutate(PathwayName = fct_lump(PathwayName, n = number_of_top_pathways, w = .$props, other_level = "Other")) %>%
  filter(!PathwayName %in% c("Metabolic pathways", "Other"))

form1 = "props_log2 ~ timepoint_numeric * growth_cat + npo_days_prop + feeds_prop_hm + AbxHoursPerHOL"
random1 = "SubjectID"

summaries_df <- run_lmer_on_props(factor_to_test = "PathwayName", props_toTest = props_toTest, s_toTest = s_toTest, form1 = form1,rep_mes_label = random1, p_cutoff =  0.05) 

summaries_df %>%
  mutate(term = str_remove(term, "growth_cat")) %>%
  pander(split.table=Inf, digits=2, caption = glue("Linear models of log2 transformed abundances of KEGG pathways"), emphasize.strong.rows = which(.$fdr<=0.05))

table_export(summaries_df, "gene_analysis_table_")

```

```{r stats1}

zeug_df <- kegg_path_props %>%
  mutate(props_log2 = log2(props)) %>%
  mutate(PathwayName = fct_lump(PathwayName, n = number_of_top_pathways, w = .$props, other_level = "Other")) %>%
  group_by(PathwayName) %>%
  summarise(total_prop = sum(props)) %>%
  ungroup() %>%
  mutate(fraction = total_prop / sum(total_prop))

total_path_abund <- zeug_df %>%
  filter(!PathwayName %in% c("Metabolic pathways", "Other")) %>%
  summarise(total_fraction = sum(fraction))

```

The top 25 pathways (excluding the general "Metabolic pathways") covered `r total_path_abund$total_fraction*100`% of all KEGG ortholog abundances.

#### Boxplot of pathway abundances

```{r boxplot}

to_select <- unique(as.character(summaries_df$PathwayName))

number_of_blocks <- ceiling(length(to_select) / 9)
chunk_it <- function(x,n) split(x, cut(seq_along(x), n, labels = FALSE))
my_blocks <- chunk_it(to_select, number_of_blocks)

for (i in 1:number_of_blocks) {
  subset <- my_blocks[[i]]

  props_toTest %>%
    filter(PathwayName %in% subset) %>%
    filter(SampleID %in% s_toTest$SampleID) %>%
    merge(s_toTest, by="SampleID") %>%
    
    mutate(PathwayName = reorder(PathwayName, -props)) %>%
    ggplot(aes(x=growth_cat, y=props, color=growth_cat)) +
    geom_boxplot(outlier.alpha = 0) +
    geom_quasirandom(dodge.width = 0.75) +
    theme_bw() +
    theme(
      axis.text.x = element_text(angle=45, vjust=1, hjust=1),
      strip.background = element_blank()) +
    #scale_color_manual(values=brewer.pal(8, "Set1")[c(2,5,1)]) +
    scale_color_brewer(palette = "Set2") +
    scale_y_continuous(trans = "log2") +
    facet_wrap(~PathwayName, scales="free_y", ncol=3) +
    labs(x="", color="growth_cat",
         y="Relative abundance",
         title = "Pathways in samples that had gene abundance differences")
  
  ggsave(here("Output",glue("{current_sub_study}_kegg_pathway_differences_{i}.pdf")), height=7, width=12, useDingbats=F)
}

```

#### Those that changed

```{r}

pathways_to_select <- summaries_df %>%
  mutate(term = str_remove(term, "growth_cat")) %>%
  filter(term %in% c("moderate to severe retardation", "mild retardation")) %>%
  pull(PathwayName) %>%
  unique()

summaries_df <- run_lmer_on_props(factor_to_test = "PathwayName", props_toTest = props_toTest, s_toTest = s_toTest, form1 = form1,rep_mes_label = random1, p_cutoff =  1)

plot_df <- summaries_df %>%
  mutate(term = str_remove(term, "growth_cat")) %>%
  filter(term %in% c("moderate to severe retardation", "mild retardation"), PathwayName %in% pathways_to_select) %>%
  mutate(CI95_lwr = round(Value - 1.96*Std.Error, 2)) %>%
  mutate(CI95_upr = round(Value + 1.96*Std.Error, 2)) %>%
  mutate(Significant = case_when(fdr < 0.05 ~ "p.adj < 0.05",
                                 fdr > 0.05 ~ "Not significant")) %>%
  arrange(Value) %>%
  mutate(PathwayName = fct_reorder(PathwayName, Value))

g <- ggplot(plot_df, aes(Value, PathwayName, color = term, shape = Significant)) +
  theme_bw() +
  geom_pointrange(aes(xmin=Value-Std.Error, xmax=Value+Std.Error), position = position_dodge(width = 0.4)) +
  geom_vline(xintercept = 0, linetype = "longdash") +
  scale_shape_manual(values = c(1,19)) +
  scale_color_manual(values=ann_colors$growth_cat[2:3]) +
  labs(y = "KEGG Pathway",
       x = "<-- Decreased vs. normal growth    Increased vs. normal growth -->",
       color = "Growth category")

plot(g)

ggsave(here("Output", glue("{current_sub_study}_kegg_pathway_pvalue_graph.pdf")), height = 5, width = 7)

```

### KEGG orthologs

Taking the top 500 KEGG orthologs and testing them

```{r message=FALSE, warning=FALSE, paged.print=TRUE, results='asis'}

top500 <- kegg_props %>%
  filter(SampleID %in% s_toTest$SampleID) %>%
  group_by(geneID) %>%
  summarise(mean_props = mean(props)) %>%
  arrange(desc(mean_props)) %>%
  slice(1:500)

kegg_props_subset <- kegg_props %>%
  select(SampleID, Ortholog = geneID, count, props) %>%
  filter(Ortholog %in% top500$geneID) %>%
  mutate(props_log2 = log2(props))

form1 = "props_log2 ~ timepoint_numeric * growth_cat + npo_days_prop + feeds_prop_hm + AbxHoursPerHOL"
random1 = "SubjectID"

# summaries_df <- run_lmer_on_props(factor_to_test = "Ortholog", props_toTest = kegg_props_subset, s_toTest = s_toTest, form1 = form1, p_cutoff = 1, rep_mes_label = random1)
# 
# save(summaries_df, file = here("Data", glue("{current_sub_study}_kegg500_changes.RDS")))

load(file = here("Data", glue("{current_sub_study}_kegg500_changes.RDS")))

to_select <- summaries_df %>%
  mutate(term = str_remove(term, "growth_cat")) %>%
  filter(term %in% c("moderate to severe retardation", "mild retardation"), fdr < 0.05, Value > 0) %>%
  arrange(desc(Value)) %>%
  slice(1:45) %>%
  pull(Ortholog) %>%
  as.character() %>%
  unique()
#actually top 32
  
plot_df <- summaries_df %>%
  mutate(term = str_remove(term, "growth_cat")) %>%
  filter(term %in% c("moderate to severe retardation", "mild retardation"), Ortholog %in% to_select) %>%
  left_join(kegg_defnitions, by = "Ortholog") %>%
  mutate(ortho_lable = paste(Ortholog, str_trunc(Definition, 30, "right"))) %>%
  mutate(CI95_lwr = round(Value - 1.96*Std.Error, 2)) %>%
  mutate(CI95_upr = round(Value + 1.96*Std.Error, 2)) %>%
  mutate(Significant = case_when(fdr < 0.05 ~ "p.adj < 0.05",
                                 fdr > 0.05 ~ "Not significant")) %>%
  arrange(Value) %>%
  mutate(ortho_lable = fct_reorder(ortho_lable, Value))

g <- ggplot(plot_df, aes(Value, ortho_lable, color = term, shape = Significant)) +
  theme_bw() +
  geom_pointrange(aes(xmin=Value-Std.Error, xmax=Value+Std.Error), position = position_dodge(width = 0.4)) +
  geom_vline(xintercept = 0, linetype = "longdash") +
  scale_shape_manual(values = c(1,19)) +
  scale_color_manual(values=ann_colors$growth_cat[2:3]) +
  labs(y = "Ortholog / Description",
       x = "<-- Decreased vs. normal growth    Increased vs. normal growth -->",
       color = "Growth category")

plot(g)

ggsave(here("Output", glue("{current_sub_study}_top_increasing_orthologs.pdf")), height = 7, width = 10)

to_select <- summaries_df %>%
  mutate(term = str_remove(term, "growth_cat")) %>%
  filter(term %in% c("moderate to severe retardation", "mild retardation"), fdr < 0.05, Value < 0) %>%
  arrange(desc(Value)) %>%
  slice(1:45) %>%
  pull(Ortholog) %>%
  as.character() %>%
  unique()
#actually top 32
  
plot_df <- summaries_df %>%
  mutate(term = str_remove(term, "growth_cat")) %>%
  filter(term %in% c("moderate to severe retardation", "mild retardation"), Ortholog %in% to_select) %>%
  left_join(kegg_defnitions, by = "Ortholog") %>%
  mutate(ortho_lable = paste(Ortholog, str_trunc(Definition, 30, "right"))) %>%
  mutate(CI95_lwr = round(Value - 1.96*Std.Error, 2)) %>%
  mutate(CI95_upr = round(Value + 1.96*Std.Error, 2)) %>%
  mutate(Significant = case_when(fdr < 0.05 ~ "p.adj < 0.05",
                                 fdr > 0.05 ~ "Not significant")) %>%
  arrange(Value) %>%
  mutate(ortho_lable = fct_reorder(ortho_lable, Value))

g <- ggplot(plot_df, aes(Value, ortho_lable, color = term, shape = Significant)) +
  theme_bw() +
  geom_pointrange(aes(xmin=Value-Std.Error, xmax=Value+Std.Error), position = position_dodge(width = 0.4)) +
  geom_vline(xintercept = 0, linetype = "longdash") +
  scale_shape_manual(values = c(1,19)) +
  scale_color_manual(values=ann_colors$growth_cat[2:3]) +
  labs(y = "Ortholog / Description",
       x = "<-- Decreased vs. normal growth    Increased vs. normal growth -->",
       color = "Growth category")

plot(g)

ggsave(here("Output", glue("{current_sub_study}_top_decreasing_orthologs.pdf")), height = 7, width = 10)

table_name <- glue("{current_sub_study}_individual_gene_changes_growth_categories.csv")

my_table <- summaries_df %>%
  mutate(term = str_remove(term, "growth_cat")) %>%
  filter(term %in% c("moderate to severe retardation", "mild retardation"), fdr < 0.05) %>%
  left_join(kegg_defnitions, by = "Ortholog") %>%
  left_join(path_annotation, by = "Ortholog")

write_csv(my_table, here("Output", table_name))

print(glue("All orthologs saved together as {table_name}"))

```

```{r stats2}

kegg_top_500 <- kegg_props %>%
  select(SampleID, Ortholog = geneID, count, props) %>%
  mutate(props_log2 = log2(props)) %>%
  mutate(Ortholog = fct_other(Ortholog, keep = top500$geneID, other_level = "Other")) %>%
  group_by(Ortholog) %>%
  summarise(total_prop = sum(props)) %>%
  ungroup() %>%
  mutate(fraction = total_prop / sum(total_prop))

total_KEGG_abund <- kegg_top_500 %>%
  filter(!Ortholog %in% c("Other")) %>%
  summarise(total_fraction = sum(fraction))


```

The top 500 KEGG terms covered `r total_KEGG_abund$total_fraction*100`% of all KEGG ortholog abundances.

### Butyrate and Aromatic amino acids

Reads from the study were aligned to protein sequences from referenced publications using Diamond. The databases were:

-(Butyrate genes)[https://mbio.asm.org/content/5/2/e00889-14]

-(Aromatic amino acids)[https://pubmed.ncbi.nlm.nih.gov/29168502]

```{r load but_ipa data}

butyrate_info <- read.delim(here::here("Data","butyrate_gene_info_full.tsv"), stringsAsFactors = F) %>%
  select(geneID="gene_name", taxon=Genome.Name, pathway_name)

# but1to3 <- read_gene_aln_results(here("Data","gene_hits", "run1to3", "butyrate_20180612")) %>%
#   left_join(butyrate_info, by=c("geneID", "taxon")) %>%
#   filter(SampleID %in% s_seq$SampleID)
# but4 <- read_gene_aln_results(here("Data","gene_hits", "run4", "butyrate_20180612")) %>%
#   left_join(butyrate_info, by=c("geneID", "taxon")) %>%
#   filter(SampleID %in% s_seq$SampleID)
# but5 <- read_gene_aln_results(here("Data","gene_hits", "run5", "butyrate_20180612")) %>%
#   left_join(butyrate_info, by=c("geneID", "taxon")) %>%
#   filter(SampleID %in% s_seq$SampleID)
# 
# but <- bind_rows(but1to3, but4, but5)
# rm(but1to3, but4, but5)
# 
# ipa1to3 <- read_gene_aln_results(here("Data","gene_hits", "run1to3", "IPA_proteins")) %>%
#   left_join(butyrate_info, by=c("geneID", "taxon")) %>%
#   filter(SampleID %in% s_seq$SampleID)
# ipa4 <- read_gene_aln_results(here("Data","gene_hits", "run4", "IPA_proteins")) %>%
#   left_join(butyrate_info, by=c("geneID", "taxon")) %>%
#   filter(SampleID %in% s_seq$SampleID)
# ipa5 <- read_gene_aln_results(here("Data","gene_hits", "run5", "IPA_proteins")) %>%
#   left_join(butyrate_info, by=c("geneID", "taxon")) %>%
#   filter(SampleID %in% s_seq$SampleID)
# 
# ipa <- bind_rows(ipa1to3, ipa4, ipa5)
# rm(ipa1to3, ipa4, ipa5)
# 
# but_ipa <- bind_rows(but, ipa) %>%
#   right_join(s_seq, by = "SampleID") %>%
#   mutate(props = count / kraken_counts) %>%
#   group_by(SampleID, geneID, database, pathway_name) %>%
#   summarize(count = sum(count), props = sum(props), .groups = "drop") %>%
#   ungroup()
# 
# # replace zeros per gene
# but_ipa %<>%
#   group_by(geneID) %>%
#   filter(sum(count) > 0) %>%
#   mutate(geneID_min = min(props[props > 0])) %>%
#   mutate(props = ifelse(props == 0, geneID_min/2, props)) %>%
#   ungroup() %>%
#   select(-geneID_min) %>%
#   droplevels()
# 
# save(but_ipa, file = here("Data", "but_ipa.RData"))

load(file = here("Data", "but_ipa.RData"))

```

#### Linear models

```{r, results='asis'}

props_toTest <- but_ipa %>%
  mutate(props_log2 = log2(props))

form1 = "props_log2 ~ timepoint_numeric * growth_cat + npo_days_prop + feeds_prop_hm + AbxHoursPerHOL"
random1 = "SubjectID"

summaries_df <- run_lmer_on_props(factor_to_test = "geneID", props_toTest = props_toTest, s_toTest = s_toTest, form1 = form1, p_cutoff = 1, rep_mes_label = random1)

summaries_df %>%
  mutate(term = str_remove(term, "growth_cat")) %>%
  filter(p.value < 0.05) %>%
  pander(split.table=Inf, digits=2, caption = glue("Linear models of log2 transformed abundances"), emphasize.strong.rows = which(.$fdr<=0.05))

table_export(summaries_df %>% filter(p.value < 0.05), "gene_analysis_table_")

```

#### Graphing these changes

```{r}

annot_df <- but_ipa %>%
  select(geneID, database, pathway_name) %>%
  unique() %>%
  mutate(database = str_remove(database, "_20180612|_proteins")) %>%
  mutate(database = ifelse(database %in% "IPA", "AAA", database))

plot_df <- summaries_df %>%
  mutate(term = str_remove(term, "growth_cat")) %>%
  filter(term %in% c("moderate to severe retardation", "mild retardation")) %>%
  left_join(annot_df, by = "geneID") %>%
  mutate(CI95_lwr = round(Value - 1.96*Std.Error, 2)) %>%
  mutate(CI95_upr = round(Value + 1.96*Std.Error, 2)) %>%
  mutate(Significant = case_when(fdr < 0.05 ~ "p.adj < 0.05",
                                 fdr > 0.05 ~ "Not significant")) %>%
  arrange(Value) %>%
  mutate(geneID = fct_reorder(geneID, Value))

g <- ggplot(plot_df, aes(Value, geneID, color = term, shape = Significant)) +
  theme_bw() +
  geom_pointrange(aes(xmin=Value-Std.Error, xmax=Value+Std.Error), position = position_dodge(width = 0.4)) +
  geom_vline(xintercept = 0, linetype = "longdash") +
  scale_shape_manual(values = c(1,19)) +
  scale_color_manual(values=ann_colors$growth_cat[2:3]) +
  facet_grid(database~"Value", scales = "free_y", space = "free_y") +
  labs(y = "geneID",
       x = "<-- Decreased vs. normal growth    Increased vs. normal growth -->",
       color = "Growth category")

plot(g)

ggsave(here("Output",glue("{current_sub_study}_gene_abundaces_but_ipa.pdf")), height=5, width=8, useDingbats=F)

```

```{r}
current_sub_study <- "All timepoints growth failure"
```

## Growth failure

Growth failure is now those babies who are the bottom 3rd percentile in terms of normal growth.

### Bray-curtis PCoA

First, we will examine overall differences using PCoA of the Bray-curtis' distances of the gene abundances between samples.

#### PCoA's

```{r fig.height=5, fig.width=5}
s_toPlot %>%
  pcoaplus(dist_in) %>%
  plot(color=growth_failure) +
    scale_color_manual(values=ann_colors$growth_fail) +
    theme_bw() +
    theme(
      axis.text=element_blank(),
      axis.ticks=element_blank(),
      panel.grid = element_blank(), 
      aspect.ratio = 1,
      strip.background = element_blank()
    ) + 
    labs(color="", title="Bray-Curtis")
```

#### Permanova

```{r, results='asis'}

s_permanova <- s_toTest %>%
  filter(!is.na(feeds_prop_hm))

summaries_df <-
  s_permanova %>%
    do(adonisplus(
      ., distmat = dist_in, formula = distmat ~ npo_days_prop + feeds_prop_hm + AbxHoursPerHOL + timepoint_numeric * growth_cat,
      sample_id_var = SampleID, rep_meas_var = SubjectID,
      shuffle = c(growth_failure = "between", timepoint_numeric = "within"), permutations=perm)) %>%
    ungroup() %>%
    mutate(metric = "Bray-Curtis") %>%
  filter(!term %in% c("Residual", "Total")) %>%
  select(metric, everything())

write_rds(summaries_df, here("Data", "kegg_bc_permanova_results_alldays_2.rds"))

summaries_df <- read_rds(here("Data", "kegg_bc_permanova_results_alldays_2.rds"))

table_export(summaries_df, "gene_analysis_table_")

summaries_df %>%
  mutate(term = str_replace(term, "timepoint_numeric", "Time")) %>%
  mutate(term = str_replace(term, "growth_failure", "Growth failure")) %>%
  pander(split.cells = Inf, digits = 2, emphasize.strong.rows = which(.$p.value<=0.05))

```

### Linear models of pathways

```{r, results='asis'}

number_of_top_pathways = 25

props_toTest <- kegg_path_props %>%
  mutate(props_log2 = log2(props)) %>%
  mutate(PathwayName = fct_lump(PathwayName, n = number_of_top_pathways, w = .$props, other_level = "Other")) %>%
  filter(!PathwayName %in% c("Metabolic pathways", "Other"))

form1 = "props_log2 ~ timepoint_numeric * growth_failure + npo_days_prop + feeds_prop_hm + AbxHoursPerHOL"
random1 = "SubjectID"

summaries_df <- run_lmer_on_props(factor_to_test = "PathwayName", props_toTest = props_toTest, s_toTest = s_toTest, form1 = form1,rep_mes_label = random1, p_cutoff =  0.05) 

summaries_df %>%
  mutate(term = str_remove(term, "growth_failure")) %>%
  pander(split.table=Inf, digits=2, caption = glue("Linear models of log2 transformed abundances of KEGG pathways"), emphasize.strong.rows = which(.$fdr<=0.05))

table_export(summaries_df, "gene_analysis_table_")

```

#### Boxplot of pathway abundances

```{r}

to_select <- unique(as.character(summaries_df$PathwayName))

number_of_blocks <- ceiling(length(to_select) / 9)
chunk_it <- function(x,n) split(x, cut(seq_along(x), n, labels = FALSE))
my_blocks <- chunk_it(to_select, number_of_blocks)

for (i in 1:number_of_blocks) {
  subset <- my_blocks[[i]]

  props_toTest %>%
    filter(PathwayName %in% subset) %>%
    filter(SampleID %in% s_toTest$SampleID) %>%
    merge(s_toTest, by="SampleID") %>%
    
    mutate(PathwayName = reorder(PathwayName, -props)) %>%
    ggplot(aes(x=growth_failure, y=props, color=growth_failure)) +
    geom_boxplot(outlier.alpha = 0) +
    geom_quasirandom(dodge.width = 0.75) +
    theme_bw() +
    theme(
      axis.text.x = element_text(angle=45, vjust=1, hjust=1),
      strip.background = element_blank()) +
    #scale_color_manual(values=brewer.pal(8, "Set1")[c(2,5,1)]) +
    scale_color_brewer(palette = "Set2") +
    scale_y_continuous(trans = "log2") +
    facet_wrap(~PathwayName, scales="free_y", ncol=3) +
    labs(x="", color="Growth failure",
         y="Relative abundance",
         title = "Pathways in samples that had gene abundance differences")
  
  ggsave(here("Output",glue("{current_sub_study}_kegg_pathway_differences_growth_failure_{i}.pdf")), height=7, width=12, useDingbats=F)
}

```

### KEGG orthologs

Taking the top 500 KEGG orthologs and testing them

```{r message=FALSE, warning=FALSE, paged.print=TRUE, results='asis'}

top500 <- kegg_props %>%
  filter(SampleID %in% s_toTest$SampleID) %>%
  group_by(geneID) %>%
  summarise(mean_props = mean(props)) %>%
  arrange(desc(mean_props)) %>%
  slice(1:500)

kegg_props_subset <- kegg_props %>%
  select(SampleID, Ortholog = geneID, count, props) %>%
  filter(Ortholog %in% top500$geneID) %>%
  mutate(props_log2 = log2(props))

form1 = "props_log2 ~ timepoint_numeric * growth_failure + npo_days_prop + feeds_prop_hm + AbxHoursPerHOL"
random1 = "SubjectID"

# summaries_df <- run_lmer_on_props(factor_to_test = "Ortholog", props_toTest = kegg_props_subset, s_toTest = s_toTest, form1 = form1, p_cutoff = 1, rep_mes_label = random1)
# 
# save(summaries_df, file = here("Data", glue("{current_sub_study}_kegg500_changes_gf.RDS")))

load(file = here("Data", glue("{current_sub_study}_kegg500_changes_gf.RDS")))

to_select <- summaries_df %>%
  mutate(term = str_remove(term, "growth_failure")) %>%
  filter(term %in% c("Yes"), fdr < 0.05, Value > 0) %>%
  arrange(desc(Value)) %>%
  slice(1:45) %>%
  pull(Ortholog) %>%
  as.character() %>%
  unique()
  
plot_df <- summaries_df %>%
  mutate(term = str_remove(term, "growth_failure")) %>%
  filter(term %in% c("Yes"), Ortholog %in% to_select) %>%
  left_join(kegg_defnitions, by = "Ortholog") %>%
  mutate(ortho_lable = paste(Ortholog, str_trunc(Definition, 30, "right"))) %>%
  mutate(CI95_lwr = round(Value - 1.96*Std.Error, 2)) %>%
  mutate(CI95_upr = round(Value + 1.96*Std.Error, 2)) %>%
  mutate(Significant = case_when(fdr < 0.05 ~ "p.adj < 0.05",
                                 fdr > 0.05 ~ "Not significant")) %>%
  arrange(Value) %>%
  mutate(ortho_lable = fct_reorder(ortho_lable, Value))

g <- ggplot(plot_df, aes(Value, ortho_lable, color = term, shape = Significant)) +
  theme_bw() +
  geom_pointrange(aes(xmin=Value-Std.Error, xmax=Value+Std.Error), position = position_dodge(width = 0.4)) +
  geom_vline(xintercept = 0, linetype = "longdash") +
  scale_shape_manual(values = c("Not significant" = 1, "p.adj < 0.05" = 19)) +
  scale_color_manual(values = ann_colors$growth_fail[2]) +
  labs(y = "Ortholog / Description",
       x = "<-- Decreased vs. no growth failure    Increased vs. no growth failure -->")

plot(g)

ggsave(here("Output", glue("{current_sub_study}_top_increasing_orthologs_gf.pdf")), height = 7, width = 10)

to_select <- summaries_df %>%
  mutate(term = str_remove(term, "growth_failure")) %>%
  filter(term %in% c("Yes"), fdr < 0.05, Value < 0) %>%
  arrange(desc(Value)) %>%
  slice(1:45) %>%
  pull(Ortholog) %>%
  as.character() %>%
  unique()
  
plot_df <- summaries_df %>%
  mutate(term = str_remove(term, "growth_failure")) %>%
  filter(term %in% c("Yes"), Ortholog %in% to_select) %>%
  left_join(kegg_defnitions, by = "Ortholog") %>%
  mutate(ortho_lable = paste(Ortholog, str_trunc(Definition, 30, "right"))) %>%
  mutate(CI95_lwr = round(Value - 1.96*Std.Error, 2)) %>%
  mutate(CI95_upr = round(Value + 1.96*Std.Error, 2)) %>%
  mutate(Significant = case_when(fdr < 0.05 ~ "p.adj < 0.05",
                                 fdr > 0.05 ~ "Not significant")) %>%
  arrange(Value) %>%
  mutate(ortho_lable = fct_reorder(ortho_lable, Value))

g <- ggplot(plot_df, aes(Value, ortho_lable, color = term, shape = Significant)) +
  theme_bw() +
  geom_pointrange(aes(xmin=Value-Std.Error, xmax=Value+Std.Error), position = position_dodge(width = 0.4)) +
  geom_vline(xintercept = 0, linetype = "longdash") +
  scale_shape_manual(values = c("Not significant" = 1, "p.adj < 0.05" = 19)) +
  scale_color_manual(values = ann_colors$growth_fail[2]) +
  labs(y = "Ortholog / Description",
       x = "<-- Decreased vs. no growth failure    Increased vs. no growth failure -->")

plot(g)

ggsave(here("Output", glue("{current_sub_study}_top_decreasing_orthologs_gf.pdf")), height = 7, width = 10)

table_name <- glue("{current_sub_study}_individual_gene_changes_growth_categories_gf.csv")

my_table <- summaries_df %>%
  mutate(term = str_remove(term, "growth_failure")) %>%
  filter(term %in% c("Yes"), fdr < 0.05) %>%
  left_join(kegg_defnitions, by = "Ortholog") %>%
  left_join(path_annotation, by = "Ortholog")

write_csv(my_table, here("Output", table_name))

print(glue("All orthologs saved together as {table_name}"))

```

### Butyrate and Aromatic amino acids

#### Linear models

```{r, results='asis'}

props_toTest <- but_ipa %>%
  mutate(props_log2 = log2(props))

form1 = "props_log2 ~ timepoint_numeric * growth_failure + npo_days_prop + feeds_prop_hm + AbxHoursPerHOL"
random1 = "SubjectID"

#for whatever reason, I'm getting a "false convergence (8)" error code in nlminb

temp_df <- tibble()

for (selected_gene in unique(but_ipa$geneID)) {

  props_toTest <- but_ipa %>%
  filter(geneID %in% selected_gene) %>%
  mutate(props_log2 = log2(props)) %>%
  inner_join(s_toTest, by = "SampleID")

  possible_error <- try(tiny_df <- tidy_lmer(nlme::lme(as.formula(form1), random = as.formula("~ 1 | SubjectID"), data=props_toTest, na.action=na.omit)) %>% filter(term != '(Intercept)'))

  tiny_df$geneID <- selected_gene

  if (class(possible_error)[1] %in% "try-error") {
    print(glue("Error with {selected_gene}"))
    tiny_df <- tibble()
  }

  temp_df <- bind_rows(temp_df, tiny_df)

}

summaries_df <- temp_df %>%
  group_by(term) %>%
  mutate(fdr = p.adjust(p.value, method="BH")) %>%
  ungroup() %>%
  select(geneID, everything())

summaries_df %>%
  mutate(term = str_remove(term, "growth_failure")) %>%
  filter(p.value < 0.05) %>%
  pander(split.table=Inf, digits=2, caption = glue("Linear models of log2 transformed abundances"), emphasize.strong.rows = which(.$fdr<=0.05))

table_export(summaries_df %>% filter(p.value < 0.05), "gene_analysis_table_")

```

#### Graphing these changes

```{r}

plot_df <- summaries_df %>%
  mutate(term = str_remove(term, "growth_failure")) %>%
  filter(term %in% c("Yes")) %>%
  left_join(annot_df, by = "geneID") %>%
  mutate(CI95_lwr = round(Value - 1.96*Std.Error, 2)) %>%
  mutate(CI95_upr = round(Value + 1.96*Std.Error, 2)) %>%
  mutate(Significant = case_when(fdr < 0.05 ~ "p.adj < 0.05",
                                 fdr > 0.05 ~ "Not significant")) %>%
  arrange(Value) %>%
  mutate(geneID = fct_reorder(geneID, Value))

g <- ggplot(plot_df, aes(Value, geneID, color = term, shape = Significant)) +
  theme_bw() +
  geom_pointrange(aes(xmin=Value-Std.Error, xmax=Value+Std.Error), position = position_dodge(width = 0.4)) +
  geom_vline(xintercept = 0, linetype = "longdash") +
  scale_shape_manual(values = c(1,19)) +
  scale_color_manual(values=ann_colors$growth_fail[2]) +
  facet_grid(database~"Value", scales = "free_y", space = "free_y") +
  labs(y = "geneID",
       x = "<-- Decreased vs. no growth failure    Increased vs. no growth failure -->",
       color = "Growth failure")

plot(g)

ggsave(here("Output",glue("{current_sub_study}_gene_abundaces_but_ipa_gf.pdf")), height=5, width=8, useDingbats=F)

```

```{r}
current_sub_study <- "0to21day"
```

# 0 to 21 day old babies

## Growth faltering

After filtering for > 0 days and < 22 days we are left with these samples per baby:

```{r fig.height=6, fig.width=8}

s_toPlot <- s_toTest <- s2 %>%
  filter(DOL < 22) %>%
  droplevels()

my_df <- s_toPlot %>%
  count(SubjectID) %>%
  arrange(desc(n)) %>%
  rename(`Number of samples` = n)

my_df %>%
  ggplot(aes(x = SubjectID, y = `Number of samples`, fill = `Number of samples`)) +
  geom_col() +
  theme_bw() +
  scale_fill_viridis_c() +
  theme(axis.text.x = element_text(angle = 90, hjust = 1, vjust = 1))

glue("With a total of {length(unique(s_toPlot$SubjectID))} babies.")

```

### Bray-curtis PCoA

First, we will examine overall differences using PCoA of the Bray-curtis' distances of the gene abundances between samples.

#### PCoA's

```{r}

dist_in <- usedist::dist_subset(bc_kegg, s_toPlot$SampleID)

ann_colors = list(
  category = setNames(brewer.pal(length(levels(s_toTest$category)), "Set2"), levels(s_toTest$category)),
  growth_cat = setNames(as.character(ghibli_palette("YesterdayMedium", direction = -1)[1:length(levels(s_toTest$growth_cat))]), levels(s_toTest$growth_cat)),
  TimePoint = setNames(viridis(length(levels(s_toTest$TimePoint))), levels(s_toTest$TimePoint)),
  growth_fail = setNames(qualpal(2)$hex, levels(s_toTest$growth_failure))
)

```

```{r fig.height=5, fig.width=5}
s_toPlot %>%
  pcoaplus(dist_in) %>%
  plot(color=TimePoint) +
    scale_color_manual(values=ann_colors$TimePoint) +
    theme_bw() +
    theme(
      axis.text=element_blank(),
      axis.ticks=element_blank(),
      panel.grid = element_blank(), 
      aspect.ratio = 1,
      strip.background = element_blank()
    ) + 
    labs(color="", title="Bray-Curtis")
```

```{r fig.height=5, fig.width=5}
s_toPlot %>%
  pcoaplus(dist_in) %>%
  plot(color=growth_cat) +
    scale_color_manual(values=ann_colors$growth_cat) +
    theme_bw() +
    theme(
      axis.text=element_blank(),
      axis.ticks=element_blank(),
      panel.grid = element_blank(), 
      aspect.ratio = 1,
      strip.background = element_blank()
    ) + 
    labs(color="", title="Bray-Curtis")
```

#### Permanova

```{r, results='asis'}

s_permanova <- s_toTest %>%
  filter(!is.na(feeds_prop_hm))

summaries_df <-
  s_permanova %>%
    do(adonisplus(
      ., distmat = dist_in, formula = distmat ~ npo_days_prop + feeds_prop_hm + AbxHoursPerHOL + timepoint_numeric * growth_cat,
      sample_id_var = SampleID, rep_meas_var = SubjectID,
      shuffle = c(growth_cat = "between", timepoint_numeric = "within"), permutations=perm)) %>%
    ungroup() %>%
    mutate(metric = "Bray-Curtis") %>%
  filter(!term %in% c("Residual", "Total")) %>%
  select(metric, everything())

write_rds(summaries_df, here("Data", "kegg_bc_permanova_results_0to21days.rds"))

summaries_df <- read_rds(here("Data", "kegg_bc_permanova_results_0to21days.rds"))

summaries_df %>%
  mutate(term = str_replace(term, "timepoint_numeric", "Time")) %>%
  mutate(term = str_replace(term, "growth_cat", "Growth category")) %>%
  pander(split.cells = Inf, digits = 2, emphasize.strong.rows = which(.$p.value<=0.05))

table_export(summaries_df, "gene_analysis_table_")

```

### Linear models of pathways

```{r, results='asis'}

number_of_top_pathways = 25

props_toTest <- kegg_path_props %>%
  mutate(props_log2 = log2(props)) %>%
  mutate(PathwayName = fct_lump(PathwayName, n = number_of_top_pathways, w = .$props, other_level = "Other")) %>%
  filter(!PathwayName %in% c("Metabolic pathways", "Other"))

form1 = "props_log2 ~ timepoint_numeric * growth_cat + npo_days_prop + feeds_prop_hm + AbxHoursPerHOL"
random1 = "SubjectID"

summaries_df <- run_lmer_on_props(factor_to_test = "PathwayName", props_toTest = props_toTest, s_toTest = s_toTest, form1 = form1,rep_mes_label = random1, p_cutoff =  0.05) 

summaries_df %>%
  mutate(term = str_remove(term, "growth_cat")) %>%
  pander(split.table=Inf, digits=2, caption = glue("Linear models of log2 transformed abundances of KEGG pathways"), emphasize.strong.rows = which(.$fdr<=0.05))

table_export(summaries_df, "gene_analysis_table_")

```

#### Boxplot of pathway abundances

```{r}

to_select <- unique(as.character(summaries_df$PathwayName))

number_of_blocks <- ceiling(length(to_select) / 9)
chunk_it <- function(x,n) split(x, cut(seq_along(x), n, labels = FALSE))
my_blocks <- chunk_it(to_select, number_of_blocks)

for (i in 1:number_of_blocks) {
  subset <- my_blocks[[i]]

  props_toTest %>%
    filter(PathwayName %in% subset) %>%
    filter(SampleID %in% s_toTest$SampleID) %>%
    merge(s_toTest, by="SampleID") %>%
    
    mutate(PathwayName = reorder(PathwayName, -props)) %>%
    ggplot(aes(x=growth_cat, y=props, color=growth_cat)) +
    geom_boxplot(outlier.alpha = 0) +
    geom_quasirandom(dodge.width = 0.75) +
    theme_bw() +
    theme(
      axis.text.x = element_text(angle=45, vjust=1, hjust=1),
      strip.background = element_blank()) +
    #scale_color_manual(values=brewer.pal(8, "Set1")[c(2,5,1)]) +
    scale_color_brewer(palette = "Set2") +
    scale_y_continuous(trans = "log2") +
    facet_wrap(~PathwayName, scales="free_y", ncol=3) +
    labs(x="", color="growth_cat",
         y="Relative abundance",
         title = "Pathways in samples that had gene abundance differences")
  
  ggsave(here("Output",glue("{current_sub_study}_kegg_pathway_differences_growth_cat_{i}.pdf")), height=7, width=12, useDingbats=F)
}

```

### KEGG orthologs

Taking the top 500 KEGG orthologs and testing them

```{r message=FALSE, warning=FALSE, paged.print=TRUE, results='asis'}

top500 <- kegg_props %>%
  filter(SampleID %in% s_toTest$SampleID) %>%
  group_by(geneID) %>%
  summarise(mean_props = mean(props)) %>%
  arrange(desc(mean_props)) %>%
  slice(1:500)

kegg_props_subset <- kegg_props %>%
  select(SampleID, Ortholog = geneID, count, props) %>%
  filter(Ortholog %in% top500$geneID) %>%
  mutate(props_log2 = log2(props))

form1 = "props_log2 ~ timepoint_numeric * growth_cat + npo_days_prop + feeds_prop_hm + AbxHoursPerHOL"
random1 = "SubjectID"

# summaries_df <- run_lmer_on_props(factor_to_test = "Ortholog", props_toTest = kegg_props_subset, s_toTest = s_toTest, form1 = form1, p_cutoff = 1, rep_mes_label = random1)
# 
# save(summaries_df, file = here("Data", glue("{current_sub_study}_kegg500_changes.RDS")))

load(file = here("Data", glue("{current_sub_study}_kegg500_changes.RDS")))

to_select <- summaries_df %>%
  mutate(term = str_remove(term, "growth_cat")) %>%
  filter(term %in% c("moderate to severe retardation", "mild retardation"), fdr < 0.05, Value > 0) %>%
  arrange(desc(Value)) %>%
  slice(1:45) %>%
  pull(Ortholog) %>%
  as.character() %>%
  unique()
#actually top 32
  
plot_df <- summaries_df %>%
  mutate(term = str_remove(term, "growth_cat")) %>%
  filter(term %in% c("moderate to severe retardation", "mild retardation"), Ortholog %in% to_select) %>%
  left_join(kegg_defnitions, by = "Ortholog") %>%
  mutate(ortho_lable = paste(Ortholog, str_trunc(Definition, 30, "right"))) %>%
  mutate(CI95_lwr = round(Value - 1.96*Std.Error, 2)) %>%
  mutate(CI95_upr = round(Value + 1.96*Std.Error, 2)) %>%
  mutate(Significant = case_when(fdr < 0.05 ~ "p.adj < 0.05",
                                 fdr > 0.05 ~ "Not significant")) %>%
  arrange(Value) %>%
  mutate(ortho_lable = fct_reorder(ortho_lable, Value))

g <- ggplot(plot_df, aes(Value, ortho_lable, color = term, shape = Significant)) +
  theme_bw() +
  geom_pointrange(aes(xmin=Value-Std.Error, xmax=Value+Std.Error), position = position_dodge(width = 0.4)) +
  geom_vline(xintercept = 0, linetype = "longdash") +
  scale_shape_manual(values = c(1,19)) +
  scale_color_manual(values=ann_colors$growth_cat[2:3]) +
  labs(y = "Ortholog / Description",
       x = "<-- Decreased vs. normal growth    Increased vs. normal growth -->",
       color = "Growth category")

plot(g)

ggsave(here("Output", glue("{current_sub_study}_top_increasing_orthologs.pdf")), height = 7, width = 10)

to_select <- summaries_df %>%
  mutate(term = str_remove(term, "growth_cat")) %>%
  filter(term %in% c("moderate to severe retardation", "mild retardation"), fdr < 0.05, Value < 0) %>%
  arrange(desc(Value)) %>%
  slice(1:45) %>%
  pull(Ortholog) %>%
  as.character() %>%
  unique()
#actually top 32
  
plot_df <- summaries_df %>%
  mutate(term = str_remove(term, "growth_cat")) %>%
  filter(term %in% c("moderate to severe retardation", "mild retardation"), Ortholog %in% to_select) %>%
  left_join(kegg_defnitions, by = "Ortholog") %>%
  mutate(ortho_lable = paste(Ortholog, str_trunc(Definition, 30, "right"))) %>%
  mutate(CI95_lwr = round(Value - 1.96*Std.Error, 2)) %>%
  mutate(CI95_upr = round(Value + 1.96*Std.Error, 2)) %>%
  mutate(Significant = case_when(fdr < 0.05 ~ "p.adj < 0.05",
                                 fdr > 0.05 ~ "Not significant")) %>%
  arrange(Value) %>%
  mutate(ortho_lable = fct_reorder(ortho_lable, Value))

g <- ggplot(plot_df, aes(Value, ortho_lable, color = term, shape = Significant)) +
  theme_bw() +
  geom_pointrange(aes(xmin=Value-Std.Error, xmax=Value+Std.Error), position = position_dodge(width = 0.4)) +
  geom_vline(xintercept = 0, linetype = "longdash") +
  scale_shape_manual(values = c(1,19)) +
  scale_color_manual(values=ann_colors$growth_cat[2:3]) +
  labs(y = "Ortholog / Description",
       x = "<-- Decreased vs. normal growth    Increased vs. normal growth -->",
       color = "Growth category")

plot(g)

ggsave(here("Output", glue("{current_sub_study}_top_decreasing_orthologs.pdf")), height = 7, width = 10)

table_name <- glue("{current_sub_study}_individual_gene_changes_growth_categories.csv")

my_table <- summaries_df %>%
  mutate(term = str_remove(term, "growth_cat")) %>%
  filter(term %in% c("moderate to severe retardation", "mild retardation"), fdr < 0.05) %>%
  left_join(kegg_defnitions, by = "Ortholog") %>%
  left_join(path_annotation, by = "Ortholog")

write_csv(my_table, here("Output", table_name))

print(glue("All orthologs saved together as {table_name}"))

```

### Butyrate and Aromatic amino acids

#### Linear models

```{r, results='asis'}

props_toTest <- but_ipa %>%
  mutate(props_log2 = log2(props))

form1 = "props_log2 ~ timepoint_numeric * growth_cat + npo_days_prop + feeds_prop_hm + AbxHoursPerHOL"
random1 = "SubjectID"

#for whatever reason, I'm getting a "false convergence (8)" error code in nlminb

temp_df <- tibble()

for (selected_gene in unique(but_ipa$geneID)) {

  props_toTest <- but_ipa %>%
  filter(geneID %in% selected_gene) %>%
  mutate(props_log2 = log2(props)) %>%
  inner_join(s_toTest, by = "SampleID")

  possible_error <- try(tiny_df <- tidy_lmer(nlme::lme(as.formula(form1), random = as.formula("~ 1 | SubjectID"), data=props_toTest, na.action=na.omit)) %>% filter(term != '(Intercept)'))

  tiny_df$geneID <- selected_gene

  if (class(possible_error)[1] %in% "try-error") {
    print(glue("Error with {selected_gene}"))
    tiny_df <- tibble()
  }

  temp_df <- bind_rows(temp_df, tiny_df)

}

summaries_df <- temp_df %>%
  group_by(term) %>%
  mutate(fdr = p.adjust(p.value, method="BH")) %>%
  ungroup() %>%
  select(geneID, everything())

summaries_df %>%
  mutate(term = str_remove(term, "growth_cat")) %>%
  filter(p.value < 0.05) %>%
  pander(split.table=Inf, digits=2, caption = glue("Linear models of log2 transformed abundances"), emphasize.strong.rows = which(.$fdr<=0.05))

table_export(summaries_df %>% filter(p.value < 0.05), "gene_analysis_table_")

```

#### Graphing these changes

```{r}

plot_df <- summaries_df %>%
  mutate(term = str_remove(term, "growth_cat")) %>%
  filter(term %in% c("moderate to severe retardation", "mild retardation")) %>%
  left_join(annot_df, by = "geneID") %>%
  mutate(CI95_lwr = round(Value - 1.96*Std.Error, 2)) %>%
  mutate(CI95_upr = round(Value + 1.96*Std.Error, 2)) %>%
  mutate(Significant = case_when(fdr < 0.05 ~ "p.adj < 0.05",
                                 fdr > 0.05 ~ "Not significant")) %>%
  arrange(Value) %>%
  mutate(geneID = fct_reorder(geneID, Value))

g <- ggplot(plot_df, aes(Value, geneID, color = term, shape = Significant)) +
  theme_bw() +
  geom_pointrange(aes(xmin=Value-Std.Error, xmax=Value+Std.Error), position = position_dodge(width = 0.4)) +
  geom_vline(xintercept = 0, linetype = "longdash") +
  scale_shape_manual(values = c(1,19)) +
  scale_color_manual(values=ann_colors$growth_cat[2:3]) +
  facet_grid(database~"Value", scales = "free_y", space = "free_y") +
  labs(y = "geneID",
       x = "<-- Decreased vs. normal growth    Increased vs. normal growth -->",
       color = "Growth category")

plot(g)

ggsave(here("Output",glue("{current_sub_study}_gene_abundaces_but_ipa.pdf")), height=5, width=8, useDingbats=F)

```

## Growth failure

Growth failure is now those babies who are the bottom 3rd percentile in terms of normal growth.

### Bray-curtis PCoA

First, we will examine overall differences using PCoA of the Bray-curtis' distances of the gene abundances between samples.

#### PCoA's

```{r fig.height=5, fig.width=5}
s_toPlot %>%
  pcoaplus(dist_in) %>%
  plot(color=growth_failure) +
    scale_color_manual(values=ann_colors$growth_fail) +
    theme_bw() +
    theme(
      axis.text=element_blank(),
      axis.ticks=element_blank(),
      panel.grid = element_blank(), 
      aspect.ratio = 1,
      strip.background = element_blank()
    ) + 
    labs(color="", title="Bray-Curtis")
```

#### Permanova

```{r, results='asis'}

s_permanova <- s_toTest %>%
  filter(!is.na(feeds_prop_hm))

summaries_df <-
  s_permanova %>%
    do(adonisplus(
      ., distmat = dist_in, formula = distmat ~ npo_days_prop + feeds_prop_hm + AbxHoursPerHOL + timepoint_numeric * growth_cat,
      sample_id_var = SampleID, rep_meas_var = SubjectID,
      shuffle = c(growth_failure = "between", timepoint_numeric = "within"), permutations=perm)) %>%
    ungroup() %>%
    mutate(metric = "Bray-Curtis") %>%
  filter(!term %in% c("Residual", "Total")) %>%
  select(metric, everything())

write_rds(summaries_df, here("Data", "kegg_bc_permanova_results_0to21days_2.rds"))

summaries_df <- read_rds(here("Data", "kegg_bc_permanova_results_0to21days_2.rds"))

table_export(summaries_df, "gene_analysis_table_")

summaries_df %>%
  mutate(term = str_replace(term, "timepoint_numeric", "Time")) %>%
  mutate(term = str_replace(term, "growth_failure", "Growth failure")) %>%
  pander(split.cells = Inf, digits = 2, emphasize.strong.rows = which(.$p.value<=0.05))

```

### Linear models of pathways

```{r, results='asis'}

number_of_top_pathways = 25

props_toTest <- kegg_path_props %>%
  mutate(props_log2 = log2(props)) %>%
  mutate(PathwayName = fct_lump(PathwayName, n = number_of_top_pathways, w = .$props, other_level = "Other")) %>%
  filter(!PathwayName %in% c("Metabolic pathways", "Other"))

form1 = "props_log2 ~ timepoint_numeric * growth_failure + npo_days_prop + feeds_prop_hm + AbxHoursPerHOL"
random1 = "SubjectID"

summaries_df <- run_lmer_on_props(factor_to_test = "PathwayName", props_toTest = props_toTest, s_toTest = s_toTest, form1 = form1,rep_mes_label = random1, p_cutoff =  0.05) 

summaries_df %>%
  mutate(term = str_remove(term, "growth_failure")) %>%
  pander(split.table=Inf, digits=2, caption = glue("Linear models of log2 transformed abundances of KEGG pathways"), emphasize.strong.rows = which(.$fdr<=0.05))

table_export(summaries_df, "gene_analysis_table_")

```

#### Boxplot of pathway abundances

```{r}

to_select <- unique(as.character(summaries_df$PathwayName))

number_of_blocks <- ceiling(length(to_select) / 9)
chunk_it <- function(x,n) split(x, cut(seq_along(x), n, labels = FALSE))
my_blocks <- chunk_it(to_select, number_of_blocks)

for (i in 1:number_of_blocks) {
  subset <- my_blocks[[i]]

  props_toTest %>%
    filter(PathwayName %in% subset) %>%
    filter(SampleID %in% s_toTest$SampleID) %>%
    merge(s_toTest, by="SampleID") %>%
    
    mutate(PathwayName = reorder(PathwayName, -props)) %>%
    ggplot(aes(x=growth_failure, y=props, color=growth_failure)) +
    geom_boxplot(outlier.alpha = 0) +
    geom_quasirandom(dodge.width = 0.75) +
    theme_bw() +
    theme(
      axis.text.x = element_text(angle=45, vjust=1, hjust=1),
      strip.background = element_blank()) +
    #scale_color_manual(values=brewer.pal(8, "Set1")[c(2,5,1)]) +
    scale_color_brewer(palette = "Set2") +
    scale_y_continuous(trans = "log2") +
    facet_wrap(~PathwayName, scales="free_y", ncol=3) +
    labs(x="", color="Growth failure",
         y="Relative abundance",
         title = "Pathways in samples that had gene abundance differences")
  
  ggsave(here("Output",glue("{current_sub_study}_kegg_pathway_differences_growth_failure_{i}.pdf")), height=7, width=12, useDingbats=F)
}

```

### KEGG orthologs

Taking the top 500 KEGG orthologs and testing them

```{r message=FALSE, warning=FALSE, paged.print=TRUE, results='asis'}

top500 <- kegg_props %>%
  filter(SampleID %in% s_toTest$SampleID) %>%
  group_by(geneID) %>%
  summarise(mean_props = mean(props)) %>%
  arrange(desc(mean_props)) %>%
  slice(1:500)

kegg_props_subset <- kegg_props %>%
  select(SampleID, Ortholog = geneID, count, props) %>%
  filter(Ortholog %in% top500$geneID) %>%
  mutate(props_log2 = log2(props))

form1 = "props_log2 ~ timepoint_numeric * growth_failure + npo_days_prop + feeds_prop_hm + AbxHoursPerHOL"
random1 = "SubjectID"

# summaries_df <- run_lmer_on_props(factor_to_test = "Ortholog", props_toTest = kegg_props_subset, s_toTest = s_toTest, form1 = form1, p_cutoff = 1, rep_mes_label = random1)
# 
# save(summaries_df, file = here("Data", glue("{current_sub_study}_kegg500_changes_gf.RDS")))

load(file = here("Data", glue("{current_sub_study}_kegg500_changes_gf.RDS")))

to_select <- summaries_df %>%
  mutate(term = str_remove(term, "growth_failure")) %>%
  filter(term %in% c("Yes"), fdr < 0.05, Value > 0) %>%
  arrange(desc(Value)) %>%
  slice(1:45) %>%
  pull(Ortholog) %>%
  as.character() %>%
  unique()
  
plot_df <- summaries_df %>%
  mutate(term = str_remove(term, "growth_failure")) %>%
  filter(term %in% c("Yes"), Ortholog %in% to_select) %>%
  left_join(kegg_defnitions, by = "Ortholog") %>%
  mutate(ortho_lable = paste(Ortholog, str_trunc(Definition, 30, "right"))) %>%
  mutate(CI95_lwr = round(Value - 1.96*Std.Error, 2)) %>%
  mutate(CI95_upr = round(Value + 1.96*Std.Error, 2)) %>%
  mutate(Significant = case_when(fdr < 0.05 ~ "p.adj < 0.05",
                                 fdr > 0.05 ~ "Not significant")) %>%
  arrange(Value) %>%
  mutate(ortho_lable = fct_reorder(ortho_lable, Value))

g <- ggplot(plot_df, aes(Value, ortho_lable, color = term, shape = Significant)) +
  theme_bw() +
  geom_pointrange(aes(xmin=Value-Std.Error, xmax=Value+Std.Error), position = position_dodge(width = 0.4)) +
  geom_vline(xintercept = 0, linetype = "longdash") +
  scale_shape_manual(values = c("Not significant" = 1, "p.adj < 0.05" = 19)) +
  scale_color_manual(values = ann_colors$growth_fail[2]) +
  labs(y = "Ortholog / Description",
       x = "<-- Decreased vs. no growth failure    Increased vs. no growth failure -->")

plot(g)

ggsave(here("Output", glue("{current_sub_study}_top_increasing_orthologs_gf.pdf")), height = 7, width = 10)

to_select <- summaries_df %>%
  mutate(term = str_remove(term, "growth_failure")) %>%
  filter(term %in% c("Yes"), fdr < 0.05, Value < 0) %>%
  arrange(desc(Value)) %>%
  slice(1:45) %>%
  pull(Ortholog) %>%
  as.character() %>%
  unique()
  
plot_df <- summaries_df %>%
  mutate(term = str_remove(term, "growth_failure")) %>%
  filter(term %in% c("Yes"), Ortholog %in% to_select) %>%
  left_join(kegg_defnitions, by = "Ortholog") %>%
  mutate(ortho_lable = paste(Ortholog, str_trunc(Definition, 30, "right"))) %>%
  mutate(CI95_lwr = round(Value - 1.96*Std.Error, 2)) %>%
  mutate(CI95_upr = round(Value + 1.96*Std.Error, 2)) %>%
  mutate(Significant = case_when(fdr < 0.05 ~ "p.adj < 0.05",
                                 fdr > 0.05 ~ "Not significant")) %>%
  arrange(Value) %>%
  mutate(ortho_lable = fct_reorder(ortho_lable, Value))

g <- ggplot(plot_df, aes(Value, ortho_lable, color = term, shape = Significant)) +
  theme_bw() +
  geom_pointrange(aes(xmin=Value-Std.Error, xmax=Value+Std.Error), position = position_dodge(width = 0.4)) +
  geom_vline(xintercept = 0, linetype = "longdash") +
  scale_shape_manual(values = c("Not significant" = 1, "p.adj < 0.05" = 19)) +
  scale_color_manual(values = ann_colors$growth_fail[2]) +
  labs(y = "Ortholog / Description",
       x = "<-- Decreased vs. no growth failure    Increased vs. no growth failure -->")

plot(g)

ggsave(here("Output", glue("{current_sub_study}_top_decreasing_orthologs_gf.pdf")), height = 7, width = 10)

table_name <- glue("{current_sub_study}_individual_gene_changes_growth_categories_gf.csv")

my_table <- summaries_df %>%
  mutate(term = str_remove(term, "growth_failure")) %>%
  filter(term %in% c("Yes"), fdr < 0.05) %>%
  left_join(kegg_defnitions, by = "Ortholog") %>%
  left_join(path_annotation, by = "Ortholog")

write_csv(my_table, here("Output", table_name))

print(glue("All orthologs saved together as {table_name}"))

```

### Butyrate and Aromatic amino acids

#### Linear models

```{r, results='asis'}

props_toTest <- but_ipa %>%
  mutate(props_log2 = log2(props))

form1 = "props_log2 ~ timepoint_numeric * growth_failure + npo_days_prop + feeds_prop_hm + AbxHoursPerHOL"
random1 = "SubjectID"

#for whatever reason, I'm getting a "false convergence (8)" error code in nlminb

temp_df <- tibble()

for (selected_gene in unique(but_ipa$geneID)) {

  props_toTest <- but_ipa %>%
  filter(geneID %in% selected_gene) %>%
  mutate(props_log2 = log2(props)) %>%
  inner_join(s_toTest, by = "SampleID")

  possible_error <- try(tiny_df <- tidy_lmer(nlme::lme(as.formula(form1), random = as.formula("~ 1 | SubjectID"), data=props_toTest, na.action=na.omit)) %>% filter(term != '(Intercept)'))

  tiny_df$geneID <- selected_gene

  if (class(possible_error)[1] %in% "try-error") {
    print(glue("Error with {selected_gene}"))
    tiny_df <- tibble()
  }

  temp_df <- bind_rows(temp_df, tiny_df)

}

summaries_df <- temp_df %>%
  group_by(term) %>%
  mutate(fdr = p.adjust(p.value, method="BH")) %>%
  ungroup() %>%
  select(geneID, everything())

summaries_df %>%
  mutate(term = str_remove(term, "growth_failure")) %>%
  filter(p.value < 0.05) %>%
  pander(split.table=Inf, digits=2, caption = glue("Linear models of log2 transformed abundances"), emphasize.strong.rows = which(.$fdr<=0.05))

table_export(summaries_df %>% filter(p.value < 0.05), "gene_analysis_table_")

```

#### Graphing these changes

```{r}

plot_df <- summaries_df %>%
  mutate(term = str_remove(term, "growth_failure")) %>%
  filter(term %in% c("Yes")) %>%
  left_join(annot_df, by = "geneID") %>%
  mutate(CI95_lwr = round(Value - 1.96*Std.Error, 2)) %>%
  mutate(CI95_upr = round(Value + 1.96*Std.Error, 2)) %>%
  mutate(Significant = case_when(fdr < 0.05 ~ "p.adj < 0.05",
                                 fdr > 0.05 ~ "Not significant")) %>%
  arrange(Value) %>%
  mutate(geneID = fct_reorder(geneID, Value))

g <- ggplot(plot_df, aes(Value, geneID, color = term, shape = Significant)) +
  theme_bw() +
  geom_pointrange(aes(xmin=Value-Std.Error, xmax=Value+Std.Error), position = position_dodge(width = 0.4)) +
  geom_vline(xintercept = 0, linetype = "longdash") +
  scale_shape_manual(values = c(1,19)) +
  scale_color_manual(values=ann_colors$growth_fail[2]) +
  facet_grid(database~"Value", scales = "free_y", space = "free_y") +
  labs(y = "geneID",
       x = "<-- Decreased vs. no growth failure    Increased vs. no growth failure -->",
       color = "Growth failure")

plot(g)

ggsave(here("Output",glue("{current_sub_study}_gene_abundaces_but_ipa_gf.pdf")), height=5, width=8, useDingbats=F)

```


```{r}
current_sub_study <- "34to37weeks"
```

# Post-menstrual age > 34 weeks and < 37 weeks

In this section, all of the statistics that include day-of-life (DOL) or hour-of-life (HOL) will instead be analyzed using post-menstrual age (PMA) which is equal to `Gestation in Weeks + (DOL/7)`. We are also going to restrict the analysis to less than 37 weeks and greater than 34 weeks with one sample per baby.

```{r}

s2 %<>%
  mutate(PMA = (DOL/7) + Gestation)

s2 %>%
  ggplot(aes(x = PMA, fill = WOL_label)) +
  geom_histogram(binwidth = 1, center = 0) +
  geom_vline(xintercept = 36, lty = 2) +
  labs(caption = "Dashed line represents 36 weeks PMA")

# check
# s2 %>%
#   filter(PMA > 36) %>%
#   select(SampleID, Zdec36_gt0.8, Zdec36_gt1.2, Zdec36_gt2, growth_cat) %>%
#   view()

```

Subset:

```{r}

s_subset2 <- read_rds(here("Data", "subset_of_babies_great34wk_less37wk.rds"))

```

After filtering for > 34 weeks and < 37 weeks PMA we are left with these samples per baby:

```{r fig.height=6, fig.width=8}

s_toPlot <- s_toTest <- s2 %>%
  filter(SampleID %in% s_subset2$SampleID) %>%
  droplevels()

#subsetting timepoint colors for nicer legends
ann_colors = list(
  category = setNames(brewer.pal(length(levels(s_toTest$category)), "Set2"), levels(s_toTest$category)),
  growth_cat = setNames(as.character(ghibli_palette("YesterdayMedium", direction = -1)[1:length(levels(s_toTest$growth_cat))]), levels(s_toTest$growth_cat)),
  TimePoint = setNames(viridis(length(levels(s_toTest$TimePoint))), levels(s_toTest$TimePoint)),
  growth_fail = setNames(qualpal(2)$hex, levels(s_toTest$growth_failure))
)
ann_colors$TimePoint <- ann_colors$TimePoint[names(ann_colors$TimePoint) %in% unique(s_toPlot$TimePoint)]

my_df <- s_toPlot %>%
  count(SubjectID) %>%
  arrange(desc(n)) %>%
  rename(`Number of samples` = n)

my_df %>%
  ggplot(aes(x = SubjectID, y = `Number of samples`, fill = `Number of samples`)) +
  geom_col() +
  theme_bw() +
  scale_fill_viridis_c() +
  theme(axis.text.x = element_text(angle = 90, hjust = 1, vjust = 1))

glue("With a total of {length(unique(s_toPlot$SubjectID))} babies.")

```

## Growth faltering

### Bray-curtis PCoA

First, we will examine overall differences using PCoA of the Bray-curtis' distances of the gene abundances between samples.

#### PCoA's

```{r}

dist_in <- usedist::dist_subset(bc_kegg, s_toPlot$SampleID)

```

```{r fig.height=5, fig.width=5}
s_toPlot %>%
  pcoaplus(dist_in) %>%
  plot(color=TimePoint) +
    scale_color_manual(values=ann_colors$TimePoint) +
    theme_bw() +
    theme(
      axis.text=element_blank(),
      axis.ticks=element_blank(),
      panel.grid = element_blank(), 
      aspect.ratio = 1,
      strip.background = element_blank()
    ) + 
    labs(color="", title="Bray-Curtis")
```

```{r fig.height=5, fig.width=5}
s_toPlot %>%
  pcoaplus(dist_in) %>%
  plot(color=growth_cat) +
    scale_color_manual(values=ann_colors$growth_cat) +
    theme_bw() +
    theme(
      axis.text=element_blank(),
      axis.ticks=element_blank(),
      panel.grid = element_blank(), 
      aspect.ratio = 1,
      strip.background = element_blank()
    ) + 
    labs(color="", title="Bray-Curtis")
```

#### Permanova

```{r, results='asis'}
s_permanova <- s_toTest %>%
  filter(!is.na(feeds_prop_hm))

summaries_df <-
  s_permanova %>%
    do(adonisplus(
      ., distmat = dist_in, formula = distmat ~ npo_days_prop + feeds_prop_hm + AbxHoursPerHOL + timepoint_numeric * growth_cat,
      sample_id_var = SampleID, rep_meas_var = SubjectID,
      shuffle = c(growth_cat = "between", timepoint_numeric = "within"), permutations=perm)) %>%
    ungroup() %>%
    mutate(metric = "Bray-Curtis") %>%
  filter(!term %in% c("Residual", "Total")) %>%
  select(metric, everything())

write_rds(summaries_df, here("Data", "kegg_bc_permanova_results_34to37weeks.rds"))

summaries_df <- read_rds(here("Data", "kegg_bc_permanova_results_34to37weeks.rds"))

summaries_df %>%
  mutate(term = str_replace(term, "timepoint_numeric", "Time")) %>%
  mutate(term = str_replace(term, "growth_cat", "Growth category")) %>%
  pander(split.cells = Inf, digits = 2, emphasize.strong.rows = which(.$p.value<=0.05))

table_export(summaries_df, "gene_analysis_table_")

```

### Linear models of pathways

```{r, results='asis'}

number_of_top_pathways = 25

props_toTest <- kegg_path_props %>%
  mutate(props_log2 = log2(props)) %>%
  mutate(PathwayName = fct_lump(PathwayName, n = number_of_top_pathways, w = .$props, other_level = "Other")) %>%
  filter(!PathwayName %in% c("Metabolic pathways", "Other"))

form1 = "props_log2 ~ timepoint_numeric * growth_cat + npo_days_prop + feeds_prop_hm + AbxHoursPerHOL"
random1 = "SubjectID"

summaries_df <- run_lmer_on_props(factor_to_test = "PathwayName", props_toTest = props_toTest, s_toTest = s_toTest, form1 = form1,rep_mes_label = random1, p_cutoff =  0.05) 

summaries_df %>%
  mutate(term = str_remove(term, "growth_cat")) %>%
  pander(split.table=Inf, digits=2, caption = glue("Linear models of log2 transformed abundances of KEGG pathways"), emphasize.strong.rows = which(.$fdr<=0.05))

table_export(summaries_df, "gene_analysis_table_")

```

#### Boxplot of pathway abundances

```{r}

to_select <- unique(as.character(summaries_df$PathwayName))

number_of_blocks <- ceiling(length(to_select) / 9)
chunk_it <- function(x,n) split(x, cut(seq_along(x), n, labels = FALSE))
my_blocks <- chunk_it(to_select, number_of_blocks)

for (i in 1:number_of_blocks) {
  subset <- my_blocks[[i]]

  props_toTest %>%
    filter(PathwayName %in% subset) %>%
    filter(SampleID %in% s_toTest$SampleID) %>%
    merge(s_toTest, by="SampleID") %>%
    
    mutate(PathwayName = reorder(PathwayName, -props)) %>%
    ggplot(aes(x=growth_cat, y=props, color=growth_cat)) +
    geom_boxplot(outlier.alpha = 0) +
    geom_quasirandom(dodge.width = 0.75) +
    theme_bw() +
    theme(
      axis.text.x = element_text(angle=45, vjust=1, hjust=1),
      strip.background = element_blank()) +
    #scale_color_manual(values=brewer.pal(8, "Set1")[c(2,5,1)]) +
    scale_color_brewer(palette = "Set2") +
    scale_y_continuous(trans = "log2") +
    facet_wrap(~PathwayName, scales="free_y", ncol=3) +
    labs(x="", color="growth_cat",
         y="Relative abundance",
         title = "Pathways in samples that had gene abundance differences")
  
  ggsave(here("Output",glue("{current_sub_study}_kegg_pathway_differences_growth_cat_{i}.pdf")), height=7, width=12, useDingbats=F)
}

```

#### Those that changed

```{r}

pathways_to_select <- summaries_df %>%
  mutate(term = str_remove(term, "growth_cat")) %>%
  filter(term %in% c("moderate to severe retardation", "mild retardation")) %>%
  pull(PathwayName) %>%
  unique()

summaries_df <- run_lmer_on_props(factor_to_test = "PathwayName", props_toTest = props_toTest, s_toTest = s_toTest, form1 = form1,rep_mes_label = random1, p_cutoff =  1)

plot_df <- summaries_df %>%
  mutate(term = str_remove(term, "growth_cat")) %>%
  filter(term %in% c("moderate to severe retardation", "mild retardation"), PathwayName %in% pathways_to_select) %>%
  mutate(term = ifelse(term %in% "mild retardation", "mild", term)) %>%
  mutate(term = ifelse(term %in% "moderate to severe retardation", "moderate to\nsevere", term)) %>%
  mutate(CI95_lwr = round(Value - 1.96*Std.Error, 2)) %>%
  mutate(CI95_upr = round(Value + 1.96*Std.Error, 2)) %>%
  mutate(Significant = case_when(fdr < 0.05 ~ "p.adj < 0.05",
                                 fdr > 0.05 ~ "Not significant")) %>%
  arrange(Value) %>%
  mutate(PathwayName = fct_reorder(PathwayName, Value))



g <- ggplot(plot_df, aes(Value, PathwayName, color = term, shape = Significant)) +
  theme_bw() +
  geom_pointrange(aes(xmin=Value-Std.Error, xmax=Value+Std.Error), position = position_dodge(width = 0.4)) +
  geom_vline(xintercept = 0, linetype = "longdash") +
  scale_shape_manual(values = c(1,19)) +
  scale_color_manual(values=c(`mild` = "#DCCA2CFF", `moderate to\nsevere` = "#6FB382FF"
)) +
  labs(y = "KEGG Pathway",
       x = "<-- Decreased vs. normal growth    Increased vs. normal growth -->",
       color = "Growth faltering")

plot(g)

ggsave(here("Output", glue("{current_sub_study}_kegg_pathway_pvalue_graph.pdf")), height = 3, width = 6)

```

### KEGG orthologs

Taking the top 500 KEGG orthologs and testing them

```{r message=FALSE, warning=FALSE, paged.print=TRUE, results='asis'}

top500 <- kegg_props %>%
  filter(SampleID %in% s_toTest$SampleID) %>%
  group_by(geneID) %>%
  summarise(mean_props = mean(props)) %>%
  arrange(desc(mean_props)) %>%
  slice(1:500)

kegg_props_subset <- kegg_props %>%
  select(SampleID, Ortholog = geneID, count, props) %>%
  filter(Ortholog %in% top500$geneID) %>%
  mutate(props_log2 = log2(props))

form1 = "props_log2 ~ timepoint_numeric * growth_cat + npo_days_prop + feeds_prop_hm + AbxHoursPerHOL"
random1 = "SubjectID"

#for whatever reason, I'm getting a "system is computationally singular: reciprocal condition number = 1.74397e-46" error code in nlme

# temp_df <- tibble()
# 
# for (kegg_gene in top500$geneID) {
# 
#   props_toTest <- kegg_props %>%
#   filter(geneID %in% kegg_gene) %>%
#   mutate(props_log2 = log2(props)) %>%
#   inner_join(s_toTest, by = "SampleID")
# 
#   possible_error <- try(tiny_df <- tidy_lmer(nlme::lme(as.formula(form1), random = as.formula("~ 1 | SubjectID"), data=props_toTest, na.action=na.omit)) %>% filter(term != '(Intercept)'))
# 
#   tiny_df$geneID <- kegg_gene
# 
#   if (class(possible_error)[1] %in% "try-error") {
#     tiny_df <- tibble()
#   }
# 
#   temp_df <- bind_rows(temp_df, tiny_df)
# 
# }
# 
# summaries_df <- temp_df %>%
#   group_by(term) %>%
#   mutate(fdr = p.adjust(p.value, method="BH")) %>%
#   ungroup() %>%
#   select(Ortholog = geneID, everything())
# 
# save(summaries_df, file = here("Data", glue("{current_sub_study}_kegg500_changes.RDS")))

load(file = here("Data", glue("{current_sub_study}_kegg500_changes.RDS")))

to_select <- summaries_df %>%
  mutate(term = str_remove(term, "growth_cat")) %>%
  filter(term %in% c("moderate to severe retardation", "mild retardation"), fdr < 0.05, Value > 0) %>%
  arrange(desc(Value)) %>%
  slice(1:45) %>%
  pull(Ortholog) %>%
  as.character() %>%
  unique()
  
plot_df <- summaries_df %>%
  mutate(term = str_remove(term, "growth_cat")) %>%
  filter(term %in% c("moderate to severe retardation", "mild retardation"), Ortholog %in% to_select) %>%
  left_join(kegg_defnitions, by = "Ortholog") %>%
  mutate(ortho_lable = paste(Ortholog, str_trunc(Definition, 30, "right"))) %>%
  mutate(CI95_lwr = round(Value - 1.96*Std.Error, 2)) %>%
  mutate(CI95_upr = round(Value + 1.96*Std.Error, 2)) %>%
  mutate(Significant = case_when(fdr < 0.05 ~ "p.adj < 0.05",
                                 fdr > 0.05 ~ "Not significant")) %>%
  arrange(Value) %>%
  mutate(ortho_lable = fct_reorder(ortho_lable, Value))

g <- ggplot(plot_df, aes(Value, ortho_lable, color = term, shape = Significant)) +
  theme_bw() +
  geom_pointrange(aes(xmin=Value-Std.Error, xmax=Value+Std.Error), position = position_dodge(width = 0.4)) +
  geom_vline(xintercept = 0, linetype = "longdash") +
  scale_shape_manual(values = c(1,19)) +
  scale_color_manual(values=ann_colors$growth_cat[2:3]) +
  labs(y = "Ortholog / Description",
       x = "<-- Decreased vs. normal growth    Increased vs. normal growth -->",
       color = "Growth category")

plot(g)

ggsave(here("Output", glue("{current_sub_study}_top_increasing_orthologs.pdf")), height = 7, width = 10)

to_select <- summaries_df %>%
  mutate(term = str_remove(term, "growth_cat")) %>%
  filter(term %in% c("moderate to severe retardation", "mild retardation"), fdr < 0.05, Value < 0) %>%
  arrange(desc(Value)) %>%
  slice(1:45) %>%
  pull(Ortholog) %>%
  as.character() %>%
  unique()
  
plot_df <- summaries_df %>%
  mutate(term = str_remove(term, "growth_cat")) %>%
  filter(term %in% c("moderate to severe retardation", "mild retardation"), Ortholog %in% to_select) %>%
  left_join(kegg_defnitions, by = "Ortholog") %>%
  mutate(ortho_lable = paste(Ortholog, str_trunc(Definition, 30, "right"))) %>%
  mutate(CI95_lwr = round(Value - 1.96*Std.Error, 2)) %>%
  mutate(CI95_upr = round(Value + 1.96*Std.Error, 2)) %>%
  mutate(Significant = case_when(fdr < 0.05 ~ "p.adj < 0.05",
                                 fdr > 0.05 ~ "Not significant")) %>%
  arrange(Value) %>%
  mutate(ortho_lable = fct_reorder(ortho_lable, Value))

g <- ggplot(plot_df, aes(Value, ortho_lable, color = term, shape = Significant)) +
  theme_bw() +
  geom_pointrange(aes(xmin=Value-Std.Error, xmax=Value+Std.Error), position = position_dodge(width = 0.4)) +
  geom_vline(xintercept = 0, linetype = "longdash") +
  scale_shape_manual(values = c(1,19)) +
  scale_color_manual(values=ann_colors$growth_cat[2:3]) +
  labs(y = "Ortholog / Description",
       x = "<-- Decreased vs. normal growth    Increased vs. normal growth -->",
       color = "Growth category")

plot(g)

ggsave(here("Output", glue("{current_sub_study}_top_decreasing_orthologs.pdf")), height = 7, width = 10)

table_name <- glue("{current_sub_study}_individual_gene_changes_growth_categories.csv")

my_table <- summaries_df %>%
  mutate(term = str_remove(term, "growth_cat")) %>%
  filter(term %in% c("moderate to severe retardation", "mild retardation"), fdr < 0.05) %>%
  left_join(kegg_defnitions, by = "Ortholog") %>%
  left_join(path_annotation, by = "Ortholog")

write_csv(my_table, here("Output", table_name))

print(glue("All orthologs saved together as {table_name}"))

```

#### Combined graph

```{r}

to_select <- summaries_df %>%
  mutate(term = str_remove(term, "growth_cat")) %>%
  filter(term %in% c("moderate to severe retardation", "mild retardation"), fdr < 0.05) %>%
  arrange(desc(fdr)) %>%
  slice(1:45) %>%
  pull(Ortholog) %>%
  as.character() %>%
  unique()
  
plot_df <- summaries_df %>%
  mutate(term = str_remove(term, "growth_cat")) %>%
  filter(term %in% c("moderate to severe retardation", "mild retardation"), Ortholog %in% to_select) %>%
  left_join(kegg_defnitions, by = "Ortholog") %>%
  mutate(ortho_lable = paste(Ortholog, str_trunc(Definition, 30, "right"))) %>%
  mutate(CI95_lwr = round(Value - 1.96*Std.Error, 2)) %>%
  mutate(CI95_upr = round(Value + 1.96*Std.Error, 2)) %>%
  mutate(Significant = case_when(fdr < 0.05 ~ "p.adj < 0.05",
                                 fdr > 0.05 ~ "Not significant")) %>%
  arrange(Value) %>%
  mutate(ortho_lable = fct_reorder(ortho_lable, Value))

g <- ggplot(plot_df, aes(Value, ortho_lable, color = term, shape = Significant)) +
  theme_bw() +
  geom_pointrange(aes(xmin=Value-Std.Error, xmax=Value+Std.Error), position = position_dodge(width = 0.4)) +
  geom_vline(xintercept = 0, linetype = "longdash") +
  scale_shape_manual(values = c(1,19)) +
  scale_color_manual(values=ann_colors$growth_cat[2:3]) +
  labs(y = "Ortholog / Description",
       x = "<-- Decreased vs. normal growth    Increased vs. normal growth -->",
       color = "Growth category")

plot(g)

ggsave(here("Output", glue("{current_sub_study}_top_changing_orthologs.pdf")), height = 7, width = 10)

```


### Butyrate and Aromatic amino acids

#### Linear models

```{r, results='asis'}

props_toTest <- but_ipa %>%
  mutate(props_log2 = log2(props))

form1 = "props_log2 ~ timepoint_numeric * growth_cat + npo_days_prop + feeds_prop_hm + AbxHoursPerHOL"
random1 = "SubjectID"

#for whatever reason, I'm getting a "false convergence (8)" error code in nlminb

temp_df <- tibble()

for (selected_gene in unique(but_ipa$geneID)) {

  props_toTest <- but_ipa %>%
  filter(geneID %in% selected_gene) %>%
  mutate(props_log2 = log2(props)) %>%
  inner_join(s_toTest, by = "SampleID")

  possible_error <- try(tiny_df <- tidy_lmer(nlme::lme(as.formula(form1), random = as.formula("~ 1 | SubjectID"), data=props_toTest, na.action=na.omit)) %>% filter(term != '(Intercept)'))

  tiny_df$geneID <- selected_gene

  if (class(possible_error)[1] %in% "try-error") {
    print(glue("Error with {selected_gene}"))
    tiny_df <- tibble()
  }

  temp_df <- bind_rows(temp_df, tiny_df)

}

summaries_df <- temp_df %>%
  group_by(term) %>%
  mutate(fdr = p.adjust(p.value, method="BH")) %>%
  ungroup() %>%
  select(geneID, everything())

summaries_df %>%
  mutate(term = str_remove(term, "growth_cat")) %>%
  filter(p.value < 0.05) %>%
  pander(split.table=Inf, digits=2, caption = glue("Linear models of log2 transformed abundances"), emphasize.strong.rows = which(.$fdr<=0.05))

table_export(summaries_df %>% filter(p.value < 0.05), "gene_analysis_table_")

```

#### Graphing these changes

```{r}

plot_df <- summaries_df %>%
  mutate(term = str_remove(term, "growth_cat")) %>%
  filter(term %in% c("moderate to severe retardation", "mild retardation")) %>%
  left_join(annot_df, by = "geneID") %>%
  mutate(CI95_lwr = round(Value - 1.96*Std.Error, 2)) %>%
  mutate(CI95_upr = round(Value + 1.96*Std.Error, 2)) %>%
  mutate(Significant = case_when(fdr < 0.05 ~ "p.adj < 0.05",
                                 fdr > 0.05 ~ "Not significant")) %>%
  arrange(Value) %>%
  mutate(geneID = fct_reorder(geneID, Value))

g <- ggplot(plot_df, aes(Value, geneID, color = term, shape = Significant)) +
  theme_bw() +
  geom_pointrange(aes(xmin=Value-Std.Error, xmax=Value+Std.Error), position = position_dodge(width = 0.4)) +
  geom_vline(xintercept = 0, linetype = "longdash") +
  scale_shape_manual(values = c(1,19)) +
  scale_color_manual(values=ann_colors$growth_cat[2:3]) +
  facet_grid(database~"Value", scales = "free_y", space = "free_y") +
  labs(y = "geneID",
       x = "<-- Decreased vs. normal growth    Increased vs. normal growth -->",
       color = "Growth category")

plot(g)

ggsave(here("Output",glue("{current_sub_study}_gene_abundaces_but_ipa.pdf")), height=5, width=8, useDingbats=F)

```


## Growth failure

Growth failure is now those babies who are the bottom 3rd percentile in terms of normal growth.

### Bray-curtis PCoA

First, we will examine overall differences using PCoA of the Bray-curtis' distances of the gene abundances between samples.

#### PCoA's

```{r fig.height=5, fig.width=5}
s_toPlot %>%
  pcoaplus(bc_kegg) %>%
  plot(color=growth_failure) +
    scale_color_manual(values=ann_colors$growth_fail) +
    theme_bw() +
    theme(
      axis.text=element_blank(),
      axis.ticks=element_blank(),
      panel.grid = element_blank(), 
      aspect.ratio = 1,
      strip.background = element_blank()
    ) + 
    labs(color="", title="Bray-Curtis")
```

#### Permanova

```{r, results='asis'}
s_permanova <- s_toTest %>%
  filter(!is.na(feeds_prop_hm))

summaries_df <-
  s_permanova %>%
    do(adonisplus(
      ., distmat = dist_in, formula = distmat ~ npo_days_prop + feeds_prop_hm + AbxHoursPerHOL + timepoint_numeric * growth_cat,
      sample_id_var = SampleID, rep_meas_var = SubjectID,
      shuffle = c(growth_failure = "between", timepoint_numeric = "within"), permutations=perm)) %>%
    ungroup() %>%
    mutate(metric = "Bray-Curtis") %>%
  filter(!term %in% c("Residual", "Total")) %>%
  select(metric, everything())

write_rds(summaries_df, here("Data", "kegg_bc_permanova_results_34to37weeks_gf.rds"))

summaries_df <- read_rds(here("Data", "kegg_bc_permanova_results_34to37weeks_gf.rds"))

table_export(summaries_df, "gene_analysis_table_")

summaries_df %>%
  mutate(term = str_replace(term, "timepoint_numeric", "Time")) %>%
  mutate(term = str_replace(term, "growth_failure", "Growth failure")) %>%
  pander(split.cells = Inf, digits = 2, emphasize.strong.rows = which(.$p.value<=0.05))

```

### Linear models of pathways

```{r, results='asis'}

number_of_top_pathways = 25

props_toTest <- kegg_path_props %>%
  mutate(props_log2 = log2(props)) %>%
  mutate(PathwayName = fct_lump(PathwayName, n = number_of_top_pathways, w = .$props, other_level = "Other")) %>%
  filter(!PathwayName %in% c("Metabolic pathways", "Other"))

form1 = "props_log2 ~ timepoint_numeric * growth_failure + npo_days_prop + feeds_prop_hm + AbxHoursPerHOL"
random1 = "SubjectID"

summaries_df <- run_lmer_on_props(factor_to_test = "PathwayName", props_toTest = props_toTest, s_toTest = s_toTest, form1 = form1,rep_mes_label = random1, p_cutoff =  0.05) 

summaries_df %>%
  mutate(term = str_remove(term, "growth_failure")) %>%
  pander(split.table=Inf, digits=2, caption = glue("Linear models of log2 transformed abundances of KEGG pathways"), emphasize.strong.rows = which(.$fdr<=0.05))

table_export(summaries_df, "gene_analysis_table_")

```

#### Boxplot of pathway abundances

```{r}

to_select <- unique(as.character(summaries_df$PathwayName))

number_of_blocks <- ceiling(length(to_select) / 9)
chunk_it <- function(x,n) split(x, cut(seq_along(x), n, labels = FALSE))
my_blocks <- chunk_it(to_select, number_of_blocks)

for (i in 1:number_of_blocks) {
  subset <- my_blocks[[i]]

  props_toTest %>%
    filter(PathwayName %in% subset) %>%
    filter(SampleID %in% s_toTest$SampleID) %>%
    merge(s_toTest, by="SampleID") %>%
    
    mutate(PathwayName = reorder(PathwayName, -props)) %>%
    ggplot(aes(x=growth_failure, y=props, color=growth_failure)) +
    geom_boxplot(outlier.alpha = 0) +
    geom_quasirandom(dodge.width = 0.75) +
    theme_bw() +
    theme(
      axis.text.x = element_text(angle=45, vjust=1, hjust=1),
      strip.background = element_blank()) +
    #scale_color_manual(values=brewer.pal(8, "Set1")[c(2,5,1)]) +
    scale_color_brewer(palette = "Set2") +
    scale_y_continuous(trans = "log2") +
    facet_wrap(~PathwayName, scales="free_y", ncol=3) +
    labs(x="", color="Growth failure",
         y="Relative abundance",
         title = "Pathways in samples that had gene abundance differences")
  
  ggsave(here("Output",glue("{current_sub_study}_kegg_pathway_differences_growth_failure_{i}.pdf")), height=7, width=12, useDingbats=F)
}

```

### KEGG orthologs

Taking the top 500 KEGG orthologs and testing them

```{r message=FALSE, warning=FALSE, paged.print=TRUE, results='asis'}

top500 <- kegg_props %>%
  filter(SampleID %in% s_toTest$SampleID) %>%
  group_by(geneID) %>%
  summarise(mean_props = mean(props)) %>%
  arrange(desc(mean_props)) %>%
  slice(1:500)

kegg_props_subset <- kegg_props %>%
  select(SampleID, Ortholog = geneID, count, props) %>%
  filter(Ortholog %in% top500$geneID) %>%
  mutate(props_log2 = log2(props))

form1 = "props_log2 ~ timepoint_numeric * growth_failure + npo_days_prop + feeds_prop_hm + AbxHoursPerHOL"
random1 = "SubjectID"

#for whatever reason, I'm getting a "Error in MEEM(object, conLin, control$niterEM) : Overfitted model!" error code in nlme

# temp_df <- tibble()
# 
# for (kegg_gene in top500$geneID) {
# 
#   props_toTest <- kegg_props %>%
#   filter(geneID %in% kegg_gene) %>%
#   mutate(props_log2 = log2(props)) %>%
#   inner_join(s_toTest, by = "SampleID")
# 
#   possible_error <- try(tiny_df <- tidy_lmer(nlme::lme(as.formula(form1), random = as.formula("~ 1 | SubjectID"), data=props_toTest, na.action=na.omit)) %>% filter(term != '(Intercept)'))
# 
#   tiny_df$geneID <- kegg_gene
# 
#   if (class(possible_error)[1] %in% "try-error") {
#     tiny_df <- tibble()
#   }
# 
#   temp_df <- bind_rows(temp_df, tiny_df)
# 
# }
# 
# summaries_df <- temp_df %>%
#   group_by(term) %>%
#   mutate(fdr = p.adjust(p.value, method="BH")) %>%
#   ungroup() %>%
#   select(Ortholog = geneID, everything())
# 
# save(summaries_df, file = here("Data", glue("{current_sub_study}_kegg500_changes_gf.RDS")))

load(file = here("Data", glue("{current_sub_study}_kegg500_changes_gf.RDS")))

to_select <- summaries_df %>%
  mutate(term = str_remove(term, "growth_failure")) %>%
  filter(term %in% c("Yes"), fdr < 0.05, Value > 0) %>%
  arrange(desc(Value)) %>%
  slice(1:45) %>%
  pull(Ortholog) %>%
  as.character() %>%
  unique()
  
plot_df <- summaries_df %>%
  mutate(term = str_remove(term, "growth_failure")) %>%
  filter(term %in% c("Yes"), Ortholog %in% to_select) %>%
  left_join(kegg_defnitions, by = "Ortholog") %>%
  mutate(ortho_lable = paste(Ortholog, str_trunc(Definition, 30, "right"))) %>%
  mutate(CI95_lwr = round(Value - 1.96*Std.Error, 2)) %>%
  mutate(CI95_upr = round(Value + 1.96*Std.Error, 2)) %>%
  mutate(Significant = case_when(fdr < 0.05 ~ "p.adj < 0.05",
                                 fdr > 0.05 ~ "Not significant")) %>%
  arrange(Value) %>%
  mutate(ortho_lable = fct_reorder(ortho_lable, Value))

g <- ggplot(plot_df, aes(Value, ortho_lable, color = term, shape = Significant)) +
  theme_bw() +
  geom_pointrange(aes(xmin=Value-Std.Error, xmax=Value+Std.Error), position = position_dodge(width = 0.4)) +
  geom_vline(xintercept = 0, linetype = "longdash") +
  scale_shape_manual(values = c("Not significant" = 1, "p.adj < 0.05" = 19)) +
  scale_color_manual(values = ann_colors$growth_fail[2]) +
  labs(y = "Ortholog / Description",
       x = "<-- Decreased vs. no growth failure    Increased vs. no growth failure -->")

plot(g)

ggsave(here("Output", glue("{current_sub_study}_top_increasing_orthologs_gf.pdf")), height = 7, width = 10)

to_select <- summaries_df %>%
  mutate(term = str_remove(term, "growth_failure")) %>%
  filter(term %in% c("Yes"), fdr < 0.05, Value < 0) %>%
  arrange(desc(Value)) %>%
  slice(1:45) %>%
  pull(Ortholog) %>%
  as.character() %>%
  unique()
  
plot_df <- summaries_df %>%
  mutate(term = str_remove(term, "growth_failure")) %>%
  filter(term %in% c("Yes"), Ortholog %in% to_select) %>%
  left_join(kegg_defnitions, by = "Ortholog") %>%
  mutate(ortho_lable = paste(Ortholog, str_trunc(Definition, 30, "right"))) %>%
  mutate(CI95_lwr = round(Value - 1.96*Std.Error, 2)) %>%
  mutate(CI95_upr = round(Value + 1.96*Std.Error, 2)) %>%
  mutate(Significant = case_when(fdr < 0.05 ~ "p.adj < 0.05",
                                 fdr > 0.05 ~ "Not significant")) %>%
  arrange(Value) %>%
  mutate(ortho_lable = fct_reorder(ortho_lable, Value))

g <- ggplot(plot_df, aes(Value, ortho_lable, color = term, shape = Significant)) +
  theme_bw() +
  geom_pointrange(aes(xmin=Value-Std.Error, xmax=Value+Std.Error), position = position_dodge(width = 0.4)) +
  geom_vline(xintercept = 0, linetype = "longdash") +
  scale_shape_manual(values = c("Not significant" = 1, "p.adj < 0.05" = 19)) +
  scale_color_manual(values = ann_colors$growth_fail[2]) +
  labs(y = "Ortholog / Description",
       x = "<-- Decreased vs. no growth failure    Increased vs. no growth failure -->")

plot(g)

ggsave(here("Output", glue("{current_sub_study}_top_decreasing_orthologs_gf.pdf")), height = 7, width = 10)

table_name <- glue("{current_sub_study}_individual_gene_changes_growth_categories_gf.csv")

my_table <- summaries_df %>%
  mutate(term = str_remove(term, "growth_failure")) %>%
  filter(term %in% c("Yes"), fdr < 0.05) %>%
  left_join(kegg_defnitions, by = "Ortholog") %>%
  left_join(path_annotation, by = "Ortholog")

write_csv(my_table, here("Output", table_name))

print(glue("All orthologs saved together as {table_name}"))

```

#### Combined graph

```{r}

to_select <- summaries_df %>%
  mutate(term = str_remove(term, "growth_failure")) %>%
  filter(term %in% c("Yes"), fdr < 0.05) %>%
  arrange(desc(fdr)) %>%
  slice(1:45) %>%
  pull(Ortholog) %>%
  as.character() %>%
  unique()
  
plot_df <- summaries_df %>%
  mutate(term = str_remove(term, "growth_failure")) %>%
  filter(term %in% c("Yes"), Ortholog %in% to_select) %>%
  left_join(kegg_defnitions, by = "Ortholog") %>%
  mutate(ortho_lable = paste(Ortholog, str_trunc(Definition, 30, "right"))) %>%
  mutate(CI95_lwr = round(Value - 1.96*Std.Error, 2)) %>%
  mutate(CI95_upr = round(Value + 1.96*Std.Error, 2)) %>%
  mutate(Significant = case_when(fdr < 0.05 ~ "p.adj < 0.05",
                                 fdr > 0.05 ~ "Not significant")) %>%
  arrange(Value) %>%
  mutate(ortho_lable = fct_reorder(ortho_lable, Value))

g <- ggplot(plot_df, aes(Value, ortho_lable, color = term, shape = Significant)) +
  theme_bw() +
  geom_pointrange(aes(xmin=Value-Std.Error, xmax=Value+Std.Error), position = position_dodge(width = 0.4)) +
  geom_vline(xintercept = 0, linetype = "longdash") +
  scale_shape_manual(values = c("Not significant" = 1, "p.adj < 0.05" = 19)) +
  scale_color_manual(values = ann_colors$growth_fail[2]) +
  labs(y = "Ortholog / Description",
       x = "<-- Decreased vs. no growth failure    Increased vs. no growth failure -->",
       color = "")

plot(g)

ggsave(here("Output", glue("{current_sub_study}_top_changing_orthologs_gf.pdf")), height = 5.5, width = 10.5)

```


### Butyrate and Aromatic amino acids

#### Linear models

```{r, results='asis'}

props_toTest <- but_ipa %>%
  mutate(props_log2 = log2(props))

form1 = "props_log2 ~ timepoint_numeric * growth_failure + npo_days_prop + feeds_prop_hm + AbxHoursPerHOL"
random1 = "SubjectID"

#for whatever reason, I'm getting a "false convergence (8)" error code in nlminb

temp_df <- tibble()

for (selected_gene in unique(but_ipa$geneID)) {

  props_toTest <- but_ipa %>%
  filter(geneID %in% selected_gene) %>%
  mutate(props_log2 = log2(props)) %>%
  inner_join(s_toTest, by = "SampleID")

  possible_error <- try(tiny_df <- tidy_lmer(nlme::lme(as.formula(form1), random = as.formula("~ 1 | SubjectID"), data=props_toTest, na.action=na.omit)) %>% filter(term != '(Intercept)'))

  tiny_df$geneID <- selected_gene

  if (class(possible_error)[1] %in% "try-error") {
    print(glue("Error with {selected_gene}"))
    tiny_df <- tibble()
  }

  temp_df <- bind_rows(temp_df, tiny_df)

}

summaries_df <- temp_df %>%
  group_by(term) %>%
  mutate(fdr = p.adjust(p.value, method="BH")) %>%
  ungroup() %>%
  select(geneID, everything())

summaries_df %>%
  mutate(term = str_remove(term, "growth_failure")) %>%
  filter(p.value < 0.05) %>%
  pander(split.table=Inf, digits=2, caption = glue("Linear models of log2 transformed abundances"), emphasize.strong.rows = which(.$fdr<=0.05))

table_export(summaries_df %>% filter(p.value < 0.05), "gene_analysis_table_")

```

#### Graphing these changes

```{r}

plot_df <- summaries_df %>%
  mutate(term = str_remove(term, "growth_failure")) %>%
  filter(term %in% c("Yes")) %>%
  left_join(annot_df, by = "geneID") %>%
  mutate(CI95_lwr = round(Value - 1.96*Std.Error, 2)) %>%
  mutate(CI95_upr = round(Value + 1.96*Std.Error, 2)) %>%
  mutate(Significant = case_when(fdr < 0.05 ~ "p.adj < 0.05",
                                 fdr > 0.05 ~ "Not significant")) %>%
  arrange(Value) %>%
  mutate(geneID = fct_reorder(geneID, Value))

g <- ggplot(plot_df, aes(Value, geneID, color = term, shape = Significant)) +
  theme_bw() +
  geom_pointrange(aes(xmin=Value-Std.Error, xmax=Value+Std.Error), position = position_dodge(width = 0.4)) +
  geom_vline(xintercept = 0, linetype = "longdash") +
  scale_shape_manual(values = c(1,19)) +
  scale_color_manual(values=ann_colors$growth_fail[2]) +
  facet_grid(database~"Value", scales = "free_y", space = "free_y") +
  labs(y = "geneID",
       x = "<-- Decreased vs. no growth failure    Increased vs. no growth failure -->",
       color = "Growth failure")

plot(g)

ggsave(here("Output",glue("{current_sub_study}_gene_abundaces_but_ipa_gf.pdf")), height=5, width=8, useDingbats=F)

```

```{r savepoint, include=FALSE}

save.image(file = here("Data", "after_gene_abundance_analysis.RData"))

```

# End of Report

```{r Generate time stamped report2, eval=FALSE, include=FALSE}

#neat: you can run this following command in the console to give your reports custom names (or date-stamp them)

rstudioapi::documentSaveAll()

script_dir <- file.path("Scripts","Rmds")
output_dir <- file.path("Output")

if (!dir.exists(here::here(output_dir))) {
  dir.create(here::here(output_dir))
}

report_fp <- here::here(script_dir, "Shotgun_gene_abundance_analysis.Rmd")
output_fp <- here::here(output_dir, paste('Shotgun_gene_abundance_analysis_', Sys.Date(), '.pdf', sep=''))

rmarkdown::render(report_fp, output_file = output_fp)

```
